### 2011-04-18 draft2.tex
before t<t' stuff

\documentclass{article}

\usepackage{stmaryrd}
\usepackage{fullpage}

\begin{document}

\section{Language}

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}

\begin{eqnarray*}
  v,D &::=& String \\
  e &::=& v \mid (e~e) \mid (D~e) \\
  &&\mid \mbox{case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)]    \\ 
  p &::=& \mbox{let } f~\vec{x} = e \\
  \Gamma &::=& (f \mapsto t),\Gamma \mid (f~\vec{x} \mapsto e), \Gamma \mid \epsilon\\
\end{eqnarray*}
$\vec{x}$ denotes a vector of variables and $\Gamma$ a context.
$t > t' \iff \{ x | x \in t' \} \subset \{ x | x \in t \}$
\section{Encoding}

We define two different translations: one for expressions and one for programs.

\subsection{Expressions}
\begin{itemize}
\item \begin{array}{rcll}
  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x^{t} \in \Gamma \\
  && \llbracket \Gamma \vdash f x = e \in t_1 \rrbracket & \mbox{ if } f~x = e \in \Gamma \\
  && \bot & \mbox{ otherwise.} 
  \end{array}
\item $\llbracket \Gamma \vdash x~e_2 \in t \rrbracket = lookup...$
\item \begin{array}{rclll}
  \llbracket \Gamma \vdash e_1 e_2 \in t \rrbracket &=&  e_1 \in t_1 \to t_2, e_2 \in t'_2 \in \Gamma & \to & \top \\
  && e_2 \in t' (\neq \any) \in \Gamma  & \to & \llbracket \Gamma \vdash e_1 \in t' \to t \rrbracket \\
  && e_1 \in t_1 \to t_2 \in \Gamma & \to & \llbracket \Gamma \vdash e_2 \in t_1 \rrbracket
  \end{array}
\item $\llbracket \Gamma \vdash e_1~e_2 \in t \rrbracket = \llbracket \Gamma \vdash e \in Ok \to t \rrbracket \land \llbracket \Gamma \vdash e_2 \in Ok \rrbracket$
\item $\llbracket \Gamma \vdash \mbox{case } e \mbox{ of } (p_i \to e_i) \in t \rrbracket = e \neq \bad \land \forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, z \mapsto \{ a \mid \top \lor z=\unr \} \vdash e_i \in t \rrbracket$
\item $\llbracket \Gamma \vdash D~e \rrbracket = e \neq \bad \leftrightarrow D~e \neq \bad$
\end{itemize}

\subsection{Programs}
\begin{itemize}
\item $\llbracket \Gamma \vdash f~x=e \in t_1 \to t_2 \rrbracket = \forall x,f^\star~\llbracket \Gamma \vdash x \in t_1 \rrbracket \to \llbracket \Gamma, x^{t_1}, f^{\star~t_1 \to t_2} \vdash e[f^\star/f] \in t_2 \rrbracket, (\Gamma, f \mapsto t_1 \to t_2)$
\item $\llbracket \Gamma \vdash f=e \in t \rrbracket = \forall f \llbracket \Gamma, f^t \vdash e \in t \rrbracket, (\Gamma, f \mapsto t)$
\item $\llbracket \Gamma \vdash f~x=e \rrbracket = \forall f \llbracket \Gamma, f~x \mapsto e \vdash e \in t \rrbracket, (\Gamma, f~x \mapsto e)$
\end{itemize}

\thispagestyle{empty}
\end{document}

### 2011-04-19 draft2.tex
Befor removing universal quant

@@
 \newcommand{\ok}{\texttt{Ok}}
+\newcommand{\mand}{\mbox{ and }}
+\newcommand{\motherwise}{\mbox{ otherwise.}}
 
 \begin{eqnarray*}
-  v,D &::=& String \\
-  e &::=& v \mid (e~e) \mid (D~e) \\
-  &&\mid \mbox{case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)]    \\ 
-  p &::=& \mbox{let } f~\vec{x} = e \\
-  \Gamma &::=& (f \mapsto t),\Gamma \mid (f~\vec{x} \mapsto e), \Gamma \mid \epsilon\\
+  x,D &::=& String & (Variables, Data)\\
+  e &::=& x \mid (f~\overline{e}) \mid (D~\overline{e}) & (Expressions) \\
+  &&\mid \mbox{case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)]   & \\ 
+  &&\mid \bad \mid \unr & \\
+  p &::=& \mbox{let } f~\overline{x} = e ::: t,p \mid \epsilon & (Programs) \\
+  \Gamma &::=& (f \mapsto t),\Gamma \mid (f~\vec{x} \mapsto e), \Gamma \mid \epsilon & (Contexts) \\
+  t &::=& \{x \mid e \} \mid x:t \to t \mid \any & (Contracts) \\
+  \phi &::=& \top \mid \bot \mid x \mid \lnot \phi \mid \phi \land \phi \mid \forall x, \phi & (FOF) \\
 \end{eqnarray*}
-$\vec{x}$ denotes a vector of variables and $\Gamma$ a context.
-$t > t' \iff \{ x | x \in t' \} \subset \{ x | x \in t \}$
+
+With the constraint that the Haskell expression in the contract must return a Bool.
+
+$\vec{x}$ denotes a vector of variables and $\Gamma$ a context and multiset (which enables multiple contracts.)
+
+$t \geq t' \iff \{ e | x \in t' \} \subset \{ e | x \in t \}$ ($t$ is more restrictive than $t'$)
 \section{Encoding}
@@
 \begin{itemize}
+\item $\llbracket \Gamma \vdash \bad \in t \rrbracket = \bot$
+\item $\llbracket \Gamma \vdash \unr \in t \rrbracket = \top$
 \item \begin{array}{rcll}
-  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x^{t} \in \Gamma \\
-  && \llbracket \Gamma \vdash f x = e \in t_1 \rrbracket & \mbox{ if } f~x = e \in \Gamma \\
+  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x \mapsto t' \in \Gamma \\
   && \bot & \mbox{ otherwise.} 
   \end{array}
-\item $\llbracket \Gamma \vdash x~e_2 \in t \rrbracket = lookup...$
-\item \begin{array}{rclll}
-  \llbracket \Gamma \vdash e_1 e_2 \in t \rrbracket &=&  e_1 \in t_1 \to t_2, e_2 \in t'_2 \in \Gamma & \to & \top \\
-  && e_2 \in t' (\neq \any) \in \Gamma  & \to & \llbracket \Gamma \vdash e_1 \in t' \to t \rrbracket \\
-  && e_1 \in t_1 \to t_2 \in \Gamma & \to & \llbracket \Gamma \vdash e_2 \in t_1 \rrbracket
+\item \begin{array}{rcll}
+  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t_1 \to \dots \to t_n \to t_c \in t_c \rrbracket &=& \bigwedge_{1 \leq i \leq n}{\llbracket \Gamma \vdash e_i \in t_i \rrbracket} \\
+  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t \in t' \rrbracket &=& \bot \mbox{ ie type mismatch.}\\
   \end{array}
-\item $\llbracket \Gamma \vdash e_1~e_2 \in t \rrbracket = \llbracket \Gamma \vdash e \in Ok \to t \rrbracket \land \llbracket \Gamma \vdash e_2 \in Ok \rrbracket$
-\item $\llbracket \Gamma \vdash \mbox{case } e \mbox{ of } (p_i \to e_i) \in t \rrbracket = e \neq \bad \land \forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, z \mapsto \{ a \mid \top \lor z=\unr \} \vdash e_i \in t \rrbracket$
-\item $\llbracket \Gamma \vdash D~e \rrbracket = e \neq \bad \leftrightarrow D~e \neq \bad$
+\item $\llbracket \Gamma \vdash D~\overline{e} \in t \rrbracket = \overline{e} \neq \bad \to \llbracket \Gamma \vdash D~\overline{e} \in \ok \rrbracket \land ...$
+\item $\llbracket \Gamma \vdash \mbox{case } e \mbox{ of } [(p_1 \to e_1),\dots,(p_n \to e_n)] \in t \rrbracket = e \neq \bad \land \forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, TODO \vdash e_i \in t \rrbracket$
 \end{itemize}
 
+$$\llbracket \Gamma \vdash e \in t \rrbracket = \phi \iff (\phi = \top \iff \Gamma,e \mbox{ satisfies } t)$$
+
 \subsection{Programs}
-\begin{itemize}
-\item $\llbracket \Gamma \vdash f~x=e \in t_1 \to t_2 \rrbracket = \forall x,f^\star~\llbracket \Gamma \vdash x \in t_1 \rrbracket \to \llbracket \Gamma, x^{t_1}, f^{\star~t_1 \to t_2} \vdash e[f^\star/f] \in t_2 \rrbracket, (\Gamma, f \mapsto t_1 \to t_2)$
-\item $\llbracket \Gamma \vdash f=e \in t \rrbracket = \forall f \llbracket \Gamma, f^t \vdash e \in t \rrbracket, (\Gamma, f \mapsto t)$
-\item $\llbracket \Gamma \vdash f~x=e \rrbracket = \forall f \llbracket \Gamma, f~x \mapsto e \vdash e \in t \rrbracket, (\Gamma, f~x \mapsto e)$
-\end{itemize}
+Assume $t = t_1 \to \dots \to t_n \to t_c$ and $\overline{x} = (x_1,\dots,x_n)$
+
+\begin{array}{rcl}
+\llbracket \Gamma \vdash \mbox{ let } f~\overline{x} = e:::t,p \rrbracket &=& \forall f^\star, \forall_{1 \leq i \leq n} x_i, \llbracket \Gamma, x_1 \mapsto t_1,\dots,x_n \mapsto t_n, f^\star \mapsto t \vdash e[f^\star/f] \in t_c \rrbracket \\
+&& \land \llbracket \Gamma, f \mapsto t \vdash p \rrbracket
+\end{array}
 

### 2011-04-26 draft2.tex
Simplifying once again...

@@
 \usepackage{stmaryrd}
+\usepackage{amsmath}
 \usepackage{fullpage}
@@
 \newcommand{\unr}{\texttt{UNR}}
+\newcommand{\unrphi}{\texttt{UNR}_\phi}
 \newcommand{\bad}{\texttt{BAD}}
+\newcommand{\badphi}{\texttt{BAD}_\phi}
 \newcommand{\any}{\texttt{Any}}
@@
   x,D &::=& String & (Variables, Data)\\
-  e &::=& x \mid (f~\overline{e}) \mid (D~\overline{e}) & (Expressions) \\
-  &&\mid \mbox{case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)]   & \\ 
+  e &::=& x \mid f(\overline{e}) \mid D(\overline{e}) & (Expressions) \\
   &&\mid \bad \mid \unr & \\
-  p &::=& \mbox{let } f~\overline{x} = e ::: t,p \mid \epsilon & (Programs) \\
+  p &::=& \mbox{let $f~\overline{x} = $ case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)] ::: t,p   & (Programs) \\ 
+  &&\mid \mbox{let } f~\overline{x} = e ::: t,p \mid \epsilon \\
   \Gamma &::=& (f \mapsto t),\Gamma \mid (f~\vec{x} \mapsto e), \Gamma \mid \epsilon & (Contexts) \\
@@
 $t \geq t' \iff \{ e | x \in t' \} \subset \{ e | x \in t \}$ ($t$ is more restrictive than $t'$)
+
+case-expression can't be nested (just as lambdas) which is simpler for the FOL translation.
+
 \section{Encoding}
 
-We define two different translations: one for expressions and one for programs.
+\subsection{Expressions}
+\subsubsection{Axioms}
+\begin{itemize}
+\item $\lVert x \rVert = x $
+\item $\lVert f(\overline{e}) \rVert = f(\lVert \overline{e} \rVert)  $
+\item $\lVert D(\overline{e}) \rVert = D(\lVert \overline{e} \rVert)  $
+\item $\lVert \bad \rVert = \badphi $ (maybe we must specify that $\badphi$ is not $\top$ or $\bot$)
+\item $\lVert \unr \rVert = \unrphi $ (same here)
+\end{itemize}
 
-\subsection{Expressions}
+\subsubsection{Contract satisfaction}
 \begin{itemize}
@@
 \item \begin{array}{rcll}
-  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x \mapsto t' \in \Gamma \\
+  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x \mapsto t \in \Gamma \\
   && \bot & \mbox{ otherwise.} 
@@
 \item \begin{array}{rcll}
-  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t_1 \to \dots \to t_n \to t_c \in t_c \rrbracket &=& \bigwedge_{1 \leq i \leq n}{\llbracket \Gamma \vdash e_i \in t_i \rrbracket} \\
-  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t \in t' \rrbracket &=& \bot \mbox{ ie type mismatch.}\\
+  \llbracket \Gamma \vdash f\overline{(e_1,\dots,e_n)} ::: t_1 \to \dots \to t_n \to t_c \in t_c \rrbracket &=& \bigwedge_{1 \leq i \leq n}{\llbracket \Gamma \vdash e_i \in t_i \rrbracket} \\
+  \llbracket \Gamma \vdash f\overline{(e_1,\dots,e_n)} ::: t \in t' \rrbracket &=& \bot \mbox{ ie type mismatch.}\\
   \end{array}
-\item $\llbracket \Gamma \vdash D~\overline{e} \in t \rrbracket = \overline{e} \neq \bad \to \llbracket \Gamma \vdash D~\overline{e} \in \ok \rrbracket \land ...$
-\item $\llbracket \Gamma \vdash \mbox{case } e \mbox{ of } [(p_1 \to e_1),\dots,(p_n \to e_n)] \in t \rrbracket = e \neq \bad \land \forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, TODO \vdash e_i \in t \rrbracket$
+\item $\llbracket \Gamma \vdash D(\overline{e}) \in t \rrbracket = (\lVert \overline{e} \rVert \neq \badphi \leftrightarrow \lVert D(\overline{e}) \rVert \neq \badphi) \land \mbox{ check $D$ as a function application}$
 \end{itemize}
 
-$$\llbracket \Gamma \vdash e \in t \rrbracket = \phi \iff (\phi = \top \iff \Gamma,e \mbox{ satisfies } t)$$
+Invariant:~$\llbracket \Gamma \vdash e \in t \rrbracket = \phi \iff (\phi = \top \iff \Gamma,e \mbox{ satisfies } t)$
+\thispagestyle{empty}
 
@@
 
-\begin{array}{rcl}
-\llbracket \Gamma \vdash \mbox{ let } f~\overline{x} = e:::t,p \rrbracket &=& \forall f^\star, \forall_{1 \leq i \leq n} x_i, \llbracket \Gamma, x_1 \mapsto t_1,\dots,x_n \mapsto t_n, f^\star \mapsto t \vdash e[f^\star/f] \in t_c \rrbracket \\
-&& \land \llbracket \Gamma, f \mapsto t \vdash p \rrbracket
+\begin{itemize}
+\item $\llbracket \Gamma \vdash \mbox{let $f(\overline{x}) = $ case } e \mbox{ of } [(p_1 \to e_1),\dots,(p_n \to e_n)] \in t \rrbracket = \lVert e \rVert \neq \badphi \land (\forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, \vec{z} \mapsto \ok \vdash e_i \in t \rrbracket \lor \lVert e' \rVert = \unrphi)$ \\
+  ($\vec{z}$ is a short hand for $(z_1,\dots,z_k)$ and $e'$ is the whole expression)
+\item \begin{array}{rcl}
+\llbracket \Gamma \vdash \mbox{ let } f(\overline{x}) = e:::t,p \rrbracket &=& \forall x_1, \dots, x_n, \{ x_1 \in t_1 \}, \dots, \{x_n \in t_n \} \\
+&& \llbracket \Gamma, x_1 \mapsto t_1,\dots,x_n \mapsto t_n, f^\star \mapsto t \vdash e[f^\star/f] \in t_c \rrbracket \\
+&& \land \llbracket \Gamma, f \mapsto t \vdash p \rrbracket\\
+\{ x \in t \} &=& t(x) \land \forall y, t(y) \iff (\llbracket \Gamma \vdash p(y) \in \{z | z = \texttt{True} \} \rrbracket)
 \end{array}
+\end{itemize}
+
+($t = \{ y | p(y) \}$ and \textt{True} refers to the Haskell boolean value (and not $\top$))
 

### 2011-04-26 draft2.tex
No more contexts.

--- 2011-04-19 draft2.tex
@@
 \newcommand{\unr}{\texttt{UNR}}
+\newcommand{\unrphi}{\texttt{UNR}_\phi}
 \newcommand{\bad}{\texttt{BAD}}
+\newcommand{\badphi}{\texttt{BAD}_\phi}
 \newcommand{\any}{\texttt{Any}}
 \newcommand{\ok}{\texttt{Ok}}
+\newcommand{\let}{\mbox{let }}
 \newcommand{\mand}{\mbox{ and }}
@@
   x,D &::=& String & (Variables, Data)\\
-  e &::=& x \mid (f~\overline{e}) \mid (D~\overline{e}) & (Expressions) \\
-  &&\mid \mbox{case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)]   & \\ 
-  &&\mid \bad \mid \unr & \\
-  p &::=& \mbox{let } f~\overline{x} = e ::: t,p \mid \epsilon & (Programs) \\
-  \Gamma &::=& (f \mapsto t),\Gamma \mid (f~\vec{x} \mapsto e), \Gamma \mid \epsilon & (Contexts) \\
+  e &::=& x \mid e~e \mid D~e \mid \bad \mid \unr & (Expressions) \\
+  p &::=& \mbox{let $f~x_1,\dots,x_n = $ case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)] ::: t,p   & (Programs) \\ 
+  &&\mid \mbox{let } f~x_1,\dots,x_n = e ::: t,p \mid \epsilon \\
   t &::=& \{x \mid e \} \mid x:t \to t \mid \any & (Contracts) \\
@@
 
-$\vec{x}$ denotes a vector of variables and $\Gamma$ a context and multiset (which enables multiple contracts.)
+$t \geq t' \iff \{ e | x \in t' \} \subset \{ e | x \in t \}$ ($t$ is more restrictive than $t'$)
 
-$t \geq t' \iff \{ e | x \in t' \} \subset \{ e | x \in t \}$ ($t$ is more restrictive than $t'$)
+case-expression can't be nested (just as lambdas) which is simpler for the FOL translation.
+
 \section{Encoding}
-
-We define two different translations: one for expressions and one for programs.
 
@@
 \begin{itemize}
-\item $\llbracket \Gamma \vdash \bad \in t \rrbracket = \bot$
-\item $\llbracket \Gamma \vdash \unr \in t \rrbracket = \top$
-\item \begin{array}{rcll}
-  \llbracket \Gamma \vdash x \in t \rrbracket &=& \top & \mbox{ if } x \mapsto t' \in \Gamma \\
-  && \bot & \mbox{ otherwise.} 
-  \end{array}
-\item \begin{array}{rcll}
-  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t_1 \to \dots \to t_n \to t_c \in t_c \rrbracket &=& \bigwedge_{1 \leq i \leq n}{\llbracket \Gamma \vdash e_i \in t_i \rrbracket} \\
-  \llbracket \Gamma \vdash f~\overline{(e_1,\dots,e_n)} ::: t \in t' \rrbracket &=& \bot \mbox{ ie type mismatch.}\\
-  \end{array}
-\item $\llbracket \Gamma \vdash D~\overline{e} \in t \rrbracket = \overline{e} \neq \bad \to \llbracket \Gamma \vdash D~\overline{e} \in \ok \rrbracket \land ...$
-\item $\llbracket \Gamma \vdash \mbox{case } e \mbox{ of } [(p_1 \to e_1),\dots,(p_n \to e_n)] \in t \rrbracket = e \neq \bad \land \forall \vec{z}~p_i[\vec{z}] = e \to \llbracket \Gamma, TODO \vdash e_i \in t \rrbracket$
+\item $\llbracket e \in \{ x | p(x) \} \rrbracket = p(e)$
+\item $\llbracket e \in t_1 \to t_2 \rrbracket = \forall x, \llbracket x \in t_1 \rrbracket \to \llbracket e~x \in t_2 \rrbracket$
 \end{itemize}
 
-$$\llbracket \Gamma \vdash e \in t \rrbracket = \phi \iff (\phi = \top \iff \Gamma,e \mbox{ satisfies } t)$$
+\thispagestyle{empty}
 
 \subsection{Programs}
-Assume $t = t_1 \to \dots \to t_n \to t_c$ and $\overline{x} = (x_1,\dots,x_n)$
+Assume $t = t_1 \to \dots \to t_n \to t_c$.
 
-\begin{array}{rcl}
-\llbracket \Gamma \vdash \mbox{ let } f~\overline{x} = e:::t,p \rrbracket &=& \forall f^\star, \forall_{1 \leq i \leq n} x_i, \llbracket \Gamma, x_1 \mapsto t_1,\dots,x_n \mapsto t_n, f^\star \mapsto t \vdash e[f^\star/f] \in t_c \rrbracket \\
-&& \land \llbracket \Gamma, f \mapsto t \vdash p \rrbracket
-\end{array}
+\begin{itemize}
+\item $\llbracket \mbox{let } f~x_1 \dots x_n = e ::: t \rrbracket = (\forall_{1 \leq i \leq n} x_i, \bigwedge_{1 \leq i \leq n} \llbracket x_i \in t_i \rrbracket~\forall f^\star \llbracket f^\star \in t \rrbracket) \to \llbracket e[f^\star/f] \in t_c \rrbracket$
+\item $\llbracket \mbox{let } f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(p_i,e_i)] \in t \rrbracket = (\forall x_i, \bigwedge_{i \leq n} \llbracket x_i \in t_i \rrbracket~\forall f^\star \llbracket f^\star \in t \rrbracket) \to (\bigwedge_{j \leq k} p_j = e \to \llbracket e_j[f^\star/f] \in t_c \rrbracket)$
+\end{itemize}
 

### 2011-05-03 draft2.tex
Brand new version! Almost complete.

@@
 
-\section{Language}
+\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
+\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
+\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
+\newcommand{\trans}[1]{\llbracket #1 \rrbracket}
 
 \newcommand{\unr}{\texttt{UNR}}
-\newcommand{\unrphi}{\texttt{UNR}_\phi}
 \newcommand{\bad}{\texttt{BAD}}
-\newcommand{\badphi}{\texttt{BAD}_\phi}
 \newcommand{\any}{\texttt{Any}}
 \newcommand{\ok}{\texttt{Ok}}
-\newcommand{\let}{\mbox{let }}
-\newcommand{\mand}{\mbox{ and }}
-\newcommand{\motherwise}{\mbox{ otherwise.}}
+
+\thispagestyle{empty}
+\section{Grammars}
+\subsection{$\lambda$-lifted haskell subset}
+\begin{eqnarray*}
+  u,e &:=& x \mid f \mid e~e \mid D \mid \bad \mid n & (Expression)\\
+  p &:=& \Delta_1 \dots \Delta_n & (Program) \\
+  \Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract) \\
+  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition) \\
+  pat &:=&D~x_1 \dots x_n & (Pattern) \\
+\end{eqnarray*}
+
+\subsection{FOL}
+\begin{eqnarray*}
+  t &:=& x \mid \mbox{app}(t_1,t_2) \mid k & (Term) \\
+  k &:=& D \mid f \mid ? \mid \bad \mid \unr & (Constant) \\
+  \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula) \\
+\end{eqnarray*}
+
+\subsection{Contracts}
+\begin{eqnarray*}
+  c &:=& x:c_1 \to c_2\\
+  &\mid& (c_1,c_2) \\
+  &\mid& \{ x \mid e \} \\
+  &\mid& \any \\
+\end{eqnarray*}
+We can consider CF as a user contract.
+
+\section{Translation}
+We define several translations: $\etrans{},\dtrans{},\strans{},\trans{}$.
+\begin{eqnarray*}
+  \etrans{} &::& Expression \to Term \\
+  \dtrans{} &::& Definition \to FOF  \\
+  \strans{} &::& Expression \to Contract \to FOF \\
+  \trans {} &::& Definition \to Contract \to FOF \\
+\end{eqnarray*}
+
+\subsection{$\etrans{}$}
+$\etrans{e}$ is a term. The translation is direct.
+
+\subsection{$\dtrans{}$}
+$\dtrans{d}$ is a first-order formula.
 
 \begin{eqnarray*}
-  x,D &::=& String & (Variables, Data)\\
-  e &::=& x \mid e~e \mid D~e \mid \bad \mid \unr & (Expressions) \\
-  p &::=& \mbox{let $f~x_1,\dots,x_n = $ case $e$ of } [(p_1 \to e_1) \dots (p_n \to e_n)] ::: t,p   & (Programs) \\ 
-  &&\mid \mbox{let } f~x_1,\dots,x_n = e ::: t,p \mid \epsilon \\
-  t &::=& \{x \mid e \} \mid x:t \to t \mid \any & (Contracts) \\
-  \phi &::=& \top \mid \bot \mid x \mid \lnot \phi \mid \phi \land \phi \mid \forall x, \phi & (FOF) \\
+\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e} & \\
+\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i}) & \\
+&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)& \\ 
+&& \land \trans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) & (\varphi) \\
 \end{eqnarray*}
 
-With the constraint that the Haskell expression in the contract must return a Bool.
+\subsection{$\strans{}$}
+$\strans{e \in c}$ is a first-order formula.
+\begin{eqnarray*}
+\strans{e \in \any} &=& true \\
+\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \mbox{CF}(\etrans{u[e/x]} \neq \bad) \land (\etrans{u[e/x]} \neq False) \\
+\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]} \\
 
-$t \geq t' \iff \{ e | x \in t' \} \subset \{ e | x \in t \}$ ($t$ is more restrictive than $t'$)
+\end{eqnarray*}
+$False$ is a data constructor here.
 
-case-expression can't be nested (just as lambdas) which is simpler for the FOL translation.
+\subsection{$\trans{}$}
+It's the final translation, which takes a function definition and its contract and returns a first-order formula\\
 
-\section{Encoding}
+$\trans{f~x_1 \dots x_n = e \in c } = \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}$
 
-\subsection{Expressions}
-\begin{itemize}
-\item $\llbracket e \in \{ x | p(x) \} \rrbracket = p(e)$
-\item $\llbracket e \in t_1 \to t_2 \rrbracket = \forall x, \llbracket x \in t_1 \rrbracket \to \llbracket e~x \in t_2 \rrbracket$
-\end{itemize}
+We'd like a typical contract-checking session to go like this:
+\begin{enumerate}
+\item Start with an empty theory $T$.
+\item Let $f~x_1 \dots x_n = e \in c$ be a function definition to check wrt contract $c$. Check (with equinox) the consistency of the theory $T' = T \cup \trans{f~x_1 \dots x_n = e \in c}$
+\item If $T'$ is consistent then let $T = \strans{f \in c} \cup T$ and go to 2.
+with the next function definition; otherwise give a counter-example and ask the user for refinement of the contracts and/or lemmas(?)
+\end{enumerate}
 
-\thispagestyle{empty}
+\section{Questions}
 
-\subsection{Programs}
-Assume $t = t_1 \to \dots \to t_n \to t_c$.
+Here are some open issues, design choices and equinox-related questions.
 
-\begin{itemize}
-\item $\llbracket \mbox{let } f~x_1 \dots x_n = e ::: t \rrbracket = (\forall_{1 \leq i \leq n} x_i, \bigwedge_{1 \leq i \leq n} \llbracket x_i \in t_i \rrbracket~\forall f^\star \llbracket f^\star \in t \rrbracket) \to \llbracket e[f^\star/f] \in t_c \rrbracket$
-\item $\llbracket \mbox{let } f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(p_i,e_i)] \in t \rrbracket = (\forall x_i, \bigwedge_{i \leq n} \llbracket x_i \in t_i \rrbracket~\forall f^\star \llbracket f^\star \in t \rrbracket) \to (\bigwedge_{j \leq k} p_j = e \to \llbracket e_j[f^\star/f] \in t_c \rrbracket)$
-\end{itemize}
+\begin{enumerate}
+\item Nested implications may lead to existential quantification, is it troublesome in equinox? (altough we don't have this case here)
+\item More generally, does equinox accept any FOF the grammar here define?
+\item In the section $\dtrans{}$, we believe replacing $\varphi$ by $\lor \trans{f x_1 \dots x_n} = \unr$ is equivalent. What's better for equinox?
+\item Is the session stuff realistic? It looks like it can have a quadratic behaviour but maybe with the right API it's ok?
 

### 2011-05-03 draft2.tex
Alternate definition from the POPL paper opaque/transp definition actually used paper

@@
 $\strans{e \in c}$ is a first-order formula.
-\begin{eqnarray*}
+\begin{eqnarray}
 \strans{e \in \any} &=& true \\
-\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \mbox{CF}(\etrans{u[e/x]} \neq \bad) \land (\etrans{u[e/x]} \neq False) \\
+\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \mbox{CF}(\etrans{u[e/x]} \neq \bad) \land \etrans{u[e/x]} \neq False) \\
 \strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]} \\
+\end{eqnarray}
+$False$ is a data constructor here.
 
-\end{eqnarray*}
-$False$ is a data constructor here.
+Remark: we follow the semantics of the POPL paper but it's a bit restrictive.
+e.g. in equation 2 we could use the alternate semantics (namely B1 in the POPL paper) : 
+$$\strans{e \in \{x \mid u \}} = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$
 
@@
 
-$\trans{f~x_1 \dots x_n = e \in c } = \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}$
+\thispagestyle{empty}
+
+\begin{eqnarray*}
+\trans{opaque(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c} \\
+\trans{transp(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c} \\
+\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to Ok)} \\
+\end{eqnarray*}
 
@@
 \item Start with an empty theory $T$.
-\item Let $f~x_1 \dots x_n = e \in c$ be a function definition to check wrt contract $c$. Check (with equinox) the consistency of the theory $T' = T \cup \trans{f~x_1 \dots x_n = e \in c}$
+\item Let $f~x_1 \dots x_n = e \in c$ be an opaque function definition to check wrt contract $c$. Check (with equinox) the consistency of the theory $T' = T \cup \trans{f~x_1 \dots x_n = e \in c}$
 \item If $T'$ is consistent then let $T = \strans{f \in c} \cup T$ and go to 2.
@@
 \item Nested implications may lead to existential quantification, is it troublesome in equinox? (altough we don't have this case here)
-\item More generally, does equinox accept any FOF the grammar here define?
+\item More generally, does equinox accept any FOF the grammar here defines?
 \item In the section $\dtrans{}$, we believe replacing $\varphi$ by $\lor \trans{f x_1 \dots x_n} = \unr$ is equivalent. What's better for equinox?
 \item Is the session stuff realistic? It looks like it can have a quadratic behaviour but maybe with the right API it's ok?
+\end{enumerate}
 

### 2011-05-04 draft2.tex
Minor fixes and prototype code.

@@
   u,e &:=& x \mid f \mid e~e \mid D \mid \bad \mid n & (Expression)\\
-  p &:=& \Delta_1 \dots \Delta_n & (Program) \\
-  \Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract) \\
-  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition) \\
-  pat &:=&D~x_1 \dots x_n & (Pattern) \\
+  p &:=& \Delta_1 \dots \Delta_n & (Program)\\
+  \Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract)\\
+  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
+  pat &:=&D~x_1 \dots x_n & (Pattern)\\
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  t &:=& x \mid \mbox{app}(t_1,t_2) \mid k & (Term) \\
-  k &:=& D \mid f \mid ? \mid \bad \mid \unr & (Constant) \\
-  \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula) \\
+  t &:=& x \mid \mbox{app}(t_1,t_2) \mid k & (Term)\\
+  k &:=& D \mid f \mid ? \mid \bad \mid \unr & (Constant)\\
+  \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
 \end{eqnarray*}
@@
   &\mid& (c_1,c_2) \\
-  &\mid& \{ x \mid e \} \\
+  &\mid& \{ x \mid e \}\\
   &\mid& \any \\
@@
 \begin{eqnarray*}
-  \etrans{} &::& Expression \to Term \\
-  \dtrans{} &::& Definition \to FOF  \\
-  \strans{} &::& Expression \to Contract \to FOF \\
-  \trans {} &::& Definition \to Contract \to FOF \\
+  \etrans{} &::& Expression \to Term\\
+  \dtrans{} &::& Definition \to FOF\\
+  \strans{} &::& Expression \to Contract \to FOF\\
+  \trans {} &::& Definition \to Contract \to FOF\\
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e} & \\
-\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i}) & \\
-&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)& \\ 
-&& \land \trans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) & (\varphi) \\
+\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e} &\\
+\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i}) &\\
+&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)&\\ 
+&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) & (\varphi)\\
 \end{eqnarray*}
@@
 \begin{eqnarray}
-\strans{e \in \any} &=& true \\
-\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \mbox{CF}(\etrans{u[e/x]} \neq \bad) \land \etrans{u[e/x]} \neq False) \\
-\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]} \\
+\strans{e \in \any} &=& true\\
+\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
+\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}\\
 \end{eqnarray}
@@
 \begin{eqnarray*}
-\trans{opaque(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c} \\
-\trans{transp(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c} \\
-\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to Ok)} \\
+\trans{opaque(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
+\trans{transp(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
+\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to \dots \to Ok)}\\
 \end{eqnarray*}

### 2011-05-04 contracts/FOL.hs

@@ -1,35 +1,37 @@
-module FOL where
+module FOL ( Term (..)
+           , Constant (..)
+           , Formula (..)
+           , foralls  
+           , apps)
+where
 
 type Variable = String
-type Constructor = String
 
-data FOL = Top
-         | Bottom
-         | And FOL FOL
-         | Or FOL FOL
-         | Not FOL
-         | Implies FOL FOL 
-         | Iff FOL FOL
-         | Forall Variable FOL
-         | Var Variable (Maybe FOL)
-         | Eq FOL FOL
-         | BAD | UNR
-         deriving (Eq,Show)
+data Term = Var Variable
+          | App Term Term
+          | Constant Constant
+          deriving (Show,Eq)                  
+       
+data Constant = Con Variable
+              | Fun Variable
+              | BAD
+              | UNR
+              deriving (Show,Eq)                  
+      
+data Formula = Forall Variable Formula
+             | Implies Formula Formula
+             | Not Formula
+             | Or Formula Formula
+             | And Formula Formula 
+             | True
+             | False
+             | Eq Term Term
+             | CF Term
+             deriving (Show,Eq)                  
 
-plug :: FOL -> FOL -> FOL
-plug (Var v Nothing) f = Var v (Just f)
-plug (Var v (Just f1)) f2 = Var v (Just (plug f1 f2))
+foralls [] f = f
+foralls (x:xs) f = Forall x (foralls xs f)
 
-
-simplify :: FOL -> FOL
-simplify (And Top x) = x
-simplify (And x Top) = x
-simplify (Implies a b) = (Not (simplify a)) `Or` simplify b
-simplify (Not (Not x)) = x
-simplify (Not x) = Not $ simplify x
-simplify (And x y) = Not $ (Not x) `Or` (Not y)
-simplify (Forall x y) = Forall x (simplify y)
-simplify (Or x y) = Or (simplify x) (simplify y)
-simplify (Var x a) = Var x a
-simplify (Eq x y) = Eq (simplify x) (simplify y)
-simplify x = x
+apps [] = error "Cannot apply nothing"
+apps [x] = x
+apps (x:xs) = x `App` (apps xs)

### 2011-05-04 contracts/Haskell.hs

@@ -1,0 +1,57 @@
+module Haskell ( Expression (..)
+               , Program (..)
+               , DefCont (..)
+               , Definition (..)
+               , Pattern (..)
+               , Contract (..)
+               , apps
+               , subst
+               , ok)
+where
+
+type Variable = String
+type Constructor = String
+
+-- Integer not implemented
+data Expression = Var Variable
+                | Fun Variable
+                | Con Constructor
+                | App Expression Expression
+                | BAD
+                deriving (Show,Eq)                  
+
+type Program = [DefCont]
+
+data DefCont = Def Definition
+             | Transp Definition Contract
+             | Opaque Definition Contract
+             deriving (Show,Eq)                  
+               
+data Definition = Let Variable [Variable] Expression
+                | LetCase Variable [Variable] Expression [(Pattern,Expression)]
+                deriving (Show,Eq)                  
+                  
+type Pattern = [Variable]
+
+data Contract = AppC Variable Contract Contract -- x : c -> c'
+              | Pair Contract Contract
+              | Pred Variable Expression   -- {x:e}
+              | Any
+              deriving (Show,Eq)                  
+
+apps [] = error "Cannot apply nothing"
+apps [x] = x
+apps (x:xs) = x `App` (apps xs)
+
+subst :: Expression -> Expression -> Variable -> Expression -- e[x/y]
+subst (Var v) x y | v == y = x
+                  | False  = Var v
+subst (Fun f) x y | f == y = x
+                  | False  = Fun f
+subst (Con c) x y | c == y = x
+                  | False  = Con c
+subst (App e1 e2) x y = App (subst e1 x y) (subst e2 x y)
+subst BAD _ _ = BAD
+
+ok :: Contract
+ok = Pred "x" (Con "True")

### 2011-05-04 contracts/Translation.hs

@@ -1,69 +1,36 @@
-import Control.Monad.State
+module Translation where
 
-import qualified Expr as E
-import qualified FOL  as F
-import Program 
-import qualified Contract as C
+import qualified Haskell as H
+import qualified FOL as F
+import Data.List (foldl')
 
-type Unique = State Int
-
-exprFol :: E.Expr -> F.FOL
-exprFol (E.Var v) = F.Var v Nothing
-exprFol (E.Con c (Just e)) = F.Var c (Just $ exprFol e)
-exprFol (E.Con c Nothing) = F.Var c Nothing
-exprFol (E.App e1 e2) = case a1 of 
-  Nothing -> F.Var v1 (Just f2)
-  Just f  -> F.Var v1 (Just $ F.plug f f2)
-  where F.Var v1 a1 = exprFol e1
-        f2 = exprFol e2
-
-exprToFol :: E.Expr -> C.Contract -> Unique F.FOL
-exprToFol e (C.Pred _ p) = return $ exprFol (p `E.App` e)
-exprToFol e (C.App t1 t2) = do
-  n <- get
-  let v = "x" ++ (show n)
-  put (n+1)
-  f1 <- exprToFol (E.Var v) t1
-  f2 <- exprToFol (e `E.App` E.Var v) t2
-  return $ F.Forall v (f1 `F.Implies` f2)
-  where v = "x0"
-exprToFol e C.Any = return $ F.Top
-
-progToFol :: Program -> Unique F.FOL
-progToFol (Let f as e t) = do
-  f1 <- exprToFol (E.Var f) t
-  f2 <- exprToFol e tc
-  f3 <- xiInti as t
-  return $ foldr (\ a f -> F.Forall a f) (f3 `F.And` (F.Forall f f1) `F.Implies` f2) as
-  where tc = C.codomain t
-progToFol (LetCase f as e pes t) = do
-  f1 <- exprToFol (E.Var f) t
-  f2 <- rhs e pes tc
-  f3 <- xiInti as t
-  return $ foldr (\ a f -> F.Forall a f) (f3 `F.And` (F.Forall f f1) `F.Implies` f2) as
-  where tc = C.codomain t
-
-rhs :: E.Expr -> [(Pattern,E.Expr)] -> C.Contract -> Unique F.FOL
-rhs e pes t = do 
-    foldr (\ (pj,ej) f -> do df <- f
[…]

### 2011-05-05 draft2.tex
Now outputs TPTP formulae, accepted by equinox
Can also output formulae in a latex format (but it's unreadable even in latex anyway)
Needs some factorisation.

@@
 \documentclass{article}
+% Test comment
 
@@
 \section{Grammars}
+
+In the following, $D$ is a data constructor, $f$ a function symbol and we consider them as strings. $n$ represents an integer.
+
 \subsection{$\lambda$-lifted haskell subset}
-\begin{eqnarray*}
+\begin{center}
+\begin{array}{rclr}
   u,e &:=& x \mid f \mid e~e \mid D \mid \bad \mid n & (Expression)\\
@@
   pat &:=&D~x_1 \dots x_n & (Pattern)\\
-\end{eqnarray*}
+\end{array}
+\end{center}
 
 \subsection{FOL}
-\begin{eqnarray*}
-  t &:=& x \mid \mbox{app}(t_1,t_2) \mid k & (Term)\\
-  k &:=& D \mid f \mid ? \mid \bad \mid \unr & (Constant)\\
+\begin{center}
+\begin{array}{rclr}
+  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
   \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
-\end{eqnarray*}
+\end{array}
+\end{center}
+CF$(t)$ holds iff $t$ satisfies $\ok$.
 
 \subsection{Contracts}
-\begin{eqnarray*}
+\begin{center}
+\begin{array}{rclr}
   c &:=& x:c_1 \to c_2\\
@@
   &\mid& \any \\
-\end{eqnarray*}
-We can consider CF as a user contract.
+\end{array}
+\end{center}
+
+Semantics of contract satisfaction:
+\begin{center}
+\begin{array}{rcl}
+  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,\unr\})\\
+  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
+  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
+  e \in \any &\iff& True  
+\end{array}
+\end{center}
 
@@
 We define several translations: $\etrans{},\dtrans{},\strans{},\trans{}$.
-\begin{eqnarray*}
+\begin{center}
+\begin{array}{rclr}
   \etrans{} &::& Expression \to Term\\
@@
   \trans {} &::& Definition \to Contract \to FOF\\
-\end{eqnarray*}
+\end{array}
+\end{center}
 
@@
 
-\begin{eqnarray*}
-\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e} &\\
-\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i}) &\\
-&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)&\\ 
-&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) & (\varphi)\\
-\end{eqnarray*}
+\begin{eqnarray}
+\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
+\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i})\\
+&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
+&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) 
+\end{eqnarray}
 
@@
 \strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
-\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}\\
+\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
 \end{eqnarray}
@@
 
-\begin{eqnarray*}
+\begin{eqnarray}
 \trans{opaque(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
 \trans{transp(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
-\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to \dots \to Ok)}\\
-\end{eqnarray*}
+\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to \dots \to Ok)}
+\end{eqnarray}
 

### 2011-05-05 contracts/Contract.hs

@@ -1,16 +1,0 @@
-module Contract where
-
-import Expr
-
-data Contract = Pred Variable Expr -- {var|expr(var)}
-              | App Contract Contract
-              | Any
-              deriving (Eq,Show)
-dummy = Pred "x" (Var "f")
-
-toList :: Contract -> [Contract]
-toList Any = [Any]
-toList c@(Pred _ _) = [c]
-toList (Contract.App c1 c2)  = c1:(toList c2)
-
-codomain = last . toList 

### 2011-05-05 contracts/Expr.hs

@@ -1,11 +1,0 @@
-module Expr where
-
-type Variable = String
-type Constructor = String
-
-data Expr = Var Variable
-          | App Expr Expr
-          | Con Constructor (Maybe Expr)
-          | BAD | UNR
-          deriving (Eq,Show)
-                  

### 2011-05-05 contracts/FOL.hs

@@ -1,22 +1,24 @@
 module FOL ( Term (..)
-           , Constant (..)
            , Formula (..)
            , foralls  
-           , apps)
-where
+           , apps
+           , toTPTP
+           , splitOnAnd
+           , toLatex
+           , toLatexDocument)
+       where
+
+import Prelude hiding (True,False)
+import Data.Char (toUpper,toLower)
 
 type Variable = String
 
 data Term = Var Variable
           | App Term Term
-          | Constant Constant
+          | Fun Variable
+          | BAD
+          | UNR
           deriving (Show,Eq)                  
-       
-data Constant = Con Variable
-              | Fun Variable
-              | BAD
-              | UNR
-              deriving (Show,Eq)                  
       
 data Formula = Forall Variable Formula
              | Implies Formula Formula
@@ -35,3 +37,46 @@ foralls (x:xs) f = Forall x (foralls xs f)
 apps [] = error "Cannot apply nothing"
 apps [x] = x
 apps (x:xs) = x `App` (apps xs)
+
+splitOnAnd ::Formula -> [Formula]
+splitOnAnd (Forall x (And f1 f2)) = splitOnAnd (Forall x f1) ++ splitOnAnd (Forall x f2)
+splitOnAnd (Forall x f) = map (Forall x) $ splitOnAnd f
+splitOnAnd (And f1 f2) = splitOnAnd f1 ++ splitOnAnd f2
+splitOnAnd f = [f]
+
+toTPTP :: Formula -> String
+toTPTP f = header ++ "\n" ++ (aux f) ++ "\n" ++ footer
+  where header = "fof(axiom,axiom,"
+        footer = ").\n"
+        aux (Forall v f) = "![ " ++ (map toUpper v) ++ " ] : (" ++ aux f ++ ")"
+        aux (Implies f1 f2) = "(" ++ aux f1 ++ ") => (" ++ aux f2 ++ ")"
+        aux (Not f) = "~(" ++ aux f ++ ")"
+        aux (Or f1 f2) = "(" ++ aux f1 ++ ") | (" ++ aux f2 ++ ")"
[…]

### 2011-05-05 contracts/Haskell.hs

@@ -6,13 +6,14 @@ module Haskell ( Expression (..)
                , Contract (..)
                , apps
                , subst
+               , substC
                , ok)
 where
 
 type Variable = String
 type Constructor = String
 
--- Integer not implemented
+-- TODO Integer not implemented
 data Expression = Var Variable
                 | Fun Variable
                 | Con Constructor
@@ -45,13 +46,19 @@ apps (x:xs) = x `App` (apps xs)
 
 subst :: Expression -> Expression -> Variable -> Expression -- e[x/y]
 subst (Var v) x y | v == y = x
-                  | False  = Var v
+                  | otherwise  = Var v
 subst (Fun f) x y | f == y = x
-                  | False  = Fun f
+                  | otherwise  = Fun f
 subst (Con c) x y | c == y = x
-                  | False  = Con c
+                  | otherwise  = Con c
 subst (App e1 e2) x y = App (subst e1 x y) (subst e2 x y)
 subst BAD _ _ = BAD
+
+substC :: Contract -> Expression -> Variable -> Contract
+substC (AppC u c1 c2) x y = AppC u (substC c1 x y) (substC c2 x y) -- TODO and if u==y?
+substC (Pair c1 c2) x y = Pair (substC c1 x y) (substC c2 x y)
+substC (Pred u e) x y = if u/=y then Pred u (subst e x y) else (Pred u e)
+substC Any _ _ = Any
 
 ok :: Contract
 ok = Pred "x" (Con "True")

### 2011-05-05 contracts/Program.hs

@@ -1,11 +1,0 @@
-module Program where
-       
-import Expr
-import Contract 
-
-data Program = Let Variable [Variable] Expr Contract
-             | LetCase Variable [Variable] Expr [(Pattern,Expr)] Contract
-             deriving (Eq,Show)
-               
-type Pattern = (Constructor,[Variable])
-

### 2011-05-05 contracts/Translation.hs

@@ -1,36 +1,79 @@
-module Translation where
+module Main where
 
 import qualified Haskell as H
 import qualified FOL as F
 import Data.List (foldl')
+import Control.Monad.State
 
 eTrans :: H.Expression -> F.Term
 dTrans :: H.Definition -> F.Formula
-sTrans :: H.Expression -> H.Contract -> F.Formula
-trans  :: H.DefCont -> H.Contract -> F.Formula
+sTrans :: H.Expression -> H.Contract -> Fresh F.Formula
+trans  :: H.DefCont    -> Fresh F.Formula
 
+type Fresh = State (String,Int)
 
 eTrans (H.Var v) = F.Var v
-eTrans (H.Fun f) = F.Constant $ F.Fun f
-eTrans (H.Con c) = F.Constant $ F.Con c
+eTrans (H.Fun f) = F.Fun f
 eTrans (H.App e1 e2) = F.App (eTrans e1) (eTrans e2)
-eTrans H.BAD = F.Constant $ F.BAD
+eTrans H.BAD = F.BAD
+eTrans (H.Con d) = F.Var d
 
 dTrans (H.Let f vs e) = F.foralls vs $ (eTransfxi f vs) `F.Eq` (eTrans e)
-dTrans (H.LetCase f vs e pes) = F.foralls vs $ ((eTrans e `F.Eq` F.Constant F.BAD) `F.Implies` (eTransfxi f vs `F.Eq` F.Constant F.BAD)) `F.And` 
-                                ((eTransfxi f vs `F.Eq` (F.Constant F.UNR)) `F.Or` (foldl' (\f (d,_) -> f `F.Or` (((F.Constant $ F.Fun "head") `F.App` eTrans e) `F.Eq` (F.Constant $ F.Con (head d)))) F.False pes))
+dTrans (H.LetCase f vs e pes) = F.foralls vs $ ((eTrans e `F.Eq` F.BAD) `F.Implies` (eTransfxi f vs `F.Eq` F.BAD)) `F.And` 
+                                ((eTransfxi f vs `F.Eq` F.UNR) `F.Or` (foldl' (\f (d,_) -> f `F.Or` (((F.Fun "head") `F.App` eTrans e) `F.Eq` (F.Var (head d)))) F.False pes))
 
-sTrans e H.Any = F.True
[…]

### 2011-05-06 draft2.tex
term translation, minor fixes

@@
 \subsection{$\etrans{}$}
-$\etrans{e}$ is a term. The translation is direct.
+$\etrans{e}$ is a term.
+
+\begin{eqnarray}
+\etrans{x} &=& x\\
+\etrans{f} &=& f\\
+\etrans{e_1~e_2} &=& app(e_1,e_2)\\
+\etrans{D} &=& D\\
+\etrans{\bad} &=& \bad\\
+\etrans{n} &=& n\\
 
@@
 && \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
-&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) 
+&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) \lor e = \bad
 \end{eqnarray}
+
+Can't we just put a XOR instead of OR in the last line?
 
@@
 \strans{e \in \any} &=& true\\
-\strans{e \in \{x \mid u \}} &=& \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
+\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
 \strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
@@
 Remark: we follow the semantics of the POPL paper but it's a bit restrictive.
-e.g. in equation 2 we could use the alternate semantics (namely B1 in the POPL paper) : 
-$$\strans{e \in \{x \mid u \}} = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$
+e.g. in equation 13 we could use the alternate semantics (namely B1 in the POPL paper) : 
+$$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$
 

### 2011-05-06 draft2.tex
Introducting data types
* Grammar updated to define data type
* D[[ ]] translation for data type, selectors...

@@
 \documentclass{article}
-% Test comment
 
@@
 \newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
+\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
 \newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
+\newcommand{\cf}[1]{\mbox{CF}(#1)}
 \newcommand{\trans}[1]{\llbracket #1 \rrbracket}
@@
 \begin{array}{rclr}
-  u,e &:=& x \mid f \mid e~e \mid D \mid \bad \mid n & (Expression)\\
-  p &:=& \Delta_1 \dots \Delta_n & (Program)\\
-  \Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract)\\
+  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
+  p &:=& \Delta,p \mid T,p \mid \epsilon & (Program)\\
+  %\Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract)\\
   d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
-  pat &:=&D~x_1 \dots x_n & (Pattern)\\
+  T &:=& \mbox{data } x~\tau_1 \dots \tau_n = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
+  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
 \end{array}
@@
 \begin{array}{rclr}
-  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
+  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D(t,\dots,t) \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
   \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
@@
 \section{Translation}
-We define several translations: $\etrans{},\dtrans{},\strans{},\trans{}$.
+We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{}$.
 \begin{center}
@@
   \dtrans{} &::& Definition \to FOF\\
+  \ttrans{} &::& Data~type  \to FOF\\
   \strans{} &::& Expression \to Contract \to FOF\\
-  \trans {} &::& Definition \to Contract \to FOF\\
+%  \trans {} &::& Definition \to Contract \to FOF\\
 \end{array}
@@
 \etrans{e_1~e_2} &=& app(e_1,e_2)\\
-\etrans{D} &=& D\\
+\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n}\\
 \etrans{\bad} &=& \bad\\
@@
 && \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
-&& \land \etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) \lor e = \bad
+&& \land (\etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) \lor e = \bad)
 \end{eqnarray}
@@
 
-\subsection{$\trans{}$}
-It's the final translation, which takes a function definition and its contract and returns a first-order formula\\
-
 \thispagestyle{empty}
 
-\begin{eqnarray}
-\trans{opaque(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
-\trans{transp(f~x_1 \dots x_n = e,c)} &=& \dtrans{f~x_1 \dots x_n = e[f_p/f]} \land \strans{f \in c} \land \strans{f_p \in c}\\
-\trans{f~x_1 \dots x_n = e} &=& \trans{opaque(f~x_1\dots x_n) = e,Ok \to \dots \to Ok)}
-\end{eqnarray}
+\section{Datatypes}
+$\ttrans{T}$ is a set of first-order formulae.
+$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.
 
-We'd like a typical contract-checking session to go like this:
-\begin{enumerate}
-\item Start with an empty theory $T$.
-\item Let $f~x_1 \dots x_n = e \in c$ be an opaque function definition to check wrt contract $c$. Check (with equinox) the consistency of the theory $T' = T \cup \trans{f~x_1 \dots x_n = e \in c}$
-\item If $T'$ is consistent then let $T = \strans{f \in c} \cup T$ and go to 2.
-with the next function definition; otherwise give a counter-example and ask the user for refinement of the contracts and/or lemmas(?)
-\end{enumerate}
+First, for each $D_i$ we introduce selectors $sel_{k,D_i}$, which are basically projections of $D_i(x_1,\dots,x_n)$ on its $k$-th composant, so that we can express that constructors are injective :
+$$S_1 := \{ \forall x_1,\dots,x_n . \bigwedge_{1 \leq i \leq n} sel_{i,D_i}(D_i(x_1,\dots,x_n)) = x_i \mid 1 \leq i \ n \}$$
 
-\section{Questions}
+For each pair of different constructors $D_i,D_j$, we state that they can never map to the same value: 
+$$S_2 := \{ \forall x_1,\dots,x_n~\forall y_1,\dots,y_k . D_i(x_1,\dots,x_n) \neq D_j(y_1,\dots,y_k) \mid 1 \leq i < j \leq n \}$$
 
-Here are some open issues, design choices and equinox-related questions.
-
-\begin{enumerate}
-\item Nested implications may lead to existential quantification, is it troublesome in equinox? (altough we don't have this case here)
-\item More generally, does equinox accept any FOF the grammar here defines?
-\item In the section $\dtrans{}$, we believe replacing $\varphi$ by $\lor \trans{f x_1 \dots x_n} = \unr$ is equivalent. What's better for equinox?
-\item Is the session stuff realistic? It looks like it can have a quadratic behaviour but maybe with the right API it's ok?
-\end{enumerate}
+Finally, we have to give crash-freeness conditions for each $D_i$:
+$$S_3 := \{ \forall x_1,\dots,x_n . (\cf{x_1} \land \dots \land \cf{x_n} \leftrightarrow \cf{D_i(x_1,\dots,x_n)}) \mid 1 \leq i \leq n \}$$
 

### 2011-05-09 draft2.tex
Parser for a toy haskell language and updates about the program translation.

@@
 
-\thispagestyle{empty}
 \section{Grammars}
@@
   u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
-  p &:=& \Delta,p \mid T,p \mid \epsilon & (Program)\\
-  %\Delta &:=& d \mid transp (d,c) \mid opaque (d,c) & (Defintion~and~contract)\\
+  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
+  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
   d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
-  T &:=& \mbox{data } x~\tau_1 \dots \tau_n = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
+  T &:=& \mbox{data } x = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
   pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
@@
 \end{center}
+
+For the moment we consider data constructors to be saturated, ie fully applied (hence the special application syntax).
 
@@
 \section{Translation}
-We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{}$.
+We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{},\trans{}$.
 \begin{center}
@@
   \strans{} &::& Expression \to Contract \to FOF\\
-%  \trans {} &::& Definition \to Contract \to FOF\\
+  \trans {} &::& Program    \to \{ FOF \}\\
 \end{array}
@@
 \subsection{$\etrans{}$}
-$\etrans{e}$ is a term.
+$\etrans{e}$ is a term, the translation is really straightforward.
 
@@
 \etrans{e_1~e_2} &=& app(e_1,e_2)\\
-\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n}\\
+\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n})\\
 \etrans{\bad} &=& \bad\\
 \etrans{n} &=& n\\
+
+\subsection{$\ttrans{}$}
+$\ttrans{T}$ is a set of first-order formulae which we break down in three parts:
+$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.
+
+First, for each $D_i$ of arity $n_i$ we introduce selectors $sel_{k,D_i}$, which are projections of $D_i(x_1,\dots,x_n_i)$ on its $k$-th composant, so that we can express that constructors are injective :
+$$S_1 := \{ \forall x_1,\dots,x_{n_i} . \bigwedge_{1 \leq j \leq n} sel_{j,D_i}(D_i(x_1,\dots,x_{n_i})) = x_j \mid 1 \leq i \leq n \}$$
+
+For each pair of different constructors $D_i,D_j$, we state that they can never map to the same value: 
+$$S_2 := \{ \forall x_1,\dots,x_{n_i}~\forall y_1,\dots,y_{n_j} . D_i(x_1,\dots,x_{n_i}) \neq D_j(y_1,\dots,y_{n_j}) \mid 1 \leq i < j \leq n \}$$
+
+Finally, we have to give crash-freeness conditions for each $D_i$:
+$$S_3 := \{ \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{D_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n \}$$
 
@@
 \begin{eqnarray}
-\dtrans{f~x_1 \dots x_n = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
-\dtrans{f~x_1 \dots x_n = \mbox{case } e \mbox{ of } [D_i~\overline{z} \mapsto e_i]} &=& \forall x_1 \dots x_n~(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i~\overline{z}} \to \etrans{f~x_1 \dots x_n} = \etrans{e_i})\\
+\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
+\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i(\overline{z})} \to \etrans{f~\overline{x}} = \etrans{e_i})\\
 && \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
-&& \land (\etrans{f~x_1 \dots x_n} = \unr \bigvee_i (\texttt{HD}(e) = D_i) \lor e = \bad)
+&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etrans{e} \neq \bad)\\
+&& ~~\to \etrans{f~x_1 \dots x_n} = \unr)
 \end{eqnarray}
 
-Can't we just put a XOR instead of OR in the last line?
+An alternative to using selectors would be to write something along the line of $\exists y_1,\dots,y_{D_i}.
+e \neq D_i(y_1,\dots,y_{D_i})$.
+It is equivalent but the skolemisation process of equinox will anyway turn those existentials in selectors functions.
+But if we pattern match $D_i$ in two different functions, we would get two different selectors (the same selector with two different names) so it's better (for the efficiency of the proof) to define selectors for each $D_i$ once in for all and use it as much as possible afterwards.
 
@@
 
-\thispagestyle{empty}
+\subsection{$\trans{}$}
+$\trans{p}$ defines the translation of a program to a theory (a set of FO formulae)
 
-\section{Datatypes}
-$\ttrans{T}$ is a set of first-order formulae.
-$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.
+\begin{eqnarray}
+\trans{\epsilon} &=& \emptyset\\
+\trans{d,p} &=& \dtrans{d} \cup \trans{p}\\
+\trans{opaque(d),p} &=& \dtrans{d} \cup \trans{p}\\
+\trans{T,p} &=& \ttrans{T} \cup \trans{p}\\
+\trans{f \in c,p} &=& \strans{f \in c} \cup \trans{p}\\
+\end{eqnarray}
 
-First, for each $D_i$ we introduce selectors $sel_{k,D_i}$, which are basically projections of $D_i(x_1,\dots,x_n)$ on its $k$-th composant, so that we can express that constructors are injective :
-$$S_1 := \{ \forall x_1,\dots,x_n . \bigwedge_{1 \leq i \leq n} sel_{i,D_i}(D_i(x_1,\dots,x_n)) = x_i \mid 1 \leq i \ n \}$$
+\section{User interaction}
 
-For each pair of different constructors $D_i,D_j$, we state that they can never map to the same value: 
-$$S_2 := \{ \forall x_1,\dots,x_n~\forall y_1,\dots,y_k . D_i(x_1,\dots,x_n) \neq D_j(y_1,\dots,y_k) \mid 1 \leq i < j \leq n \}$$
+The user provides a program $p$ which consists of functions definition (either opaque or transparent), data types definition and claims that functions satisfies contracts.
 
-Finally, we have to give crash-freeness conditions for each $D_i$:
-$$S_3 := \{ \forall x_1,\dots,x_n . (\cf{x_1} \land \dots \land \cf{x_n} \leftrightarrow \cf{D_i(x_1,\dots,x_n)}) \mid 1 \leq i \leq n \}$$
+Then, for each function definition $d_f := f~x_1 \dots x_n = e$ of a function $f$ that has to satisfy the set of contracts $C_f$, we construct the context $C = p \setminus ( d_f \cup C_f )$, which is basically the program $p$ without the definition of $f$ and the contracts it has to satisfy.
 
-\thispagestyle{empty}
+We then want to check (with equinox) that:
+ $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$
+
+Which we rewrite as:
+ $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$
+
+\section{Remarks}
+\begin{itemize}
+\item What to do with mutually recursive functions?
+\item Explore the variants translations (cf variants of the popl paper)
+\end{itemize}
+
 \end{document}

### 2011-05-09 contracts/Haskell.hs

@@ -1,9 +1,11 @@
 module Haskell ( Expression (..)
+               , DataType (..)
                , Program (..)
-               , DefCont (..)
                , Definition (..)
                , Pattern (..)
                , Contract (..)
+               , DefGeneral (..)
+               , ContSat (..)
                , apps
                , subst
                , substC
@@ -19,19 +21,24 @@ data Expression = Var Variable
                 | Con Constructor
                 | App Expression Expression
                 | BAD
-                deriving (Show,Eq)                  
+                deriving (Show,Eq)          
 
-type Program = [DefCont]
+type Program = [DefGeneral]
+data DefGeneral = Def Definition
+                | DataType DataType
+                | ContSat ContSat
+                deriving (Eq,Show)
 
-data DefCont = Def Definition
-             | Transp Definition Contract
-             | Opaque Definition Contract
+data ContSat = Satisfies Variable Contract
              deriving (Show,Eq)                  
                
 data Definition = Let Variable [Variable] Expression
                 | LetCase Variable [Variable] Expression [(Pattern,Expression)]
                 deriving (Show,Eq)                  
                   
+data DataType = Data Variable [(Variable,Int)] -- Data constructors + arity
+              deriving (Eq,Show)
+
 type Pattern = [Variable]
 
 data Contract = AppC Variable Contract Contract -- x : c -> c'

### 2011-05-09 contracts/Translation.hs

@@ -8,7 +8,8 @@ import Control.Monad.State
 eTrans :: H.Expression -> F.Term
 dTrans :: H.Definition -> F.Formula
 sTrans :: H.Expression -> H.Contract -> Fresh F.Formula
-trans  :: H.DefCont    -> Fresh F.Formula
+tTrans :: H.DataType -> [F.Formula]
+trans  :: H.DefGeneral    -> Fresh F.Formula
 
 type Fresh = State (String,Int)
 
@@ -35,35 +36,16 @@ sTrans e (H.AppC x c1 c2) = do
   f2 <- sTrans (H.App e (H.Var freshX)) c2'
   return $ F.Forall freshX (f1 `F.Implies` f2)
 
-trans (H.Def d) = trans (H.Opaque d (okFromd d))
-trans (H.Transp (H.Let f vs e) c) = do
-  (s,k) <- get
-  put (s,k+1)
-  let freshF = H.Var $ s++(show k) 
-  return $ dTrans (H.Let f vs $ H.subst e freshF f) `F.And` (evalState (sTrans (H.Var f) c) (f,0)) `F.And` (evalState (sTrans freshF c) (f++"_",0))
-trans (H.Transp (H.LetCase f vs e pes) c) = do
-  (s,k) <- get
-  put (s,k+1)
-  let freshF = H.Var $ s++(show k) 
-  return $ dTrans (H.LetCase f vs (H.subst e freshF f) pes) `F.And` (evalState (sTrans (H.Var f) c) (f,0)) `F.And` (evalState (sTrans freshF c) (f++"_",0))
-trans (H.Opaque (H.Let f vs e) c) = do
-  (s,k) <- get
-  put (s,k+1)
-  let freshF = H.Var $ s++(show k) 
-  return $ dTrans (H.Let f vs $ H.subst e freshF f) `F.And` (evalState (sTrans (H.Var f) c) (f,0)) `F.And` (evalState (sTrans freshF c) (f++"_",0))
-trans (H.Opaque (H.LetCase f vs e pes) c) = do
-  (s,k) <- get
-  put (s,k+1)
-  let freshF = H.Var $ s++(show k) 
-  return $ dTrans (H.LetCase f vs (H.subst e freshF f) pes) `F.And` (evalState (sTrans (H.Var f) c) (f,0)) `F.And` (evalState (sTrans freshF c) (f++"_",0))
+tTrans (H.Data d dns) = undefined
 
[…]

### 2011-05-09 contracts/grammar.hs

@@ -1,0 +1,533 @@
+{-# OPTIONS_GHC -w #-}
+{-# OPTIONS -cpp #-}
+module Main where
+import Data.Char
+import Haskell
+import qualified Data.Array as Happy_Data_Array
+import qualified System.IO as Happy_System_IO
+import qualified System.IO.Unsafe as Happy_System_IO_Unsafe
+import qualified Debug.Trace as Happy_Debug_Trace
+
+-- parser produced by Happy Version 1.18.6
+
+data HappyAbsSyn t4 t5 t6 t7 t8 t9
+	= HappyTerminal (Token)
+	| HappyErrorToken Int
+	| HappyAbsSyn4 t4
+	| HappyAbsSyn5 t5
+	| HappyAbsSyn6 t6
+	| HappyAbsSyn7 t7
+	| HappyAbsSyn8 t8
+	| HappyAbsSyn9 t9
+
+happyActOffsets :: Happy_Data_Array.Array Int Int
+happyActOffsets = Happy_Data_Array.listArray (0,38) ([1,1,41,38,37,40,35,33,1,-4,34,34,39,32,36,0,31,30,0,-4,27,29,0,28,28,0,0,26,25,16,-4,-3,4,24,22,0,0,0,0
+	])
+
+happyGotoOffsets :: Happy_Data_Array.Array Int Int
+happyGotoOffsets = Happy_Data_Array.listArray (0,38) ([12,23,0,0,0,0,0,0,8,21,20,19,0,18,0,0,0,0,0,15,0,17,0,14,13,0,0,0,11,0,9,6,0,6,-1,0,0,0,0
+	])
+
+happyDefActions :: Happy_Data_Array.Array Int Int
+happyDefActions = Happy_Data_Array.listArray (0,38) ([-4,0,0,0,0,0,0,-3,-4,0,-12,-12,0,-12,0,-6,0,0,-2,0,0,-10,-11,0,-5,-13,-7,0,0,0,0,0,-9,-14,-10,-15,-16,-8
+	])
+
+happyCheck :: Happy_Data_Array.Array Int Int
[…]

### 2011-05-11 contracts/FOL.hs
Ugly and the formulas for datatypes aren't working.
But we're not so far...

@@ -22,6 +22,7 @@ data Term = Var Variable
       
 data Formula = Forall Variable Formula
              | Implies Formula Formula
+             | Iff Formula Formula
              | Not Formula
              | Or Formula Formula
              | And Formula Formula 
@@ -50,6 +51,7 @@ toTPTP f = header ++ "\n" ++ (aux f) ++ "\n" ++ footer
         footer = ").\n"
         aux (Forall v f) = "![ " ++ (map toUpper v) ++ " ] : (" ++ aux f ++ ")"
         aux (Implies f1 f2) = "(" ++ aux f1 ++ ") => (" ++ aux f2 ++ ")"
+        aux (Iff f1 f2) = "(" ++ aux f1 ++ ") <=> (" ++ aux f2 ++ ")"
         aux (Not f) = "~(" ++ aux f ++ ")"
         aux (Or f1 f2) = "(" ++ aux f1 ++ ") | (" ++ aux f2 ++ ")"
         aux (And f1 f2) = "(" ++ aux f1 ++ ") & (" ++ aux f2 ++ ")"
@@ -57,9 +59,10 @@ toTPTP f = header ++ "\n" ++ (aux f) ++ "\n" ++ footer
         aux False = "$false"
         aux (Eq t1 t2) = "((" ++ auxTerm t1 ++ ") = (" ++ auxTerm t2 ++ "))"
         aux (CF t) = "(" ++ auxTerm t ++ ")"
-        auxTerm (Var v) = map toLower v
+        
+        auxTerm (Var v) = v
         auxTerm (App t1 t2) = "(" ++ auxTerm t1 ++ "(" ++ auxTerm t2 ++ "))"
-        auxTerm (Fun f) = map toLower f
+        auxTerm (Fun f) = f
         auxTerm BAD = "bad"
         auxTerm UNR = "unr"
 

### 2011-05-11 contracts/Translation.hs

@@ -1,15 +1,17 @@
-module Main where
+module Translation where
 
 import qualified Haskell as H
 import qualified FOL as F
 import Data.List (foldl')
 import Control.Monad.State
 
+import System.IO.Unsafe
+
 eTrans :: H.Expression -> F.Term
 dTrans :: H.Definition -> F.Formula
 sTrans :: H.Expression -> H.Contract -> Fresh F.Formula
 tTrans :: H.DataType -> [F.Formula]
-trans  :: H.DefGeneral    -> Fresh F.Formula
+trans  :: H.DefGeneral    -> [F.Formula]
 
 type Fresh = State (String,Int)
 
@@ -36,27 +38,24 @@ sTrans e (H.AppC x c1 c2) = do
   f2 <- sTrans (H.App e (H.Var freshX)) c2'
   return $ F.Forall freshX (f1 `F.Implies` f2)
 
-tTrans (H.Data d dns) = undefined
+b = b
 
-trans = undefined
+tTrans (H.Data d dns) = s1 ++ s2 ++ s3
+  where s1 = map (\(d,a) -> F.foralls (xs a) $ (foldl (\f x -> ((F.Var ("sel" ++ x ++ d) `F.App` (foldl (\f y -> f `F.App` F.Var y) (F.Var d) (xs a))) `F.Eq` F.Var x) `F.And` f) F.True (xs a))) dns
+        s2 = map (\((dx,ax),(dy,ay)) -> F.foralls (xs ax)  $ F.foralls (ys ay) $  F.Not $ (foldl (\ f x -> f `F.App` F.Var x)  (F.Var dx) (xs ax)) `F.Eq` (foldl (\ f x -> f `F.App` F.Var x)  (F.Var dy) (ys ay))) [(a,b) | a <- dns, b <- dns, a /= b]
+        s3 = map (\(d,a) -> F.foralls (xs a) $ cfEq d (xs a)) dns
+        xs a = map (\n -> "X"++(show n)) [1..a]
+        ys a = map (\n -> "Y"++(show n)) [1..a]
+        cfEq d xs = foldl (\f x -> F.CF (F.Var x) `F.And` f) F.True (xs) `F.Iff`
+                      (F.CF $ foldl (\f x -> f `F.App` (F.Var x)) (F.Var d)  xs)
+        
+
+trans (H.Def d) = [dTrans d]
+trans (H.DataType t) = tTrans t
[…]

### 2011-05-11 contracts/grammar.hs

@@ -1,16 +1,11 @@
 {-# OPTIONS_GHC -w #-}
-{-# OPTIONS -cpp #-}
-module Main where
+module Grammar where
 import Data.Char
 import Haskell
-import qualified Data.Array as Happy_Data_Array
-import qualified System.IO as Happy_System_IO
-import qualified System.IO.Unsafe as Happy_System_IO_Unsafe
-import qualified Debug.Trace as Happy_Debug_Trace
 
 -- parser produced by Happy Version 1.18.6
 
-data HappyAbsSyn t4 t5 t6 t7 t8 t9
+data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12
 	= HappyTerminal (Token)
 	| HappyErrorToken Int
 	| HappyAbsSyn4 t4
@@ -19,192 +14,615 @@ data HappyAbsSyn t4 t5 t6 t7 t8 t9
 	| HappyAbsSyn7 t7
 	| HappyAbsSyn8 t8
 	| HappyAbsSyn9 t9
+	| HappyAbsSyn10 t10
+	| HappyAbsSyn11 t11
+	| HappyAbsSyn12 t12
 
-happyActOffsets :: Happy_Data_Array.Array Int Int
-happyActOffsets = Happy_Data_Array.listArray (0,38) ([1,1,41,38,37,40,35,33,1,-4,34,34,39,32,36,0,31,30,0,-4,27,29,0,28,28,0,0,26,25,16,-4,-3,4,24,22,0,0,0,0
-	])
+action_0 (14) = happyShift action_3
+action_0 (21) = happyShift action_4
+action_0 (4) = happyGoto action_5
+action_0 (6) = happyGoto action_6
+action_0 _ = happyReduce_3
 
-happyGotoOffsets :: Happy_Data_Array.Array Int Int
-happyGotoOffsets = Happy_Data_Array.listArray (0,38) ([12,23,0,0,0,0,0,0,8,21,20,19,0,18,0,0,0,0,0,15,0,17,0,14,13,0,0,0,11,0,9,6,0,6,-1,0,0,0,0
-	])
+action_1 (14) = happyShift action_3
+action_1 (21) = happyShift action_4
+action_1 (6) = happyGoto action_2
+action_1 _ = happyFail
 
-happyDefActions :: Happy_Data_Array.Array Int Int
[…]

### 2011-05-11 test/list.hs

@@ -1,0 +1,6 @@
+data List = Nil 0 
+          | Cons 2;;
+
+data Int = Zero 0
+         | Succ 1;;
+    

### 2011-05-11 contracts/Translation.hs
Data constructors should translate to the good FOF now

@@ -2,29 +2,51 @@ module Translation where
 
 import qualified Haskell as H
 import qualified FOL as F
-import Data.List (foldl')
 import Control.Monad.State
 
 import System.IO.Unsafe
 
-eTrans :: H.Expression -> F.Term
-dTrans :: H.Definition -> F.Formula
-sTrans :: H.Expression -> H.Contract -> Fresh F.Formula
-tTrans :: H.DataType -> [F.Formula]
-trans  :: H.DefGeneral    -> [F.Formula]
+-- Type signatures:
+-- eTrans :: H.Expression -> F.Term
+-- dTrans :: H.Definition -> F.Formula
+-- sTrans :: H.Expression -> H.Contract -> Fresh F.Formula
+-- tTrans :: H.DataType -> Fresh [F.Formula]
+-- trans  :: H.DefGeneral -> [F.Formula]
 
 type Fresh = State (String,Int)
 
+
+
+
+
+-- Expression
+------------
+
+eTrans :: H.Expression -> F.Term
 eTrans (H.Var v) = F.Var v
 eTrans (H.Fun f) = F.Fun f
 eTrans (H.App e1 e2) = F.App (eTrans e1) (eTrans e2)
 eTrans H.BAD = F.BAD
 eTrans (H.Con d) = F.Var d
 
+
+
+-- Definition
+-------------
+
+dTrans :: H.Definition -> F.Formula
 dTrans (H.Let f vs e) = F.foralls vs $ (eTransfxi f vs) `F.Eq` (eTrans e)
 dTrans (H.LetCase f vs e pes) = F.foralls vs $ ((eTrans e `F.Eq` F.BAD) `F.Implies` (eTransfxi f vs `F.Eq` F.BAD)) `F.And` 
-                                ((eTransfxi f vs `F.Eq` F.UNR) `F.Or` (foldl' (\f (d,_) -> f `F.Or` (((F.Fun "head") `F.App` eTrans e) `F.Eq` (F.Var (head d)))) F.False pes))
+                                ((eTransfxi f vs `F.Eq` F.UNR) `F.Or` (foldl (\f (d,_) -> f `F.Or` (((F.Fun "head") `F.App` eTrans e) `F.Eq` (F.Var (head d)))) F.False pes))
 
+
+
+
+
+-- Contract satisfaction
+------------------------
+
[…]

### 2011-05-11 contracts/FOL.hs
Translations should be correct. Now plumbing with tptp

@@ -49,7 +49,7 @@ toTPTP :: Formula -> String
 toTPTP f = header ++ "\n" ++ (aux f) ++ "\n" ++ footer
   where header = "fof(axiom,axiom,"
         footer = ").\n"
-        aux (Forall v f) = "![ " ++ (map toUpper v) ++ " ] : (" ++ aux f ++ ")"
+        aux (Forall v f) = "![ " ++ (map id v) ++ " ] : (" ++ aux f ++ ")"
         aux (Implies f1 f2) = "(" ++ aux f1 ++ ") => (" ++ aux f2 ++ ")"
         aux (Iff f1 f2) = "(" ++ aux f1 ++ ") <=> (" ++ aux f2 ++ ")"
         aux (Not f) = "~(" ++ aux f ++ ")"

### 2011-05-11 contracts/Translation.hs

@@ -3,6 +3,7 @@ module Translation where
 import qualified Haskell as H
 import qualified FOL as F
 import Control.Monad.State
+import Control.Applicative
 
 import System.IO.Unsafe
 
@@ -15,10 +16,6 @@ import System.IO.Unsafe
 
 type Fresh = State (String,Int)
 
-
-
-
-
 -- Expression
 ------------
 
@@ -29,6 +26,9 @@ eTrans (H.App e1 e2) = F.App (eTrans e1) (eTrans e2)
 eTrans H.BAD = F.BAD
 eTrans (H.Con d) = F.Var d
 
+-- A little helper function for later
+eTransfxi f vs = eTrans $ H.apps (H.Fun f:map H.Var vs)
+
 
 
 -- Definition
@@ -36,9 +36,15 @@ eTrans (H.Con d) = F.Var d
 
 dTrans :: H.Definition -> F.Formula
 dTrans (H.Let f vs e) = F.foralls vs $ (eTransfxi f vs) `F.Eq` (eTrans e)
-dTrans (H.LetCase f vs e pes) = F.foralls vs $ ((eTrans e `F.Eq` F.BAD) `F.Implies` (eTransfxi f vs `F.Eq` F.BAD)) `F.And` 
-                                ((eTransfxi f vs `F.Eq` F.UNR) `F.Or` (foldl (\f (d,_) -> f `F.Or` (((F.Fun "head") `F.App` eTrans e) `F.Eq` (F.Var (head d)))) F.False pes))
-
+dTrans (H.LetCase f vs e pes) = 
+  F.foralls vs $ (F.foralls zs $ (foldl (\fo (pi,ei)-> F.And fo $ ((eTrans e) `F.Eq` (eTrans (H.apps $ H.Var (head pi) : map H.Var zs))) `F.Implies` 
+                                                                        ((eTransfxi f vs) `F.Eq` eTrans ei)) F.True pes)) `F.And` 
+  ((eTrans e `F.Eq` F.BAD) `F.Implies` (eTransfxi f vs `F.Eq` F.BAD)) `F.And`  -- Eq 10
+  (((F.Not $ eTrans e `F.Eq` F.BAD) `F.And` (foldl (\f (d,a) -> f `F.And` (F.Not $ F.Eq (eTrans e) $ F.apps (F.Var d:(sels d a)))) F.True context)) `F.Implies` -- Eq 11
[…]

### 2011-05-11 test/list.hs

@@ -3,4 +3,17 @@ data List = Nil 0
 
 data Int = Zero 0
          | Succ 1;;
+
+length x = case x of
+  | Nil -> Zero
+  | Cons a b -> Succ (length b);;
     
+zero x = case x of
+  | Nil -> True
+  | Cons a b -> False;;
+    
+f x = case x of 
+    | Nil -> Nil
+    | Cons a b -> Nil;;
+    
+f ::: a:{x : True} -> {y : zero (length y)};;

### 2011-05-11 contracts/test.hs
Utility for testing

@@ -1,0 +1,17 @@
+module Main where
+  
+import Grammar
+import Translation
+import Control.Monad
+import FOL
+import System.Environment
+
+
+fulltest f = do
+  s <- readFile f
+  forM_ (map toTPTP $ (haskell $ lexer s) >>= trans) putStrLn
+  
+main = do
+  f:_ <- getArgs
+  fulltest f
+  

### 2011-05-13 contracts/Haskell.hs
Associativity issues fixed

@@ -47,9 +47,8 @@ data Contract = AppC Variable Contract Contract -- x : c -> c'
               | Any
               deriving (Show,Eq)                  
 
-apps [] = error "Cannot apply nothing"
-apps [x] = x
-apps (x:xs) = x `App` (apps xs)
+apps xs = foldl1 (\w x -> App w x) xs
+
 
 subst :: Expression -> Expression -> Variable -> Expression -- e[x/y]
 subst (Var v) x y | v == y = x

### 2011-05-13 contracts/Parser.hs

@@ -5,7 +5,7 @@ import Haskell
 
 -- parser produced by Happy Version 1.18.6
 
-data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12
+data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12 t13
 	= HappyTerminal (Token)
 	| HappyErrorToken Int
 	| HappyAbsSyn4 t4
@@ -17,340 +17,327 @@ data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12
 	| HappyAbsSyn10 t10
 	| HappyAbsSyn11 t11
 	| HappyAbsSyn12 t12
+	| HappyAbsSyn13 t13
 
-action_0 (14) = happyShift action_3
-action_0 (22) = happyShift action_4
+action_0 (15) = happyShift action_3
+action_0 (23) = happyShift action_4
 action_0 (4) = happyGoto action_5
 action_0 (6) = happyGoto action_6
 action_0 _ = happyReduce_3
 
-action_1 (14) = happyShift action_3
-action_1 (22) = happyShift action_4
+action_1 (15) = happyShift action_3
+action_1 (23) = happyShift action_4
 action_1 (6) = happyGoto action_2
 action_1 _ = happyFail
 
-action_2 (25) = happyShift action_7
+action_2 (26) = happyShift action_7
 action_2 _ = happyFail
 
-action_3 (22) = happyShift action_11
+action_3 (23) = happyShift action_11
 action_3 _ = happyFail
 
-action_4 (17) = happyShift action_9
-action_4 (22) = happyShift action_10
+action_4 (18) = happyShift action_9
+action_4 (23) = happyShift action_10
 action_4 (10) = happyGoto action_8
 action_4 _ = happyReduce_27
 
-action_5 (30) = happyAccept
+action_5 (31) = happyAccept
 action_5 _ = happyFail
 
-action_6 (25) = happyShift action_7
+action_6 (26) = happyShift action_7
 action_6 _ = happyReduce_2
 
-action_7 (14) = happyShift action_3
-action_7 (22) = happyShift action_4
+action_7 (15) = happyShift action_3
+action_7 (23) = happyShift action_4
 action_7 (4) = happyGoto action_19
 action_7 (6) = happyGoto action_6
 action_7 _ = happyReduce_3
 
-action_8 (15) = happyShift action_18
+action_8 (16) = happyShift action_18
[…]

### 2011-05-13 contracts/Translation.hs

@@ -23,7 +23,7 @@ type Fresh = State (String,Int)
 eTrans :: H.Expression -> F.Term
 eTrans (H.Var v) = F.Var v
 eTrans (H.Fun f) = F.Fun f
-eTrans (H.App e1 e2) = F.App (eTrans e1) (eTrans e2)
+eTrans (H.App e1 e2) = F.App (eTrans e1) $ (eTrans e2)
 eTrans H.BAD = F.BAD
 eTrans (H.Con d) = F.Var d
 

### 2011-05-13 test/associativity.hs

@@ -1,0 +1,9 @@
+data List = Nil 0 
+          | Cons 2;;
+
+head x y z = case x of
+  | Nil -> BAD
+  | Cons a b -> Cons a b c;;
+    
+id x y z = x y z;;
+bl a b c = (a b) c;;

### 2011-05-13 contracts/Translation.hs
Minor bug

@@ -36,7 +36,9 @@ eTransfxi f vs = eTrans $ H.apps (H.Fun f:map H.Var vs)
 -------------
 
 dTrans :: H.Definition -> F.Formula
-dTrans (H.Let f vs e) = F.foralls vs $ (eTransfxi f vs) `F.Eq` (eTrans e)
+dTrans (H.Let f vs e) = F.foralls vs' $ (eTransfxi f vs') `F.Eq` (eTrans e')
+  where vs' = map (map toUpper) vs
+        e'  = foldl (\ e v -> H.subst e (H.Var $ map toUpper v) v) e vs
 dTrans (H.LetCase f vs e pes) = 
   F.foralls vs' $ (F.foralls zs $ (foldl (\fo (pi,ei)-> F.And fo $ ((eTrans e') `F.Eq` (eTrans (H.apps $ H.Var (head pi) : map H.Var (take (fromJust $ lookup (head pi) context) zs)))) `F.Implies` 
                                                                         ((eTransfxi f vs') `F.Eq` eTrans ei)) F.True pes')) `F.And` 

### 2011-05-13 contracts/Parser.hs
Parser fix

@@ -96,9 +96,10 @@ action_18 _ = happyFail
 
 action_19 _ = happyReduce_1
 
-action_20 _ = happyReduce_34
+action_20 _ = happyReduce_35
 
 action_21 (23) = happyShift action_23
+action_21 (29) = happyShift action_41
 action_21 (11) = happyGoto action_40
 action_21 _ = happyReduce_7
 
@@ -107,7 +108,7 @@ action_22 (23) = happyShift action_23
 action_22 (29) = happyShift action_25
 action_22 (11) = happyGoto action_20
 action_22 (12) = happyGoto action_39
-action_22 _ = happyReduce_31
+action_22 _ = happyReduce_32
 
 action_23 _ = happyReduce_28
 
@@ -144,200 +145,223 @@ action_29 _ = happyReduce_25
 
 action_30 _ = happyReduce_10
 
-action_31 (24) = happyShift action_48
+action_31 (24) = happyShift action_51
 action_31 _ = happyFail
 
-action_32 (23) = happyShift action_47
+action_32 (23) = happyShift action_50
 action_32 _ = happyFail
 
 action_33 (17) = happyShift action_22
 action_33 (23) = happyShift action_23
 action_33 (29) = happyShift action_25
 action_33 (11) = happyGoto action_20
-action_33 (12) = happyGoto action_46
+action_33 (12) = happyGoto action_49
 action_33 _ = happyFail
 
-action_34 (22) = happyShift action_45
+action_34 (22) = happyShift action_48
 action_34 _ = happyFail
 
 action_35 (20) = happyShift action_15
 action_35 (23) = happyShift action_16
 action_35 (29) = happyShift action_17
-action_35 (13) = happyGoto action_44
+action_35 (13) = happyGoto action_47
 action_35 _ = happyFail
 
 action_36 (23) = happyShift action_23
-action_36 (11) = happyGoto action_43
+action_36 (29) = happyShift action_41
+action_36 (30) = happyShift action_46
+action_36 (11) = happyGoto action_45
 action_36 _ = happyFail
 
 action_37 (17) = happyShift action_22
 action_37 (23) = happyShift action_23
 action_37 (29) = happyShift action_25
 action_37 (11) = happyGoto action_20
-action_37 (12) = happyGoto action_42
-action_37 _ = happyReduce_31
+action_37 (12) = happyGoto action_44
+action_37 _ = happyReduce_32
 
[…]

### 2011-05-13 contracts/FOL.hs
Associativity bug fixed (once again)

@@ -34,9 +34,8 @@ data Formula = Forall Variable Formula
 foralls [] f = f
 foralls (x:xs) f = Forall x (foralls xs f)
 
-apps [] = error "Cannot apply nothing"
-apps [x] = x
-apps (x:xs) = App x $ apps xs
+apps xs = foldl1 (\w x -> App w x) xs
+
 
 splitOnAnd :: Formula -> [Formula]
 splitOnAnd (Forall x (And f1 f2)) = splitOnAnd (Forall x f1) ++ splitOnAnd (Forall x f2)

### 2011-05-13 contracts/Translation.hs
Sel bu

@@ -97,7 +97,7 @@ s1D (d,a) = do
   (s,k) <- get
   put (s,k+1)
   let xs = map (\n -> s++"_"++(show n)) [1..a]
-  return $ F.foralls xs $ foldl (\f (x,k) -> f `F.And` (F.Eq (F.Var x) $ F.apps $ F.Var ("sel_"++(show k)++"_"++d) : F.Var d : map F.Var xs)) F.True (zip xs [1..a])
+  return $ F.foralls xs $ foldl (\f (x,k) -> f `F.And` (F.Eq (F.Var x) $ F.App (F.Var ("sel_"++(show k)++"_"++d)) (F.apps $ F.Var d : map F.Var xs))) F.True (zip xs [1..a])
 
 
 s2 :: H.DataType -> Fresh [F.Formula]

### 2011-05-13 draft2.tex
Ugly and the formulas for datatypes aren't working.
But we're not so far...
tex update & parser fix

@@
   \dtrans{} &::& Definition \to FOF\\
-  \ttrans{} &::& Data~type  \to FOF\\
+  \ttrans{} &::& Data~type  \to \{ FOF \}\\
   \strans{} &::& Expression \to Contract \to FOF\\
@@
 \subsection{$\ttrans{}$}
-$\ttrans{T}$ is a set of first-order formulae which we break down in three parts:
-$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3$.
+$\ttrans{T}$ is a set of first-order formulae which we break down in four parts:
+$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3 \cup S_4$.
 
@@
 
-Finally, we have to give crash-freeness conditions for each $D_i$:
+Then, we have to give crash-freeness conditions for each $D_i$:
 $$S_3 := \{ \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{D_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n \}$$
+
+Finally, we have to say that none of the $D_i$ is unreachable:
+$$S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$$
 
@@
 \begin{eqnarray}
-\trans{\epsilon} &=& \emptyset\\
+\trans{\epsilon} &=& \{ \cf{\unr}, \lnot \cf{\bad}, \forall f,x~\cf{f} \land \cf{x} \implies \cf{app(f,x)} \}\\
 \trans{d,p} &=& \dtrans{d} \cup \trans{p}\\

### 2011-05-13 contracts/Parser.hs

@@ -596,7 +596,7 @@ happyReduction_27  =  HappyAbsSyn10
 happyReduce_28 = happySpecReduce_1  11 happyReduction_28
 happyReduction_28 (HappyTerminal (TokenVar happy_var_1))
 	 =  HappyAbsSyn11
-		 (if isUpper $ head happy_var_1 then Con (map toLower happy_var_1) else Var happy_var_1
+		 (if isUpper $ head happy_var_1 then Con (map toLower happy_var_1) else Var (map toLower happy_var_1)
 	)
 happyReduction_28 _  = notHappyAtAll 
 

### 2011-05-13 contracts/Translation.hs

@@ -85,8 +85,8 @@ sTrans e (H.AppC x c1 c2) = do
 --------------------
 
 tTrans :: H.DataType -> Fresh [F.Formula]
-tTrans d = liftM3 (+++) (s1 d) (s2 d) (s3 d)
-  where (+++) a b c = a ++ b ++ c
+tTrans d = liftM4 (++++) (s1 d) (s2 d) (s3 d) (return $ s4 d)
+  where (++++) a b c d = a ++ b ++ c ++ d
 
 s1 :: H.DataType -> Fresh [F.Formula]
 s1 (H.Data _ dns) = sequence $ map s1D dns
@@ -124,7 +124,8 @@ s3D (d,a) = do
   let xs = map (\n -> s++(show k)++"_"++(show n)) [1..a]
   return $ F.foralls xs $ F.Iff (F.CF $ F.apps $ F.Var d : map F.Var xs) (foldl (\f x -> f `F.And` F.CF (F.Var x)) F.True xs)
 
-
+s4 :: H.DataType -> [F.Formula]
+s4 (H.Data _ dns) = map (\(d,a) -> F.Not (F.Var d `F.Eq` F.UNR)) dns
 
 
 
@@ -135,7 +136,7 @@ s3D (d,a) = do
 trans  :: H.DefGeneral -> [F.Formula]
 trans (H.Def d) = [dTrans d]
 trans (H.DataType t) = evalState (tTrans t) ("Dtype",0)
-trans (H.ContSat (H.Satisfies v c)) = map F.Not [evalState (sTrans (H.Var v) c) ("Zcont",0)] ++ [(F.Forall "F" $ F.Forall "X" $ (F.And (F.CF $ F.Var "X") (F.CF $ F.Var "F")) `F.Implies` (F.CF $ (F.App (F.Var "F") (F.Var "X")))),F.Not $ F.CF $ F.BAD]
+trans (H.ContSat (H.Satisfies v c)) = map F.Not [evalState (sTrans (H.Var v) c) ("Zcont",0)] ++ [(F.Forall "F" $ F.Forall "X" $ (F.And (F.CF $ F.Var "X") (F.CF $ F.Var "F")) `F.Implies` (F.CF $ (F.App (F.Var "F") (F.Var "X")))),F.Not $ F.CF F.BAD,F.CF F.UNR]
 
 okFromd :: H.Definition -> H.Contract
 okFromd (H.Let _ vs _) = foldl (\c _ -> H.AppC "dummy" c H.ok) H.ok vs

### 2011-05-19 contracts/Translation.hs
Working version

@@ -20,12 +20,12 @@ type Fresh = State (String,Int)
 -- Expression
 -------------
 
-eTrans :: H.Expression -> F.Term
-eTrans (H.Var v) = F.Var v
-eTrans (H.Fun f) = F.Fun f
-eTrans (H.App e1 e2) = F.App (eTrans e1) $ (eTrans e2)
-eTrans H.BAD = F.BAD
-eTrans (H.Con d) = F.Var d
+eTrans :: H.Expression -> F.Term F.Variable
+eTrans (H.Var v) = F.Var $ F.Regular v
+eTrans (H.Fun f) = F.Var $ F.Regular f
+eTrans (H.App e1 e2) = F.App [eTrans e1,eTrans e2] -- TODO modifier H.App
+eTrans H.BAD = F.Var $ F.BAD
+eTrans (H.Con d) = F.Var $ F.Regular d
 
 -- A little helper function for later
 eTransfxi f vs = eTrans $ H.apps (H.Fun f:map H.Var vs)
@@ -35,36 +35,34 @@ eTransfxi f vs = eTrans $ H.apps (H.Fun f:map H.Var vs)
 -- Definition
 -------------
 
-dTrans :: H.Definition -> F.Formula
-dTrans (H.Let f vs e) = F.foralls vs' $ (eTransfxi f vs') `F.Eq` (eTrans e')
-  where vs' = map (map toUpper) vs
-        e'  = foldl (\ e v -> H.subst e (H.Var $ map toUpper v) v) e vs
+dTrans :: H.Definition -> F.Formula (F.Term F.Variable)
+dTrans (H.Let f vs e) = F.Forall vvs $ F.Eq (F.App (eTrans (H.Var f):vvs)) (eTrans e)
+  where vvs = map (F.Var . F.Regular) vs
 dTrans (H.LetCase f vs e pes) = 
-  F.foralls vs' $ (F.foralls zs $ (foldl (\fo (pi,ei)-> F.And fo $ ((eTrans e') `F.Eq` (eTrans (H.apps $ H.Var (head pi) : map H.Var (take (fromJust $ lookup (head pi) context) zs)))) `F.Implies` 
-                                                                        ((eTransfxi f vs') `F.Eq` eTrans ei)) F.True pes')) `F.And` 
[…]

### 2011-05-19 draft2.tex
Clarification

@@
 \end{center}
-CF$(t)$ holds iff $t$ satisfies $\ok$.
+We give the following equations to define crash-freeness:
+$$\cf{\unr}, \lnot \cf{\bad}, \forall f,x~\cf{f} \land \cf{x} \implies \cf{app(f,x)}$$
+$$\forall D, \forall t_i~1 \leq i \leq n, \bigwedge_{1 \leq i \leq n} \cf{t_i} \implies \cf{D(t_1,\dots,t_n)}$$
+It does not follow exactly the semantics given by the popl paper
 
@@
 \etrans{n} &=& n\\
+As usual, we use left associativity.
 
@@
 
+We have to give some semantics to $\cf{}$, $\bad$ and $\unr$, which are given below.
+
 \begin{eqnarray}

### 2011-05-25 draft2.tex
Paper update

--- 2011-05-13 draft2.tex
@@
 
+In the calculus, every definition is toplevel ($\lambda$-lifted) and there isn't any nested case.
+
 For the moment we consider data constructors to be saturated, ie fully applied (hence the special application syntax).
+That makes defining $\dtrans{}$ less painful, because we avoid a quadratic number (in the arity of the data contrustors) of axioms. cf to $S_4$ below, for example.
 
@@
 \end{center}
-CF$(t)$ holds iff $t$ satisfies $\ok$.
+We always give the following equations to define crash-freeness:
+$$\cf{\unr}, \lnot \cf{\bad}, \forall f,x~\cf{f} \land \cf{x} \implies \cf{app(f,x)}$$
+Note that it's only the equation we \textit{always} give, but there some other equations that will arise from the translations below.
+It does not follow exactly the semantics given by the popl paper, but we hope it'll be a sufficient approximation!
 
@@
 
-Semantics of contract satisfaction:
+Semantics of contract satisfaction (exactly the same as in the POPL paper):
 \begin{center}
@@
 Finally, we have to say that none of the $D_i$ is unreachable:
-$$S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$$
+$$S_4 := \{ D_i(x_1,\dots,x_n) = \unr \to x_1 = \unr \lor \dots \lor x_n = \unr \mid 1 \leq i \leq n \}$$
+
+NB: We first started with $S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$, but we couldn't prove that the (add x y), defined with peano integers, satisfied ($x=0 \to y=0 \to add x y=0$).
+It always also bizarre to state that the function was $\unr$ but not its application.
 
@@
 
+Equation (10) is not restrictive even in the case of laziness because $e$ is always evaluated. (TODO, but not always fully evaluated...)
+
+Equation (11) and (12) state that if we have a type mismatch than the code is $\unr$. It should never occur because we're assuming that the code has already been typechecked.
+
 \subsection{$\strans{}$}
@@
 Remark: we follow the semantics of the POPL paper but it's a bit restrictive.
-e.g. in equation 13 we could use the alternate semantics (namely B1 in the POPL paper) : 
+e.g. in equation 14 we could use the alternate semantics (namely B1 in the POPL paper) : 
 $$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$
@@
 $\trans{p}$ defines the translation of a program to a theory (a set of FO formulae)
+
+We have to give some semantics to $\cf{}$, $\bad$ and $\unr$, which are given below.
 
@@
 
-\section{Remarks}
-\begin{itemize}
-\item What to do with mutually recursive functions?
-\item Explore the variants translations (cf variants of the popl paper)
-\end{itemize}
+\section{Example}
+Consider the following program (the integer after a data constructor is its arity):
+\begin{verbatim}
+data List = Nil 0
+          | Cons 2;;
+
+data Bool = True 0
+          | False 0;;
+
+copy x = case x of
+  | Nil -> x
+  | Cons a b -> Cons a (copy b);;
+
+copy ::: a:{x : True} -> {y:True};;
+\end{verbatim}
+The contract states that if we give to copy a crash-free argument $x$, copy will return a crash-free result $y$.
+Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it. TODO (or can we?)
+
+\begin{verbatim}
+data Nat = Zero 0
+         | Succ 1;;
+
+data Bool = True 0
+          | False 0;;
+
+add x y = case x of
+        | Zero -> y
+        | Succ z -> Succ (add z y);;
+
+isZero x = case x of
+         | Zero -> True
+         | Succ z -> False;;
+
+not x = case x of
+      | True -> False
+      | False -> True;;
+
+add ::: a:{x : True} -> {y:True} -> {z: True};;             -- C1
+add ::: a:{x : isZero x} -> {y:isZero y} -> {z: True};;     -- C2
+add ::: a:{x : isZero x} -> {y:isZero y} -> {z: isZero z};; -- C3
+\end{verbatim}
+
+As of today, we can prove C1 and C2 but not C3.
+Note that we can prove C2 only since the modification of $S_4$ in $\dtrans{}$.
+
+In the definition of $add$, if we replace $Succ (add z y)$ by $Succ (add y z)$, we can't prove anything which isn't very surprising because we're to do the recursion on the first variable, not on the second one.
+All the more so that they don't have the same contract.
 

### 2011-05-25 contracts/Translation.hs
UNR update (S4)

@@ -43,8 +43,8 @@ dTrans (H.LetCase f vs e pes) =
   where vvs = map (F.Var . F.Regular) vs
         arities = map (\p -> (head p, length $ tail p)) $ map fst pes :: [(String,Int)]
         zs = [F.Var $ F.Regular $ "Zdef" ++ show x | x <- [1..(foldl1 max [snd y | y <- arities])]]
-        eq9 = [(eTrans e `F.Eq` (F.App ((F.Var $ F.Regular $ head pi):(take (length pi - 1) [ z | (v,z) <- zip (tail pi) zs ])))) `F.Implies` (F.App (eTrans (H.Var f):vvs) `F.Eq` (eTrans (zedify ei pi))) | (pi,ei) <- pes]
-        eq10 = (eTrans e `F.Eq` (F.Var F.BAD)) `F.Implies` F.Not (F.App (eTrans (H.Var f):vvs) `F.Eq` F.Var F.BAD)
+        eq9 = [(eTrans e `F.Eq` (F.App ((F.Var $ F.Regular $ head pi):(take (length pi - 1) [ z | (v,z) <- zip (tail pi) zs ])))) `F.Implies` (F.App (eTrans (H.Var f):vvs) `F.Eq` (F.Weak $ eTrans (zedify ei pi))) | (pi,ei) <- pes]
+        eq10 = (eTrans e `F.Eq` (F.Var F.BAD)) `F.Implies` (F.App (eTrans (H.Var f):vvs) `F.Eq` F.Var F.BAD)
         eq11 = (F.And $ (F.Not $ eTrans e `F.Eq` F.Var F.BAD):bigAndSel ) `F.Implies` eq12
         eq12 = (F.App (eTrans (H.Var f):vvs) `F.Eq` F.Var F.UNR)
         bigAndSel = [F.Not $ eTrans e `F.Eq` F.Weak (F.App ((F.Var (F.Regular di)):[F.App [(F.Var . F.Regular) ("sel_"++(show i)++"_"++di),eTrans e] | i <- [1..ai]])) | (di,ai) <- arities]
@@ -61,7 +61,7 @@ test = (H.LetCase "head" ["xyz"] (H.Var "xyz") [(["nil"],H.BAD),(["cons","a","b"
 sTrans :: H.Expression -> H.Contract -> Fresh (F.Formula (F.Term F.Variable))
 sTrans e H.Any = return F.True
 
[…]

### 2011-05-25 draft2.tex
Paper update

@@
 For the moment we consider data constructors to be saturated, ie fully applied (hence the special application syntax).
-That makes defining $\dtrans{}$ less painful, because we avoid a quadratic number (in the arity of the data contrustors) of axioms. cf to $S_4$ below, for example.
+That makes defining $\dtrans{}$ less painful, because we avoid a quadratic number (in the arity of the data contrustors) of axioms. cf $S_4$ below, for example.
 
@@
 \etrans{n} &=& n\\
+\end{eqnarray}
+
 As usual, we use left associativity.
@@
 
-First, for each $D_i$ of arity $n_i$ we introduce selectors $sel_{k,D_i}$, which are projections of $D_i(x_1,\dots,x_n_i)$ on its $k$-th composant, so that we can express that constructors are injective :
+First, for each $D_i$ of arity $n_i$ we introduce selectors $sel_{k,D_i}$, which are projections of $D_i(x_1,\dots,x_n_i)$ on its $k$-th component:
 $$S_1 := \{ \forall x_1,\dots,x_{n_i} . \bigwedge_{1 \leq j \leq n} sel_{j,D_i}(D_i(x_1,\dots,x_{n_i})) = x_j \mid 1 \leq i \leq n \}$$
@@
 $$S_3 := \{ \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{D_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n \}$$
+Note that we have an equivalence in the case of a data constructor whereas we only have an (direct) implication for functions.
+It's because a function, due to laziness, could not use all its arguments.
+For example, $fst (1,BAD)$ is crash-free but its arguments aren't.
 
@@
 $$S_4 := \{ D_i(x_1,\dots,x_n) = \unr \to x_1 = \unr \lor \dots \lor x_n = \unr \mid 1 \leq i \leq n \}$$
-
-NB: We first started with $S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$, but we couldn't prove that the (add x y), defined with peano integers, satisfied ($x=0 \to y=0 \to add x y=0$).
-It always also bizarre to state that the function was $\unr$ but not its application.
+NB: We first started with $S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$, but we couldn't prove that the (add x y), defined with peano integers, satisfied ($x=0 \to y=0 \to add~x~y=0$).
+It's also bizarre to state that the function was $\unr$ but not its application.
 

### 2011-06-08 draft2.tex
PDF Update

@@
 \section{Example}
+
+\subsection{Semantics}
 Consider the following program (the integer after a data constructor is its arity):
@@
 The contract states that if we give to copy a crash-free argument $x$, copy will return a crash-free result $y$.
-Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it. TODO (or can we?)
+Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it. TODO
 
+\subsection{Arguments order}
 \begin{verbatim}
@@
 
-add ::: a:{x : True} -> {y:True} -> {z: True};;             -- C1
-add ::: a:{x : isZero x} -> {y:isZero y} -> {z: True};;     -- C2
-add ::: a:{x : isZero x} -> {y:isZero y} -> {z: isZero z};; -- C3
+add ::: a:{x : isZero x} -> {y:isZero y} -> {z: isZero z};;
 \end{verbatim}
-
-As of today, we can prove C1 and C2 but not C3.
-Note that we can prove C2 only since the modification of $S_4$ in $\dtrans{}$.
 
@@
 
+A lemma stating (+) commutativity would be helpful here.
+
+\subsection{Multiple contracts - Take 1}
+We suppose defined datatypes for Nat and Bool, just as in the example above.
+
+\begin{verbatim}
+add a b = case a of
+  | Zero -> b
+  | Succ x -> Succ (add x b);;
+
+mult a b = case a of
+  | Zero -> Zero
+  | Succ x -> add b (mult x b);;
+
+notZero x = case x of
+  | Zero -> False
+  | Succ a -> True;;
+
+notZero ::: a:{x:True} -> {y: True};;
+add  ::: a:{x: True} -> b:{y: True} -> {z: True};;
+add  ::: a:{x: notZero x} -> b:{y: True} -> {z: True};;
+mult ::: a:{x: notZero x} -> b:{y: notZero y} -> {z: True};;
+\end{verbatim}
+
+Equinox diverges when we try to check mult's contract.
+
+The reason stems from that $mult~(Succ~x)~b$ calls recursively
+$mult~x~b$ The induction hypothesis says that if $a$ is nonzero and
+$b$ is nonzero then the result is $\ok$. But in this case $x$ may be
+zero, so the induction hypothesis does not apply immediately. Now, the
+theorem prover could do a split on $x$ (either Zero or Succ$\_$) and
+in one case the result should follow from the induction hypothesis,
+but in the other case we are stuck! Because we don't actually know
+that $multp~0~b$ is $\ok$ ($multp$ is the recursive version of the
+function, whose contract we assumed but no further defining
+equations!)
+
+A possible fix, which is only possible when the function is total is
+to write $$f ::: a_1:\{x \mid p_1(x) \} \to \cdots \to a_n:\{x \mid
+p_n(x) \}$$ as $$f ::: \ok \to \cdots \to \ok \to \{x \mid
+if~(\bidWedge_{i < n} p_i(a_i))~then~p_n(x)~else~True)$$ And know we
+can prove $mult$'s contract, but it's not very satisfying.
+
 \end{document}

### 2011-06-09 draft2.tex
pdf update

@@
 \newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
+\newcommand{\etranst}[1]{\mathcal{E}_t \llbracket #1 \rrbracket}
+\newcommand{\etransf}[1]{\mathcal{E}_{f} \llbracket #1 \rrbracket}
 \newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
@@
 
-\subsection{$\lambda$-lifted haskell subset}
+\subsection{$\mathcal{H}$: Haskell subset}
+Our source language $\mathcal{H}$ is defined below. Note that $\lambda$ abstractions and case expressions can appear anywhere in the code.
+
+assert $(e,c)$ in $e'$ means that if $e$ does not satisfy the contract $c$ then the compiler should return $\BAD$ and otherwise execute $e'$. % TODO return bad? Not really...
+
+$e$ `satisfies` $c$ is a boolean expression which is True iff the expression $e$ satisfies the contract $c$ in the current context. % TODO we talk about cotnract satisfaction in the first paragraph!
+
 \begin{center}
 \begin{array}{rclr}
-  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
+  u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
+  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] &\\
   p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
   \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
-  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
-  T &:=& \mbox{data } x = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
+  d &:=& f~x_1 \dots x_n ::: c = e \mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
+  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
+  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
   pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
@@
 
-In the calculus, every definition is toplevel ($\lambda$-lifted) and there isn't any nested case.
+\subsection{$\mathcal{H'}$: $\lambda$-lifted haskell subset}
+This language is as expressive as $\mathcal{H}$ but contains fewer constructs. Compared to $\mathcal{H}$ the restrictions are:
+\begin{itemize}
+  \item No $\lambda$, ie every $\lambda$ is lifted to a function definition at the toplevel
+  \item No case expressions which are lifted out to the toplevel (ie pattern matching is only possible for a function declaration)
+  \item No $assert$ which are syntactic sugar for case expressions
+\end{itemize} % TODO describe preprocessing.
+From now on, expressions will always refer to $\mathcal{H'}$ expressions.
+
+\begin{center}
+\begin{array}{rclr}
+  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c & (Expression)\\
+  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
+  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
+  d &:=& f~x_1 \dots x_n ::: c = e\mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
+  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
+  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
+  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
+\end{array}
+\end{center}
 
@@
 That makes defining $\dtrans{}$ less painful, because we avoid a quadratic number (in the arity of the data contrustors) of axioms. cf $S_4$ below, for example.
+
+\subsection{From $\mathcal{H}$ to $\mathcal{H'}$}
+\subsubsection{Step 1: $assert$}
+We translate every expression $e = \mbox{assert } (e_c,c) \mbox{ in } e'$ in $\mbox{case } e_c `satisfies` c \mbox{ of } True -> e' | False -> \bad$.
+\subsubsection{Step 2: $\lambda-lifting$}
+GHC's API can do that.
+\subsubsection{Step 3: $case$ lifting}
+...
 
@@
 \begin{array}{rclr}
-  \etrans{} &::& Expression \to Term\\
-  \dtrans{} &::& Definition \to FOF\\
+  \etrans{} &::& Expression \to (Term,FOF)\\
+  \dtrans{} &::& Definition \to \{ FOF \}\\
   \ttrans{} &::& Data~type  \to \{ FOF \}\\
-  \strans{} &::& Expression \to Contract \to FOF\\
+  \strans{} &::& Expression \to Contract \to \{ FOF \}\\
   \trans {} &::& Program    \to \{ FOF \}\\
@@
 \subsection{$\etrans{}$}
-$\etrans{e}$ is a term, the translation is really straightforward.
+$\etrans{e}$ is a pair (term,formula), the only tricky translations is
+for satisfies.  We write $\etranst{}$ and $\etransf{}$ the projection
+of $\etrans{}$ on the first and second component.
 
 \begin{eqnarray}
-\etrans{x} &=& x\\
-\etrans{f} &=& f\\
-\etrans{e_1~e_2} &=& app(e_1,e_2)\\
-\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n})\\
-\etrans{\bad} &=& \bad\\
-\etrans{n} &=& n\\
+\etrans{x} &=& (x,\emptyset)\\
+\etrans{f} &=& (f,\emptyset)\\
+\etrans{e_1~e_2} &=& (app(\etranst{e_1},\etranst{e_2}),\etransf{e_1} \cup \etransf{e_2})\\
+\etrans{D(e_1,\dots,e_n)} &=& (D(\etranst{e_1},\dots,\etranst{e_n}),\cup_{1 \leq i \leq n} \etransf{e_i})\\
+\etrans{\bad} &=& (\bad,\emptyset)\\
+\etrans{n} &=& (n,\emptyset)\\
+\etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
 \end{eqnarray}
 
+TODO: take car of the FV in c for satisfies.
 As usual, we use left associativity.
@@
 \subsection{$\ttrans{}$}
-$\ttrans{T}$ is a set of first-order formulae which we break down in four parts:
-$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3 \cup S_4$.
+$\ttrans{T}$ is a set of first-order formulae which we break down in five parts:
+$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3 \cup S_4 \cup S_5$.
 
@@
 
-Finally, we have to say that none of the $D_i$ is unreachable:
+We also have to say that none of the $D_i$ is unreachable:
 $$S_4 := \{ D_i(x_1,\dots,x_n) = \unr \to x_1 = \unr \lor \dots \lor x_n = \unr \mid 1 \leq i \leq n \}$$
@@
 
+Finally, we give contracts to each data constructors.
+It is the contract $c$ specified by the user.
+If no contract is given for a specific data constructor, then we assume that the contract is $\ok \to \dots \to \ok$.
+We write the contract $c$ as $c_1 \to \dots \to c_{n_i}}$
+$$S_5 := \{ \forall x_1,\dots,x_{n_i} \strans{D_i(x_1,\dots,x_{n_i}) \in c} \leftrightarrow \bigwedge_{1 \leq i \leq n_i} \strans{x_i \in c_i} \mid 1 \leq i \leq n \}$$
+
 \subsection{$\dtrans{}$}
-$\dtrans{d}$ is a first-order formula.
+$\dtrans{d}$ is a set first-order formula.
 
 \begin{eqnarray}
-\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
-\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i(\overline{z})} \to \etrans{f~\overline{x}} = \etrans{e_i})\\
-&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
-&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etrans{e} \neq \bad)\\
-&& ~~\to \etrans{f~x_1 \dots x_n} = \unr)
+\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etranst{f~x_1 \dots x_n} = \etranst{e} \cup \etransf{e}\\
+\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etranst{e} = \etranst{D_i(\overline{z})} \to \etranst{f~\overline{x}} = \etranst{e_i})\\
+&& \land \etranst{e} = \bad \to \etranst{f~x_1 \dots x_n} = \bad)\\ 
+&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etranst{e} \neq \bad)\\
+&& ~~\to \etranst{f~x_1 \dots x_n} = \unr)\\
+&& \cup \etransf{e}
 \end{eqnarray}
+
+Note that in those equations, there is no $\etransf{f~x_1 \dots x_n}$ because it is always the empty set.
 
@@
 \subsection{$\strans{}$}
-$\strans{e \in c}$ is a first-order formula.
+$\strans{e \in c}$ is a set of first-order formulae.
 \begin{eqnarray}
 \strans{e \in \any} &=& true\\
-\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
+\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
+&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
 \strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
@@
 e.g. in equation 14 we could use the alternate semantics (namely B1 in the POPL paper) : 
-$$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$
+$$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False) \cup \etransf{e} \cup \etransf{u[e/x]}$$
 
@@
 We then want to check (with equinox) that:
- $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$
+ $$\trans{C} \cup \{ \dtrans{f~x_1 \dots x_n = e[f^\star/f]}, \dtrans{f^\star~x_1 \dots x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$
 
 Which we rewrite as:
- $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$
+ $$\trans{C} \cup \{ \dtrans{f~x_1 \dots x_n = e[f^\star/f]}, \dtrans{f^\star~x_1 \dots x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$
 

### 2011-06-09 contracts/Haskell.hs
Code update, no sematics change (just types)

@@ -9,7 +9,8 @@ module Haskell ( Expression (..)
                , apps
                , subst
                , substC
-               , ok)
+               , ok
+               , okContract)
 where
 
 type Variable = String
@@ -36,7 +37,7 @@ data Definition = Let Variable [Variable] Expression
                 | LetCase Variable [Variable] Expression [(Pattern,Expression)]
                 deriving (Show,Eq,Ord)                  
                   
-data DataType = Data Variable [(Variable,Int)] -- Data constructors + arity
+data DataType = Data Variable [(Variable,Int,Contract)] -- Data constructors + arity + contract
               deriving (Eq,Show,Ord)
 
 type Pattern = [Variable]
@@ -68,3 +69,6 @@ substC Any _ _ = Any
 
 ok :: Contract
 ok = Pred "dummy" (Con "True")
+
+okContract 0 = ok
+okContract n = AppC "okDummy" (okContract $ n-1) ok

### 2011-06-09 contracts/Parser.hs

@@ -39,7 +39,7 @@ action_3 _ = happyFail
 action_4 (18) = happyShift action_9
 action_4 (23) = happyShift action_10
 action_4 (10) = happyGoto action_8
-action_4 _ = happyReduce_27
+action_4 _ = happyReduce_31
 
 action_5 (31) = happyAccept
 action_5 _ = happyFail
@@ -64,16 +64,16 @@ action_9 _ = happyFail
 
 action_10 (23) = happyShift action_10
 action_10 (10) = happyGoto action_13
-action_10 _ = happyReduce_27
+action_10 _ = happyReduce_31
 
 action_11 (23) = happyShift action_10
 action_11 (10) = happyGoto action_12
-action_11 _ = happyReduce_27
+action_11 _ = happyReduce_31
 
 action_12 (16) = happyShift action_29
 action_12 _ = happyFail
 
-action_13 _ = happyReduce_26
+action_13 _ = happyReduce_30
 
 action_14 _ = happyReduce_9
 
@@ -96,7 +96,7 @@ action_18 _ = happyFail
 
 action_19 _ = happyReduce_1
 
-action_20 _ = happyReduce_35
+action_20 _ = happyReduce_39
 
 action_21 (23) = happyShift action_23
 action_21 (29) = happyShift action_41
@@ -108,9 +108,9 @@ action_22 (23) = happyShift action_23
 action_22 (29) = happyShift action_25
 action_22 (11) = happyGoto action_20
 action_22 (12) = happyGoto action_39
-action_22 _ = happyReduce_32
+action_22 _ = happyReduce_36
 
-action_23 _ = happyReduce_28
+action_23 _ = happyReduce_32
 
 action_24 (17) = happyShift action_22
 action_24 (23) = happyShift action_23
@@ -141,7 +141,7 @@ action_28 _ = happyFail
 action_29 (23) = happyShift action_31
 action_29 (29) = happyShift action_32
 action_29 (9) = happyGoto action_30
-action_29 _ = happyReduce_25
+action_29 _ = happyReduce_29
 
 action_30 _ = happyReduce_10
 
@@ -178,7 +178,7 @@ action_37 (23) = happyShift action_23
 action_37 (29) = happyShift action_25
 action_37 (11) = happyGoto action_20
 action_37 (12) = happyGoto action_44
-action_37 _ = happyReduce_32
+action_37 _ = happyReduce_36
 
 action_38 (23) = happyShift action_23
[…]

### 2011-06-09 contracts/Translation.hs

@@ -89,8 +89,8 @@ s1 :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
 s1 (H.Data _ dns) = sequence $ map s1D dns
 
 -- It's the set S1 but for only one data constructor
-s1D :: (String,Int) -> Fresh (F.Formula (F.Term F.Variable))
-s1D (d,a) = do
+s1D :: (String,Int,H.Contract) -> Fresh (F.Formula (F.Term F.Variable))
+s1D (d,a,c) = do
   (s,k) <- get
   put (s,k+1)
   let xs = map (\n -> s++"_"++(show n)) [1..a]
@@ -101,8 +101,8 @@ s2 :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
 s2 (H.Data _ dns) = sequence $ map s2D [(a,b) | a <- dns, b <- dns, a < b]
 
 -- It's S2 for a pair of data constructors.
-s2D :: ((String,Int),(String,Int)) -> Fresh (F.Formula (F.Term F.Variable))
-s2D ((d1,a1),(d2,a2)) = do
+s2D :: ((String,Int,H.Contract),(String,Int,H.Contract)) -> Fresh (F.Formula (F.Term F.Variable))
+s2D ((d1,a1,c1),(d2,a2,c2)) = do
   (s,k) <- get
   put (s,k+2)
   let xs1 = map (\n -> s++(show k)++"_"++(show n)) [1..a1]
@@ -114,8 +114,8 @@ s3 :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
 s3 (H.Data _ dns) = sequence $ map s3D dns
 
 -- It's S3 but only for one data constructor
-s3D :: (String,Int) -> Fresh (F.Formula (F.Term F.Variable))
-s3D (d,a) = do
+s3D :: (String,Int,H.Contract) -> Fresh (F.Formula (F.Term F.Variable))
+s3D (d,a,c) = do
   (s,k) <- get
   put (s,k+1)
   let xs = map (\n -> s++(show k)++"_"++(show n)) [1..a]
@@ -127,8 +127,8 @@ s4 :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
 --s4 (H.Data _ dns) = map (\(d,a) -> F.Not ((F.Var . F.Regular) d `F.Eq` F.Var F.UNR)) dns
[…]

### 2011-06-17 talk.tex
Talk final version

\documentclass{beamer}
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{graphicx}

\usetheme{Warsaw}

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\etranst}[1]{\mathcal{E}_t \llbracket #1 \rrbracket}
\newcommand{\etransf}[1]{\mathcal{E}_{f} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
\newcommand{\cf}[1]{\mbox{CF}(#1)}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}

\title{Contracts for Haskell}   
\author{Simon (PJ), Dimitrios, Koen, Charles-Pierre} 
\date{}

\begin{document}

\frame{\titlepage}

\section{Goals}  
\frame{
\frametitle{Goals}
We would like to write things like:\\
\ \\
\texttt{head ::: \{ x | not (null x) \} -> \ok\\
  head xs = case xs of\\
  | [] -> \bad \\
  | x:\_ -> x
}\\
\ \\ \ \\

\texttt{map ::: (Ok -> Ok) -> xs:Ok ->\\
  \quad \quad \quad \quad \quad \quad \{ ys | length ys = length xs \} }
\ \\ \ \\
Note: $\ok$ is just syntactic sugar for $\{ x \mid True \}$
}

\frame{
  \frametitle{Contract syntax}
  \begin{center}
  \begin{array}{rclr}
        c &:=& \{ x \mid e \}\\
        &\mid& \any \\
        &\mid& (c,c) \\
        &\mid& x{:}c_1 \to c_2\\
      \end{array}
  \end{center}
      \begin{itemize}
      \item Predicates should be any boolean Haskell expression.
      \item Everything (even errors!) satisfies $\any$\\
        \texttt{ fst ::: ($\ok,\any$) -> $\ok$ }
      \item Named arguments: $x$ is any value satisfying $c_1$.
      \end{itemize}
}

\frame{\frametitle{Crash-freeness}

  Morally, an expression $e$ is crash-free (cf) iff any \bad-free
  context in which we execute the expression in does not yield \bad.\\
  \ \\

  Laziness makes things tricky:
  \begin{itemize}
  \item \texttt{Just BAD}: not CF
  \item \texttt{(3,error ''foo'')}: not CF
  \item \texttt{fst (3,error ''foo'')}: \alert{CF}
  \end{itemize}
  \ \\
  And now to the operational semantics of contracts!
}

\frame{\frametitle{Contract semantics}
\begin{array}{rcl}
  e :::\{x \mid p \} &\iff& e \mbox{ diverges or } \\
  && (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
  e ::: x:t_1 \to t_2 &\iff& \mbox{for all } e_1, e_1 ::: t_1 \mbox{ implies } (e~e_1) ::: t_2[e_1/x]\\
  e ::: (t_1,t_2) &\iff& e \mbox{ diverges or } \\
  && (e \to^\star (e_1,e_2) \mbox{ and } e_1 ::: t_1 \mbox{ and } e_2 ::: t_2)\\
  e ::: \any &\iff& True  
\end{array}
}

\frame{\frametitle{Contract checking}
  Two roads: runtime checking and static checking.
  \begin{itemize}
  \item Runtime: Findler \& Felleisen, many others ... 
  \item The road less taken: static checking
  \end{itemize}
  Our simple plan: 
  \begin{center}
    Haskell + Contracts $\to$ First-order logic $\to$ Automated Theorem Prover (equinox)
  \end{center}

  Equinox is Koen's automated theorem prover. It's a FO theorem
  prover with equality and is under active development.

}

\section{How we do it}

\frame{\frametitle{Our language}
  Ok, not really Haskell. Wlog we make these asumptions:
  \begin{itemize}
  \item Top-level only pattern-matching 
  \item $\lambda$-lifting (top-level function definitions)
  \item Saturated data constructors (l.o.g. here but only for the sake of clarity)
  \end{itemize}
  From now on, we only consider well-typed expressions.
}

\frame{\frametitle{Our syntax}
  \begin{array}{rcl}
    u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad\\
    mod &:=& def,mod \mid tdecl,mod \mid f ::: c,mod \mid \epsilon\\
    def &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)]\\
    pat &:=& D(x_1,\dots,x_n)\\
    tdecl &:=& \mbox{data } T = dcons\\
    dcons &:=& \epsilon \mid D ; dcons \mid D ::: c ; dcons
  \end{array}
}

\frame{\frametitle{The big picture} Recall that $mod := def,mod \mid
  T,mod \mid f ::: c,mod \mid \epsilon$; a module is a list of
  function definitions, data definitions and claims that functions
  satisfy their contracts.
}

\section{The plan}
\frame{\frametitle{The big picture}
  \includegraphics[scale=0.4]{plan.pdf}
}

\frame{\frametitle{What we need}
  \begin{block}{Theorem (unproven)}
    $T_p \vdash T_c \implies f ::: c$
  \end{block}
  Unfortunately, so far we're only believers.

  Given this theorem we can use Koen's theorem prover to decide if
  $T_p \vdash T_c$.

}

\section{Translations}
\frame{\frametitle{Expressions become first-order logic terms}
  \begin{eqnarray*}
    \etrans{x} &=& x\\
    \etrans{f} &=& f\\
    \etrans{e_1~e_2} &=& app(\etrans{e_1},\etrans{e_2})\\
    \etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n})\\
    \etrans{\bad} &=& \bad
  \end{eqnarray*}
  \begin{itemize}
  \item Looks reasonable... But what's this ``app'' ?
  \item Why not $\etrans{e_1~e_2} = e_1(e_2)$ with $e_1$ being a FO function?
  \item To keep things first-order...
  \item Because we will need to quantify over an arbitrary
    expression (that includes functions)
  \end{itemize}
}

\frame{\frametitle{Formulae in the preamble}  
  We gave an (informal) definition of crash-freeness with the
  semantics of contract satisfaction. But here, we use CF as a logic
  predicate.

  We always add to our theory the following formulae, only
  approximating the intended semantics:
  $$ \forall f,x.~\cf{f} \land \cf{x} \implies \cf{app(f,x)} $$
  $$ \lnot \cf{\bad} $$
}

\frame{\frametitle{Translating modules}
  \begin{center}
    \begin{array}{rclr}
      \etrans{} &::& Expression \to Term\\
      \dtrans{} &::& Definition \to \{ FOF \}\\
      \ttrans{} &::& Data~type~declaration  \to \{ FOF \}\\
      \strans{} &::& Expression \to Contract \to \{ FOF \}\\
    \end{array}
  \end{center}

}

\frame{\frametitle{$\dtrans{}$ -- Definition translation (easy case)}
  Let $f$ be defined as $f~x = (x,x)$. It gives rise to the formula $\forall x.~app(f,x) = \mbox{Pair}(x,x)$.

  \begin{itemize}
  \item In the formula $f$ is an uninterpreted constant symbol.
  \item Pair is a function in FO logic.\\ \alert{We will give formulae for it soon!}
  \end{itemize}

  In general:
  $$\dtrans{f~x~y = e} = \forall x~y.~\etrans{f~x~y} = \etrans{e}$$
}

\subsection{Data constructors}
\frame{\frametitle{$\ttrans{}$ -- Data declaration translation}
  $\ttrans{\mobx{data } T = D_1 \mid \dots \mid D_n} = ?$

  Reasonable requirements:
  \begin{itemize}
  \item Injectivity
  \item CF conditions
  \end{itemize}

  Let's give some details...
}

\frame{\frametitle{$\ttrans{}$ -- Data constructor injectivity}
  $\ttrans{\mobx{data } T = D_1 \mid D_2} = ?$

  Assume the following arities: $D_1$:2, $D_2$:1.

  $$ \forall x_1,x_2 .~\forall y .~ D_1(x_1,x_2) \neq D_2(y) $$

  Do this for every pair of constructors!
}

\frame{\frametitle{$\ttrans{}$ -- Data constructor crash-freeness}
  $\ttrans{\mobx{data } T = D_1 \mid D_2} = ?$

  $$ \forall x_1,x_2.~\cf{x_1} \land \cf{x_2} \leftrightarrow \cf{D_1(x_1,x_2)} $$
  $$ \forall y.~\cf{y} \leftrightarrow \cf{D_2(y)}$$

  Values of type $T$ are crash-free iff they are created with crash-free values.

  \begin{itemize}
  \item $\rightarrow$ also true for cf function application
  \item $\leftarrow$  not true even for cf functions\\ (recall that \texttt{fst (3,$\bad$)} is crash-free)
  \end{itemize}
}

\subsection{Pattern matching}
\frame{\frametitle{$\dtrans{}$ -- The case of case}
  \texttt{
    head xs = case xs of\\
    ~~| Nil -> $\bad$\\
    ~~| Cons(y,ys) -> y
  }
  translates to:
  \begin{center}
  \begin{array}{rl}
    $       &app(head,Nil) = \bad\\
     \land &\forall y,ys.~app(head,Cons(y,ys)) = y\\
     \land & app(head,\bad) = \bad\\
     \land & \forall xs.~xs \neq Nil \land (\forall a~b.~xs \neq Cons(a,b))\\
     & ~~~\land xs \neq \bad \implies app(head,xs) = \unr$
  \end{array}
  \end{center}

  What is $\unr$? What is it for?
}
\subsection{Unreachability}
\frame{\frametitle{$\strans{}$ -- Contract satisfaction and the role of $\unr$}
  How do we translate that $e ::: c$?
  \begin{eqnarray*}
    \strans{e \in \any} &=& true\\
    \strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad\\
    &&~\land \etrans{u[e/x]} \neq False)\\
    \strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
  \end{eqnarray*}

  \begin{block}{$\unr$}
  Used to represent divergence or things that should not happen.

  $$fst ::: \ok \to \ok \mbox{ and }3 ::: \ok$$
  Thus, $fst~3$ should satisfy $\ok$!

  But it cannot happen (it's ill-typed), so we map $fst~3$ to a
  special value which satisfies $\ok$. (FOL is untyped!)
  \end{block}
}

\subsection{Recursion} 
\frame{\frametitle{$\dtrans{}$ -- Recursive functions: Take 1}
  Assume that we want to prove the claim $add ::: c$ where $c=\ok \to \ok \to \ok$.\\
  \texttt{add~x~y = case x of\\
    | Zero -> y\\
    | Succ a -> Succ (add a y)
  }\\

  Naively, we would do as usual: ask the theorem prover if
  $$\dtrans{add=e} \vdash \strans{add,c}$$
  But $add$ is recursive, so we would not go very far!

  \alert{Fix:} define a function $add\_rec$ that we use as IH. We now ask the theorem prover:
  $$\dtrans{add=e[add/add\_rec]},\strans{add\_rec,c} \vdash \strans{add,c}$$
}

\frame{\frametitle{$\dtrans{}$ -- Recursive functions: Take 1}
  That yields the following formula:
  \begin{center}
    \begin{array}{rl}
      $\forall x~y.&app(app(add,Zero),y) = y\\
      \land & \forall a.app(app(add,Succ(a)),y) = Succ(app(app(\alert{add\_rec},a),y))\\
      \land & app(app(add,\bad),y) = \bad\\
      \land & x \neq Zero \land (\forall c.~x \neq Succ(c))\\
      & ~~~\land x \neq \bad \implies app(app(add,x),y) = \unr$
    \end{array}
  \end{center}
  \ \\
  How about: $app(app(add,x),\bad) = \bad$?
}

\frame{\frametitle{$\dtrans{}$ -- Recursive functions: Take 2}
  A more complicated example: $mult ::: c$ where 
  $$c = \{x \mid x \neq 0\} \to \{y \mid y \neq 0\} \to \{z \mid z \neq 0\}$$

  \texttt{
    mult x y = case x of\\
    ~~| Zero -> Zero\\
    ~~| Succ a -> add (mult a y) y
    }

  $$\dtrans{mult=e[mult/mult\_rec]},\strans{mult\_rec,c} \not \vdash \strans{mult,c}$$
}

\frame{\frametitle{$\dtrans{}$ -- Recursive functions: Take 2}
  \texttt{
    mult x y = case x of\\
    ~~| Zero -> Zero\\
    ~~| Succ a -> add (mult a y) y
    }
  $$\dtrans{mult=e[mult/mult\_rec]},\strans{mult\_rec,c} \not \vdash \strans{mult,c}$$
  Why? Because the IH is too restrictive!\\
  \alert{Fix:}\\
  \begin{array}{rcl}
    $\dtrans{mult=e[mult/mult\_rec]}$&&\\
    $\strans{mult\_rec,c} & \vdash & \strans{mult,c}$\\\
    $\dtrans{mult\_rec=e[mult/mult\_rec]}$&&\\
  \end{array}
}

\frame{\frametitle{Questions}
  \begin{itemize}
  \item \emph{Proof that $T_p \vdash T_c \implies f ::: c$}
  \item Speed-up the theorem prover (by giving strategies)
  \item The user should be able to specify lemmas. How?
  \item Hoare Logic view of contracts (partial spec) vs. liquid types (complete spec)
  \item Contracts on datatypes
  \item Not all contracts expressible as Haskell predicates. Need special operator: \texttt{`satisfies` c}?
  \end{itemize}
}

\end{document}

### 2011-06-22 contracts.tex
pdf update

--- 2011-06-09 draft2.tex
@@
 \usepackage{fullpage}
+\usepackage{graphicx}
 
@@
 \begin{array}{rcl}
-  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,\unr\})\\
+  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
   e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
@@
 
+A stupid fix was to provide defining equations for $multp$ (the same
+as for $mult$). Indeed, we could prove that $mult$ satisfied its
+contract, but it was trivially true...
+
+Current idea is to unfold the definition a fixed number of times.
+
+\section{De-appification and removed CFness}
+\subsection{De-appification}
+Instead of always using the $app$ function when doing function
+application, we now use directly the function name if it is fully
+applied. For example, if $f$ is of arity 2, $f x y$ is now translated
+to $f(x,y)$ instead of $app(app(f,x),y)$.  To deal with partially
+applied functions, we can use $f\_ptr$ and some extra axioms.
+
+This transformation makes things much faster for equinox, especially
+given that in most cases, function are totally applied.
+
+\subsection{Removing CFness}
+In the POPL paper, CF was introduced in the contract satisfaction
+semantics because it was required in order to define the
+wrapping. Given that we do not use any wrapping, we can remove it from
+the semantics.
+
+In fact, it's probably better to decouple crash-freeness and predicate
+satisfaction.
+
+Here the new version of the semantics : $e \in \{x \mid p \} \iff e
+\mbox{ diverges or } (p[e/x] \not \to^\star \{\bad,False\})$
+
+But it leads to some tricky cases:
+
+Assume a reasonable definition of $add$. Then, this contract does not
+hold.
+$$add ::: a:\{ x \mid True \} \to \{ y \mid True \} \to \{ z \mid z
+\geq a \}$$
+
+Why? Because now that we have removed crash-freeness, we can have $x$
+and $y$ instantiated to $\bad$. And then, $ add~x~y \geq z $ becomes
+$\bad$. Or the semantics say that the predicate should not evaluate to
+$\bad$.
+
+A possible fix is to provide, directly in the source language a $cf$
+predicate, so that one can write contract like:
+
+$$add ::: a:\{ x \mid cf(x) \} \to \{ y \mid cf(y) \} \to \{ z \mid z
+\geq a \}$$
+
+Note that in this setting $\{ x \mid True \} = \any$ and $\{ x \mid
+cf(x) \} = \ok$.
+
+\subsection{Dimitrios point}
+Dimitrios isn't convinced that we keep the consistency of the whole
+module when checking one function after the other.
+
+Assume that we have a module that contains two functions defintion $f$
+and $g$ and two contracts : $f ::: Any -> Ok$ and $g ::: Any ->
+Ok$. We assume that those contract do not hold, for example if $f$ is
+$not$ and $g$ is $id$.
+
+Now, we want to check $f$'s contract. We take out $f$'s contract,
+translate the function definition for $f$ and $g$ and add the contract
+satisfation formulae for $g$'s contract.
+
+But, given that $g$'s contract does not hold, we can derive $\bot$ and
+then prove that $f$'s contract hold.
+
+Now the user is happy, and does the same thing for $g$'s contract. It
+holds too, given that we can also derive $\bot$ for $f$'s contract.
+
+Finally, the user is really happy, but in fact he has proved nothing.
+
+I think that it's not an actual issue, if we only give to equinox
+useful formulae. Ie, if $f$ does not use $g$ we do not include $g$ in
+the translation. Obviously it slows down the proof and, breaks the
+consistency.
+
+Consider the following graph, an arrow from $a$ to $b$ meaning that
+$b$ uses $a$ in its definition.
+
+\begin{center}
+\includegraphics[scale=0.5]{flow.png}
+\end{center}
+
+We (D and I) believe that if we check a function by only giving the
+formulae coming from the ancestors, then we do not break consistency.
+Eg, to check $a$'s contract we need to give the contract and
+definition for $f$ and $g$ too, but not for $h$.
+
 \end{document}

### 2011-06-30 contracts.tex

@@
 \end{center}
+
 We always give the following equations to define crash-freeness:
@@
   &\mid& \{ x \mid e \}\\
-  &\mid& \any \\
+  &\mid& \{ !x \mid e \}\\
 \end{array}
@@
 \begin{array}{rcl}
-  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
+  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
+  e \in \{!x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
   e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
   e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
-  e \in \any &\iff& True  
 \end{array}
 \end{center}
+
+\section{CF-ness and unreachability}
+\subsection{CF-ness}
+In the POPL paper, CF was introduced in the contract satisfaction
+semantics because it was required in order to define the
+wrapping. Given that we do not use any wrapping, we can make a
+separation between crash-freeness and predicate satisfaction now, as
+shown in the contracts semantics, with the special syntax.
+
+It leads to some tricky cases:
+
+Assume a reasonable definition of $add$. Then, this contract does not
+hold.
+$$add ::: a:\{ x \mid True \} \to \{ y \mid True \} \to \{ z \mid z
+\geq a \}$$
+
+Why? Because now that we have removed crash-freeness, we can have $x$
+and $y$ instantiated to $\bad$. And then, $ add~x~y \geq z $ becomes
+$\bad$. Or the semantics say that the predicate should not evaluate to
+$\bad$.
+
+Now, the correct contract is:
+$$add ::: a:\{ !x \mid cf(x) \} \to \{ !y \mid cf(y) \} \to \{ !z \mid z
+\geq a \}$$
 
@@
 
-Equation (10) is not restrictive even in the case of laziness because $e$ is always evaluated. (TODO, but not always fully evaluated...)
+Equation (10) is not restrictive even in the case of laziness because $e$ is always evaluated.
 
@@
 \begin{eqnarray}
-\strans{e \in \any} &=& true\\
 \strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
@@
 
+\subsection{Checking the contracts in the right order}
+Here is a little example showing that we should be careful about we
+belong to a theory.
+
+Assume that we have a module that contains two functions defintion $f$
+and $g$ and two contracts : $f ::: Any \to Ok$ and $g ::: Any \to
+Ok$. We assume that those contract do not hold, for example if $f$ is
+$not$ and $g$ is $id$.
+
+Now, we want to check $f$'s contract. We take out $f$'s contract,
+translate the function definition for $f$ and $g$ and add the contract
+satisfation formulae for $g$'s contract.
+
+But, given that $g$'s contract does not hold, we can derive $\bot$ and
+then prove that $f$'s contract hold.
+
+Now the user is happy, and does the same thing for $g$'s contract. It
+holds too, given that we can also derive $\bot$ for $f$'s contract.
+
+Finally, the user is really happy, but in fact he has proven nothing.
+
+The solution is to only include the formulae that belongs to a
+functions that are actually used. For example, to prove $a$'s
+contract, we should only include $f$ and $g$ formulae.
+
+\begin{center}
+\includegraphics[scale=0.5]{flow.png}
+\end{center}
+
 \section{Example}
@@
 
-copy ::: a:{x : True} -> {y:True};;
+copy ::: a:{!x : True} -> {!y:True};;
 \end{verbatim}
 The contract states that if we give to copy a crash-free argument $x$, copy will return a crash-free result $y$.
-Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it. TODO
+Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it.
 
@@
 
-add ::: a:{x : isZero x} -> {y:isZero y} -> {z: isZero z};;
+add ::: a:{!x : isZero x} -> {!y:isZero y} -> {!z: isZero z};;
 \end{verbatim}
@@
 
-notZero ::: a:{x:True} -> {y: True};;
-add  ::: a:{x: True} -> b:{y: True} -> {z: True};;
-add  ::: a:{x: notZero x} -> b:{y: True} -> {z: True};;
-mult ::: a:{x: notZero x} -> b:{y: notZero y} -> {z: True};;
+notZero ::: a:{!x:True} -> {!y: True};;
+add  ::: a:{!x: True} -> b:{!y: True} -> {!z: True};;
+add  ::: a:{!x: notZero x} -> b:{!y: True} -> {!z: True};;
+mult ::: a:{!x: notZero x} -> b:{!y: notZero y} -> {!z: True};;
 \end{verbatim}
@@
 
-\subsection{Removing CFness}
-In the POPL paper, CF was introduced in the contract satisfaction
-semantics because it was required in order to define the
-wrapping. Given that we do not use any wrapping, we can remove it from
-the semantics.
-
-In fact, it's probably better to decouple crash-freeness and predicate
-satisfaction.
-
-Here the new version of the semantics : $e \in \{x \mid p \} \iff e
-\mbox{ diverges or } (p[e/x] \not \to^\star \{\bad,False\})$
-
-But it leads to some tricky cases:
-
-Assume a reasonable definition of $add$. Then, this contract does not
-hold.
-$$add ::: a:\{ x \mid True \} \to \{ y \mid True \} \to \{ z \mid z
-\geq a \}$$
-
-Why? Because now that we have removed crash-freeness, we can have $x$
-and $y$ instantiated to $\bad$. And then, $ add~x~y \geq z $ becomes
-$\bad$. Or the semantics say that the predicate should not evaluate to
-$\bad$.
-
-A possible fix is to provide, directly in the source language a $cf$
-predicate, so that one can write contract like:
-
-$$add ::: a:\{ x \mid cf(x) \} \to \{ y \mid cf(y) \} \to \{ z \mid z
-\geq a \}$$
-
-Note that in this setting $\{ x \mid True \} = \any$ and $\{ x \mid
-cf(x) \} = \ok$.
-
-\subsection{Dimitrios point}
-Dimitrios isn't convinced that we keep the consistency of the whole
-module when checking one function after the other.
-
-Assume that we have a module that contains two functions defintion $f$
-and $g$ and two contracts : $f ::: Any -> Ok$ and $g ::: Any ->
-Ok$. We assume that those contract do not hold, for example if $f$ is
-$not$ and $g$ is $id$.
-
-Now, we want to check $f$'s contract. We take out $f$'s contract,
-translate the function definition for $f$ and $g$ and add the contract
-satisfation formulae for $g$'s contract.
-
-But, given that $g$'s contract does not hold, we can derive $\bot$ and
-then prove that $f$'s contract hold.
-
-Now the user is happy, and does the same thing for $g$'s contract. It
-holds too, given that we can also derive $\bot$ for $f$'s contract.
-
-Finally, the user is really happy, but in fact he has proved nothing.
-
-I think that it's not an actual issue, if we only give to equinox
-useful formulae. Ie, if $f$ does not use $g$ we do not include $g$ in
-the translation. Obviously it slows down the proof and, breaks the
-consistency.
-
-Consider the following graph, an arrow from $a$ to $b$ meaning that
-$b$ uses $a$ in its definition.
-
-\begin{center}
-\includegraphics[scale=0.5]{flow.png}
-\end{center}
-
-We (D and I) believe that if we check a function by only giving the
-formulae coming from the ancestors, then we do not break consistency.
-Eg, to check $a$'s contract we need to give the contract and
-definition for $f$ and $g$ too, but not for $h$.
-
 \end{document}

### 2011-06-30 contracts/FOL.hs
deappification fix

@@ -27,7 +27,9 @@ data Term a = Var a
 
 instance Show a => Show (Term a) where
   show (Var v) = show v
-  show (App v) = show v
+  show (App []) = error "Cannot apply nothing"
+  show (App [t]) = show t
+  show (App ts) = "app(" ++ show (App (init ts)) ++ "," ++ show (last ts) ++ ")"
   show (Weak v) = show v
   show (FullApp f []) = show f
   show (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"

### 2011-06-30 contracts/Haskell.hs

@@ -23,6 +23,7 @@ data Expression = Var Variable
                 | App Expression Expression
                 | FullApp Variable [Expression]
                 | Sat Expression Contract -- e `satisfies` c --> True iff e \in c
+                | CF Expression
                 | BAD
                 deriving (Show,Eq,Ord)          
 
@@ -47,6 +48,8 @@ type Pattern = [Variable]
 data Contract = AppC Variable Contract Contract -- x : c -> c'
               | Pair Contract Contract
               | Pred Variable Expression   -- {x:e}
+              | And Contract Contract
+              | Or  Contract Contract
               | Any
               deriving (Show,Eq,Ord)                  
 
@@ -62,11 +65,14 @@ subst (App e1 e2) x y = App (subst e1 x y) (subst e2 x y)
 subst (FullApp f es) x y = let Var x' = (subst (Var f) x y) in 
   FullApp x' $ map (\e -> subst e x y) es
 subst BAD _ _ = BAD
+subst (CF e) x y = CF (subst e x y)
 
 substC :: Contract -> Expression -> Variable -> Contract
 substC (AppC u c1 c2) x y = AppC u (substC c1 x y) (substC c2 x y) -- TODO and if u==y?
 substC (Pair c1 c2) x y = Pair (substC c1 x y) (substC c2 x y)
 substC (Pred u e) x y = if u/=y then Pred u (subst e x y) else (Pred u e)
+--substC (And cs) x y = And $ map (\c -> substC c x y) cs
+--substC (Or cs) x y = Or $ map (\c -> substC c x y) cs
 substC Any _ _ = Any
 
 ok :: Contract

### 2011-06-30 contracts.tex

@@
   u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
-  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] &\\
+  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] \mid \cf{e}&\\
   p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
@@
 \begin{array}{rclr}
-  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c & (Expression)\\
+  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \cf{e} & (Expression)\\
   p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
@@
   &\mid& \{ x \mid e \}\\
-  &\mid& \{ !x \mid e \}\\
+  &\mid& c_1 \land c_2 \\
+  &\mid& c_1 \lor c_2 \\
 \end{array}
@@
   e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
-  e \in \{!x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
   e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
   e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
+  e \in c_1 \land c_2 &\iff& e \in c_1 \mbox{ and } e \in c_2\\
+  e \in c_1 \lor c_2 &\iff& e \in c_1 \mbox{ or } e \in c_2\\
 \end{array}
@@
 \etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
+\etrans{\cf{e}} &=& \cf{e}\\
 \end{eqnarray}
@@
 \begin{eqnarray}
-\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
+\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
 && \cup \etransf{e} \cup \etransf{u[e/x]}\\
-\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
+\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}\\
+\strans{e \in c_1 \land c_2} &=& \strans{e \in c_1} \land \strans{e \in c_2}\\
+\strans{e \in c_1 \lor c_2} &=& \strans{e \in c_1} \lor \strans{e \in c_2}\\
 \end{eqnarray}

### 2011-06-30 contracts/Translation.hs
ptr

@@ -49,7 +49,7 @@ dTrans (H.Let f vs e) = do
   return $ [F.Forall vvs $ F.Eq (F.FullApp (F.Regular f) vvs) (F.Weak $ et),fptr1,fptr2]
   where vvs = map (F.Var . F.Regular) vs
         fptr1 = F.Forall vvs $ F.Iff (F.CF (F.FullApp (F.Regular f) vvs)) (F.CF $ F.Var $ F.Regular (f++"_ptr"))
-        fptr2 = F.Forall vvs $ F.Eq (F.FullApp (F.Regular (f++"_ptr")) vvs) (F.App $ (F.Var . F.Regular) f : vvs)
+        fptr2 = F.Forall vvs $ F.Eq (F.FullApp (F.Regular (f)) vvs) (F.App $ (F.Var . F.Regular) (f++"_ptr") : vvs)
 
 dTrans (H.LetCase f vs e pes) = do
   et <- eTrans e
@@ -66,7 +66,7 @@ dTrans (H.LetCase f vs e pes) = do
       eq12 = (F.FullApp (F.Regular f) vvs `F.Eq` F.Var F.UNR)
       bigAndSel = [F.Not $ et `F.Eq` F.Weak (F.FullApp (F.Regular di) [F.FullApp (F.Regular ("sel_"++(show i)++"_"++di)) [et] | i <- [1..ai]]) | (di,ai) <- arities]
       fptr1 = F.Forall vvs $ F.Iff (F.CF (F.FullApp (F.Regular f) vvs)) (F.CF $ F.Var $ F.Regular (f++"_ptr"))
-      fptr2 = F.Forall vvs $ F.Eq (F.FullApp (F.Regular (f ++ "_ptr")) vvs) (F.App $ (F.Var . F.Regular) f : vvs)
+      fptr2 = F.Forall vvs $ F.Eq (F.FullApp (F.Regular f) vvs) (F.App $ (F.Var . F.Regular) (f++"_ptr") : vvs)
   return $ [F.Forall (vvs ++ zs) $ F.And (eq9++[eq10,eq11]),fptr1,fptr2]
 
 test = (H.LetCase "head" ["xyz"] (H.Var "xyz") [(["nil"],H.BAD),(["cons","a","b"],H.Var "a")])

### 2011-07-05 contracts/FOL.hs
removing upperify

@@ -1,5 +1,4 @@
-module FOL
-       where
+module FOL where
 
 import Prelude hiding (True,False)
 import Data.Char (toUpper)
@@ -18,26 +17,20 @@ instance Show Variable where
   show BAD = "bad"
   show UNR = "unr"
 
-
 data Term a = Var a
             | App [Term a]
+            | FullApp a [Term a]
             | Weak (Term a)
             deriving (Eq)
 
 instance Show a => Show (Term a) where
   show (Var v) = show v
-  show (App v) = show v
+  show (App []) = error "Cannot apply nothing"
+  show (App [t]) = show t
+  show (App ts) = "app(" ++ show (App (init ts)) ++ "," ++ show (last ts) ++ ")"
   show (Weak v) = show v
-      
-instance Functor Term where
-  fmap f (Var v) = Var (f v)
-  fmap f (App terms) = App (map (fmap f) terms)
-  
-instance F.Foldable Term where
-  foldr f z (Var v) = f v z
-  foldr f z (App []) = z
-  foldr f z (App (Var v:as)) = f v (F.foldr f z (App as))
-  foldr f z (App (App as1:as2)) = F.foldr f (F.foldr f z (App as1)) (App as2)
+  show (FullApp f []) = show f
+  show (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"
 
 
 data Formula a = Forall [a] (Formula a)
@@ -51,33 +44,6 @@ data Formula a = Forall [a] (Formula a)
                | Eq a a
                | CF a
                deriving (Show,Eq)                  
-
-instance F.Foldable Formula where
-  foldr = undefined
-  
-instance Traversable Formula where
-  traverse f (Forall vs fof) = undefined
-  traverse f (Implies fof1 fof2) = Implies <$> (traverse f fof1) <*> (traverse f fof2)
-  traverse f (Iff fof1 fof2) = Iff <$> (traverse f fof1) <*> (traverse f fof2)
-  traverse f (Not fof) = Not <$> traverse f fof
-  traverse f (Or fofs) = Or <$> sequenceA (map (traverse f) fofs)
[…]

### 2011-07-05 contracts/FOL.hs
au point ou j'en suis

@@ -1,3 +1,5 @@
+{-# LANGUAGE DeriveFunctor #-}
+
 module FOL where
 
 import Prelude hiding (True,False)
@@ -43,7 +45,7 @@ data Formula a = Forall [a] (Formula a)
                | False
                | Eq a a
                | CF a
-               deriving (Show,Eq)                  
+               deriving (Show,Eq,Functor)
   
 splitOnAnd :: Formula a -> [Formula a]
 splitOnAnd (Forall xs (And fs)) = map (Forall xs) fs
@@ -68,26 +70,38 @@ simplify f = filter (/= True) $ splitOnAnd (removeConstants f)
 
 extractVR (Var (Regular x)) = x
 
+upperIfy :: [String] -> Formula (Term Variable) -> Formula (Term Variable)
+upperIfy c (Forall xs f) = Forall (map (Var . Regular . map toUpper . extractVR) xs) (upperIfy c' f)
+  where c' = (map extractVR xs)++c
+upperIfy c f = fmap (auxUpper c) f
+
+auxUpper :: [String] -> Term Variable -> Term Variable
+auxUpper c (Var (Regular v)) = if v `elem` c then (Var . Regular) (map toUpper v) else Var $ Regular v
+auxUpper c (Var v) = Var v
+auxUpper c (App ts) = App $ map (auxUpper c) ts
+auxUpper c (FullApp x ts) = FullApp x (map (auxUpper c) ts) -- TODO think about it
+auxUpper c (Weak t) = Weak $ auxUpper c t
+
 toTPTP :: Formula (Term Variable) -> String
-toTPTP f = header ++ "\n" ++ (go f) ++ "\n" ++ footer
+toTPTP f = header ++ "\n" ++ (aux $ upperIfy [] f) ++ "\n" ++ footer
   where header = "fof(axiom,axiom,"
         footer = ").\n"
-        go (Forall xs f) = "! " ++ show xs ++ "  : (" ++ go f ++ ")"
-        go (Implies f1 f2) = "(" ++ go f1 ++ ") => (" ++ go f2 ++ ")"
-        go (Iff f1 f2) = "(" ++ go f1 ++ ") <=> (" ++ go f2 ++ ")"
-        go (Not (Eq t1 t2)) =  goTerm t1 ++ " != " ++ goTerm t2
-        go (Not f) = "~(" ++ go f ++ ")"
[…]

### 2011-07-05 contracts/Script.hs

@@ -11,6 +11,9 @@ import System.IO
 toTheory f c = do
   s <- readFile f
   return $ trans (haskell $ lexer s) c >>= simplify >>= toTPTP
+blo f c = do
+  s <- readFile f
+  return $ (haskell $ lexer s) 
 
 main = do
   files:c:_ <- getArgs
@@ -18,3 +21,4 @@ main = do
   cts <- toTheory files c
   writeFile (files ++ ".tptp") cts
   
+test = toTheory "test.hs" "gt"

### 2011-07-05 contracts.tex

@@
   u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
-  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] \mid \cf{e}&\\
+  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] \mid \cf{e} \mid f(e,\dots,e)&\\
   p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
@@
 \begin{array}{rclr}
-  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \cf{e} & (Expression)\\
+  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \cf{e} \mid f(e,\dots,e) & (Expression)\\
   p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
@@
 \begin{array}{rclr}
-  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D(t,\dots,t) \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
+  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D(t,\dots,t) \mid f(t,\dots,t) \mid f \mid n \mid \bad \mid \unr & (Term)\\
   \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
@@
 \etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
-\etrans{\cf{e}} &=& \cf{e}\\
+\etrans{f(e_1,\dots,e_n)} &=& f(\etranst{e_1},\dots,\etranst{e_n})\\
+\etrans{\cf{e}} &=& (True,\cf{\eTranst{e}})\\
 \end{eqnarray}
@@
 \dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etranst{f~x_1 \dots x_n} = \etranst{e} \cup \etransf{e}\\
+&\cup& (\forall x_1 \dots x_n. (\bigwedge \cf{x_i} \rightarrow \cf{f(x_1,\dots,x_n)}) \leftrightarrow \cf{f\_ptr}\\
+&\cup& (\forall x_1 \dots x_n. (f(x_1,\dots,x_n) = app(...app(f\_ptr,x_1)...,x_n)))\\
 \dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etranst{e} = \etranst{D_i(\overline{z})} \to \etranst{f~\overline{x}} = \etranst{e_i})\\
@@
 && ~~\to \etranst{f~x_1 \dots x_n} = \unr)\\
-&& \cup \etransf{e}
+&& \cup \etransf{e}\\
+&\cup& (\forall x_1 \dots x_n. (\bigwedge \cf{x_i} \rightarrow \cf{f(x_1,\dots,x_n)}) \leftrightarrow \cf{f\_ptr}\\
+&\cup& (\forall x_1 \dots x_n. (f(x_1,\dots,x_n) = app(...app(f\_ptr,x_1)...,x_n)))\\
 \end{eqnarray}
@@
 
-Equation (11) and (12) state that if we have a type mismatch than the code is $\unr$. It should never occur because we're assuming that the code has already been typechecked.
+Equation (11) and (12) state that if we have a type mismatch than the code is $\unr$. It should never occur because we're assuming that the code has already been typechecked. %TODO eq number
+
+Finally, we have to relate $f\_ptr$ to $f$ and define when we can say that $f$ is crash-free.
 

### 2011-07-05 contracts/Haskell.hs
auto fullapp

@@ -1,3 +1,5 @@
+{-# LANGUAGE DeriveFunctor #-}
+
 module Haskell ( Expression (..)
                , DataType (..)
                , Program (..)
@@ -7,6 +9,9 @@ module Haskell ( Expression (..)
                , DefGeneral (..)
                , ContSat (..)
                , apps
+               , arities  
+               , appifyExpr
+               , appify
                , subst
                , substC
                , toList
@@ -17,64 +22,84 @@ where
 type Variable = String
 type Constructor = String
 
--- TODO Integer not implemented
 data Expression = Var Variable
-                | Con Constructor
                 | App Expression Expression
                 | FullApp Variable [Expression]
-                | Sat Expression Contract -- e `satisfies` c --> True iff e \in c
+                | Sat Expression (Contract Expression) -- e `satisfies` c --> True iff e \in c
                 | CF Expression
                 | BAD
-                deriving (Show,Eq,Ord)          
+                deriving (Show,Eq,Ord)
 
-type Program = [DefGeneral]
-data DefGeneral = ContSat ContSat
-                | Def Definition
-                | DataType DataType
-                deriving (Eq,Show,Ord)
+type Program = [DefGeneral Expression]
+data DefGeneral a = ContSat (ContSat a)
+                  | Def (Definition a)
+                  | DataType (DataType a)
+                  deriving (Eq,Show,Ord,Functor)
 
-data ContSat = Satisfies Variable Contract
-             deriving (Show,Eq,Ord)                  
+data ContSat a = Satisfies Variable (Contract a)
+               deriving (Show,Eq,Ord,Functor)
                
-data Definition = Let Variable [Variable] Expression
-                | LetCase Variable [Variable] Expression [(Pattern,Expression)]
-                deriving (Show,Eq,Ord)                  
+data Definition a = Let Variable [Variable] a
+                  | LetCase Variable [Variable] a [(Pattern,a)]
+                  deriving (Show,Eq,Ord,Functor)
                   
-data DataType = Data Variable [(Variable,Int,Contract)] -- Data constructors + arity + contract
-              deriving (Eq,Show,Ord)
-
[…]

### 2011-07-05 contracts/Parser.hs

@@ -105,9 +105,9 @@ action_19 _ = happyReduce_1
 
 action_20 _ = happyReduce_40
 
-action_21 (24) = happyShift action_50
-action_21 (30) = happyShift action_51
-action_21 (11) = happyGoto action_49
+action_21 (24) = happyShift action_23
+action_21 (30) = happyShift action_49
+action_21 (11) = happyGoto action_48
 action_21 _ = happyReduce_7
 
 action_22 (18) = happyShift action_22
@@ -115,10 +115,9 @@ action_22 (24) = happyShift action_23
 action_22 (30) = happyShift action_25
 action_22 (33) = happyShift action_26
 action_22 (11) = happyGoto action_20
-action_22 (12) = happyGoto action_48
+action_22 (12) = happyGoto action_47
 action_22 _ = happyReduce_37
 
-action_23 (30) = happyShift action_47
 action_23 _ = happyReduce_32
 
 action_24 (18) = happyShift action_22
@@ -176,29 +175,29 @@ action_33 _ = happyReduce_29
 
 action_34 _ = happyReduce_10
 
-action_35 (25) = happyShift action_64
+action_35 (25) = happyShift action_60
 action_35 _ = happyFail
 
-action_36 (24) = happyShift action_63
+action_36 (24) = happyShift action_59
 action_36 _ = happyFail
 
 action_37 (34) = happyShift action_31
 action_37 (35) = happyShift action_32
-action_37 _ = happyReduce_47
+action_37 _ = happyReduce_46
 
 action_38 (34) = happyShift action_31
 action_38 (35) = happyShift action_32
-action_38 _ = happyReduce_48
+action_38 _ = happyReduce_47
 
 action_39 (18) = happyShift action_22
 action_39 (24) = happyShift action_23
 action_39 (30) = happyShift action_25
 action_39 (33) = happyShift action_26
 action_39 (11) = happyGoto action_20
-action_39 (12) = happyGoto action_62
+action_39 (12) = happyGoto action_58
 action_39 _ = happyFail
 
-action_40 (23) = happyShift action_61
+action_40 (23) = happyShift action_57
 action_40 (34) = happyShift action_31
 action_40 (35) = happyShift action_32
 action_40 _ = happyFail
@@ -206,23 +205,23 @@ action_40 _ = happyFail
 action_41 (21) = happyShift action_15
 action_41 (24) = happyShift action_16
 action_41 (30) = happyShift action_17
-action_41 (14) = happyGoto action_60
[…]

### 2011-07-05 contracts/Script.hs

@@ -21,5 +21,4 @@ main = do
   cts <- toTheory files c
   writeFile (files ++ ".tptp") cts
   
-test = toTheory "test.hs" "gt"
-
+test = toTheory "test.hs" "add"

### 2011-07-05 contracts/Translation.hs

@@ -15,7 +15,7 @@ type Fresh = State (String,Int,[F.Formula (F.Term F.Variable)])
 -- Expression
 -------------
 
-eTrans :: H.Expression -> Fresh (F.Term F.Variable)
+--eTrans :: H.Expression -> Fresh (F.Term F.Variable)
 eTrans (H.Var v) = return $ (F.Var $ F.Regular v)
 eTrans (H.App e1 e2) = do 
   t1 <- eTrans e1
@@ -25,7 +25,6 @@ eTrans (H.FullApp f es) = do
   ts <- sequence $ map eTrans es
   return $ F.FullApp (F.Regular f) ts
 eTrans H.BAD = return $ F.Var $ F.BAD
-eTrans (H.Con d) = return $ F.Var $ F.Regular d
 eTrans (H.Sat e c) = do 
   ts <- sTrans (H.Var "x") c
   te <- eTrans $ H.FullApp "satC" [H.Var "x"]
@@ -43,7 +42,7 @@ eTrans (H.CF e) = do
 -- Definition
 -------------
 
-dTrans :: H.Definition -> Fresh [F.Formula (F.Term F.Variable)]
+--dTrans :: H.Definition -> Fresh [F.Formula (F.Term F.Variable)]
 dTrans (H.Let f vs e) = do
   et <- eTrans e
   return $ [F.Forall vvs $ F.Eq (F.FullApp (F.Regular f) vvs) (F.Weak $ et),fptr1,fptr2,fptr3]
@@ -79,7 +78,7 @@ test = (H.LetCase "head" ["xyz"] (H.Var "xyz") [(["nil"],H.BAD),(["cons","a","b"
 -- Contract satisfaction
 ------------------------
 
-sTrans :: H.Expression -> H.Contract -> Fresh (F.Formula (F.Term F.Variable))
+--sTrans :: H.Expression -> H.Contract -> Fresh (F.Formula (F.Term F.Variable))
 sTrans e H.Any = return F.True
 
 sTrans e (H.Pred x u) =  do
@@ -117,15 +116,15 @@ sTrans e (H.Or c1 c2) = do
 -- Data constructors
 --------------------
 
-tTrans :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
+--tTrans :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
[…]

### 2011-08-15 paper.tex
Paper update

\documentclass[preprint]{sigplanconf}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage[pdftex]{graphicx}

\begin{document}

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\cfc}{\texttt{CF}}

\title{Static contract checking for Haskell}
\maketitle

\section{Introduction}

Our approach to static contract-checking is to translate source code
to a first-order logic theory and then use an automated theorem prover
to check the consistency of the theory.

Consider:
\begin{verbatim}
data List a = Nil | Cons a (List a)

notnull x = case x of
  Nil -> False
  Cons(x,y) -> True

head ::: (CF && {x | notnull x }) -> CF
head xs = case xs of
  Nil -> BAD
  Cons(x,y) -> x
\end{verbatim}

First, we need to encode the List structure.

We start by stating that $Nil$ and $Cons$ can never be equal:
$$\forall a,b.~Cons(a,b) \neq Nil$$

Then, we must state that $Nil$ never crashes (ie cannot be evaluated
to an exception) and that $Cons(x,y)$ crashes iff either $x$ or $y$
crashes. The statement $x$ crashes is encoded by the term $\cf{x}$.

$$\cf{Nil}$$
$$\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}$$

We also say some stuff about unreachibility but I can't think of a
good way to explain it right now.
$$\forall y,ys.~Cons(y,ys) \neq \unr $$
$$Nil \neq \unr $$

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
$$\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y $$
$$\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys $$

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.
$$\forall xs.~ xs=Nil \implies notnull(xs) = false$$
$$\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true$$

We also need to specify the translation of calls to $notnull$ with
$\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
$$\forall xs.~xs=\bad \to notnull(xs) = \bad$$

Finally, we say that one call $notnull$ with a an argument which is
not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
$$\land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
$$ notnull(X) = \unr$$

The translation of $head$ follows the same pattern:
$$\forall xs.~ x=Nil \implies head(xs) = \bad$$
$$\forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y$$

$$\forall xs.~xs=\bad \to head(xs) = \bad$$

$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
$$ \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
$$ head(X) = \unr$$

We now have translated the source code. Let us call all those formulae
the theory $T$. We translate separatly the contract:
$$ \phi := \forall xs.~\cf{xs} \land not(null(xs)) = true \to \cf{head(xs)} $$

Now that we translated everything to first-order logic, we can ask the
theorem prover if the theory formed by those formulae is
consistent, ie if $T \vdash \phi$.

Intuitivly, $T$ is consistent (ie $T \not \vdash \bot$), because each
formula serves a specific purpose. Now, assume that $xs$ satisfies
$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
because we have $\cf{xs}$ and $\not \cf{\bad}$. The constraint
$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
some $y$ and $ys$. But the translation of $notnull$ is total, ie the
function is totally defined. This implies that there exist $y$ and
$ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can now
derive $\cf{y}$ and $\cf{ys}$. But $head(xs) = y$ in our setting, and
$y$ is crash-free. QED.

\section{Languages}
\subsection{$\hprime:~\lambda$-calculus variant}

The syntax of $\mathcal{H}$ is defined in \ref{hprime-stx}. A module
is a list of toplevel definitions, claims that functions statisfy
contracts and data definitions.

\begin{itemize}
\item There's no $\lambda$-abstraction, because we can always lift
  them to toplevel declaration.
\item We do not allow nested case expressions, because once again, we
  can always lift them to the toplevel.
\item Until \label{ho} we will only consider full application of
  functions ($f(x,y)$), in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

\begin{figure}[h]
  \centering
  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \\
    &\mid& f \in c & \\
    &\mid& f~\vec{x}~=~e & \\
    &\mid& \hspace{-0.2cm}
       \begin{array}{rcl}
         f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1(\vec{x}) \to e_1 \mid \cdots \mid K_n(\vec{x}) \to e_n \\
       \end{array} & \\
  \end{array}

  \begin{array}{rclr}
    x,y,f,g,a,b & \in & \mbox{Variables} \\
    T &\in& \mbox{Type Constructors} \\
    K &\in& \mbox{Data Constructors} \\
  \end{array}

  \begin{array}{rclr}
    e &\in& \mbox{Expressions} & \\
    e &::=& x & \\
    &\mid& \bad & \\
    &\mid& e~e & \\
    &\mid& f(e,\dots,e) & \\
    &\mid& K(e,\dots,e) & \\
  \end{array}

\caption{Syntax of the language $\hprime$}
\label{hprime-stx}
\end{figure}

\subsection{Contracts}
Contract syntax is described in figure \ref{cont-stx}. The predicates
we use in our contracts can be any boolean $\hprime$ expression. We
only consider pairs of contract for simplicity, although there is no
issue with generalisation to arbitrary tuples.

\begin{figure}[h]
  \label{cont-stx}
  \centering 
  \begin{array}{rclr}
  c &:=& x:c \to c\\
  &\mid& (c,c) \\
  &\mid& c \land c \\
  &\mid& c \lor c \\
  &\mid& \{ x \mid p \}\\
  &\mid& \cfc \\
  \end{array}
  \caption{Contract syntax}
\end{figure}

We give the semantics of contract by defining ``$e$ satisfies $t$'',
written $e \in t$ in figure \ref{cont-smt}. Note that this definition
doesn't yield any operative way to check that an expression actually
meets the specification given by its contract.

\begin{figure}[h]
  \label{cont-smt}
  \centering
  \begin{array}{rcl}
    e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
    e \in x:c_1 \to c_2 &\iff& \forall e_1 \in c_1, (e~e_1) \in c_2[e_1/x]\\
    e \in (c_1,c_2) &\iff& e \mbox{ diverges or }\\
    &&  (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in c_1, e_2 \in c_2)\\
    e \in c_1 \land c_2 &\iff& e \in c_1 \mbox{ and } e \in c_2 \\
    e \in c_1 \lor c_2 &\iff& e \in c_1 \mbox{ or } e \in c_2 \\
    e \in \cfc &\iff& e \mbox{ is crash-free}
  \end{array}
  \caption{Semantics of contract satisfaction}
\end{figure}

\subsection{Crash-freeness}
Note that $\cfc$ represents two things: it can be a contract, as in $f
\in \cfc$ or a special formula in first-order logic $\cf{f}$.

We use $\bad$ to signal that something has gone wrong in the program :
it has crashed.

\begin{definition}[Crash]
A closed term $e$ crashed iff $e \to^* \bad$.
\end{definition}

\begin{definition}[Diverges]
A closed expression $e$ diverges iff either $e \to^* \unr$ or there is
no value $val$ such that $e \to^* val$
\end{definition}

\begin{definition}[Syntactic safety]
A (possibly open) expression $e$ is syntactically safe iff $\bad \not
\in_s e$. Similarly a context \cal C is syntactically safe iff $\bad
\not \in_s \mathcal{C}$.
\end{definition}

The notation $\bad \not \in e$ means that $\bad$ does not appear
anywhere in $e$, similarly for $\bad \not \in_s \mathcal{C}$. For
example, $Just 3$ is syntactically safe whereas $Just \bad$ is not.

\begin{definition}[Crash-free]
An expression $e$ is said to be crash-free iff 
$$\forall \mathcal{C}.~\bad \not \in_s \mathcal{C} \mbox{ and } \vdash
\mathcal{C} \llbracket e \rrbracket :: () \mathcal{C} \not \to^* \bad$$
\end{definition}
The notation $\mathcal{C} \llbracket e \rrbracket :: ()$ means that
$\mathcal{C} \llbracket e \rrbracket$ is closed and well-typed.  Note
that there are crash-free expression that are not syntactically safe,
for example $fst~(1,\bad)$.

\subsection{BAD and UNR}
Consider the following piece of code:
\begin{verbatim}
a = 0 + True

b ::: CF
b = undefined

c = error "foo"
\end{verbatim}

\begin{itemize}
\item $a$ is ill-typed
\item $b$'s implementation is not correct wrt its contract
\item $c$ goes through the whole toolchain (compiler, typechecker, contractchecker)
\end{itemize}

One thing to notice is that $a$ and $b$ are things that ``sould not
happen'' but are caught statically whereas $c$ should not happen but
can only be dealt with dynamically.

We can now define two types of problematic expressions: those that
cannot happen during a run of the program and those that can.
Expressions of the first type are called unreachable (and equated to
the special value $\unr$ in our first-order theory), whilst
expressions of the latter type are called bad (and equated to the
special value $\bad$).

We said earlier that we only considered syntactically correct and
well-typed programs as input. That implies that the ``$a$'' case
cannot happen. But given that our first-order logic is not typed, the
theorem prover may decide to instanciate a variable with an ill-typed
value! In order to prevent this, we will need to encode some basic
type-checking mecanism directly in our first-order theory.

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.
There are two different symbols for implication : $\to$ and
$\implies$. Altough they share the same semantics, their operational behaviour differs, as explained in \ref{guarded-pat}.

\begin{figure}
  \label{fol-stx}
  \centering
  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr & \\
    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \implies \phi \mid \phi \to \phi \mid \phi \leftrightarrow \phi & \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array}
  \caption{First-order logic syntax}
\end{figure}

\section{Translations}

\subsection{$\etrans{}$ -- Expressions}
Our most basic translation is from expressions in $\hprime$ to terms
in first-order logic. Given this translation we will be able to
translate definitions, data types and contracts to first-order
formulae. It is described in \ref{etrans}.

\begin{figure}
\label{etrans}
\begin{eqnarray*}
  \etrans{x} &=& x\\
  \etrans{f(e_1,\dots,e_n)} &=& f(\etrans{e_1},\dots,\etrans{e_n})\\
  \etrans{K(e_1,\dots,e_n)} &=& K(\etrans{e_1},\dots,\etrans{e_n})\\
  \etrans{\bad} &=& \bad
\end{eqnarray*}
\caption{$\etrans{}$ -- Expression translation}
\end{figure}

\subsection{$\dtrans{}$ -- Definitions}
We give in \ref{dtrans-reg} and \ref{dtrans-case} the two translations
of function definitions.

\ref{dtrans-reg} gives the translation of function not defined by
pattern matching, which is really easy: we just have to state the
equality between le left-hand side and the right-hand side.

Translating definitions that use pattern-matching is more challenging
and is described in \ref{dtrans-case}.

The first line says that when applied to an argument that matches a
pattern of the case expression, we should equate the function call to
the corresponding expression.

The second line states that if the pattern-matching failed or if we
pattern-matched on $\bad$ then the result should be $\unr$.

\begin{figure}
  \label{dtrans-reg}
  $$\dtrans{f(x_1,\dots,x_n) = e} = Y = f(X_1,\dots,X_n) \implies Y = \dtrans{e}$$
  \caption{$\dtrans{}$ -- Regular definitions}
\end{figure}

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_i(\overline{z})] = e_i } = \bigwedge_{1 \leq i \leq n} Y = f(K_i(x_1,\dots,x_{K_i})) \implies Y = \etrans{e_i} $$
  $$ \cup f(X_1,\dots,X_n) = Y \implies Y = \unr \lor \bigvee_{1 \leq i \leq n} \etrans{e} = K_i(sel_{1,K_i}(\etrans{e}),\dots,sel_{n_K,K_i}(\etrans{e})) \lor \etrans{e} = \bad $$
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

\subsection{$\ttrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in \ref{ttrans}
$$\ttrans{\mbox{data } T = K_1, \dots, K_n} = S_1 \cup S_2 \cup S_3 \cup S_4$$

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i)$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  being crash-free or not doesn't matter.
  \end{itemize}

\item $(S_4)$ None of the $K_i$ is unreachable.
\item One may want to also state that if $\vec{x} \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\cf{\vec{x}} \to \cf{K_i(\vec{x})}$.
\end{itemize}

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \bigwedge_{1 \leq i \leq n} sel_{i,K_j}(K_j(X_1,\dots,X_n)) = Y \implies Y = X_i  \mid 1 \leq j \leq n$$
  $$ S_2 := \forall x_1,\dots,x_{n_i}~\forall y_1,\dots,y_{n_j} . K_i(x_1,\dots,x_{n_i}) \neq K_j(y_1,\dots,y_{n_j}) \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{K_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n$$
  $$ S_4 := \forall x_1,\dots,x_n.~K_i(x_1,\dots,x_n) \neq \unr \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

\subsection{$\strans{}$ -- Contracts}
We give in \ref{strans} the translation of contract satisfaction.

\begin{figure*}
$$ \strans{f \in c} = \top \to f \in c \lor \bot $$

Given $r$ a fresh predicate and $a$ a fresh variable,
\begin{eqnarray*}
  p \to f \in \{x \mid b(x) \} \lor q &=& p \land f=X \implies b(X)=true \lor b(X) = \unr \lor q  \\
  p \to f \in x:c_1 \to c_2(x) \lor q &=&  p \to f(X) \in c_2(X) \lor r(X), r(X) \to X \not \in c_1 \lor q \\
  p \to f \in \cfc \lor q &=& p \land \forall x_1,\dots,x_n. \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
\end{eqnarray*}

\begin{eqnarray*}
  p \to f \not \in \{x \mid b(x) \} \lor q &=&  p \land f=X \implies b(X)=false \lor b(X) = \bad \lor q  \\
  p \to f \not \in x:c_1 \to c_2(x) \lor q &=&  p \to a \in c_1 \lor q, p \to f(a) \not \in c_2(a) \lor q \\
  p \to f \not \in \cfc \lor q &=&  p \land \forall x_1,\dots,x_n. \lnot \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
\end{eqnarray*}
\caption{$\strans{}$ -- Contract translation}
\label{strans}
\end{figure*}

\section{$\trans{}$ -- Checking a module}

\subsection{Prelude}
There are some formulae that should always be included in our FO theory.

We need to state that $\bad$ is not crash-free with the formula:
$\lnot \cf{\bad}$.

Plus we need to give formulae for the boolean datatype and for
 unreachability. Strictly speaking, we can omit them and just add the
following lines to source files:
\begin{verbatim}
data UNR = UNR
data Bool = True | False
\end{verbatim}
But given that those datatypes are used by our translation, we can
just directly include their translation every time we translate a
module.

\subsection{Contract checking -- Non-recursive case}
Input: a module $M$ that consists of a list of definitions, datatypes,
contracts and a contract $c$ for a non-recursive function $f$ this is
defined in $M$.

We say that the function implementation is correct wrt to its contract
iff $$\trans{M} \vdash \strans{f \in c}$$

\subsection{Contract checking -- Recursive case}
If the function $f=e$ is recursive, then we ask the theorem prover the
following:

$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \strans{f_p \in c} \vdash \strans{f \in c}$$

Where $M - f $ means the content of the module $M$ without $f$'s
definition and $f$'s contract. TODO Stress that it's not always enough
and that we may have to unroll several times!

\subsection{Module checking}
A module is a collection of function definitions, data definitions and
contracts. What we want to do is to check that functions satisfy their
contract(s).

\subsubsection{Naive example}
Here is a little example showing that we should be careful about which
formulae should belong to a theory.

Assume that we have a module that contains two functions defintion $f$
and $g$ and two contracts : $f \in \cfc$ and $g \in \cfc$. We assume
that those contracts do not hold, for example if $f$ is $head$ and $g$
is $last$.

First, we want to check $f$'s contract. So we ask the theorem prover
if
$$ \dtrans{f}, \dtrans{g}, \strans{g} \vdash \strans{f} $$

But, given that $g$'s contract does not hold, we can derive $\bot$ and
then prove that $f$'s contract hold.

For the same reason, we can prove that $g$ contract's holds, when in
fact it doesn't.

Finally, the user thinks he's done, but in fact he has proven nothing.

\subsubsection{The proper way to check a module}
Consider the following situation, where $a$'s definition relies on $f$
and $g$.
\begin{center}
\includegraphics[scale=0.5]{flow.png}
\end{center}

We should only include formulae that belongs to functions that are
actually used. For example, to prove $a$'s contract, we should only
include $f$ and $g$ translations, and we would ask equinox:
$$ \dtrans{f},\dtrans{g},\dtrans{a},\strans{f},\strans{g} \vdash
\strans{a}$$

\section{Correctness of the translation}
For the translation to be useful, $\trans{M} \vdash \strans{f \in c} $
should imply that $f \in c$.

\section{Guarded patterns}
It is our first encounter with the guarded-pattern operator
($\implies$). 
It syntax is:
$$ s_1 = t_1 \land \dots \land s_k = t_k \implies v_1 = w_1 \lor \dots \lor v_n = w_n $$

$s_i,t_i,v_i,w_i$ are terms, and all the variables on the right-hand
side should appear somewhere on the left-hand side.

The semantics are: if during the proof by the theorem prover, we have
equations that match the LHS than the theorem prover should strongly
consider the equations on the RHS. Those semantics are sound (we can
interpret $\implies$ as normal implication) but if modifies the
operational behabviour of Equinox, because it tells it very finely
when to use equations and how to instantiate them.

\section{Higher-orderness}
\label{ho}
Our current translation only considers fully applied functions and
first-order functions. For example, we cannot translate any contract
for $map$ because it would involve quantifying over a function, a
thing that is not first-order.

There is a possible workaround, which involves the ``app'' function in
first-order logic. Assume we have a function $f$ that is not fully
applied somewhere in a module. We create the term $f\_ptr$ which
relates to $f$ by the equations given in \ref{ho-fig}

This way, we can emulate quantification over function by
quantification on their $ptr$ counterpart.

\begin{figure*}
\label{ho-fig}
$$\etrans{e_1~e_2} = app(e_1,e_2)$$
$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
$$ \cf{f\_ptr} \iff \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \implies \cf{f(x_1,\dots,x_n)}$$
$$\forall f\_ptr,x~\cf{f\_ptr} \land \cf{x} \implies \cf{app(f\_ptr,x)}$$
\caption{Encoding of higher-orderness}
\end{figure*}

\section{Experiments}
That's how we roll.

\begin{figure*}
\label{comparison}
\begin{center}
    \begin{tabular}{l|c|c|c|c|c}
      Problem & Equinox & Equinox (+ weak) & SPASS & Vampire & E \\
      \hline
      Add.hs & 0.25 & 0.08 & 0.04 & 0.12 & 0.05\\
      BinaryTree.hs & 0.45 & 0.2 & 0.04 & 0.01 & 0.04 \\
      Branch.hs & 0.27 & 0.40 & 0.04 & 0.01 & 0.03 \\
      Copy.hs & 0.86 & 0.09 & 0.03 & 0.01 & 184.3 \\
      Head.hs & 0.32 & 0.29 & 0.03 & 0.03 & 4.2 \\
      Implies.hs & 3.24 & 0.32 & 0.06 & 0.02 & 0.11 \\
      Map.hs & 2.47 & 0.14 & 0.92 & 1.02 & $>$300 \\
      Mult.hs & $>$300 & 0.41 & 0.05 & 0.22 & 11.71 \\
      Multgt.hs & $>$300 & 1.24 & 0.62 & 1.31 & $>$300 \\
      NatEq.hs & 203.12 & 0.33 & 0.02 & 0.03 & 0.343 \\
      Odd.hs & 0.42 & 1.17 & 0.06 & 0.03 & $>$300 \\
      Reverse.hs & 72.32 & 0.12 & 0.05 & 0.02 & 0.038 \\
      Simple.hs & 0.07 & 0.04 & 0.01 & 0.01 & 0.022 \\
      Test.hs & 7.76 & 2.86 & 0.08 & 0.05 & $>$300 \\
      Test2.hs & 5.63 & 0.09 & 0.07 & 0.01 & 1.02 \\
    \end{tabular}
\end{center}
\caption{Comparison (in seconds) with other theorem provers}
\end{figure*}

\end{document}

### 2011-08-17 paper.tex
introduction

@@
 We start by stating that $Nil$ and $Cons$ can never be equal:
-$$\forall a,b.~Cons(a,b) \neq Nil$$
+\begin{equation*}
+\forall a,b.~Cons(a,b) \neq Nil
+\end{equation*}
 
@@
 
-$$\cf{Nil}$$
-$$\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}$$
+\begin{equation*}
+\cf{Nil}
+\end{equation*}
+\begin{equation}
+  \label{cfcons}
+\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}
+\end{equation}
 
@@
 good way to explain it right now.
-$$\forall y,ys.~Cons(y,ys) \neq \unr $$
-$$Nil \neq \unr $$
+\begin{equation*}
+\forall y,ys.~Cons(y,ys) \neq \unr 
+\end{equation*}
+\begin{equation*}
+Nil \neq \unr 
+\end{equation*}
 
@@
 necessary, but it will be handy:
-$$\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y $$
-$$\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys $$
+\begin{equation**}
+\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
+\end{equation**}
+\begin{equation*}
+\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
+\end{equation*}
 
@@
 logic.
-$$\forall xs.~ xs=Nil \implies notnull(xs) = false$$
-$$\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true$$
+\begin{equation*}
+\forall xs.~ xs=Nil \implies notnull(xs) = false
+\end{equation*}
+
+\begin{equation}
+  \label{defnotnull}
+\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true
+\end{equation}
 
@@
 $\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
-$$\forall xs.~xs=\bad \to notnull(xs) = \bad$$
+\begin{equation*}
+\forall xs.~xs=\bad \to notnull(xs) = \bad
+\end{equation*}
 
@@
 not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
-$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
-$$\land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
-$$ notnull(X) = \unr$$
+\begin{align}
+  \label{unrnotnull}
+  \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
+  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
+  notnull(X) = \unr 
+\end{align}
 
 The translation of $head$ follows the same pattern:
-$$\forall xs.~ x=Nil \implies head(xs) = \bad$$
-$$\forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y$$
+\begin{align}
+  \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
+  \forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
+    \forall xs.~xs=\bad \to head(xs) = \bad \notag
+\end{align}
 
-$$\forall xs.~xs=\bad \to head(xs) = \bad$$
-
-$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
-$$ \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
-$$ head(X) = \unr$$
+\begin{align*}
+ \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
+ \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
+ head(X) = \unr 
+\end{algin*}
 
@@
 the theory $T$. We translate separatly the contract:
-$$ \phi := \forall xs.~\cf{xs} \land not(null(xs)) = true \to \cf{head(xs)} $$
+\begin{equation*}
+  \phi := \forall xs.~\cf{xs} \land notnull(xs) = true \to \cf{head(xs)} 
+\end{equation*}
 
@@
 
-Intuitivly, $T$ is consistent (ie $T \not \vdash \bot$), because each
+Intuitively, $T$ is consistent (ie $T \not \vdash \bot$), because each
 formula serves a specific purpose. Now, assume that $xs$ satisfies
 $\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
-because we have $\cf{xs}$ and $\not \cf{\bad}$. The constraint
+because we have $\cf{xs}$ and $\lnot \cf{\bad}$. The constraint
 $notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
-some $y$ and $ys$. But the translation of $notnull$ is total, ie the
-function is totally defined. This implies that there exist $y$ and
-$ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can now
-derive $\cf{y}$ and $\cf{ys}$. But $head(xs) = y$ in our setting, and
-$y$ is crash-free. QED.
+some $y$ and $ys$. But $notnull$ is totally defined, because of
+\eqref{unrnotnull}. This implies (by \eqref{defnotnull}) that there exist
+$y$ and $ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can
+now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head(xs) =
+y$ because of \eqref{defhead}, and $y$ is crash-free, so we can finally
+derive $\cf{head(xs)}$. QED.
 

### 2011-08-17 paper.tex
New version with

@@
 
+\newcommand{\tot}{\leftrightarrow}
+
 \newcommand{\unr}{\texttt{UNR}}
@@
 \newcommand{\hprime}{\mathcal{H}'}
+\newcommand{\cfc}{\texttt{CF}}
 \newcommand{\cf}[1]{\texttt{CF}(#1)}
-\newcommand{\cfc}{\texttt{CF}}
+\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
+\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}
 
@@
 We use first-order logic with equality, defined in figure \ref{fol-stx}.
-There are two different symbols for implication : $\to$ and
-$\implies$. Altough they share the same semantics, their operational behaviour differs, as explained in \ref{guarded-pat}.
 
@@
     \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
-    && \mid \phi \land \phi \mid \phi \implies \phi \mid \phi \to \phi \mid \phi \leftrightarrow \phi & \\
+    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi & \\
     \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
@@
   \label{dtrans-reg}
-  $$\dtrans{f(x_1,\dots,x_n) = e} = Y = f(X_1,\dots,X_n) \implies Y = \dtrans{e}$$
+  $$\dtrans{f(\vec{x}) = e}$$
+  $$ \forall \vec{x},y.~ \min{y} \land y = f(\vec{x}) \to y = \etrans{e}$$
   \caption{$\dtrans{}$ -- Regular definitions}
@@
   \label{dtrans-case}
-  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_i(\overline{z})] = e_i } = \bigwedge_{1 \leq i \leq n} Y = f(K_i(x_1,\dots,x_{K_i})) \implies Y = \etrans{e_i} $$
-  $$ \cup f(X_1,\dots,X_n) = Y \implies Y = \unr \lor \bigvee_{1 \leq i \leq n} \etrans{e} = K_i(sel_{1,K_i}(\etrans{e}),\dots,sel_{n_K,K_i}(\etrans{e})) \lor \etrans{e} = \bad $$
+  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
+  \begin{center}
+    \begin{array}{rccccccl}
+      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
+                                               &&&&&& \cdots & \\
+                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
+                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
+                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
+    \end{array}
+    \end{center}
   \caption{$\dtrans{}$ -- Case definitions}
@@
 \item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i)$
+  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
   on its $k$-th component.
@@
   \label{ttrans}
-  $$ S_1 := \bigwedge_{1 \leq i \leq n} sel_{i,K_j}(K_j(X_1,\dots,X_n)) = Y \implies Y = X_i  \mid 1 \leq j \leq n$$
-  $$ S_2 := \forall x_1,\dots,x_{n_i}~\forall y_1,\dots,y_{n_j} . K_i(x_1,\dots,x_{n_i}) \neq K_j(y_1,\dots,y_{n_j}) \mid 1 \leq i < j \leq n$$
-  $$ S_3 := \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{K_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n$$
-  $$ S_4 := \forall x_1,\dots,x_n.~K_i(x_1,\dots,x_n) \neq \unr \mid 1 \leq i \leq n$$
+  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
+  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
+  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
+  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
   \caption{$\ttrans{}$ -- Data type translation}
@@
 We give in \ref{strans} the translation of contract satisfaction.
+$true$ refers to the translation to a term of the data constructor
+\texttt{True} in $\hprime$, not to the actual true value.
 
 \begin{figure*}
-$$ \strans{f \in c} = \top \to f \in c \lor \bot $$
-
-Given $r$ a fresh predicate and $a$ a fresh variable,
 \begin{eqnarray*}
-  p \to f \in \{x \mid b(x) \} \lor q &=& p \land f=X \implies b(X)=true \lor b(X) = \unr \lor q  \\
-  p \to f \in x:c_1 \to c_2(x) \lor q &=&  p \to f(X) \in c_2(X) \lor r(X), r(X) \to X \not \in c_1 \lor q \\
-  p \to f \in \cfc \lor q &=& p \land \forall x_1,\dots,x_n. \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
+  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
+  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
+  \strans{e ::: \cfc} &=& \cf{e}
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  p \to f \not \in \{x \mid b(x) \} \lor q &=&  p \land f=X \implies b(X)=false \lor b(X) = \bad \lor q  \\
-  p \to f \not \in x:c_1 \to c_2(x) \lor q &=&  p \to a \in c_1 \lor q, p \to f(a) \not \in c_2(a) \lor q \\
-  p \to f \not \in \cfc \lor q &=&  p \land \forall x_1,\dots,x_n. \lnot \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
+  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
+  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
+  \strans{e :/: \cfc} &=& ~\cf{e}
 \end{eqnarray*}
+
+$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
+$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
+$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$
+
 \caption{$\strans{}$ -- Contract translation}
@@
 
-\section{Guarded patterns}
-It is our first encounter with the guarded-pattern operator
-($\implies$). 
-It syntax is:
-$$ s_1 = t_1 \land \dots \land s_k = t_k \implies v_1 = w_1 \lor \dots \lor v_n = w_n $$
-
-$s_i,t_i,v_i,w_i$ are terms, and all the variables on the right-hand
-side should appear somewhere on the left-hand side.
-
-The semantics are: if during the proof by the theorem prover, we have
-equations that match the LHS than the theorem prover should strongly
-consider the equations on the RHS. Those semantics are sound (we can
-interpret $\implies$ as normal implication) but if modifies the
-operational behabviour of Equinox, because it tells it very finely
-when to use equations and how to instantiate them.
-
 \section{Higher-orderness}
@@
 Our current translation only considers fully applied functions and
-first-order functions. For example, we cannot translate any contract
-for $map$ because it would involve quantifying over a function, a
-thing that is not first-order.
+first-order functions. For example, so far, we cannot give any
+contract to $map$ because it would involve quantifying over a
+function, a thing that is not first-order.
 
-There is a possible workaround, which involves the ``app'' function in
-first-order logic. Assume we have a function $f$ that is not fully
-applied somewhere in a module. We create the term $f\_ptr$ which
-relates to $f$ by the equations given in \ref{ho-fig}
+There is a possible workaround, which involves the ``app'' function
+defined in our first-order logic. Assume we have a function $f$ that
+is not fully applied somewhere in a module. We create the term
+$f\_ptr$ which relates to $f$ by the equations given in \ref{ho-fig}
 
-This way, we can emulate quantification over function by
-quantification on their $ptr$ counterpart.
+This way, we can emulate quantification over function by quantifying
+on their $ptr$ counterpart.
 
@@
 $$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
-$$ \cf{f\_ptr} \iff \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \implies \cf{f(x_1,\dots,x_n)}$$
-$$\forall f\_ptr,x~\cf{f\_ptr} \land \cf{x} \implies \cf{app(f\_ptr,x)}$$
+$$ \cf{f\_ptr} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{f(x_1,\dots,x_n)}$$
+$$\forall f\_ptr,x.~\cf{f\_ptr} \land \cf{x} \to \cf{app(f\_ptr,x)}$$
 \caption{Encoding of higher-orderness}

### 2011-08-17 contracts/FOL.hs
Paper fix (Koen)

@@ -25,6 +25,7 @@ data MetaTerm a = Var a
                 | App [MetaTerm a]
                 | FullApp a [MetaTerm a]
                 | Weak (MetaTerm a)
+                | Min (MetaTerm a)
                 deriving (Eq,Functor)
 
 instance Show a => Show (MetaTerm a) where
@@ -33,6 +34,7 @@ instance Show a => Show (MetaTerm a) where
   show (App [t]) = show t
   show (App ts) = "app(" ++ show (App (init ts)) ++ "," ++ show (last ts) ++ ")"
   show (Weak v) = show v
+  show (Min v) = show v
   show (FullApp f []) = show f
   show (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"
 
@@ -107,6 +109,7 @@ auxUpper c (Var v) = Var v
 auxUpper c (App ts) = App $ map (auxUpper c) ts
 auxUpper c (FullApp x ts) = FullApp x (map (auxUpper c) ts) -- TODO think about it
 auxUpper c (Weak t) = Weak $ auxUpper c t
+auxUpper c (Min t) = Min $ auxUpper c t
 
 toTPTP :: Formula -> String
 toTPTP f = header ++ "\n" ++ (go $ upperIfy [] f) ++ "\n" ++ footer
@@ -131,6 +134,7 @@ toTPTP f = header ++ "\n" ++ (go $ upperIfy [] f) ++ "\n" ++ footer
         goTerm (FullApp f []) = show f
         goTerm (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"
         goTerm (Weak t) = "$weak(" ++ goTerm t ++")"
+        goTerm (Min t) = "$min(" ++ goTerm t ++")"
 
 
 
@@ -142,14 +146,16 @@ appifyF a fs = map (fmap go) fs
           Just n -> Var (Regular $ v ++ "_ptr")
           Nothing -> Var $ Regular v
         go (App ts) = App (map go ts)
-        go (FullApp f ts) = FullApp f (map go ts)
+        go (FullApp f ts) = FullApp f (map go ts) 
         go (Weak t) = Weak $ go t
+        go (Min t) = Min $ go t
         go t = t
         trim = filter (\s -> case s of H.Fun _ _ -> True; _ -> False)
         
 removeWeakAnnotations :: Formula -> Formula
 removeWeakAnnotations = fmap go
[…]

### 2011-08-17 contracts/Translation.hs

@@ -1,6 +1,5 @@
 module Translation where
 
-import Debug.Trace
 import qualified Haskell as H
 import qualified FOL as F
 import FOL (MetaFormula(..))

### 2011-08-17 contracts/FOL.hs
+ prefixWithAorC (but probably not the good semantics)

@@ -86,9 +86,9 @@ simplify f = filter (/= Top) $ splitOnAnd $ removeConstants f
 extractVR (Var (Regular x)) = x
 
 
--- The TPTP format says that only quantified variables should be uppercase
--- So we make everything lowercase (it's supposed to be already done)
--- And upperIfy makes the good variables uppercase.
+-- The TPTP spec says that only quantified variables should be
+-- uppercase So we make everything lowercase (it's supposed to be
+-- already done) And upperIfy makes the good variables uppercase.
 upperIfy :: [String] -> Formula -> Formula
 upperIfy c (Forall xs f) = Forall (map (Var . Regular . map toUpper . extractVR) xs) (upperIfy c' f)
   where c' = (map extractVR xs)++c

### 2011-08-17 contracts/Haskell.hs

@@ -2,11 +2,12 @@
 
 module Haskell where
 
+import Data.Maybe (isJust)
+
 type Variable = String
 type Constructor = String
 
 -- All this meta stuff is just here to allow me to derive functors for free.
-
 type DataType = MetaDataType Expression
 type Expression = MetaExpression Variable
 type DefGeneral = MetaDefGeneral Expression
@@ -70,7 +71,15 @@ appify :: Program -> Program
 appify p = map (fmap $ appifyExpr a) p 
   where a = arities p
 
-
+prefixWithAorC :: Program -> Program
+prefixWithAorC p = (map $ fmap $ fmap prefixVar) p
+  where as = arities p
+        prefixVar v = case lookupT v as of
+          Nothing -> v
+          Just _  -> if v `isConstructor` as then "c_"++v else "a_"++v
+        isConstructor v = any (\f -> case f of Fun _ _ -> False ; _ -> True)
+        
+lookupT :: String -> [Type Variable] -> Maybe Int
 lookupT v [] = Nothing
 lookupT v (Fun f n:as)  = if f == v  then Just n else lookupT v as
 lookupT v (Cons f n:as) =  if f == v  then Just n else lookupT v as

### 2011-08-17 contracts/Translation.hs

@@ -70,7 +70,6 @@ dTrans (H.LetCase f vs e pes) = do
 sTrans e H.Any = return [Top]
 
 sTrans e (H.Pred x u) =  do
-  a <- liftM arities get
   let  u' = H.subst e x u
   et' <- eTrans e 
   ut' <- eTrans u'
@@ -81,9 +80,10 @@ sTrans e (H.Pred x u) =  do
   return $ [F.And $ [F.Or [(et :=: F.Var F.UNR) ,F.And [F.Var F.BAD :/=: ut' , ut' :/=: (F.Var $ F.Regular "false")]]]] -- The data constructor False.
 
 sTrans e (H.AppC x c1 c2) = do
-  S s k a <- get
-  put $ S s (k+1) a
-  let freshX = s++(show k) 
+  s <- get
+  let k = count s
+  put $ s {count = k + 1}
+  let freshX = (prefix s)++(show $ k) 
       c2' = H.substC (H.Var freshX) x c2
   [f1] <- sTrans (H.Var freshX) c1
   [f2] <- case e of 
@@ -178,7 +178,7 @@ isToCheck fs (H.ContSat (H.Satisfies f _)) = f `elem` fs
 isToCheck _ _                              = False
 
 trans :: H.Program -> [H.Variable] -> [F.Formula]
-trans ds fs = evalState (go fs (H.appify ds)) (S "Z" 0 (H.arities ds))
+trans ds fs = evalState (go fs ((H.appify) ds)) (S "Z" 0 (H.arities ds))
   where go fs ds = do 
           let (toCheck,regDefs) = partition (isToCheck fs) ds
               recVar x = x ++ "p"

### 2011-08-17 contracts/FOL.hs
min version, first take. Shouldn't work

@@ -25,7 +25,6 @@ data MetaTerm a = Var a
                 | App [MetaTerm a]
                 | FullApp a [MetaTerm a]
                 | Weak (MetaTerm a)
-                | Min (MetaTerm a)
                 deriving (Eq,Functor)
 
 instance Show a => Show (MetaTerm a) where
@@ -34,7 +33,6 @@ instance Show a => Show (MetaTerm a) where
   show (App [t]) = show t
   show (App ts) = "app(" ++ show (App (init ts)) ++ "," ++ show (last ts) ++ ")"
   show (Weak v) = show v
-  show (Min v) = show v
   show (FullApp f []) = show f
   show (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"
 
@@ -52,6 +50,7 @@ data MetaFormula a = Forall [a] (MetaFormula a)
                    | a :=: a
                    | a :/=: a
                    | CF a
+                   | Min a
                    deriving (Show,Eq,Functor)
 
 
@@ -100,6 +99,7 @@ upperIfy c Bottom = Bottom
 upperIfy c (t1 :=: t2) = (auxUpper c t1) :=: (auxUpper c t2)
 upperIfy c (t1 :/=: t2) = (auxUpper c t1) :/=: (auxUpper c t2)
 upperIfy c (CF t) = CF (auxUpper c t)
+upperIfy c (Min t) = Min (auxUpper c t)
 upperIfy c (And fs) = And (map (upperIfy c) fs)
 upperIfy c (Or fs) = Or (map (upperIfy c) fs)
   
@@ -109,7 +109,6 @@ auxUpper c (Var v) = Var v
 auxUpper c (App ts) = App $ map (auxUpper c) ts
 auxUpper c (FullApp x ts) = FullApp x (map (auxUpper c) ts) -- TODO think about it
 auxUpper c (Weak t) = Weak $ auxUpper c t
-auxUpper c (Min t) = Min $ auxUpper c t
 
 toTPTP :: Formula -> String
 toTPTP f = header ++ "\n" ++ (go $ upperIfy [] f) ++ "\n" ++ footer
@@ -127,6 +126,7 @@ toTPTP f = header ++ "\n" ++ (go $ upperIfy [] f) ++ "\n" ++ footer
         go (t1 :=: t2) = goTerm t1 ++ " = " ++ goTerm t2
         go (t1 :/=: t2) = goTerm t1 ++ " != " ++ goTerm t2
[…]

### 2011-08-17 contracts/Haskell.hs

@@ -30,6 +30,7 @@ data MetaDefGeneral a = ContSat (MetaContSat a)
 
 
 data MetaContSat a = Satisfies Variable (MetaContract a)
+                   | DoesntSatisfy Variable (MetaContract a)
                    deriving (Show,Eq,Functor,Ord)
                
 data MetaDefinition a = Let Variable [Variable] a
@@ -44,7 +45,6 @@ data MetaContract a = AppC Variable (MetaContract a) (MetaContract a) -- x : c -
                     | And (MetaContract a) (MetaContract a)
                     | Or  (MetaContract a) (MetaContract a)
                     | CF
-                    | Any
                     deriving (Show,Eq,Functor,Ord)
 
 data Type a = Fun a Int
@@ -126,7 +126,6 @@ substC x y (Pred u e)     = if u/=y then Pred u (subst x y e) else (Pred u e)
 substC x y (And c1 c2)    = And (substC x y c1) (substC x y c2)
 substC x y (Or c1 c2)     = Or (substC x y c1) (substC x y c2)
 substC x y CF             = CF 
-substC _ _ Any            = Any
 
 
 -- FIXME: it's needed to compile but utterly useless it's needed

### 2011-08-17 contracts/Translation.hs

@@ -34,7 +34,9 @@ eTrans H.BAD = return $ F.Var F.BAD
 
 dTrans (H.Let f vs e) = do
   et <- eTrans e
-  return $ [F.Forall vvs $ (F.FullApp (F.Regular f) vvs) :=: (F.Weak $ et),fptr1,fptr2,fptr3]
+  s <- get
+  let freshT = F.Var $ F.Regular $ prefix s ++ (show $ count s)
+  return $ [F.Forall (freshT:vvs) $ F.And [F.Min freshT, freshT :=: F.FullApp (F.Regular f) vvs ] :=>: freshT :=: (F.Weak $ et),fptr1,fptr2,fptr3]
   where vvs = map (F.Var . F.Regular) vs
         fptr1 = (F.Forall vvs $ (F.And [F.CF v | v <- vvs]) :=>: F.CF (F.FullApp (F.Regular f) vvs)) :<=>: (F.CF $ F.Var $ F.Regular (f++"_ptr"))
         fptr2 = F.Forall vvs $ (F.FullApp (F.Regular f) vvs) :=: (F.Weak (F.App $ (F.Var . F.Regular) (f++"_ptr") : vvs))
@@ -43,22 +45,28 @@ dTrans (H.Let f vs e) = do
 dTrans (H.LetCase f vs e pes) = do
   et <- eTrans e
   ft <- eTrans $ H.Var f
+  s <- get
+  let freshT = F.Var $ F.Regular $ prefix s ++ (show $ count s)
   let zedify ei pi = foldl (\e (v,z) -> H.subst (H.Var $ extractVR z) v e) ei (take (length (tail pi)) $ zip (tail pi) zs)
       extractVR (F.Var (F.Regular v)) = v 
       arities = map (\p -> (head p, length $ tail p)) $ map fst pes :: [(String,Int)]
       zs = [F.Var $ F.Regular $ "Zdef" ++ show x | x <- [1..(foldl1 max [snd y | y <- arities])]]
   tpieis <- sequence [eTrans (zedify ei pi) | (pi,ei) <- pes]
   let vvs = map (F.Var . F.Regular) vs
[…]

### 2011-08-19 paper.tex
Paper fix (Koen)
min version, first take. Shouldn't work
Paper update, mostly
Ttrans (datatype translation) becomes Ktrans (because we call the data constructors K anyway)
Strans (contract translation) becomes Ctrans (because s stood for satisfies but it's not obvious)

@@
 \newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
-\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
-\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
+\newcommand{\ktrans}[1]{\mathcal{K} \llbracket #1 \rrbracket}
+\newcommand{\ctrans}[1]{\mathcal{C} \llbracket #1 \rrbracket}
 \newcommand{\trans}[1]{\llbracket #1 \rrbracket}
@@
 \newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
-\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}
+\renewcommand{\min}[1]{\mbox{min}(#1)}
 
@@
 \section{Translations}
+For an overview of the different translations we define, see \ref{trans}
+
+\begin{figure}
+  \label{trans}
+  \begin{center}
+  \begin{array}{rcl}
+    \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
+    \dtrans{f~\vec{x} = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
+    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
+    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
+  \end{array}
+  \end{center}
+  \caption{Translations}
+\end{figure}
 
@@
   \begin{center}
-    \begin{array}{rccccccl}
-      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
-                                               &&&&&& \cdots & \\
-                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
-                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
-                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
+    \begin{array}{rclcl}
+      \forall \vec{a},y.~\min{y} \land f(\vec{a}) = y &\to& (\multicolumn{3}{c}{\min{e} \land (e = \bad \lor \bigvee_i \forall \vec{x}.~e = K_i(\vec{x}) \lor y = \unr)} \\      
+      &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& y = \etrans{e_1} \\
+      &\land& & \cdots & \\
+      &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& y = \etrans{e_n}) \\
     \end{array}
@@
 
-\subsection{$\ttrans{}$ -- Datatypes}
-We break down the translation for datatypes in four parts, described in \ref{ttrans}
-$$\ttrans{\mbox{data } T = K_1, \dots, K_n} = S_1 \cup S_2 \cup S_3 \cup S_4$$
+\subsection{$\ktrans{}$ -- Datatypes}
+We break down the translation for datatypes in four parts, described in \ref{ktrans}
+$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 
 \begin{itemize}
-\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
+\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
   $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
   on its $k$-th component.
-\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
+\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, we state that they can
   never map to the same value.
-\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
+\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
   Notice that we have a equivalence.
@@
 
-\item $(S_4)$ None of the $K_i$ is unreachable.
+\item $(\Phi_4)$ None of the $K_i$ is unreachable.
 \item One may want to also state that if $\vec{x} \neq \bad$ then
@@
 \begin{figure*}
-  \label{ttrans}
-  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
-  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
-  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
-  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
-  \caption{$\ttrans{}$ -- Data type translation}
+  \label{ktrans}
+  $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
+  where
+  \begin{center}
+    \begin{array}{rcl}
+      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
+      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land K_i(\vec{x}) = a \land K_j(\vec{x}) = a) \\
+      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~\min{a} \land a = K_i(\vec{x}) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})}) \\
+      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
+    \end{array}
+  \end{center}
+  \caption{$\ktrans{}$ -- Data type translation}
 \end{figure*}
 
-\subsection{$\strans{}$ -- Contracts}
-We give in \ref{strans} the translation of contract satisfaction.
+\subsection{$\ctrans{}$ -- Contracts}
+We give in \ref{ctrans} the translation of contract satisfaction.
 $true$ refers to the translation to a term of the data constructor
@@
 
+Note that we define the translation of $f \in c$ and of $f \not \in
+c$. We have to do that because one is not the negation of the other,
+even though $\lnot (f \not \in c)$ implies $f \in c$.
+
 \begin{figure*}
 \begin{eqnarray*}
-  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
-  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
-  \strans{e ::: \cfc} &=& \cf{e}
+  \ctrans{e \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{e} = \unr)\\
+  \ctrans{e \in x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{e(x)}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{\etrans{e(x)} \in c_2(x)})\\
+  \ctrans{e \in \cfc} &=& \cf{\etrans{e}}
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
-  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
-  \strans{e :/: \cfc} &=& ~\cf{e}
+  \ctrans{e \not \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = false \lor \etrans{e} = \bad)\\
+  \ctrans{e \not \in x:c_1 \to c_2(x)} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{\etrans{e(x)} \not \in c_2(x)}\\
+  \ctrans{e \not \in \cfc} &=& \lnot \cf{\etrans{e}}
 \end{eqnarray*}
 
-$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
-$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
-$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$
+\begin{center}
+\begin{array}{rcl}
+ \ctrans{e \in c_1 \&\& c_2} &=& \ctrans{e \in c_1} \land \ctrans{e \in c_2} \\
+ \ctrans{e \not \in c_1 \&\& c_2} &=& \ctrans{e \not \in c_1} \lor \ctrans{e \not \in c_2}\\
+ \ctrans{e \in c_1 || c_2} &=& \ctrans{e \in c_1} \lor \ctrans{e \in c_2} \\
+  \ctrans{e \not \in c_1 || c_2} &=& \ctrans{e \not \in c_1} \land \ctrans{e \not \in c_2}\\
+ \ctrans{(a,b) \in (c_1,c_2)}   &=& \ctrans{a \in c_1} \land \ctrans{b \in c_2} \\
+ \ctrans{(a,b) \not \in (c_1,c_2)}   &=& \ctrans{a \not \in c_1} \lor \ctrans{b \not \in c_2} \\
+\end{array}
+\end{center}
 
-\caption{$\strans{}$ -- Contract translation}
-\label{strans}
+\caption{$\ctrans{}$ -- Contract translation}
+\label{ctrans}
 \end{figure*}
@@
 We say that the function implementation is correct wrt to its contract
-iff $$\trans{M} \vdash \strans{f \in c}$$
+iff $$\trans{M} \vdash \ctrans{f \in c}$$
 
@@
 
-$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \strans{f_p \in c} \vdash \strans{f \in c}$$
+$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \ctrans{f_p \in c} \vdash \ctrans{f \in c}$$
 
@@
 if
-$$ \dtrans{f}, \dtrans{g}, \strans{g} \vdash \strans{f} $$
+$$ \dtrans{f}, \dtrans{g}, \ctrans{g} \vdash \ctrans{f} $$
 
@@
 include $f$ and $g$ translations, and we would ask equinox:
-$$ \dtrans{f},\dtrans{g},\dtrans{a},\strans{f},\strans{g} \vdash
-\strans{a}$$
+$$ \dtrans{f},\dtrans{g},\dtrans{a},\ctrans{f},\ctrans{g} \vdash
+\ctrans{a}$$
 
 \section{Correctness of the translation}
-For the translation to be useful, $\trans{M} \vdash \strans{f \in c} $
+For the translation to be useful, $\trans{M} \vdash \ctrans{f \in c} $
 should imply that $f \in c$.

### 2011-08-19 contracts/Translation.hs

@@ -61,7 +61,7 @@ dTrans (H.LetCase f vs e pes) = do
 
 --      eq3 = (F.And $ (et :/=: F.Var F.BAD):bigAndSel ) :=>: eq4
       eq4 = (F.FullApp (F.Regular f) vvs :=: F.Var F.UNR)
-      bigAndSel = [et :/=: F.Weak (F.FullApp (F.Regular di) [F.FullApp (F.Regular ("sel_"++(show i)++"_"++di)) [et] | i <- [1..ai]]) | (di,ai) <- arities]
+--      bigAndSel = [et :/=: F.Weak (F.FullApp (F.Regular di) [F.FullApp (F.Regular ("sel_"++(show i)++"_"++di)) [et] | i <- [1..ai]]) | (di,ai) <- arities]
       fptr1 = (F.Forall vvs $ (F.And [F.CF v | v <- vvs]) :=>: F.CF (F.FullApp (F.Regular f) vvs)) :<=>: (F.CF $ F.Var $ F.Regular (f++"_ptr"))
       fptr2 = F.Forall vvs $ (F.FullApp (F.Regular f) vvs) :=: (F.Weak (F.App $ (F.Var . F.Regular) (f++"_ptr") : vvs))
       fptr3 = F.Forall vvs $ (F.FullApp (F.Regular (f ++ "p")) vvs) :=: (F.Weak (F.App $ (F.Var . F.Regular) (f++"p_ptr") : vvs))
@@ -230,8 +230,6 @@ trans ds fs = evalState (go fs ((H.appify) ds)) (S "Z" 0 (H.arities ds))
   where go fs ds = do 
           booleanPrelude <- tTrans $ H.Data "Bool" [("true",0,H.ok),("false",0,H.ok)]
           let prelude = [(F.Forall (map (F.Var . F.Regular) ["F","X"]) $ (F.And [F.CF $ F.Var $ F.Regular "X", F.CF $ F.Var $ F.Regular "F"]) :=>: (F.CF $ (F.App [(F.Var $ F.Regular "F"), (F.Var $ F.Regular "X")])))] ++ booleanPrelude
-                        --   ,F.Not $ F.CF $ F.Var $ F.BAD,F.CF $ F.Var $ F.UNR,(F.Var $ F.Regular "false") :/=: (F.Var $ F.Regular "true")
[…]

### 2011-08-22 paper.tex
paper fix

--- 2011-08-17 paper.tex
@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{amssymb}
@@
 
+\newtheorem{definition}{Definition}
+
 \newcommand{\tot}{\leftrightarrow}
@@
 \newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
-\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}
+\renewcommand{\min}[1]{\mbox{min}(#1)}
 
 \title{Static contract checking for Haskell}
+\authorinfo{Simon, Koen, Dimitrios, Charles-Pierre}
 \maketitle
@@
 necessary, but it will be handy:
-\begin{equation**}
+\begin{equation*}
 \forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
-\end{equation**}
+\end{equation*}
 \begin{equation*}
@@
  head(X) = \unr 
-\end{algin*}
+\end{align*}
 
@@
   can always lift them to the toplevel.
-\item Until \label{ho} we will only consider full application of
+\item Until \ref{ho} we will only consider full application of
   functions ($f(x,y)$), in order to remove clutter. Dealing with
@@
   \centering
-  \begin{array}{rclr}
+  \[  \begin{array}{rclr}
     mod &:=& def_1,\dots,def_n &\\
@@
        \end{array} & \\
-  \end{array}
+  \end{array} \]
 
-  \begin{array}{rclr}
+  \[  \begin{array}{rclr}
     x,y,f,g,a,b & \in & \mbox{Variables} \\
@@
     K &\in& \mbox{Data Constructors} \\
-  \end{array}
+  \end{array} \]
 
-  \begin{array}{rclr}
+  \[  \begin{array}{rclr}
     e &\in& \mbox{Expressions} & \\
@@
     &\mid& K(e,\dots,e) & \\
-  \end{array}
+  \end{array} \]
 
@@
   \centering 
-  \begin{array}{rclr}
+  \[  \begin{array}{rclr}
   c &:=& x:c \to c\\
@@
   &\mid& \cfc \\
-  \end{array}
+  \end{array} \]
   \caption{Contract syntax}
@@
   \centering
-  \begin{array}{rcl}
+  \[  \begin{array}{rcl}
     e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
@@
     e \in \cfc &\iff& e \mbox{ is crash-free}
-  \end{array}
+  \end{array} \]
   \caption{Semantics of contract satisfaction}
@@
   \centering
-  \begin{array}{rcl}
+  \[  \begin{array}{rcl}
     v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
-    && \mid \bad \mid \unr & \\
+    && \mid \bad \mid \unr \\
     \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
-    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi & \\
+    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi  \\
     \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
-  \end{array}
+  \end{array} \]
   \caption{First-order logic syntax}
@@
   \begin{center}
-    \begin{array}{rccccccl}
+    \[  \begin{array}{rccccccl}
       \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
@@
                                                &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
-                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
-    \end{array}
+                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{e} \land (e=\bad \lor \bigvee_i e = \weak{K_i(\vec{x})} \lor y = \unr)
+    \end{array} \]
     \end{center}
@@
 \item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
+  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
   on its $k$-th component.
@@
   $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
-  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
-  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
+  $$ S_2 := \forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land K_i(\vec{x}) = a \land K_j(\vec{x}) = a) \mid 1 \leq i < j \leq n$$
+  $$ S_3 := \forall \vec{x},a.~\min{a} \land a = K_i(\vec{x}) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})}) \mid 1 \leq i \leq n$$
   $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
@@
 \begin{eqnarray*}
-  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
-  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
-  \strans{e ::: \cfc} &=& \cf{e}
+  \strans{e \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
+  \strans{e \in x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x \not \in c_1} \lor  \strans{\etrans{e(x)} \in c_2(x)})\\
+  \strans{e \in \cfc} &=& \cf{e}
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
-  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
-  \strans{e :/: \cfc} &=& ~\cf{e}
+  \strans{e \not \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = false \lor \etrans{b(e)} = \bad)\\
+  \strans{e \not \in x:c_1 \to c_2(x)} &=& \exists x.~ \strans{x \in c_1} \land  \strans{\etrans{e(x)} \not \in c_2(x)}\\
+  \strans{e \not \in \cfc} &=& \lnot \cf{e}
 \end{eqnarray*}
 
-$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
-$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
-$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$
+\begin{center}
+\[  \begin{array}{rcl}
+ \strans{e \in c_1 \&\& c_2} &=& \strans{e \in c_1} \land \strans{e \in c_2} \\
+ \strans{e \not \in c_1 \&\& c_2} &=& \strans{e \not \in c_1} \lor \strans{e \not \in c_2}\\
+ \strans{e \in c_1 || c_2} &=& \strans{e \in c_1} \lor \strans{e \in c_2} \\
+  \strans{e \not \in c_1 || c_2} &=& \strans{e \not \in c_1} \land \strans{e \not \in c_2}\\
+ \strans{(a,b) \in (c_1,c_2)}   &=& \strans{a \in c_1} \land \strans{b \in c_2} \\
+ \strans{(a,b) \not \in (c_1,c_2)}   &=& \strans{a \not \in c_1} \lor \strans{b \not \in c_2} \\
+\end{array} \]
+\end{center}
 

### 2011-08-22 paper.tex
paper update

--- 2011-08-19 paper.tex
@@
 
+\newtheorem{definition}{Definition}
+
 \newcommand{\unr}{\texttt{UNR}}
@@
 \title{Static contract checking for Haskell}
+\authorinfo{Simon, Koen, Dimitrios, Charles-Pierre}
 \maketitle
@@
 
-The syntax of $\mathcal{H}$ is defined in \ref{hprime-stx}. A module
+The syntax of $\mathcal{H}$ is defined in figure~\ref{hprime-stx}. A module
 is a list of toplevel definitions, claims that functions statisfy
@@
   can always lift them to the toplevel.
-\item Until \label{ho} we will only consider full application of
+\item Until section~\ref{ho} we will only consider full application of
   functions ($f(x,y)$), in order to remove clutter. Dealing with
@@
   \centering
+  \[
   \begin{array}{rclr}
@@
 \begin{figure}[h]
-  \label{cont-stx}
  \centering 
-  \begin{array}{rclr}
+  \[  \begin{array}{rclr}
   c &:=& x:c \to c\\
@@
   &\mid& \cfc \\
-  \end{array}
+  \end{array} \]
   \caption{Contract syntax}
+  \label{cont-stx}
 \end{figure}
@@
 \begin{figure}[h]
-  \label{cont-smt}
  \centering
-  \begin{array}{rcl}
+  \[  \begin{array}{rcl}
     e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
@@
     e \in \cfc &\iff& e \mbox{ is crash-free}
-  \end{array}
+  \end{array} \]
   \caption{Semantics of contract satisfaction}
+  \label{cont-smt}
 \end{figure}
@@
 \begin{figure}
+ \centering
+  \[  \begin{array}{rcl}
+    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
+    && \mid \bad \mid \unr \\
+    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
+    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi \\
+    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
+  \end{array} \]
+  \caption{First-order logic syntax}
   \label{fol-stx}
-  \centering
-  \begin{array}{rcl}
-    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
-    && \mid \bad \mid \unr & \\
-    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
-    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi & \\
-    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
-  \end{array}
-  \caption{First-order logic syntax}
 \end{figure}
@@
 \section{Translations}
-For an overview of the different translations we define, see \ref{trans}
+For an overview of the different translations we define, see figure~\ref{trans}
 
 \begin{figure}
-  \label{trans}
  \begin{center}
-  \begin{array}{rcl}
+  \[  \begin{array}{rcl}
     \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
@@
     \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
-  \end{array}
+  \end{array} \]
   \end{center}
   \caption{Translations}
+  \label{trans}
 \end{figure}
@@
 translate definitions, data types and contracts to first-order
-formulae. It is described in \ref{etrans}.
+formulae. It is described in figure~\ref{etrans}.
 
 \begin{figure}
-\label{etrans}
 \begin{eqnarray*}
@@
 \caption{$\etrans{}$ -- Expression translation}
+\label{etrans}
 \end{figure}
@@
 \subsection{$\dtrans{}$ -- Definitions}
-We give in \ref{dtrans-reg} and \ref{dtrans-case} the two translations
+We give in figure~\ref{dtrans-reg} and figure~\ref{dtrans-case} the two translations
 of function definitions.
 
-\ref{dtrans-reg} gives the translation of function not defined by
+figure~\ref{dtrans-reg} gives the translation of function not defined by
 pattern matching, which is really easy: we just have to state the
@@
 Translating definitions that use pattern-matching is more challenging
-and is described in \ref{dtrans-case}.
+and is described in figure~\ref{dtrans-case}.
 
@@
 \begin{figure}
-  \label{dtrans-reg}
  $$\dtrans{f(\vec{x}) = e}$$
@@
   \caption{$\dtrans{}$ -- Regular definitions}
+  \label{dtrans-reg}
 \end{figure}
@@
 \begin{figure*}
-  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
-    \begin{array}{rclcl}
-      \forall \vec{a},y.~\min{y} \land f(\vec{a}) = y &\to& (\multicolumn{3}{c}{\min{e} \land (e = \bad \lor \bigvee_i \forall \vec{x}.~e = K_i(\vec{x}) \lor y = \unr)} \\      
+   \[  \begin{array}{rclcl}
+     \forall \vec{a},y.~\mbox{min}(y) \land f(\vec{a}) = y &\rightarrow&
+     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
+       \bigvee_i \forall \vec{x}.~e = K_i(\vec{x}) \lor y = \texttt{UNR})} \\
      &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& y = \etrans{e_1} \\
@@
      &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& y = \etrans{e_n}) \\
-    \end{array}
+   \end{array} \]
  \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
+ \label{dtrans-case}
 \end{figure*}
@@
 \subsection{$\ktrans{}$ -- Datatypes}
-We break down the translation for datatypes in four parts, described in \ref{ktrans}
+We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
 $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
@@
 \item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
+  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
   on its $k$-th component.
@@
 \begin{figure*}
-  \label{ktrans}
  $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
@@
   \begin{center}
-    \begin{array}{rcl}
+    \[  \begin{array}{rcl}
       \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
@@
       \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
-    \end{array}
+    \end{array} \]
   \end{center}
   \caption{$\ktrans{}$ -- Data type translation}
+  \label{ktrans}
 \end{figure*}
@@
 \subsection{$\ctrans{}$ -- Contracts}
-We give in \ref{ctrans} the translation of contract satisfaction.
+We give in figure~\ref{ctrans} the translation of contract satisfaction.
 $true$ refers to the translation to a term of the data constructor
@@
 \begin{center}
-\begin{array}{rcl}
+\[  \begin{array}{rcl}
  \ctrans{e \in c_1 \&\& c_2} &=& \ctrans{e \in c_1} \land \ctrans{e \in c_2} \\
@@
  \ctrans{(a,b) \not \in (c_1,c_2)}   &=& \ctrans{a \not \in c_1} \lor \ctrans{b \not \in c_2} \\
-\end{array}
+\end{array} \]
 \end{center}
@@
 is not fully applied somewhere in a module. We create the term
-$f\_ptr$ which relates to $f$ by the equations given in \ref{ho-fig}
+$f\_ptr$ which relates to $f$ by the equations given in figure~\ref{ho-fig}
 
@@
 \begin{figure*}
-\label{ho-fig}
 $$\etrans{e_1~e_2} = app(e_1,e_2)$$
@@
 \caption{Encoding of higher-orderness}
+\label{ho-fig}
 \end{figure*}
@@
 \begin{figure*}
-\label{comparison}
 \begin{center}
@@
 \caption{Comparison (in seconds) with other theorem provers}
+\label{comparison}
 \end{figure*}

### 2011-08-23 paper.tex
Add ruleform macro and use it in Fig 6
Paper update BNF

@@
     \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
-    \dtrans{f~\vec{x} = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
+    \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
     \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
@@
 \subsection{$\dtrans{}$ -- Definitions}
-We give in figure~\ref{dtrans-reg} and figure~\ref{dtrans-case} the two translations
-of function definitions.
+We give in figure~\ref{dtrans} the two translations of function
+definitions.
 
@@
 Translating definitions that use pattern-matching is more challenging
-and is described in figure~\ref{dtrans-case}.
+and is described in figure~\ref{dtrans}.
 
@@
 
-\begin{figure}
- $$\dtrans{f(\vec{x}) = e}$$
-  $$ \forall \vec{x},y.~ \min{y} \land y = f(\vec{x}) \to y = \etrans{e}$$
-  \caption{$\dtrans{}$ -- Regular definitions}
-  \label{dtrans-reg}
-\end{figure}
-
 \begin{figure*}
- $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
- \begin{center}
-   \[  \begin{array}{rclcl}
+\[
+\begin{array}{rcl}
+  \dtrans{\mbox{data } T = K_1, \dots, K_n} &=& \ktrans{\mbox{data } T = K_1, \dots, K_n} \\
+  \dtrans{f \in c} &=& \ctrans{f \in c} \\
+  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},y.~ \min{y} \land y =
+  f(\vec{x}) \to y = \etrans{e} \\
+   \raisebox{ 0.5 cm }{$\dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
      \forall \vec{a},y.~\mbox{min}(y) \land f(\vec{a}) = y &\rightarrow&
@@
      &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& y = \etrans{e_n}) \\
-   \end{array} \]
- \end{center}
- \caption{$\dtrans{}$ -- Case definitions}
- \label{dtrans-case}
+   \end{array} 
+\end{array}
+\]
+\caption{$\dtrans{}$ -- Defintions translation}
+ \label{dtrans}
 \end{figure*}
@@
  $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
-  where
+ \hspace{5.6cm}where
   \begin{center}

### 2011-08-23 paper.tex
Now using Simon's nice ruleform for translations

@@
 
+\def\ruleform#1{{\setlength{\fboxrule}{1.2pt}\fbox{\normalsize $#1$}}}
 \newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
@@
     def &\in& \mbox{Definition} & \\
-    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \\
-    &\mid& f \in c & \\
+    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \mbox{Data definition}\\
+    &\mid& f \in c & \mbox{Contract claim}\\
     &\mid& f~\vec{x}~=~e & \\
-    &\mid& \hspace{-0.2cm}
+    &\raisebox{ 0.18 cm }{$\mid$}& \multicolumn{2}{c}{
       \begin{array}{rcl}
-         f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
-         && K_1(\vec{x}) \to e_1 \mid \cdots \mid K_n(\vec{x}) \to e_n \\
-       \end{array} & \\
+        \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
+         && K_1(\vec{x}_1) \to e_1 \mid \cdots \mid K_n(\vec{x}_n) \to e_n \\
+       \end{array} 
+       } \\
   \end{array} \]
@@
 \begin{figure}
-\begin{eqnarray*}
+$$
+\begin{array}{rcl}
+\multicolumn{3}{c}{\ruleform{\etrans{e} = t}} \\[2.5mm]
  \etrans{x} &=& x\\
@@
   \etrans{\bad} &=& \bad
-\end{eqnarray*}
+\end{array}
+$$
 \caption{$\etrans{}$ -- Expression translation}
@@
 \begin{array}{rcl}
+  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
   \dtrans{\mbox{data } T = K_1, \dots, K_n} &=& \ktrans{\mbox{data } T = K_1, \dots, K_n} \\
@@
 \begin{figure*}
+  $$ \ruleform{\ktrans{data~def} = \Phi} $$
 $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
@@
 \begin{figure*}
+  $$ \ruleform{\ctrans{contract} = \Phi} $$
 \begin{eqnarray*}
@@
 \begin{center}
-\includegraphics[scale=0.5]{flow.png}
+\includegraphics[scale=0.3]{flow.png}
 \end{center}

### 2011-08-26 contracts/Check.hs
paper update + cosmetic fix

@@ -76,8 +76,9 @@ hasNoContract f prog = and $ (flip map) prog $ \d -> case d of
   _ -> True
 
 
--- the distinction between one or several functions to check at the same time is 
--- not stricly necessary but it gives a nicer output to the user
+-- the distinction between one or several functions to check at the
+-- same time is not stricly necessary but it gives a nicer output to
+-- the user
 check :: Program -> [Variable] -> Conf -> [Variable] -> IO Bool
 check prog [] cfg _ = error "There should be at least one definition!"
 check prog [f] cfg checkedDefs | f `hasNoContract` prog = return True

### 2011-08-26 contracts/Translation.hs

@@ -111,7 +111,7 @@ sTrans e (H.CF) = do
 -----------------------
 
 tTrans :: H.DataType -> Fresh [F.Formula]
-tTrans d = concat <$> (sequence [s1 d,s2 d,s3 d,s4 d])
+tTrans d = concat <$> sequence [s1 d,s2 d,s3 d,s4 d]
 
 --s1 :: H.DataType -> Fresh [F.Formula (F.Term F.Variable)]
 s1 (H.Data _ dns) = sequence $ map s1D dns
@@ -186,7 +186,7 @@ trans ds fs = evalState (go fs ((H.appify) ds)) (S "Z" 0 (H.arities ds))
   where go fs ds = do 
           let (toCheck,regDefs) = partition (isToCheck fs) ds
               recVar x = x ++ "p"
-          a <- fmap arities get
+          a <- arities <$> get
           regFormulae <- forM regDefs $ \d -> case d of
             H.DataType t                -> tTrans t
             H.Def d                     -> dTrans d

### 2011-10-19 paper.tex
paper update + cosmetic fix
Added a Latex annotations package (trackchanges.sty) and some examples.
I'll now erase the examples and get to work annotating the paper.  The
trackchanges is from http://trackchanges.sourceforge.net/ and includes a
Python based gui that I did not try or include here.

@@
 \usepackage[pdftex]{graphicx}
+
+% Margin notes (ugly and complicated)
+%% \usepackage[draft]{fixme}
+%% \fxusetheme{color}
+%% %\fxusetheme{signature}
+%% %\fxuselayouts{pdfmargin}
+%% %\fxusetargetlayout{color}
+%% \FXRegisterAuthor{nc}{anc}{NC}
+
+% Margin notes round two (easier and better looking)
+%
+% http://trackchanges.sourceforge.net/help_stylefile.html
+\usepackage[inline]{trackchanges} % [margins] is also good
+\addeditor{nc}
 
@@
 Our approach to static contract-checking is to translate source code
-to a first-order logic theory and then use an automated theorem prover
-to check the consistency of the theory.
-
+to a first-order \note[nc]{this is a note} logic theory and
+ then use an \annote[ncc]{this is an annote}{automated theorem prover}
+\add[nc]{new text} to check the \remove[nc]{consistency} of the theory.
+\change[nc]{Old text}{New text}.
 Consider:
@@
 
-Now that we translated everything to first-order logic, we can ask the
+\ncnote*{Thi is  a note}{Now that we translated} everything to first-order logic, we can ask the
 theorem prover if the theory formed by those formulae is
@@
 \begin{eqnarray*}
-  \ctrans{e \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{e} = \unr)\\
-  \ctrans{e \in x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{e(x)}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{\etrans{e(x)} \in c_2(x)})\\
-  \ctrans{e \in \cfc} &=& \cf{\etrans{e}}
+  \ctrans{e \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = true \lor \etrans{e} = \unr)\\
+  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{\etrans{e~x}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
+  \ctrans{e \in \cfc} &=& \min{\etrans{e}} \land \cf{\etrans{e}}
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  \ctrans{e \not \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = false \lor \etrans{e} = \bad)\\
-  \ctrans{e \not \in x:c_1 \to c_2(x)} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{\etrans{e(x)} \not \in c_2(x)}\\
-  \ctrans{e \not \in \cfc} &=& \lnot \cf{\etrans{e}}
+  \ctrans{e \not \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = false \lor \etrans{e} = \bad)\\
+  \ctrans{e \not \in x:c_1 \to c_2} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{e~x \not \in c_2}\\
+  \ctrans{e \not \in \cfc} &=& \min{\etrans{e}} \land \lnot \cf{\etrans{e}}
 \end{eqnarray*}

### 2011-10-19 paper.tex
Paper: removed annote examples and updated figures 7 and 8.
Added Simon's parens, made more uniform, and added questions.

@@
 Our approach to static contract-checking is to translate source code
-to a first-order \note[nc]{this is a note} logic theory and
- then use an \annote[ncc]{this is an annote}{automated theorem prover}
-\add[nc]{new text} to check the \remove[nc]{consistency} of the theory.
-\change[nc]{Old text}{New text}.
+to a first-order logic theory and
+ then use an automated theorem prover
+to check the consistency of the theory.
+
 Consider:
@@
   \dtrans{\mbox{data } T = K_1, \dots, K_n} &=& \ktrans{\mbox{data } T = K_1, \dots, K_n} \\
-  \dtrans{f \in c} &=& \ctrans{f \in c} \\
-  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},y.~ \min{y} \land y =
-  f(\vec{x}) \to y = \etrans{e} \\
-   \raisebox{ 0.5 cm }{$\dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
-     \forall \vec{a},y.~\mbox{min}(y) \land f(\vec{a}) = y &\rightarrow&
+  \dtrans{f \in c} &=& \ctrans{f \in c}\\
+  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
+  f(\vec{x})) \to a = \etrans{e} \\
+   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
+     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
      \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
-       \bigvee_i \forall \vec{x}.~e = K_i(\vec{x}) \lor y = \texttt{UNR})} \\
-     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& y = \etrans{e_1} \\
+       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
+     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& a = \etrans{e_1} \\
      &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& y = \etrans{e_n}) \\
+     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n}} \\
    \end{array} 
@@
  \label{dtrans}
+ \note[nc]{??? Why do we introduce the extra quantified variable $a$,
+   representing the result (``answer''?), instead of just using the result directly
+   (which is logically equivalent)? Same question for the next
+   Figure.  Does this make a difference to Equinox?}
 \end{figure*}
@@
     \[  \begin{array}{rcl}
-      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
-      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land K_i(\vec{x}) = a \land K_j(\vec{x}) = a) \\
-      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~\min{a} \land a = K_i(\vec{x}) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})}) \\
+      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
+      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
+      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a\text{\note[nc]{This was $K_i(\vec{x})$.
+Typo, right?}}}) \\
       \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 

### 2011-10-19 paper.tex
Paper: made a full comments pass through section 1.
Ouch!

@@
 
-Consider:
+Consider:\note[nc]{Unlike the POPL '09 paper, $CF(e)$ is not implied by $e \in \{x|p\}$.}
 \begin{verbatim}
@@
 Then, we must state that $Nil$ never crashes (ie cannot be evaluated
-to an exception) and that $Cons(x,y)$ crashes iff either $x$ or $y$
-crashes. The statement $x$ crashes is encoded by the term $\cf{x}$.
+to an exception) and that $Cons(x,y)$ \add[nc]{never} crashes iff \change[nc]{either}{both} $x$ \change[nc]{or}{and} $y$
+\add[nc]{never} crash. The statement $x$ \add[nc]{never} crashes is encoded by the term $\cf{x}$\note[nc]{Made more precise in section \ref{sec:cf}}.
 
@@
 
-We also say some stuff about unreachibility but I can't think of a
-good way to explain it right now.
+We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
+  CF(fst(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
+can't think of a good way to explain it right now.
 \begin{equation*}
@@
 Nil \neq \unr 
-\end{equation*}
+\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\Phi_4$) say that constructors are ``lazy'' or ``total''.}
 
-Finally, we define projections for $Cons$. It is not strictly
-necessary, but it will be handy:
+Finally, we define projections for $Cons$. \annote[nc]{It is not
+  strictly necessary, but it will be handy:}{??? Why is it handy? It
+  doesn't seem to play any role in formal part of this paper (besides
+  being specified in axioms $\Phi_1$ (projections)). In the current
+  implementation it's used to avoid introducing existentially
+  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
+  functions defined by case expressions. Equation \eqref{unrnotnull}
+  below uses it that way. ??? Does Equinox care?}
 \begin{equation*}
@@
 \forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
-\end{equation*}
+\end{equation*}\note[nc]{???
+Why not the simpler \[\forall y,ys.~sel_{2,Cons}(Cons(y,ys)) = ys\]?
+Same question in Figure \ref{dtrans}.
+(Equinox does have a built in notion of equality, e.g.
+running equinox on
+\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
+ results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
+%% \[
+%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
+%%                  \text{\texttt{| equinox /dev/stdin}}}}
+%% \]
 
@@
 $False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
-logic.
+logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
+  text.
+The lowercase $true$ and $false$ is an artifact of the implementation, i.e.
+Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.
+Moreover, the lower case here is inconsistent with the rest of the presentation, e.g.
+$Cons$ and $Nil$ above.}
 \begin{equation*}
 \forall xs.~ xs=Nil \implies notnull(xs) = false
-\end{equation*}
+\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
+\[
+\forall notnull(Nil) = false
+\]?
+The answer is that this is an expansion of the $etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.
+I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.
+In this special case we could remove the implication, but not in general.
+The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}
 
@@
   \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
-  notnull(X) = \unr 
+  notnull(xs) = \unr 
 \end{align}
@@
   \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
-  \forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
+  \forall xs,y,ys.~ xs=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
     \forall xs.~xs=\bad \to head(xs) = \bad \notag
-\end{align}
+\end{align}\note[nc]{??? Why not use the same project+construct trick from the last equation?  I.e., why not
+\[
+\forall xs.~ xs=Cons(sel_{Cons,1}(xs),sel_{Cons,2}(xs)) \implies head(xs) = sel_{Cons,1}(xs)
+\]?  Because it's ugly?  Would Equinox care?}
 
@@
  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
- head(X) = \unr 
+ head(xs) = \unr 
 \end{align*}
@@
 
-\ncnote*{Thi is  a note}{Now that we translated} everything to first-order logic, we can ask the
-theorem prover if the theory formed by those formulae is
-consistent, ie if $T \vdash \phi$.
+Now that we translated everything to first-order logic, we can ask the
+theorem prover if the theory formed by those formulae \change[nc]{is
+consistent}{entails the contract}, ie if $T \vdash \phi$.
 
 Intuitively, $T$ is consistent (ie $T \not \vdash \bot$), because each
-formula serves a specific purpose. Now, assume that $xs$ satisfies
+formula serves a specific purpose.\note[nc]{
+
+  A more precise consistency argument by exhibiting a model for $T$:
+  in the model we define $\min{}$ to be the trivial predicate,
+  i.e. $\forall x. \min{x}$.  So, we can drop all the $\min{}$'s.  The
+  model for the resulting formulas is a language with the syntax used
+  in this paper ($\mathcal{H}'$ below).
+
+  The carrier consists of all closed (i.e. variable free / grounded)
+  programs that can be written using the functions and constructors in
+  $T$, using the expression syntax of $\mathcal{H}'$ below, and
+  evaluated as described in the following paragraphs.  We assume $T$
+  comes from a ``well-typed'' source program $P$, i.e. s.t. the
+  Haskell program corresponding to $P$ is well-typed.
+
+  The semantics are what you would expect for ``well-typed''
+  terminating expressions (again, ``well-typed'' in the sense that the
+  corresponding Haskell expression is well-typed in an environment
+  including $P$'s decls). For example, \texttt{Cons(Z,Nil)} is
+  ``well-typed'' but \texttt{Cons(Nil,Z)} is ``ill-typed'' in a the
+  context of a program declaring List and Nat types.
+
+  In general, scrutinizing a \unr{} yields \unr, as does scrutinizing
+  a terminating expression with the wrong constructor head
+  (e.g. \texttt{tail(S Z) =} \unr{}). Scrutinizing an ill-typed
+  expression with a constructor head of the right type evaluates as
+  specified by the pattern (e.g. \texttt{tail(Cons(Nil,Z)) = Z}),
+  unless there is no corresponding pattern (``non-exhaustive
+  pattern''), in which case the result is \bad{}.  NB: wrong-type and
+  no-corresponding-pattern are treated very differently.  The
+  intuition is that the former never happens in a well-typed program.
+
+  Applying \unr{} or \bad{} as a function results in \unr{} or \bad{},
+  respectively.
+
+  Term constructors have disjoint ranges, and are total.
+
+  I think that covers functions and datatypes, but it remains to argue
+  that the axioms for contracts are consistent \ldots.
+
+  \ldots but maybe this is tricky?  Modeling $CF(e)$ seems easy,
+  i.e. just ask the oracle used above for evaluation if $e$ is \bad{}
+  or not.  But recursion seems problematic, since we get to assume in
+  $T$ that the contract holds in the recursive position (Section
+  \ref{sec:recursiveContracts}), and so it seems we need to argue that
+  \emph{all} expressable contracts are consistent.
+
+  Here's a flawed contradiction ``proof'': consider
+  \[
+  e = e ::: CF \&\& \neg CF.
+  \]
+  Now, $T$ will contain $CF(ep) \land \neg CF(ep)$, and were hosed.
+  This is flawed because there is no $\neg$ in the annotated
+  contracts.
+
+  OK, and actually this might be easy: we simply interpret all
+  recursive function symbols (the $fp$'s for recursive $f$'s) as
+  \unr{}, and use the fact that forall $x$, $\unr{}~x = \unr$.  Now,
+  an easy induction on contract translations (Figure \ref{ctrans})
+  gives \[ \emptyset \vdash \ctrans{\unr \in c}
+  \]
+  for all contracts $c$, and we're done!  Well, this is really just
+  the base case, where all contracts in $T$ are for otherwise
+  undefined recursion symbols ($fp$'s), but the inductive case is just
+  as easy: given a consistent $T$, which may include contracts for
+  defined symbols, we can always extend it to a consistent $T'$ by
+  adjoining constracts for undefined symbols.
+
+}
+
+Now, assume that $xs$ satisfies
 $\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
-because we have $\cf{xs}$ and $\lnot \cf{\bad}$. The constraint
+because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{Obviously true, given the definition of $CF$, but it's worth noting that this plays a key part in the axiomitization of $CF$, which is ommited in this draft (so far ...)}.
+The constraint
 $notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
@@
     def &\in& \mbox{Definition} & \\
-    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \mbox{Data definition}\\
+    def &:=& \texttt{data } T = K_1 a_1 \mid \cdots \mid K_n a_n& \mbox{Data definition}\\
     &\mid& f \in c & \mbox{Contract claim}\\
@@
 
-\subsection{Crash-freeness}
+\subsection{Crash-freeness}\label{sec:cf}
 Note that $\cfc$ represents two things: it can be a contract, as in $f
@@
 
-\subsection{Contract checking -- Recursive case}
+\subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}
 If the function $f=e$ is recursive, then we ask the theorem prover the

### 2011-10-20 paper.tex
Paper: made a pass through section 2, with some section 1 fixes along the way.

@@
 % http://trackchanges.sourceforge.net/help_stylefile.html
+%
+% NB: if you want to add a note in math mode, you need a trick:
+%
+% $ ... \text{\note{$<math note>$}} ... $
 \usepackage[inline]{trackchanges} % [margins] is also good
@@
 
-Then, we must state that $Nil$ never crashes (ie cannot be evaluated
-to an exception) and that $Cons(x,y)$ \add[nc]{never} crashes iff \change[nc]{either}{both} $x$ \change[nc]{or}{and} $y$
-\add[nc]{never} crash. The statement $x$ \add[nc]{never} crashes is encoded by the term $\cf{x}$\note[nc]{Made more precise in section \ref{sec:cf}}.
+Then, we state that $Nil$ can cannot cause a crash, and that
+$Cons(x,y)$ cannot cause a crash iff both its arguments cannot. The
+statement ``$x$ cannot cause a crash'' is encoded by the formula
+$\cf{x}$.\note[nc]{I just rewrote this, but it still sounds weird: in
+  $head(Nil) = \bad$, who ``caused'' the crash? The precise definition
+  of $\cf$ in Section \ref{sec:cf}, in terms of not leading to crashes
+  when inserted into syntactically \bad-free environments,  is good, but
+  how to state it informally?}
 
@@
 \[
-\forall notnull(Nil) = false
+notnull(Nil) = false
 \]?
@@
   an easy induction on contract translations (Figure \ref{ctrans})
-  gives \[ \emptyset \vdash \ctrans{\unr \in c}
-  \]
-  for all contracts $c$, and we're done!  Well, this is really just
+  gives \[ \emptyset \vdash \ctrans{\unr \in c} \]\note[nc]{
+
+    Actually, this was not true!  We need an axiom saying $\cf{\unr}$.
+    There was already such an axiom in the implementation, but it
+    wasn't mentioned in this draft.  So, the correct statement is more
+    like
+    \[ \text{The Prelude} \vdash \ctrans{\unr \in c} \]
+
+  } for all contracts $c$, and we're done!  Well, this is really just
   the base case, where all contracts in $T$ are for otherwise
@@
 $\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
-because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{Obviously true, given the definition of $CF$, but it's worth noting that this plays a key part in the axiomitization of $CF$, which is ommited in this draft (so far ...)}.
-The constraint
+because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{
+
+Follows from the definition of $CF$, but it's worth noting that this
+plays a key part in the axiomitization of $CF$.
+
+}. The constraint
 $notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
@@
 
-The syntax of $\mathcal{H}$ is defined in figure~\ref{hprime-stx}. A module
+The syntax of $\mathcal{H}'$ is defined in figure~\ref{hprime-stx}. A module
 is a list of toplevel definitions, claims that functions statisfy
@@
 \item Until section~\ref{ho} we will only consider full application of
-  functions ($f(x,y)$), in order to remove clutter. Dealing with
+  functions ($f(x,y)$),\note[nc]{
+
+Not really true: the definition of satisfaction for arrow contracts
+$x:c_1 \to c_2$ already includes partial applications.  We could add a
+special cases for each function arity, e.g.
+\begin{eqnarray*}
+&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
+&&      :=\forall x_1, \ldots, x_n. \min{\etrans{f(x_1, \ldots, x_n)}} \\
+&&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
+              \lor \ctrans{f(x_1, \ldots, x_n) \in c}
+\end{eqnarray*}
+for functions of arity $n$, to make the claim true, but that would add
+more clutter.  This may be related to my confusion where removing the
+$f_{ptr}$ stuff caused problems where no higher-orderness was
+involved?
+
+}, in order to remove clutter. Dealing with
   partial application is not hard but a bit cumbersome.
@@
 only consider pairs of contract for simplicity, although there is no
-issue with generalisation to arbitrary tuples.
+issue with generalisation to arbitrary tuples.\note[nc]{
+
+So, ``tuples'' refers to the the term constructor, not the type
+constructor, and the generalization is e.g.
+\[
+e \in Cons(c_1,c_2) \iff e \text{ diverges or } 
+      (e \to^\star Cons(e_1,e_2) \mbox{ and } e_1 \in c_1, e_2 \in c_2)
+\]
+?
+}
 
@@
 
-We give the semantics of contract by defining ``$e$ satisfies $t$'',
-written $e \in t$ in figure \ref{cont-smt}. Note that this definition
+We give the semantics of contract by defining ``$e$ satisfies $c$'',
+written $e \in c$ in the logic (figure \ref{cont-smt}), and 
+\texttt{e ::: c} in code. Note that this definition
 doesn't yield any operative way to check that an expression actually
@@
    \caption{Semantics of contract satisfaction}
+
+\note[nc]{
+
+      ??? Does $\unr$ satisfy any contract? It's supposed to in the
+      model sketched in the intro, and does intuitively, since we
+      don't do termination analysis.  But now we can't prove the arrow
+      case for $\unr$ in general, i.e.  $\ctrans{\unr \in x:c_1 \to
+        c_2}$ is not provable for arbitrary $c_1$ and $c_2$ (I don't
+      think).  Two solutions: (1) add the axiom $\forall x. \unr~x =
+      \unr$ (the model uses this); (2) add the premise ``$e$ diverges
+      or'' to the arrow case.
+
+      Now, $\unr \in CF$ is intuitively true, for the Definition
+      \ref{def:cf}, but to actually prove it it seems we need a
+      semantics for evaluation that includes $\forall x. \unr~x =
+      \unr$.  In the translation, $\ctrans{\unr \in CF}$ is provable,
+      but only because of an axiom saying $\cf{\unr}$ :P
+
+}
   \label{cont-smt}
@@
 Note that $\cfc$ represents two things: it can be a contract, as in $f
-\in \cfc$ or a special formula in first-order logic $\cf{f}$.
+\in \cfc$ or a predicate in first-order logic, as in the formula $\cf{f}$.
 
 We use $\bad$ to signal that something has gone wrong in the program :
-it has crashed.
+it has crashed. (NB: looping is not crashing.)
 
 \begin{definition}[Crash]
-A closed term $e$ crashed iff $e \to^* \bad$.
+A closed term $e$ \emph{crashed} iff $e \to^* \bad$.\note[nc]{Again, the evaluation semantics are not given, but we assume that a pattern match failure is a crash.}
 \end{definition}
@@
 \begin{definition}[Diverges]
-A closed expression $e$ diverges iff either $e \to^* \unr$ or there is
+A closed expression $e$ \emph{diverges} iff either $e \to^* \unr$ or there is
 no value $val$ such that $e \to^* val$
@@
 \begin{definition}[Syntactic safety]
-A (possibly open) expression $e$ is syntactically safe iff $\bad \not
-\in_s e$. Similarly a context \cal C is syntactically safe iff $\bad
+A (possibly open) expression $e$ is \emph{syntactically safe} iff $\bad \not
+\in_s e$. Similarly a context $\mathcal{C}$ is \emph{syntactically safe} iff $\bad
 \not \in_s \mathcal{C}$.
@@
 
-The notation $\bad \not \in e$ means that $\bad$ does not appear
+The notation $\bad \notin_s e$ means that $\bad$ does not appear
 anywhere in $e$, similarly for $\bad \not \in_s \mathcal{C}$. For
-example, $Just 3$ is syntactically safe whereas $Just \bad$ is not.
+example, $Just ~3$ is syntactically safe whereas $Just ~\bad$ is not.
 
-\begin{definition}[Crash-free]
-An expression $e$ is said to be crash-free iff 
-$$\forall \mathcal{C}.~\bad \not \in_s \mathcal{C} \mbox{ and } \vdash
-\mathcal{C} \llbracket e \rrbracket :: () \mathcal{C} \not \to^* \bad$$
+\begin{definition}[Crash-free]\label{def:cf}
+An expression $e$ is said to be \emph{crash-free} iff 
+for all $\mathcal{C}$, $\bad \not \in_s \mathcal{C}$ and $\emptyset \vdash
+\mathcal{C} \llbracket e \rrbracket :: ()$ implies $\mathcal{C} \llbracket e \rrbracket \not \to^* \bad$
 \end{definition}
-The notation $\mathcal{C} \llbracket e \rrbracket :: ()$ means that
+The notation $\emptyset \vdash \mathcal{C} \llbracket e \rrbracket :: ()$ means that
 $\mathcal{C} \llbracket e \rrbracket$ is closed and well-typed.  Note
-that there are crash-free expression that are not syntactically safe,
-for example $fst~(1,\bad)$.
+that there are crash-free expressions that are not syntactically safe,
+for example $fst(Nil,\bad)$.
 
@@
 
-One thing to notice is that $a$ and $b$ are things that ``sould not
+One thing to notice is that $a$ and $b$ are things that ``should not
 happen'' but are caught statically whereas $c$ should not happen but
-can only be dealt with dynamically.
+can only be dealt with dynamically.\note[nc]{
+I find this example confusing: \texttt{error} calls should be just as
+easy to find as \texttt{undefined} calls statically.}
 
@@
 Expressions of the first type are called unreachable (and equated to
-the special value $\unr$ in our first-order theory), whilst
+the special value $\unr$ in our first-order theory)\note[nc]{
+%
+In particular, ill-typed expressions are $\unr$, not $\bad$
+%
+}, whilst
 expressions of the latter type are called bad (and equated to the
@@
 theorem prover may decide to instanciate a variable with an ill-typed
-value! In order to prevent this, we will need to encode some basic
-type-checking mecanism directly in our first-order theory.
+value! \change[nc]{In order to prevent this, we will need to encode some basic
+type-checking mecanism directly in our first-order theory.}{This is OK, because the axioms are designed so that scrutinizing an ill-typed constructor-head results in $\unr$.
+On the other hand, some ill-typed expressions are not $\unr$, e.g.
+we can prove that $tail(Cons(Nil,Z)) = Z$.}
 
 \subsection{First-order logic with equality}
-We use first-order logic with equality, defined in figure \ref{fol-stx}.
+We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e.
+forall $phi$, $\forall x,y.x=y \to phi \iff phi[y/x]$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
 
@@
     && \mid \bad \mid \unr \\
-    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
+    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
     && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi \\
@@
   \caption{First-order logic syntax}
+  \note[nc]{The $\Phi$ are unnecessary and don't make things simpler, because a set of formulas is equivalent to the conjunction of its contents.
+XXX, TODO: drop $\Phi$ and replace all $\bigcup$ with $\bigwedge$.}
   \label{fol-stx}
@@
      &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n}} \\
+     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n} \\%removed } here ...
    \end{array} 
@@
 $\lnot \cf{\bad}$.
+
+\add[nc]{We also state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
+
+This is already present in the implementation, and assumed in the
+consistency proof.}
 

### 2011-10-20 paper.tex
Paper: partial update of section 3.

@@
 
-figure~\ref{dtrans-reg} gives the translation of function not defined by
+figure~\ref{dtrans} gives the translation of function not defined by
 pattern matching, which is really easy: we just have to state the
@@
 Translating definitions that use pattern-matching is more challenging
-and is described in figure~\ref{dtrans}.
-
-The first line says that when applied to an argument that matches a
-pattern of the case expression, we should equate the function call to
-the corresponding expression.
-
-The second line states that if the pattern-matching failed or if we
-pattern-matched on $\bad$ then the result should be $\unr$.
+and is described in figure~\ref{dtrans}.  The first line says that
+when applied to an argument that matches a pattern of the case
+expression, we should equate the function call to the corresponding
+expression. The second line states that if the pattern-matching failed
+or if we pattern-matched on $\bad$ then the result should be
+$\unr$.\note[nc]{This text does not agree with the figure.  In
+  particular, the axiom says nothing about the result of scrutinizing
+  $\bad$.  I think the correct behavior is that scrutinizing $\bad$
+  results in $\bad$, \emph{not} $\unr$.}
 
@@
   \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
-  \dtrans{\mbox{data } T = K_1, \dots, K_n} &=& \ktrans{\mbox{data } T = K_1, \dots, K_n} \\
+  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
   \dtrans{f \in c} &=& \ctrans{f \in c}\\
@@
 We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
-$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
+$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 
@@
   $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
-  on its $k$-th component.
-\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, we state that they can
+  on its $k$-th component, $k = 1\text{ to }a_i$.
+\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
   never map to the same value.
@@
     constructor, the resulting value is crash-free.
-  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
+  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
+    ``type'' is used intuitively here: it means the constructor head is of type $T$.
+    } is crash-free implies
     that every value packed in it is crash-free. Recall that one can
     define projection on any argument of a value of type $t$. So if
-    the k-th argument of $t$ is not crash-free, then the k-th
+    the $k$-th argument of $t$ is not crash-free, then the $k$-th
     projection is a crash-free context that throws an expression that
@@
 
-\item $(\Phi_4)$ None of the $K_i$ is unreachable.
-\item One may want to also state that if $\vec{x} \neq \bad$ then
+\item $(\Phi_4)$ Now full application of a constructor $K_i$ is unreachable or crashing.
+\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
   $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
-  $\cf{\vec{x}} \to \cf{K_i(\vec{x})}$.
+  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
 \end{itemize}

### 2011-10-21 paper.tex
Paper: done editing section 3, and mostly done overall.
Added extensive notes about min, including the beginnings of an alternative translation
that seems more natural to me.

@@
 
-Consider:\note[nc]{Unlike the POPL '09 paper, $CF(e)$ is not implied by $e \in \{x|p\}$.}
+Consider:\note[nc]{Unlike the POPL '09 paper, $\cf{e}$ is not implied by $e \in \{x|p\}$.}
 \begin{verbatim}
@@
 \item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
+  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
   on its $k$-th component, $k = 1\text{ to }a_i$.
@@
 
-\item $(\Phi_4)$ Now full application of a constructor $K_i$ is unreachable or crashing.
+\item $(\Phi_4)$ No full application of a constructor $K_i$ is unreachable or crashing.
 \item One may want to also state that if for all $j$, $x_j \neq \bad$ then
@@
 We give in figure~\ref{ctrans} the translation of contract satisfaction.
-$true$ refers to the translation to a term of the data constructor
-\texttt{True} in $\hprime$, not to the actual true value.
+\annote[nc]{$true$ refers to the translation to a term of the data constructor
+\texttt{True} in $\hprime$, not to the actual true value.}{
+Again, we should use $True$ (or \texttt{True}) in the translation; the
+lowercase is an artifact of the concrete TPTP syntax.
+}
+\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
+c$. We have to do that because one is not the negation of the other,
+even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{
 
-Note that we define the translation of $f \in c$ and of $f \not \in
-c$. We have to do that because one is not the negation of the other,
-even though $\lnot (f \not \in c)$ implies $f \in c$.
+The contract translation, $\ctrans{}$, introduces a new predicate,
+$\min{}$, and contract non-satisfaction, $f \notin c$.  The formula
+$\min{e}$, short for ``(a)m in(terested) in $e$'' (?), is used to
+guide the proof search, by restricting proofs to follow the evaluation
+order when instantiating hypotheses (Is this meaningful? Is it even
+correct?).  To understand the translation intuitively, simply ignore
+all $\min{}$'s (I.e., assume $\forall x. \min{x}$).  The
+non-satisfaction form, $f \notin c$, is introduced to make the
+$\min{}$'s appear positively in translations in negative/contravariant
+positions; Ignoring $\min{}$'s, we have $\ctrans{f \notin c} \equiv
+\lnot \ctrans{f \in c}.$ (Well, after I change the definitions of
+$\ctrans{}$ for $e \in \{x | p\}$ and $e \notin \{x|p\}$ \ldots)
+
+Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
+$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
+  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
+\to \cfc$.  If we define 
+\protect \[ 
+\ctrans{g \in x:c1 \to c2} :=
+\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
+\]
+then we get stuck, because we must prove $\min{Z}$ to satisfy
+$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
+definition).  On the other hand, if we define
+\protect \[
+\ctrans{g \in x:c1 \to c2} 
+:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
+\]
+then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
+$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$).
+
+The irrelevance of $\lnot \min{Z}$ above leads to the question, can we
+do away with the $\min{}$'s altogether in such positions?  It turns
+out the answer is ``No'', when we consider proving $g ~(g ~Z) \in
+\cfc$, but I have some ideas for an alternative translation where the
+answer is ``Yes''.  But first $g ~(g ~Z)$: to use $g$'s contract
+assumption we need to prove $\lnot \min{g ~Z} \lor \cf{g ~Z}$, and to
+prove $\cf{g ~Z}$ we in turn need to use $g$'s contract again.  To use
+$g$'s contract again we need $\min{g ~Z}$, and so $\min{}$ is relevant
+again.  The solution is to invoke excluded middle on $\min{g ~Z}$:
+either $\min{g ~Z}$, and we're done, or $\lnot \min{g ~Z}$, in which
+case we satisfy $g$'s preconditions \emph{without} showing $g$'s
+argument to be crash free.  This seems a little surprising to me,
+making a translation with no $\min{}$ here even more appealing.
+
+Here's a start at such a translation.  I
+think it's more natural to think about variance, so write
+$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
+a system that works for these $\cfc$ examples is
+\protect \begin{eqnarray*}
+\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
+\ctrans{x \in \cfc}^- &:=& \cf{x}\\
+\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
+  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
+\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
+  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
+\end{eqnarray*}
+Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
+it's natural to think of contract assumptions in the theory as
+negative, and contract goals as positive.  An example is helpful:
+\protect \begin{eqnarray*}
+\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
+&&\text{whereas}\\
+\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
+\end{eqnarray*}
+Now, given $\ctrans{g \in \cfc \to \cfc}$, we may 
+want to prove $\ctrans{g ~(g ~Z) \in \cfc}$ 
+from $\ctrans{g \in \cfc \to \cfc}^-$ (note the negative).
+So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
+This time the proof proceeds more naturally (no EM).  
+From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
+$\min{g ~Z} \to \cf{g ~Z}$.
+But that follows by instantiating $g$'s contract again, and 
+then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.
+
+Notice the assymetry between the positive and negative arrow
+translations: the positive case has no $\min{f ~x}$ assumption.  I
+left this $\min{}$ out because it's already in the 
+conclusion of the $\cf$ example, but I certainly haven't
+worked enough examples to justify it.  Moreover, this fragment
+(no $e \in \{x|p\}$ yet) already fails to have a standard
+property of bidirectional systems: that you can check/verify what you assume/use.
+In other words, we can't prove that, forall $e$ and $c$,
+\protect \[
+  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
+\]
+The trouble seems to come down to needing to prove $\min{e ~x} \implies
+\min{x}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
+assumption to the positive arrow translation to recover the
+ ``negative implies positive'' property.
+
+Going back to the current translation, we consider some of it's
+metatheoretic properties.
+\protect \begin{enumerate}
+\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
+  translation, in the sense that $\ctrans{e \in c}$ should imply
+  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
+  predicate case so that the implication holds (e.g. by more accurately
+  modeling the operation semantics of contracts), and add a symmetric 
+  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
+  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
+  not imply $\ctrans{x \in \cfc}$.
+\item The ``fancy'' $\min{}$-ified translation should 
+  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
+  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
+  plus that axiom.  In symbols, we have
+  \protect \[
+  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
+  \]  I.e., ``\min{} is sound for `naive'''.
+\item We'd like to have a result  like
+  \protect \[
+  \text{Given some validity or consistency assumptions on $\Gamma$ and $e \in c$}, 
+    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
+  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
+  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
+  completeness theorem without some work, since we're purposefully 
+  restricting the proof search to follow
+  evaulation strategy, and it's far from obvious that all proofs can proceed this way.
+
+\end{enumerate}
+
+Concerns about the current translation
+\protect \begin{enumerate}
+\item The changes needed to make the first property above hold (a $\min{}$ in the negative
+  arrow case and symmetric predicate contract cases).
+  I had trouble proving
+  \protect \[
+  (\lambda x. \text{case $x$ of $True \to True$; $False \to \bad$}) 
+  \in \{x|x==True\} \to \cfc
+  \]
+  , after adding $p = \unr$ to $\ctrans{x \in \{x|p\}}$, but I need to think about it more.
+
+\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
+  it's impossble to prove $\ctrans{True \in \cfc}$, because you can't prove 
+  $\min{True}$ a priori.  On the other hand, we don't need any extra 
+  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
+  very simple examples (checked on the simple
+  example that $id \in \cfc \to \cfc$).
+
+\item A related question, is what do we feed to Equinox?  Equinox is a 
+  refutation prover, so to prove $\Gamma \vdash \phi$ we ask if 
+  $\Gamma,\lnot \phi \vdash \bot$.  So, for us, when $\phi$ is $\ctrans{e \in c}$, do
+  give Equinox $\Gamma,\lnot \ctrans{e \in c}$ or $\Gamma, \ctrans{e \notin c}$.  I think
+  we want the former, since ignoring Equinox our goal is $\Gamma \vdash \ctrans{e \in c}$.
+\end{enumerate}
+}
 

### 2011-10-24 paper.tex
Paper: completed the new ctrans and related it to old ctrans.
The new ctrans is nearly the same as the old, if viewed the right way, once
we fix the old trans in the ways we had already wanted.  I also added an alternate
dtrans.

@@
 $\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
-definition).  On the other hand, if we define
+definition).  (UPDATE: Actually, this is probably just more evidence
+that $\ctrans{e \in \cfc}$ is defined incorrectly: we were already
+worried about that $\min{}$, because it stops us from proving
+$\cf{Z}$, even when no arrows are involved).  On the other hand, if we
+define the arrow case as in Figure~\ref{ctrans}, namely
 \protect \[
@@
 then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
-$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$).
+$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$. UPDATE: when you
+notice that that implies $\min{Z} \to \cf{Z}$, (and the reverse follows from EM), the next 
+paragraph might become a little dubious.  I still think the second form 
+is more intuitive.).
 
@@
 \end{eqnarray*}
-Now, given $\ctrans{g \in \cfc \to \cfc}$, we may 
-want to prove $\ctrans{g ~(g ~Z) \in \cfc}$ 
-from $\ctrans{g \in \cfc \to \cfc}^-$ (note the negative).
+Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
+want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
 So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
@@
 From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
-$\min{g ~Z} \to \cf{g ~Z}$.
+$\min{g ~Z} \to \cf{g ~Z}$ (UPDATE: this is actually the same as 
+before, up to using EM to rewrite the implication).
 But that follows by instantiating $g$'s contract again, and 
@@
 \]
-The trouble seems to come down to needing to prove $\min{e ~x} \implies
-\min{x}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
+The trouble seems to come down to needing to prove $\min{e ~x}$ implies 
+$\min{e}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
 assumption to the positive arrow translation to recover the
- ``negative implies positive'' property.
-
-Going back to the current translation, we consider some of it's
+ ``negative implies positive'' property. (UPDATE: the symmetric $\min{}$
+may also be necessary when reasoning about functions that return functions: if
+$f ~x = g$ then $f ~x ~y = g ~y$, but we'd need a $\min{f ~x}$ to unfold $f$'s def).
+
+Going back to the current translation, we consider some of its
 metatheoretic properties.
@@
   \]  I.e., ``\min{} is sound for `naive'''.
-\item We'd like to have a result  like
+\item We'd like to have a result  like: given some validity or consistency 
+  assumptions on $\Gamma$ and $e \in c$,
   \protect \[
-  \text{Given some validity or consistency assumptions on $\Gamma$ and $e \in c$}, 
     \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
@@
   very simple examples (checked on the simple
-  example that $id \in \cfc \to \cfc$).
+  example that $id \in \cfc \to \cfc$). (UPDATE: removing the $\min{x} \land$ in $cfc$, 
+  or changing it to $\min{x} \to$, is probably the solution).
 
@@
 }
+
+\note[nc]{Here's a more fleshed out alternate translation, and a
+  comparison with the current translation.}
+\small
+\begin{verbatim}
+new translation
+===============
+
+ -- you can assume both min()s when proving e in {x|p} 
+[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
+ -- you need to prove min(e) but then get min(p[e/x])
+[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
+ -- could also write this as 'e /= unr -> p[e/x] in {true,unr}'
+[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}
+
+[[e in cf]]+ := min(e) -> cf(e)
+-- i think we can add a 'min(e) ->' here without losing anything.
+[[e in cf]]- :=           cf(e)
+
+-- should add a 'min(e x) ->' here as discussed above.
+[[e in x:c1 -> c2]]+
+  := forall x.             [[x in c1]]- -> [[e x in c2]]+   
+[[e in x:c1 -> c2]]-
+  := forall x. min(e x) -> [[x in c1]]+ -> [[e x in c2]]-
+
+old translation (written in implication style)
+==============================================
+
+[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
+not[[e notin {x|p}]] := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)
+
+[[e in cf]]       := min(e) /\ cf(e)
+not[[e notin cf]] := min(e) -> cf(e)
+
+[[e in x:c1 -> c2]] 
+  := forall x. min(e x) -> not[[x notin c1]] -> [[e x in c2]]
+not[[e notin x:c1 -> c2]]
+  := forall x.             [[x in c1]] -> not[[e x notin c2]]
+
+updated cf and {x|p} rules for old translation
+==============================================
+
+[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
+not[[e notin {x|p}]] := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})
+
+-- what we actually discussed was adding a top level min(e) assumption, but
+-- removing the min(e) goal has a similar effect.
+[[e in cf]]       :=           cf(e)
+not[[e notin cf]] := min(e) -> cf(e)
+\end{verbatim}
+\note[nc]{And what do we see?
+If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
+then the ``new'' translation looks almost the same,
+except it has more $\min{}$'s in the predicate rules.  So, have we learned
+anything? Probably:
+\protect \begin{enumerate}
+  \item We're more confident about our suggested changes to the old translation,
+        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
+  \item The new translation is more intuitive / we might understand how Koen came up
+        with the old translation.
+  \item We can answer the question about what we give as input to Equinox: for the
+        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
+        That is consistent with giving Equinox $\in$ axioms.  
+        What to give as the goal is maybe
+        less obvious, since Equinox is a refutation prover, but consider: ignoring the
+        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
+        But
+        that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.  
+        And going back
+        to the old translation, the analogy is 
+        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
+\end{enumerate}
+
+The predicate case for the new translation came from an example:
+}
+\begin{verbatim}
+try:
+
+[[e in {x|p} ]]+ = min(e) -> min(p[e/x]) -> [[e in {x|p}]]-
+[[e in {x|p} ]]- = e = unr \/ p[e/x] in {true, unr}
+
+example:
+
+e := (\x.case x of True -> True) -- so e false = bad
+c := {x|x==True} -> CF
+[[e in c]]+ = forall x. min(e x) -> [[x in {x|x==True}]]- -> [[e x in CF]]+
+            = forall x. min(e x) -> (x = unr \/ p in {true,unr})
+                                 -> (min(e x) -> CF(e x))
+
+min(e x) -> min(x) /\ (
+         \/ (x = bad /\ (e x) = bad) 
+         \/ (x = True /\ (e x) = true) 
+         \/ (x = False /\ (e x) = bad)
+ -- better to add the redundant x /= bad, true, false here?
+         \/ (e x) = unr
+         )
+
+if x = unr then (e x) = unr and so CF(unr) and done
+
+if x /= unr /\ p = true then we're stuck, since we can't 
+unroll the definition of ==true in p
+
+so, change the defs:
+
+[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
+[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
+[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}
+
+starting again, we have min(x) from D[[]] so we get
+
+min(p) /\ (x = unr \/ p in {true,unr})
+
+now
+
+ - x = unr gives (e x) = unr as before
+
+ - for x/=unr and p in {true,unr} we unroll def (==true) 
+   using min(p), for p = ((==True) x):
+
+   min((==True) x) -> min(x) /\ ( -- e' is (==True) here
+         \/ (x = bad /\ (e' x) = bad) 
+         \/ (x = True /\ (e' x) = true) 
+         \/ (x = False /\ (e' x) = bad)
+         \/ (e' x) = unr
+         )
+
+for p = unr, we get x notin {bad,true,false}, 
+and hence (e x) = unr, and we're done.
+
+for p = true, we get x = true, 
+and hence (e x) = true, and we're done.
+\end{verbatim}
+\note[nc]{That example also has an alternate version of the
+ definition translation for functions defined by case
+matching.  It's a little more concise than the old translation.
+}
+\begin{verbatim}
+In any case, another question: 
+does equinox distinguish between D[[]] like
+
+  (x = Nil /\ f x = eNil)
+  \/
+  (exists y,ys. x = Cons(y,ys) /\ f x = eCons)
+
+and D[[]] like
+
+  (x = Nil /\ f x = eNil)
+  \/
+  (x = Cons(cons_1 x,cons_2 x) /\ f x = eCons[cons_1 x/y, cons_2 x/ys])
+
+the latter is nice because we don't need to invent names (consider
+
+  f x = case x of
+          K x -> x
+
+we can't write
+
+  (exists x. x = K x /\ f x = x)
+
+since the two x's equated are not the same!). 
+the former is nice because it's more concise?
+\end{verbatim}
 

### 2011-10-26 paper.tex
Paper: formatting.
Paper: expanded prelude and contract translation sections, including figures.
Put the new contract translation presentation in a figure, and wrote a few paragraphs
explaining the motivation behind the min()s.

@@
 \renewcommand{\min}[1]{\mbox{min}(#1)}
+
+\newcommand{\designChoice}{(DESIGN CHOICE/OPTIONAL)}
 
@@
 \]?
-The answer is that this is an expansion of the $etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.
+The answer is that this is an expansion of the $\etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.
 I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.
@@
   \label{defnotnull}
-\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true
+\forall xs,y,ys.~ xs=Cons(y,ys) \implies notnull(xs) = true
 \end{equation}
@@
 We also need to specify the translation of calls to $notnull$ with
-$\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
+$\bad$ values, to encode the fact that $notnull(\bad) = \bad$.
 \begin{equation*}
@@
 We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e.
-forall $phi$, $\forall x,y.x=y \to phi \iff phi[y/x]$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
+forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
+\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.}
 
@@
   \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
-    ``type'' is used intuitively here: it means the constructor head is of type $T$.
+    ``type'' is used loosely here: it means the constructor head is of type $T$.
     } is crash-free implies
@@
       \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
-      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a\text{\note[nc]{This was $K_i(\vec{x})$.
-Typo, right?}}}) \\
+      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
       \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
@@
 
-\subsection{$\ctrans{}$ -- Contracts}
+\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
+\note[nc]{%
+Figure~\ref{fig:newCTrans} contains the new min contract translation, and
+Figure~\ref{fig:naiveCTrans} contains the naive contract translation
+with no $\min$'s.  The min translation reduces to the naive translation when
+all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.
+
+The min translation has two forms, $\ctrans{}^-$ for 
+assumptions/axioms and $\ctrans{}^+$
+for conclusions/theorems.  The choice of minus and 
+plus comes from the negative and positive
+positions in an implication: $- \to +$.  The goal of the $\min$'s is to limit the
+applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ktrans{}$), 
+in order to prune 
+the search space.  So, $\min$'s
+appear as obligations (the premises in implications) 
+in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
+and the translation is designed so that $\min{e}$ is introduced whenever
+$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
+obligations that prevent using the axioms on ``uninteresting'' terms.
+
+For example, consider $\ctrans{e \in \cfc}^+$ and $\ctrans{e \in \cfc}^-$
+in Figure~\ref{fig:newCTrans}.  In the plus-translation, $\min{e}$ is a
+premise that can be used to prove $\cf{e}$, whereas in the minus-translation, either
+there is no $\min{e}$, or $\min{e}$ is an obligation that must be
+satisfied before $\cf{e}$ may be concluded.  In the plus-translation, we
+get interest ($\min{e}$) in $e$ because we are trying to prove something about $e$,
+whereas in the minus-translation, we might have to prove we're interested in $e$ before
+we can conclude that $e$ is crash free (this latter $\min{e}$ obligation is
+a design choice because it's not clear why we'd want to constrain the use
+of $\cf{e}$: we'd need to prove $\min{e}$ anyway before instantiating 
+another axiom at $e$).
+
+As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
+in Figure~\ref{dtrans}.  Definition translations are axioms, 
+so the the initial $\min{f(\vec{x})}$
+is an obligation: we aren't allowed to 
+``unfold the definition'' of the function unless we prove
+that we're interested in a particular application.  
+Once that obligation is satisfied, interest ($\min{e}$) in the
+scrutinee of the case match may be concluded, because the function
+must inspect the scrutinee to compute the result.
+
+Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
+Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
+assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
+we need to prove we're interested in $e$.  The $\min{e}$ constraint
+here is not optional, because in addition to concluding 
+$e \neq \unr \to p[e/x] \in \{True,\unr\}$, we also conclude $\min{p[e/x]}$.
+The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
+$p[e/x] \in \{True,\unr\}$, and because it can be used to unfold definitions,
+we guard it with $\min{e}$.
+}
+\begin{figure*}
+\begin{verbatim}
+[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
+[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
+[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}
+
+[[e in cf]]+ :=  min(e) ->  cf(e)
+[[e in cf]]- := [min(e) ->] cf(e)
+
+-- 'min's optional if a
+-- 'forall f,x. min(f x) -> min(f)' axiom
+-- is added to prelude.
+[[e in x:c1 -> c2]]+
+  := forall x. [min(e x) ->] [[x in c1]]- -> [[e x in c2]]+   
+[[e in x:c1 -> c2]]-
+  := forall x. [min(e x) ->] [[x in c1]]+ -> [[e x in c2]]-
+\end{verbatim}
+\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
+of optional parts, are enclosed in single square brackets.}
+\label{fig:newCTrans}
+\end{figure*}
+
+\begin{figure*}
+\begin{verbatim}
+[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
+[[e in cf]]         := cf(e)
+[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
+\end{verbatim}
+\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
+\label{fig:naiveCTrans}
+\end{figure*}
+
 We give in figure~\ref{ctrans} the translation of contract satisfaction.
@@
 
-[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
-not[[e notin {x|p}]] := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)
+[[e in {x|p}]]       
+  := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
+not[[e notin {x|p}]] 
+  := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)
 
@@
 
-[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
-not[[e notin {x|p}]] := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})
-
--- what we actually discussed was adding a top level min(e) assumption, but
+[[e in {x|p}]]       
+  := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
+not[[e notin {x|p}]] 
+  := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})
+
+-- what we actually discussed was 
+-- adding a top level min(e) assumption, but
 -- removing the min(e) goal has a similar effect.
@@
 c := {x|x==True} -> CF
-[[e in c]]+ = forall x. min(e x) -> [[x in {x|x==True}]]- -> [[e x in CF]]+
-            = forall x. min(e x) -> (x = unr \/ p in {true,unr})
+[[e in c]]+ = forall x. min(e x) 
+              -> [[x in {x|x==True}]]- -> [[e x in CF]]+
+            = forall x. min(e x) 
+              -> (x = unr \/ p in {true,unr})
               -> (min(e x) -> CF(e x))
@@
   \/
-  (x = Cons(cons_1 x,cons_2 x) /\ f x = eCons[cons_1 x/y, cons_2 x/ys])
-
-the latter is nice because we don't need to invent names (consider
+  (x = Cons(cons_1 x,cons_2 x)
+   /\ f x = eCons[cons_1 x/y, cons_2 x/ys])
+
+the latter is nice because we don't need 
+to invent names (consider
 
@@
 \subsection{Prelude}
-There are some formulae that should always be included in our FO theory.
-
-We need to state that $\bad$ is not crash-free with the formula:
+Figure~\ref{fig:prelude} summarizes the common axioms we add to all
+translations.  We 
+\begin{itemize}
+  \item state that $\bad$ is not crash-free with the formula:
 $\lnot \cf{\bad}$.
 
-\add[nc]{We also state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
-
+  \item \add[nc]{state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
 This is already present in the implementation, and assumed in the
@@
 
-Plus we need to give formulae for the boolean datatype and for
- unreachability. Strictly speaking, we can omit them and just add the
-following lines to source files:
+  \item give formulae for the boolean datatype, equivalent to
+translating
 \begin{verbatim}
-data UNR = UNR
-data Bool = True | False
+data Bool = True 0 | False 0
 \end{verbatim}
-But given that those datatypes are used by our translation, we can
-just directly include their translation every time we translate a
-module.
+
+  \item \designChoice state that interest in an application implies
+  interest in the function: $\forall f,x. \min{f ~x} \to \min{f}$.  This axiom can
+  be used to eliminate some $\min{}$'s in the translations for arrow contracts.
+
+  \item \designChoice state that applying $\bad$ and $\unr$ as functions
+  results in $\bad$ and $\unr$, respecitively: \[
+  \forall x. \min{\bad ~x} \to \bad ~x = \bad \land \min{\unr ~x} \to \unr ~x = \unr.
+  \]
+
+  \item \designChoice state that applying a $\cfc$ expression can be used like a
+  $\cfc \to \cfc$ function: $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$.
+  Simon suggested this to allow one to write, e.g., $f ::: \cfc$ instead of 
+  $f ::: \cfc \to \cfc$.
+\end{itemize}
+
+\begin{figure}\label{fig:prelude}
+  \begin{itemize}
+    \item $\lnot \cf{\bad}$
+    \item $\cf{\unr}$
+    \item $\dtrans{data~Bool = True~0 | False~0}$
+    \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$
+    \item \designChoice\\ $\forall x. \min{\bad ~x} \to \bad ~x = \bad$ 
+    \item \designChoice\\ $\forall x. \min{\unr ~x} \to \unr ~x = \unr$
+    \item \designChoice\\ $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$
+    \item 
+  \end{itemize}
+  \caption{Prelude -- Axioms common to all translations.}
+\end{figure}
 

### 2011-10-26 contracts/Haskell.hs
Variable names and formatting changes.

@@ -101,12 +101,14 @@ substs :: [(Expression, Variable)] -> Expression -> Expression
 substs [] e = e
 substs ((x,y):xys) e = substs xys $ subst x y e
 
-subst :: Expression -> Variable -> Expression -> Expression -- e[x/y]
-subst x y (Var v) | v         == y = x
+-- 'subst e1 y e2' = e2[e1/y]
+subst :: Expression -> Variable -> Expression -> Expression
+subst e y (Var v) | v == y    = e
                   | otherwise = Var v
-subst x y (App e1 e2)         = App (subst x y e1) (subst x y e2)
-subst x y (FullApp f es)      = let Var x' = (subst x y (Var f)) in FullApp x' $ map (\e -> subst x y e) es
-subst x y BAD                 = BAD
+subst e y (App e1 e2)         = App (subst e y e1) (subst e y e2)
+subst e y (FullApp f es)      = let Var f' = (subst e y (Var f))
+                                in FullApp f' $ map (subst e y) es
+subst _ _ BAD                 = BAD
 
 
 substsC :: [(Expression,Variable)] -> Contract -> Contract

### 2011-10-26 contracts/FOL.hs
Finished folding the uppercasing of variables into the translation.
Also made the dTrans for functions defined by pattern matching less obfuscated.
Might be a good time to switch to better solution to fresh variables and substitution,
e.g. use the unbound library.

Still need to untangle the f_ptr and f_p stuff ...

@@ -5,7 +5,7 @@ module FOL where
 import qualified Haskell as H
 import Debug.Trace
 import Data.Char (toUpper)
-import Data.List (intersperse)
+import Data.List (intercalate)
 
 type Term = MetaTerm Variable
 type Formula = MetaFormula Term
@@ -32,7 +32,7 @@ instance Show a => Show (MetaTerm a) where
   show (App [t]) = show t
   show (App ts) = "app(" ++ show (App (init ts)) ++ "," ++ show (last ts) ++ ")"
   show (FullApp f []) = show f
-  show (FullApp f as) = show f ++ "(" ++ (concat $ intersperse "," $ map show as) ++ ")"
+  show (FullApp f as) = show f ++ "(" ++ (intercalate "," $ map show as) ++ ")"
 
 
 infix 7 :<=>:
@@ -79,34 +79,8 @@ removeConstants f = f
 
 simplify f = filter (/= Top) $ splitOnAnd $ removeConstants f
 
-extractVR (Var (Regular x)) = x
-
-
--- The TPTP spec says that only quantified variables should be
--- uppercase So we make everything lowercase (it's supposed to be
--- already done) And upperIfy makes the good variables uppercase.
-upperIfy :: [String] -> Formula -> Formula
-upperIfy c (Forall xs f) = Forall (map (Var . Regular . map toUpper . extractVR) xs) (upperIfy c' f)
-  where c' = (map extractVR xs)++c
-upperIfy c (f1 :<=>: f2) = (upperIfy c f1) :<=>: (upperIfy c f2)
-upperIfy c (f1 :=>: f2) = (upperIfy c f1) :=>: (upperIfy c f2)
-upperIfy c (Not f) = Not (upperIfy c f)
-upperIfy c Top = Top
-upperIfy c Bottom = Bottom
-upperIfy c (t1 :=: t2) = (auxUpper c t1) :=: (auxUpper c t2)
-upperIfy c (t1 :/=: t2) = (auxUpper c t1) :/=: (auxUpper c t2)
-upperIfy c (CF t) = CF (auxUpper c t)
-upperIfy c (And fs) = And (map (upperIfy c) fs)
-upperIfy c (Or fs) = Or (map (upperIfy c) fs)
-  
-auxUpper :: [String] -> Term -> Term
[…]

### 2011-10-26 contracts/Translation.hs

@@ -86,6 +86,7 @@ dTrans (H.Let f vs e) = do
         fptr3 = F.Forall vs' $ (F.FullApp (F.Regular (f ++ "p")) vs') :=: (F.App $ (F.Var . F.Regular) (f++"p_ptr") : vs')
 -}
 
+-- Recall that the patterns 'pes' has the form [([Variable],Expression)].
 dTrans (H.LetCase f vs e pes) = do
   let vs'  = map makeVar vs
       -- Uppercasified variables wrapped in Haskell constructors.
@@ -94,11 +95,25 @@ dTrans (H.LetCase f vs e pes) = do
   -- Uppercasify any quantified vars in e before translation.
   et <- eTrans $ sub e
   ft <- eTrans $ H.Var f
-  let zedify ei pi = foldl (\e (v,z) -> H.subst (fVar2HVar z) v e) ei (take (length (tail pi)) $ zip (tail pi) zs)
-      arities = map (\p -> (head p, length $ tail p)) $ map fst pes :: [(String,Int)]
-      zs = [F.Var $ F.Regular $ "Zdef" ++ show x | x <- [1..(foldl1 max [snd y | y <- arities])]]
-  tpieis <- sequence [eTrans (zedify ei pi) | (pi,ei) <- pes]
-  let eq1 = [(et :=: (F.FullApp (F.Regular $ head pi) (take (length pi - 1) [ z | (v,z) <- zip (tail pi) zs ]))) :=>: (F.FullApp (F.Regular f) vs' :=: tpiei) | ((pi,ei),tpiei) <- zip pes tpieis]
+  --pes' <- map
+  -- A Pattern is a [Var], e.g. 'Cons x xs' ==> ['Cons','x','xs'], so
+  -- the 'tail' of a pattern is the variables.  The 'arities' below would
+  -- be simpler if Pattern were (Var,[Var]), e.g. 'Cons x xs' ==> ('Cons',['x','xs']).
+  let -- e.g. [('Cons',2),('Nil',0)]
+      arities = [(c, length xs) | ((c:xs),_) <- pes] :: [(String,Int)]
+      -- A list of "fresh" variables. Length chosen to be large enough
+      -- to substitute for any constructors pattern variables.
+      zs = makeVars (maximum $ map snd arities) "Zdef"
+      -- Variables wrapped in Haskell constructors.
+      zsH = map fVar2HVar zs
+      -- NB: must substitute pattern variables before definition variables,
[…]

### 2011-10-27 paper.tex
Paper: added two questions and fixed some syntax to correspond to implementation.

@@
 \begin{verbatim}
-data List a = Nil | Cons a (List a)
+data List = Nil 0 | Cons 2
 
@@
   Nil -> False
-  Cons(x,y) -> True
+  Cons x y -> True
 
@@
   Nil -> BAD
-  Cons(x,y) -> x
+  Cons x y -> x
 \end{verbatim}
@@
   $$ \ruleform{\ktrans{data~def} = \Phi} $$
-$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
+$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
  \hspace{5.6cm}where
@@
     \[  \begin{array}{rcl}
-      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
+      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
+$\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
       \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
@@
 Where $M - f $ means the content of the module $M$ without $f$'s
-definition and $f$'s contract. TODO Stress that it's not always enough
-and that we may have to unroll several times!
+definition and $f$'s contract. XXX, TODO Stress that it's not always enough
+and that we may have to unroll several times!\note[nc]{??? Why? Need an example here.}
 

### 2011-10-27 contracts/FOL.hs
Abstracted name generation for "abstract recursive" and "pointer" functions.

@@ -106,7 +106,7 @@ toTPTP f = header ++ "\n" ++ go f ++ "\n" ++ footer
 appifyF :: [H.Type H.Variable] -> [Formula] -> [Formula]
 appifyF a fs = map (fmap go) fs
   where go (Var (Regular v)) = case H.lookupT v (trim a) of
-          Just n -> Var (Regular $ v ++ "_ptr")
+          Just n -> Var (Regular $ H.makePtr v)
           Nothing -> Var $ Regular v
         go (App ts) = App (map go ts)
         go (FullApp f ts) = FullApp f (map go ts) 

### 2011-10-27 contracts/Haskell.hs

@@ -52,12 +52,18 @@ data Type a = Fun a Int
             | Cons a Int
             deriving (Eq,Show)
 
+-- Make a name for abstract recursive occurences of a function
+makeRec f = f ++ "_rec"
+
+-- Make a name for the curried ("pointer") version of a function
+makePtr f = f ++ "_ptr"
+
 apps xs = foldl1 App xs
 
 -- returns the arities of data constructors and functions
 arities :: Program -> [Type Variable]
 arities x = go x >>= \s -> case s of
-  Fun f n -> [Fun f n, Fun (f++"p") n]
+  Fun f n -> [Fun f n, Fun (makeRec f) n]
   d -> [d]
   where go [] = []
         go (Def d:ds) = go2 d:go ds
@@ -83,7 +89,7 @@ appifyExpr a e = go a 1 e []
   where go a count g@(App (Var v) e) acc = case lookupT v a of
           Just n -> if count == n 
                     then FullApp v (e':acc)
-                    else apps (App (Var $ v ++ "_ptr") e':acc)
+                    else apps (App (Var $ makePtr v) e':acc)
           Nothing -> apps (App (Var v) e':acc)
           where e' = go a 1 e []
         go a count g@(App e1 e2) acc = go a (count+1) e1 (acc++[go a 1 e2 []])
@@ -91,7 +97,7 @@ appifyExpr a e = go a 1 e []
         go a count BAD acc = BAD
         go a count (Var v) acc = case lookupT v a of
           Just 0 -> Var v
-          Just n -> Var $ v ++ "_ptr"
+          Just n -> Var $ makePtr v
           Nothing -> Var v
 
 

### 2011-10-27 contracts/Translation.hs

@@ -47,6 +47,12 @@ makeSel k d = "sel_"++show k++"_"++unquote d where
   -- in the parser.  I.e., *not* have single quotes yet at this point.
   unquote ('\'':cs) = init cs
 
+{- Moved to Haskell.hs to avoid module import cyle -}
+-- Make a name for abstract recursive occurences of a function
+-- makeRec f = f ++ "_rec"
+-- Make a name for the curried ("pointer") version of a function
+-- makePtr f = f ++ "_ptr"
+
 -- Expression translation
 -------------------------
 
@@ -81,9 +87,9 @@ dTrans (H.Let f vs e) = do
 {-
         -- fptri are equations defining functions relatively to their app counterparts.
         -- eg that app(app(f_ptr,x),y) = f(x,y)
-        fptr1 = (F.Forall vs' $ (F.And [F.CF v | v <- vs']) :=>: F.CF (F.FullApp (F.Regular f) vs')) :<=>: (F.CF $ F.Var $ F.Regular (f++"_ptr"))
-        fptr2 = F.Forall vs' $ (F.FullApp (F.Regular f) vs') :=: (F.App $ (F.Var . F.Regular) (f++"_ptr") : vs')
-        fptr3 = F.Forall vs' $ (F.FullApp (F.Regular (f ++ "p")) vs') :=: (F.App $ (F.Var . F.Regular) (f++"p_ptr") : vs')
+        fptr1 = (F.Forall vs' $ (F.And [F.CF v | v <- vs']) :=>: F.CF (F.FullApp (F.Regular f) vs')) :<=>: (F.CF $ F.Var $ F.Regular $ makePtr f)
+        fptr2 = F.Forall vs' $ (F.FullApp (F.Regular f) vs') :=: (F.App $ (F.Var . F.Regular) (makePtr f) : vs')
+        fptr3 = F.Forall vs' $ (F.FullApp (F.Regular (H.makeRec f)) vs') :=: (F.App $ (F.Var . F.Regular) (makePtr $ makeRec f) : vs')
 -}
 
 -- Recall that the patterns 'pes' has the form [([Variable],Expression)].
@@ -121,9 +127,9 @@ dTrans (H.LetCase f vs e pes) = do
[…]

### 2011-10-27 contracts/FOL.hs
Finished removing _ptr support by adding "appification" support for contracts.
The _ptr support is now completely optional for examples that only use
full application. "appification" is the process of recognizing full
applications and treating them specially.

Interestingly, there are now some improvements to the tests:

ok/{multgt.hs,nateqsymmetry.hs}: now pass
diverges/{badder.hs,extrasimple.hs}: now fail by being satisfiable
  (this is good: these tests assert CF(BAD))

The remaining regressions, compared with the master branch, are due
to no _ptr support:

ok/{add.hs,implies.hs,test.hs}: n

@@ -20,7 +20,9 @@ instance Show Variable where
   show BAD = "bad"
   show UNR = "unr"
 
-
+-- XXX, ???: Is there any reason we don't just use
+-- Haskell.MetaExpression ? That would simplify things (e.g. appifyF
+-- could use appifyExpr instead of duplicating it).
 data MetaTerm a = Var a
                 | App [MetaTerm a]
                 | FullApp a [MetaTerm a]
@@ -79,8 +81,13 @@ removeConstants f = f
 
 simplify f = filter (/= Top) $ splitOnAnd $ removeConstants f
 
+
 toTPTP :: Formula -> String
 toTPTP f = header ++ "\n" ++ go f ++ "\n" ++ footer
+  -- XXX, MAYBE TODO: add better header or comments. Right now the first "axiom"
+  -- below is the name of the axiom.  The TPTP format also allows optional 4th and
+  -- 5th fields for comments.  Would be nice to see each formula labeled with its
+  -- type or source, to make debugging the generated .tptp file easier.
   where header = "fof(axiom,axiom,"
         footer = ").\n"
         go (Forall xs f) = "! " ++ show xs ++ "  : (" ++ go f ++ ")"
@@ -102,13 +109,25 @@ toTPTP f = header ++ "\n" ++ go f ++ "\n" ++ footer
         goTerm (FullApp f []) = show f
         goTerm (FullApp f as) = show f ++ "(" ++ (intercalate "," $ map show as) ++ ")"
 
--- takes a program and a list of arities for each definition
+-- takes formulas and a list of arities for each definition
+-- and returns those formulas using "full application" whereever possible
 appifyF :: [H.Type H.Variable] -> [Formula] -> [Formula]
 appifyF a fs = map (fmap go) fs
-  where go (Var (Regular v)) = case H.lookupT v (trim a) of
-          Just n -> Var (Regular $ H.makePtr v)
-          Nothing -> Var $ Regular v
-        go (App ts) = App (map go ts)
-        go (FullApp f ts) = FullApp f (map go ts) 
-        go t = t
-        trim = filter (\s -> case s of H.Fun _ _ -> True; _ -> False)
+  -- XXX, REFACTOR: copied from Haskell.hs appifyExpr.  This is a
+  -- separate function because the Haskell and FOL use different
+  -- expression types, but this seems completely unnecessary.
+  where go :: Term -> Term
+        go (App (App es1 : es2)) = go (App (es1++es2))
[…]

### 2011-10-27 contracts/Haskell.hs

@@ -85,20 +85,28 @@ lookupT v (Cons f n:as) =  if f == v  then Just n else lookupT v as
 
 -- takes a program and a list of arities for each definition
 -- returns the same program but using full application wherever possible
-appifyExpr a e = go a 1 e []
-  where go a count g@(App (Var v) e) acc = case lookupT v a of
-          Just n -> if count == n 
-                    then FullApp v (e':acc)
-                    else apps (App (Var $ makePtr v) e':acc)
-          Nothing -> apps (App (Var v) e':acc)
-          where e' = go a 1 e []
-        go a count g@(App e1 e2) acc = go a (count+1) e1 (acc++[go a 1 e2 []])
-        go a count (FullApp v es) acc = FullApp v $ map (\e -> go a 1 e []) es
-        go a count BAD acc = BAD
-        go a count (Var v) acc = case lookupT v a of
+appifyExpr :: [Type Variable] -> Expression -> Expression
+appifyExpr a e = go e []
+  -- 'a' is the arities, 'args' is the arguments to the enclosing applications.
+  where go (App (Var v) e) args = case lookupT v a of
+          Just n -> if length args' == n
+                    then FullApp v args'
+                    else apps (App (Var $ makePtr v) e' : args)
+          Nothing -> apps (App (Var v) e' : args)
+          where e' = go e []
+                args' = e':args
+
+        go (App e1 e2) args = go e1 (args++[go e2 []])
+
+        -- There should be no enclosing applications in these case, so no args.
+        go (FullApp v es) [] = FullApp v $ map (\e -> go e []) es
+        go (Var v) [] = case lookupT v a of
           Just 0 -> Var v
-          Just n -> Var $ makePtr v
+          Just n -> Var $ makePtr v -- XXX, ??? BUG: Won't this make 'f_ptr_ptr' ???
           Nothing -> Var v
+        -- XXX, ??? REMOVE: this is a weird place to simplify using
+        -- the reduction semantics for 'BAD'.
+        go BAD _ = BAD
 
 
 -- Bunch of substitution utility

### 2011-10-27 contracts/Translation.hs

@@ -223,6 +223,10 @@ isToCheck fs (H.Def (H.LetCase f _ _ _))   = f `elem` fs
 isToCheck fs (H.ContSat (H.Satisfies f _)) = f `elem` fs
 isToCheck _ _                              = False
 
+-- XXX, ???: "appification" is an optimization.  Do we need it? It
+-- also makes the generated formulas simpler, but at the cost of
+-- complicating the generation code.  Would like to see how much if
+-- any it speeds up Equinox in practice.
 trans :: H.Program -> [H.Variable] -> [F.Formula]
 trans ds fs = evalState (go fs ((H.appify) ds)) (S "Z" 0 (H.arities ds))
   where go fs ds = do
@@ -230,22 +234,32 @@ trans ds fs = evalState (go fs ((H.appify) ds)) (S "Z" 0 (H.arities ds))
               recSubst  = H.substs  recVars
               recSubstC = H.substsC recVars
               recVars = zip (map recVar fs) fs
-              recVar x = H.Var $ H.makeRec x
+              recVar = H.Var . H.makeRec
           a <- arities <$> get
           regFormulae <- forM regDefs $ \d -> case d of
             H.DataType t                -> tTrans t
             H.Def d                     -> dTrans d
             H.ContSat (H.Satisfies x y) -> F.appifyF a <$> cTrans (H.Var x) y
-          speFormulae <- forM toCheck $ \d -> case d of
+          checkFormulae <- forM toCheck $ \d -> case d of
             H.DataType t                 -> error "No contracts for datatypes yet!"
             H.Def (H.Let f xs e)         -> dTrans $ H.Let     f xs (recSubst e)
             H.Def (H.LetCase f xs e pes) -> dTrans $ H.LetCase f xs (recSubst e)
                                                        [(p,(recSubst e)) | (p,e) <- pes]
+            -- For contract satisfaction 'f ::: c', generate
+            -- 1) an assumption that the recursive occurances 'f_rec' satisfy the contract
+            -- 2) the negation of 'f ::: c'
+            -- XXX, ???: should 'f' be allowed to occur in 'c' at all? The 'recSubstC' below
+            -- indicates this is expected.
             H.ContSat (H.Satisfies x y)  -> do
-              contP   <- F.appifyF a <$> cTrans (recVar x) (recSubstC y)
+              contRec <- F.appifyF a <$> cTrans (recVar x) (recSubstC y)
+              -- XXX, ???: why not 'recSubstC y' instead of plain 'y'?
[…]

### 2011-10-27 paper.tex
Wibble.

@@
   \[
-  e = e ::: CF \&\& \neg CF.
+  e = e ::: CF \&\& \lnot CF.
   \]
-  Now, $T$ will contain $CF(ep) \land \neg CF(ep)$, and were hosed.
-  This is flawed because there is no $\neg$ in the annotated
+  Now, $T$ will contain $CF(ep) \land \lnot CF(ep)$, and were hosed.
+  This is flawed because there is no $\lnot$ in the annotated
   contracts.
@@
   \begin{itemize}
-  \item $\leftarrow$: if we pack crash-free values in a data
+  \item $\gets$: if we pack crash-free values in a data
     constructor, the resulting value is crash-free.
-  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
+  \item $\to$: a value $t$ of type $T$\note[nc]{
     ``type'' is used loosely here: it means the constructor head is of type $T$.

### 2011-11-06 paper.tex
Paper: Removed the expression translation and fixed some typos.
The expression translation was the identity in the paper, and is recently
also the identity in the implementation.

@@
 \def\ruleform#1{{\setlength{\fboxrule}{1.2pt}\fbox{\normalsize $#1$}}}
-\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
 \newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
@@
 \newcommand{\ok}{\texttt{Ok}}
-\newcommand{\hprime}{\mathcal{H}'}
+\newcommand{\hprime}{\ensuremath{\mathcal{H}'}}
 \newcommand{\cfc}{\texttt{CF}}
@@
 \]?
-The answer is that this is an expansion of the $\etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.
+The answer is that this is an expansion of the $e = K_i(\vec{x_i}) \to a = e_i$ part of the $\dtrans{}$ rule for functions defined by cases.
 I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.
 In this special case we could remove the implication, but not in general.
-The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}
+The $a = e_i$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}
 
@@
 &&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
-&&      :=\forall x_1, \ldots, x_n. \min{\etrans{f(x_1, \ldots, x_n)}} \\
+&&      :=\forall x_1, \ldots, x_n. \min{f(x_1, \ldots, x_n)} \\
 &&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
@@
   \[  \begin{array}{rcl}
-    \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
     \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
@@
 
-\subsection{$\etrans{}$ -- Expressions}
-Our most basic translation is from expressions in $\hprime$ to terms
-in first-order logic. Given this translation we will be able to
-translate definitions, data types and contracts to first-order
-formulae. It is described in figure~\ref{etrans}.
-
-\begin{figure}
-$$
-\begin{array}{rcl}
-\multicolumn{3}{c}{\ruleform{\etrans{e} = t}} \\[2.5mm]
- \etrans{x} &=& x\\
-  \etrans{f(e_1,\dots,e_n)} &=& f(\etrans{e_1},\dots,\etrans{e_n})\\
-  \etrans{K(e_1,\dots,e_n)} &=& K(\etrans{e_1},\dots,\etrans{e_n})\\
-  \etrans{\bad} &=& \bad
-\end{array}
-$$
-\caption{$\etrans{}$ -- Expression translation}
-\label{etrans}
-\end{figure}
+\subsection{Expressions}
+Expressions are the same in \hprime and FOL, so no translation is
+necessary.
 
@@
   \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
-  f(\vec{x})) \to a = \etrans{e} \\
+  f(\vec{x})) \to a = e \\
    \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
@@
        \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
-     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& a = \etrans{e_1} \\
+     &\land& \forall \vec{x_1}.~e = K_1(\vec{x_1}) &\to& a = e_1 \\
      &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n} \\%removed } here ...
+     &\land& \forall \vec{x_n}.~e = K_n(\vec{x_n}) &\to& a = e_n \\
    \end{array} 
@@
     \[  \begin{array}{rcl}
-      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
+      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
 $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
-      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
+      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
       \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
@@
 \begin{eqnarray*}
-  \ctrans{e \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = true \lor \etrans{e} = \unr)\\
-  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{\etrans{e~x}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
-  \ctrans{e \in \cfc} &=& \min{\etrans{e}} \land \cf{\etrans{e}}
+  \ctrans{e \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = true \lor e = \unr)\\
+  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{e~x} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
+  \ctrans{e \in \cfc} &=& \min{e} \land \cf{e}
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-  \ctrans{e \not \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = false \lor \etrans{e} = \bad)\\
+  \ctrans{e \not \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = false \lor e = \bad)\\
   \ctrans{e \not \in x:c_1 \to c_2} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{e~x \not \in c_2}\\
-  \ctrans{e \not \in \cfc} &=& \min{\etrans{e}} \land \lnot \cf{\etrans{e}}
+  \ctrans{e \not \in \cfc} &=& \min{e} \land \lnot \cf{e}
 \end{eqnarray*}
@@
 \begin{figure*}
-$$\etrans{e_1~e_2} = app(e_1,e_2)$$
+$$e_1~e_2 = app(e_1,e_2)$$
 $$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$

### 2011-11-06 paper.tex
Paper: removed $\Phi$, the union of formulas, and replaced with conjunction.

@@
 Nil \neq \unr 
-\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\Phi_4$) say that constructors are ``lazy'' or ``total''.}
+\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phi_\text{lazy}$) say that constructors are ``lazy'' or ``total''.}
 
@@
   doesn't seem to play any role in formal part of this paper (besides
-  being specified in axioms $\Phi_1$ (projections)). In the current
+  being specified in axioms $\phi_\text{projection}$). In the current
   implementation it's used to avoid introducing existentially
@@
     \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
-    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi \\
-    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
+    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
   \end{array} \]
   \caption{First-order logic syntax}
-  \note[nc]{The $\Phi$ are unnecessary and don't make things simpler, because a set of formulas is equivalent to the conjunction of its contents.
-XXX, TODO: drop $\Phi$ and replace all $\bigcup$ with $\bigwedge$.}
   \label{fol-stx}
@@
   \[  \begin{array}{rcl}
-    \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
-    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
-    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
+    \dtrans{def} &\to& \phi~(\mbox{formulae}) \\
+    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formulae}) \\
+    \ctrans{f \in c} &\to& \phi~(\mbox{formulae}) \\
   \end{array} \]
@@
 \begin{array}{rcl}
-  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
+  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
   \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
@@
 We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
-$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
+$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 
 \begin{itemize}
-\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
+\item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
   $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
   on its $k$-th component, $k = 1\text{ to }a_i$.
-\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
+\item $(\phi_\text{disjoint})$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
   never map to the same value.
-\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
+\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
   Notice that we have a equivalence.
@@
 
-\item $(\Phi_4)$ No full application of a constructor $K_i$ is unreachable or crashing.
+\item $(\phi_\text{lazy})$ No full application of a constructor $K_i$ is unreachable or crashing.
 \item One may want to also state that if for all $j$, $x_j \neq \bad$ then
@@
   $$ \ruleform{\ktrans{data~def} = \Phi} $$
-$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
+$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
  \hspace{5.6cm}where
@@
     \[  \begin{array}{rcl}
-      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
+      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
 $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
-      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
-      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
-      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
+      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
+      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
+      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
     \end{array} \]
@@
 \begin{figure*}
-  $$ \ruleform{\ctrans{contract} = \Phi} $$
+  $$ \ruleform{\ctrans{contract} = \phi} $$
 \begin{eqnarray*}

### 2011-11-06 paper.tex
Paper: now uses 'True' and 'False', and removed discussion of 'true' is 'True'.

@@
 
+%% We could use \texttt{true} and \texttt{false} here, to be more
+%% consistent with \bad and \unr, but then we should also use
+%% \texttt{Nil}, \texttt{Cons}, etc.
+\newcommand{\true}{True}
+\newcommand{\false}{False}
 \newcommand{\unr}{\texttt{UNR}}
@@
 
-Now we translate the $null$ function. Note that the symbols $true$ and
-$false$ are the representation of the data constructors $True$ and
-$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
-logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
-  text.
-The lowercase $true$ and $false$ is an artifact of the implementation, i.e.
-Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.
-Moreover, the lower case here is inconsistent with the rest of the presentation, e.g.
-$Cons$ and $Nil$ above.}
+Now we translate the $null$ function.
 \begin{equation*}
-\forall xs.~ xs=Nil \implies notnull(xs) = false
+\forall xs.~ xs=Nil \implies notnull(xs) = \false
 \end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
 \[
-notnull(Nil) = false
+notnull(Nil) = \false
 \]?
@@
   \label{defnotnull}
-\forall xs,y,ys.~ xs=Cons(y,ys) \implies notnull(xs) = true
+\forall xs,y,ys.~ xs=Cons(y,ys) \implies notnull(xs) = \true
 \end{equation}
@@
 \begin{equation*}
-  \phi := \forall xs.~\cf{xs} \land notnull(xs) = true \to \cf{head(xs)} 
+  \phi := \forall xs.~\cf{xs} \land notnull(xs) = \true \to \cf{head(xs)} 
 \end{equation*}
@@
 Now, assume that $xs$ satisfies
-$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
+$\cf{xs}$ and $notnull(xs) = \true$.  We can derive that $xs \neq \bad$
 because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{
@@
 }. The constraint
-$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
+$notnull(xs) = \true$ doesn't directly imply that $xs = Cons(y,ys)$ for
 some $y$ and $ys$. But $notnull$ is totally defined, because of
@@
   \[  \begin{array}{rcl}
-    e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
+    e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,\false\}\\
     e \in x:c_1 \to c_2 &\iff& \forall e_1 \in c_1, (e~e_1) \in c_2[e_1/x]\\
@@
 here is not optional, because in addition to concluding 
-$e \neq \unr \to p[e/x] \in \{True,\unr\}$, we also conclude $\min{p[e/x]}$.
+$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
 The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
-$p[e/x] \in \{True,\unr\}$, and because it can be used to unfold definitions,
+$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
 we guard it with $\min{e}$.
@@
 We give in figure~\ref{ctrans} the translation of contract satisfaction.
-\annote[nc]{$true$ refers to the translation to a term of the data constructor
-\texttt{True} in $\hprime$, not to the actual true value.}{
-Again, we should use $True$ (or \texttt{True}) in the translation; the
-lowercase is an artifact of the concrete TPTP syntax.
-}
 \remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
@@
   \protect \[
-  (\lambda x. \text{case $x$ of $True \to True$; $False \to \bad$}) 
-  \in \{x|x==True\} \to \cfc
+  (\lambda x. \text{case $x$ of $\true \to \true$; $\false \to \bad$})
+  \in \{x|x==\true\} \to \cfc
   \]
@@
 \item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
-  it's impossble to prove $\ctrans{True \in \cfc}$, because you can't prove 
-  $\min{True}$ a priori.  On the other hand, we don't need any extra 
+  it's impossble to prove $\ctrans{\true \in \cfc}$, because you can't prove
+  $\min{\true}$ a priori.  On the other hand, we don't need any extra
   $\min{}$ assumption in the arrow case, since it's already assumed, at least in
@@
 \begin{eqnarray*}
-  \ctrans{e \not \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = false \lor e = \bad)\\
+  \ctrans{e \not \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = \false \lor e = \bad)\\
   \ctrans{e \not \in x:c_1 \to c_2} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{e~x \not \in c_2}\\
@@
     \item $\cf{\unr}$
-    \item $\dtrans{data~Bool = True~0 | False~0}$
+    \item $\dtrans{data~Bool = \true~0 | \false~0}$
     \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$

### 2011-11-06 paper.tex
Paper: updated expression langauge; limit 'full' application to axioms.
Added 'UNR' and function symbols to the expression language, and made
most examples use partial application.  Axioms still use full application.
The point is that full application is only relevant as an optimization
of generated formulas to make the theorem prover more easily recognize
when a formula applies to a particular term. E.g., we think it's easier
for the prover to match on 'f(x1,...,xn)' then 'f `app` x1 `app` ... `app` xn'.

@@
 \newcommand{\hprime}{\ensuremath{\mathcal{H}'}}
+% Full function application, e.g. $\full{f}(x_1,\ldots,x_n)$. Contrast
+% with partial application, e.g. $f~x_1 \ldots x_n$
+\newcommand{\full}[1]{\hat{#1}}
 \newcommand{\cfc}{\texttt{CF}}
 \newcommand{\cf}[1]{\texttt{CF}(#1)}
-\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
 \renewcommand{\min}[1]{\mbox{min}(#1)}
@@
 \begin{equation*}
-\forall a,b.~Cons(a,b) \neq Nil
+\forall a,b.~Cons~a~b \neq Nil
 \end{equation*}
@@
 Then, we state that $Nil$ can cannot cause a crash, and that
-$Cons(x,y)$ cannot cause a crash iff both its arguments cannot. The
+$Cons~x~y$ cannot cause a crash iff both its arguments cannot. The
 statement ``$x$ cannot cause a crash'' is encoded by the formula
 $\cf{x}$.\note[nc]{I just rewrote this, but it still sounds weird: in
-  $head(Nil) = \bad$, who ``caused'' the crash? The precise definition
+  $head~Nil = \bad$, who ``caused'' the crash? The precise definition
   of $\cf$ in Section \ref{sec:cf}, in terms of not leading to crashes
@@
   \label{cfcons}
-\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}
+\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons~a~b}
 \end{equation}
@@
 We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
-  CF(fst(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
+  CF(fst~(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
 can't think of a good way to explain it right now.
 \begin{equation*}
-\forall y,ys.~Cons(y,ys) \neq \unr 
+\forall y,ys.~Cons~y~ys \neq \unr
 \end{equation*}
@@
 \begin{equation*}
-\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
+\forall xs,y,ys.~sel_{1,Cons}~Cons~y~ys = xs \implies xs = y 
 \end{equation*}
 \begin{equation*}
-\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
+\forall xs,y,ys.~sel_{2,Cons}~Cons~y~ys = xs \implies xs = ys 
 \end{equation*}\note[nc]{???
-Why not the simpler \[\forall y,ys.~sel_{2,Cons}(Cons(y,ys)) = ys\]?
+Why not the simpler \[\forall y,ys.~sel_{2,Cons}~Cons~y~ys = ys\]?
 Same question in Figure \ref{dtrans}.
@@
 \begin{equation*}
-\forall xs.~ xs=Nil \implies notnull(xs) = \false
+\forall xs.~ xs=Nil \implies notnull~xs = \false
 \end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
 \[
-notnull(Nil) = \false
+notnull~Nil = \false
 \]?
-The answer is that this is an expansion of the $e = K_i(\vec{x_i}) \to a = e_i$ part of the $\dtrans{}$ rule for functions defined by cases.
+The answer is that this is an expansion of the $e = K_i~\vec{x_i} \to a = e_i$ part of the $\dtrans{}$ rule for functions defined by cases.
 I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.
@@
   \label{defnotnull}
-\forall xs,y,ys.~ xs=Cons(y,ys) \implies notnull(xs) = \true
+\forall xs,y,ys.~ xs=Cons~y~ys \implies notnull~xs = \true
 \end{equation}
@@
 We also need to specify the translation of calls to $notnull$ with
-$\bad$ values, to encode the fact that $notnull(\bad) = \bad$.
+$\bad$ values, to encode the fact that $notnull~\bad = \bad$.
 \begin{equation*}
-\forall xs.~xs=\bad \to notnull(xs) = \bad
+\forall xs.~xs=\bad \to notnull~xs = \bad
 \end{equation*}
@@
   \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
-  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
+  \land xs \neq Cons~(sel_{1,Cons}~xs)~(sel_{2,Cons}~xs) \to  \notag \\
   notnull(xs) = \unr 
@@
 \begin{align}
-  \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
-  \forall xs,y,ys.~ xs=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
-    \forall xs.~xs=\bad \to head(xs) = \bad \notag
+  \forall xs.~ x=Nil \implies head~xs = \bad \notag\\
+  \forall xs,y,ys.~ xs=Cons~y~ys \implies head~xs = y     \label{defhead} \\
+    \forall xs.~xs=\bad \to head~xs = \bad \notag
 \end{align}\note[nc]{??? Why not use the same project+construct trick from the last equation?  I.e., why not
 \[
-\forall xs.~ xs=Cons(sel_{Cons,1}(xs),sel_{Cons,2}(xs)) \implies head(xs) = sel_{Cons,1}(xs)
+\forall xs.~ xs=Cons~(sel_{Cons,1}~xs)~(sel_{Cons,2}~xs) \implies head~xs = sel_{Cons,1}~xs
 \]?  Because it's ugly?  Would Equinox care?}
@@
  \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
- \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
- head(xs) = \unr 
+ \land xs \neq Cons~(sel_{1,Cons}~xs)~(sel_{2,Cons}~xs) \to  \notag \\
+ head~xs = \unr
 \end{align*}
@@
 \begin{equation*}
-  \phi := \forall xs.~\cf{xs} \land notnull(xs) = \true \to \cf{head(xs)} 
+  \phi := \forall xs.~\cf{xs} \land notnull~xs = \true \to \cf{head~xs} 
 \end{equation*}
@@
   corresponding Haskell expression is well-typed in an environment
-  including $P$'s decls). For example, \texttt{Cons(Z,Nil)} is
-  ``well-typed'' but \texttt{Cons(Nil,Z)} is ``ill-typed'' in a the
+  including $P$'s decls). For example, \texttt{Cons~Z~Nil} is
+  ``well-typed'' but \texttt{Cons~Nil~Z} is ``ill-typed'' in a the
   context of a program declaring List and Nat types.
@@
   a terminating expression with the wrong constructor head
-  (e.g. \texttt{tail(S Z) =} \unr{}). Scrutinizing an ill-typed
+  (e.g. \texttt{tail~(S Z) =} \unr{}). Scrutinizing an ill-typed
   expression with a constructor head of the right type evaluates as
-  specified by the pattern (e.g. \texttt{tail(Cons(Nil,Z)) = Z}),
+  specified by the pattern (e.g. \texttt{tail~(Cons~Nil~Z) = Z}),
   unless there is no corresponding pattern (``non-exhaustive
@@
 Now, assume that $xs$ satisfies
-$\cf{xs}$ and $notnull(xs) = \true$.  We can derive that $xs \neq \bad$
+$\cf{xs}$ and $notnull~xs = \true$.  We can derive that $xs \neq \bad$
 because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{
@@
 }. The constraint
-$notnull(xs) = \true$ doesn't directly imply that $xs = Cons(y,ys)$ for
+$notnull~xs = \true$ doesn't directly imply that $xs = Cons~y~ys$ for
 some $y$ and $ys$. But $notnull$ is totally defined, because of
 \eqref{unrnotnull}. This implies (by \eqref{defnotnull}) that there exist
-$y$ and $ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can
-now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head(xs) =
+$y$ and $ys$ such that $xs = Cons~y~ys$. Recalling $\cf{xs}$, we can
+now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head~xs =
 y$ because of \eqref{defhead}, and $y$ is crash-free, so we can finally
-derive $\cf{head(xs)}$. QED.
+derive $\cf{head~xs}$. QED.
 
@@
 \item There's no $\lambda$-abstraction, because we can always lift
-  them to toplevel declaration.
+  them to a toplevel declaration.
 \item We do not allow nested case expressions, because once again, we
@@
 \item Until section~\ref{ho} we will only consider full application of
-  functions ($f(x,y)$),\note[nc]{
-
-Not really true: the definition of satisfaction for arrow contracts
+  functions ($\full{f}(x,y)$),\note[nc]{
+
+Not true: the definition of satisfaction for arrow contracts
 $x:c_1 \to c_2$ already includes partial applications.  We could add a
@@
 &&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
-&&      :=\forall x_1, \ldots, x_n. \min{f(x_1, \ldots, x_n)} \\
+&&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
 &&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
-              \lor \ctrans{f(x_1, \ldots, x_n) \in c}
+              \lor \ctrans{\full{f}(x_1, \ldots, x_n) \in c}
 \end{eqnarray*}
 for functions of arity $n$, to make the claim true, but that would add
-more clutter.  This may be related to my confusion where removing the
-$f_{ptr}$ stuff caused problems where no higher-orderness was
-involved?
-
+more clutter.
 }, in order to remove clutter. Dealing with
@@
         \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
-         && K_1(\vec{x}_1) \to e_1 \mid \cdots \mid K_n(\vec{x}_n) \to e_n \\
+         && K_1~\vec{x}_1 \to e_1 \mid \cdots \mid K_n~\vec{x}_n \to e_n \\
        \end{array}
@@
   \[  \begin{array}{rclr}
-    x,y,f,g,a,b & \in & \mbox{Variables} \\
+    x,y,g,a,b & \in & \mbox{Variables} \\
     T &\in& \mbox{Type Constructors} \\
     K &\in& \mbox{Data Constructors} \\
+    f &\in& \text{Function Symbols} \\
   \end{array} \]
@@
     e &\in& \mbox{Expressions} & \\
-    e &::=& x & \\
-    &\mid& \bad & \\
+    e &::=& x \mid f \mid K \mid \bad \mid \unr & \\
+%    &
     &\mid& e~e & \\
-    &\mid& f(e,\dots,e) & \\
-    &\mid& K(e,\dots,e) & \\
+    &\mid& \full{f}(e,\dots,e) & \\
+    &\mid& \full{K}(e,\dots,e) & \\
   \end{array} \]
@@
 issue with generalisation to arbitrary tuples.\note[nc]{
-
+%
+  XXX: Tuple contracts are not actually supported in the
+  implementation. Should they be discussed here?
+%
 So, ``tuples'' refers to the the term constructor, not the type
@@
 \[
-e \in Cons(c_1,c_2) \iff e \text{ diverges or } 
-      (e \to^\star Cons(e_1,e_2) \mbox{ and } e_1 \in c_1, e_2 \in c_2)
+e \in Cons~c_1~c_2 \iff e \text{ diverges or } 
+      (e \to^\star Cons~e_1~e_2 \mbox{ and } e_1 \in c_1, e_2 \in c_2)
 \]
@@
 The notation $\bad \notin_s e$ means that $\bad$ does not appear
-anywhere in $e$, similarly for $\bad \not \in_s \mathcal{C}$. For
+anywhere in $e$\note[nc]{,including the definitions of the function symbols used in $e$}, similarly for $\bad \not \in_s \mathcal{C}$. For
 example, $Just ~3$ is syntactically safe whereas $Just ~\bad$ is not.
@@
 On the other hand, some ill-typed expressions are not $\unr$, e.g.
-we can prove that $tail(Cons(Nil,Z)) = Z$.}
+we can prove that $tail~(Cons~Nil~Z) = Z$.}
 
@@
 forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
-\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.}
+\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} The expression language in our logic is \hprime.
 
@@
   \[  \begin{array}{rcl}
-    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
-    && \mid \bad \mid \unr \\
-    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
+    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t) \mid \min{t} \\
     && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
@@
   \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
-  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
+  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ktrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
   \dtrans{f \in c} &=& \ctrans{f \in c}\\
-  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
-  f(\vec{x})) \to a = e \\
-   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
-     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
+  \dtrans{f~\vec{x} = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
+  \full{f}(\vec{x})) \to a = e \\
+   \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
+     \forall \vec{x},a.~(\mbox{min}(a) \land a = \full{f}(\vec{x})) &\to&
      \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
-       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
-     &\land& \forall \vec{x_1}.~e = K_1(\vec{x_1}) &\to& a = e_1 \\
+       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
+     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& a = e_1 \\
      &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~e = K_n(\vec{x_n}) &\to& a = e_n \\
+     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& a = e_n \\
    \end{array} 
@@
 We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
-$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+$$\ktrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 
@@
 \item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
+  $sel_{k,K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
   on its $k$-th component, $k = 1\text{ to }a_i$.
-\item $(\phi_\text{disjoint})$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
-  never map to the same value.
+\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
 \item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
@@
 \item One may want to also state that if for all $j$, $x_j \neq \bad$ then
-  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
-  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
+  $\full{K}_i(\vec{x}) \neq \bad$. It is already implied by the fact that
+  $\left(\forall j. \cf{x_j}\right) \to \cf{\full{K}_i(\vec{x})}$.
 \end{itemize}
@@
   $$ \ruleform{\ktrans{data~def} = \Phi} $$
-$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+$$\ktrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
  \hspace{5.6cm}where
@@
     \[  \begin{array}{rcl}
-      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
+      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
 $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
-      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
-      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
-      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
+      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
+      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
+      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
     \end{array} \]
@@
 in Figure~\ref{dtrans}.  Definition translations are axioms, 
-so the the initial $\min{f(\vec{x})}$
+so the the initial $\min{\full{f}(\vec{x})}$
 is an obligation: we aren't allowed to 
@@
 
+\newcommand{\rec}[1]{#1_\text{rec}}
 \subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}
@@
 
-$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \ctrans{f_p \in c} \vdash \ctrans{f \in c}$$
+$$\trans{M - f}, \dtrans{f=e[f/\rec{f}]}, \ctrans{\rec{f} \in c} \vdash \ctrans{f \in c}$$
 
@@
 \begin{figure*}
-$$e_1~e_2 = app(e_1,e_2)$$
-$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
-$$ \cf{f\_ptr} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{f(x_1,\dots,x_n)}$$
-$$\forall f\_ptr,x.~\cf{f\_ptr} \land \cf{x} \to \cf{app(f\_ptr,x)}$$
+$$ \forall x_1,\dots,x_n.~\full{f}(x_1,\dots,x_n) = f~x_1~x_2~\dots~x_n$$
+$$ \cf{f} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{\full{f}(x_1,\dots,x_n)}$$
+$$\forall g,x.~\cf{f} \land \cf{x} \to \cf{g~x}$$
 \caption{Encoding of higher-orderness}

### 2011-11-06 contracts/ThmProver.hs
Trying Vampire and updated vampire support accordingly.

@@ -27,11 +27,11 @@ fof = Theory {
       , fileExtension = "tptp"
       , footer = ""
       }
-smtlib :: Theory
-smtlib = Theory {
+smt2 :: Theory
+smt2 = Theory {
            showFormula = F.toSMTLIB
          , header = F.showDefsSMTLIB
-         , fileExtension = "smtlib"
+         , fileExtension = "smt2"
          , footer = "(check-sat)"
          }
                     
@@ -46,9 +46,14 @@ provers = [ ("equinox", T "equinox"
                         ("Proof found" `isInfixOf`)
                         fof
             )
-          , ("vampire", T "vampire_lin64"
+          , ("vampire32", T "vampire_lin32"
                           ["--mode", "casc" ,"--input_file"]
-                          ("Refutation" `isInfixOf`)
+                          ("SZS status Unsatisfiable" `isInfixOf`)
+                          fof
+            )
+          , ("vampire64", T "vampire_lin64"
+                          ["--mode", "casc" ,"--input_file"]
+                          ("SZS status Unsatisfiable" `isInfixOf`)
                           fof
             )
           , ("E", T "eprover"
@@ -59,6 +64,6 @@ provers = [ ("equinox", T "equinox"
           , ("z3", T "z3"
                      ["-nw","-smt2"]
                      ("unsat" `isInfixOf`)
-                     smtlib
+                     smt2
             )
           ]

### 2011-11-08 contracts/Haskell.hs
Added support for simple 'import's in example files.
Imports are resolved relative the file containing them.  Imports are
paths, enclosed in double quotes.

@@ -54,6 +54,7 @@ data MetaExpression v = Named v
 data MetaDefGeneral a = ContSat (MetaContSat a)
                       | Def (MetaDefinition a)
                       | DataType (MetaDataType a)
+                      | Import FilePath
                       deriving (Eq,Show,Functor,Ord)
 
 -- No contracts for constructors! So, Name, not Named here.

### 2011-11-08 contracts/Parser.hs

@@ -5,7 +5,7 @@ import Haskell
 
 -- parser produced by Happy Version 1.18.6
 
-data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12 t13 t14 t15 t16 t17 t18 t19 t20 t21 t22 t23 t24
+data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12 t13 t14 t15 t16 t17 t18 t19 t20 t21 t22 t23 t24 t25 t26
 	= HappyTerminal (Token)
 	| HappyErrorToken Int
 	| HappyAbsSyn4 t4
@@ -29,314 +29,337 @@ data HappyAbsSyn t4 t5 t6 t7 t8 t9 t10 t11 t12 t13 t14 t15 t16 t17 t18 t19 t20 t
 	| HappyAbsSyn22 t22
 	| HappyAbsSyn23 t23
 	| HappyAbsSyn24 t24
+	| HappyAbsSyn25 t25
+	| HappyAbsSyn26 t26
 
-action_0 (25) = happyShift action_5
-action_0 (33) = happyShift action_6
-action_0 (4) = happyGoto action_7
+action_0 (27) = happyShift action_5
+action_0 (28) = happyShift action_6
+action_0 (37) = happyShift action_7
+action_0 (4) = happyGoto action_8
 action_0 (7) = happyGoto action_2
 action_0 (17) = happyGoto action_3
 action_0 (19) = happyGoto action_4
-action_0 _ = happyReduce_30
+action_0 _ = happyReduce_31
 
-action_1 (25) = happyShift action_5
-action_1 (33) = happyShift action_6
+action_1 (27) = happyShift action_5
+action_1 (28) = happyShift action_6
+action_1 (37) = happyShift action_7
 action_1 (7) = happyGoto action_2
 action_1 (17) = happyGoto action_3
 action_1 (19) = happyGoto action_4
 action_1 _ = happyFail
 
-action_2 (36) = happyShift action_14
+action_2 (40) = happyShift action_17
+action_2 (21) = happyGoto action_16
 action_2 _ = happyFail
 
-action_3 (25) = happyShift action_5
-action_3 (33) = happyShift action_6
+action_3 (27) = happyShift action_5
+action_3 (28) = happyShift action_6
+action_3 (37) = happyShift action_7
 action_3 (7) = happyGoto action_2
 action_3 (17) = happyGoto action_3
[…]

### 2011-11-08 contracts/ThmProver.hs

@@ -46,6 +46,11 @@ provers = [ ("equinox", T "equinox"
                         ("Proof found" `isInfixOf`)
                         fof
             )
+          -- I can't locate any vampire usage docs, and '-h' and
+          -- '--help' don't work :P From the CASC competition page I
+          -- found that '-t <time>' can be used to time limit vampire.
+          -- Experience shows that, by default, vampire has a 60
+          -- second time limit.
           , ("vampire32", T "vampire_lin32"
                           ["--mode", "casc" ,"--input_file"]
                           ("SZS status Unsatisfiable" `isInfixOf`)

### 2011-11-08 contracts/Parser.hs
Added support for blank-lines-as-separators in example files.
Now you can omit the ';;' separators when there is a blank line.  This is
not smart about layout, e.g. a blank line in a function will break it.
Since we don't have 'let', nested 'case', 'where', or lambda, I doubt this
will be an issue.

@@ -820,7 +820,7 @@ lexPath cs = TokenPath p : lexer rest
   where (p,'"':rest) = span (/='"') cs
 
 -- Lex blank lines as separators.
-tryLexBlankLine cs = 
+tryLexBlankLine cs =
   if all isSpace line
   then TokenSep : lexer cs'
   else lexer cs

### 2011-11-08 paper.tex

@@
 \newcommand{\hprime}{\ensuremath{\mathcal{H}'}}
+
 % Full function application, e.g. $\full{f}(x_1,\ldots,x_n)$. Contrast
@@
 \newcommand{\full}[1]{\hat{#1}}
+% Recursive anonymous versions of functions.
+\newcommand{\rec}[1]{#1_\text{rec}}
+
 \newcommand{\cfc}{\texttt{CF}}
@@
   \[  \begin{array}{rcl}
-    \dtrans{def} &\to& \phi~(\mbox{formulae}) \\
-    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formulae}) \\
-    \ctrans{f \in c} &\to& \phi~(\mbox{formulae}) \\
+    \dtrans{def} &\to& \phi~(\mbox{formula}) \\
+    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
+    \ctrans{f \in c} &\to& \phi~(\mbox{formula}) \\
   \end{array} \]
@@
   \begin{itemize}
-  \item $\gets$: if we pack crash-free values in a data
-    constructor, the resulting value is crash-free.
+  \item $\gets$: if we pack crash-free values in a data constructor,
+    the resulting value is crash-free. \note[nc]{We use the contract
+    translation for arrows, and this translation is principled.}
   \item $\to$: a value $t$ of type $T$\note[nc]{
@@
     projection is a crash-free context that throws an expression that
-    is not crash-free.
+    is not crash-free. \note[nc]{XXX: This translation is ad-hoc.}
 
@@
   being crash-free or not doesn't matter.
+
+  \note[nc]{Things become more interesting when we add $\min$'s.  For
+    the $\gets$ direction, we are asserting that $K_i \in \cfc^{a_i} \to
+    \cfc$.  Since we are asserting, it's a minus translation, and so
+    the arguments are translated as goals, i.e. we get $\ctrans{x \in
+      \cfc}^+$, which is $\min{x} \to \cf{x}$.
+
+  Those $\min$ assumptions on $x$ are very important, as the following
+  example illustrates.  Consider proving that addition, defined by
+  recursion on the first argument, satisfies $\cfc^2 \to \cfc$. When
+  considering $add~(Succ~x)~y$ we need to show
+  $\cf{Succ~(\rec{add}~x~y)}$ from $\min{Succ~(\rec{add}~x~y)}$ and
+  $\ctrans{\rec{add} \in \cfc^2 \to \cfc}$.  The arrow contract for
+  $\rec{add}$ has a $\min{\rec{add}~x~y}$ constraint, but we only know
+  $\min$ for the $Succ$ of that.  Hence, the $\min{x}$ assumption,
+  from $\ctrans{x \in \cfc}^+$ in $\ctrans{Succ \in \cfc \to
+  \cfc}^-$, saves the day.}
   \end{itemize}
@@
       \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
-      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
+      \phi_\text{cf} &=& 
+        \bigwedge_{1 \leq i \leq n}~\ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
+          \land \forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) 
+                  \to \cf{a} \to (\bigwedge_{1 \leq j \leq a_i} \cf{x_j}) \\
       \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
@@
 }
+
+\begin{figure*}
+\begin{verbatim}
+[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
+[[e in cf]]         := cf(e)
+[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
+\end{verbatim}
+\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
+\label{fig:naiveCTrans}
+\end{figure*}
+
 \begin{figure*}
@@
 \label{fig:newCTrans}
-\end{figure*}
-
-\begin{figure*}
-\begin{verbatim}
-[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
-[[e in cf]]         := cf(e)
-[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
-\end{verbatim}
-\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
-\label{fig:naiveCTrans}
 \end{figure*}
@@
 
-\newcommand{\rec}[1]{#1_\text{rec}}
 \subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}

### 2011-11-08 contracts/Parser.hs
Revert "Added support for blank-lines-as-separators in example files."
This reverts commit 5cefdeaea3204f7918de8221ff4c8a540177999f.

I realize now that I want to write imports and contracts without
intervening blank lines.  Omitting ';;' causes trouble in these cases,
which is confusing, so I'm reverting the blank-lines-as-separators
support.

@@ -820,7 +820,7 @@ lexPath cs = TokenPath p : lexer rest
   where (p,'"':rest) = span (/='"') cs
 
 -- Lex blank lines as separators.
-tryLexBlankLine cs =
+tryLexBlankLine cs = 
   if all isSpace line
   then TokenSep : lexer cs'
   else lexer cs

### 2011-11-09 contracts/Translation.hs
Added support for labeled formulas.  Labels appear in generated theory.
Now the axiom names give some indication of where they came from.
Still have long ways to go in terms of useful names, but now the
machinery is there.  Joint implementation with Dimitrios: blame him
for any bugs :)

@@ -84,7 +84,7 @@ dPtr f vs = if null vs then Top else F.And [eq1Min]--,eq2Min]
 -- | Same as dPtr, but for recursive occurances of f.
 dPtrRec f@(Var v) = dPtr $ Rec v
 
-dTrans :: H.Definition -> Fresh [F.Formula]
+dTrans :: H.Definition -> Fresh [F.LabeledFormula]
 dTrans (H.Let f vs e) = do
   eT <- eTrans e
   let vsN = map (Named . Var) vs
@@ -94,8 +94,8 @@ dTrans (H.Let f vs e) = do
       eq    = F.Forall vs $ full :=: eT
       eqMin = F.Forall vs $ F.Min full :=>: (full :=: eT)
 
-  return [eqMin
-         ,dPtr fV vs, dPtrRec fV vs]
+  return $ map (F.LabeledFormula f)
+             [eqMin ,dPtr fV vs, dPtrRec fV vs]
 
 
 -- Recall that the patterns 'pes' has the form [([Variable],Expression)].
@@ -151,7 +151,8 @@ dTrans (H.LetCase f vs e pes) = do
       eqs = (eq1++[eq2,eq3])
       eqs'    = map (F.Forall (vs++zs)) eqs
       eqs'Min = map (\eq -> F.Forall (vs++zs) $ F.Min full :=>: F.And [F.Min eT,eq]) eqs
-  return $ dPtr fV vs : dPtrRec fV vs : eqs'Min
+
+  return $ map (F.LabeledFormula f) $ dPtr fV vs : dPtrRec fV vs : eqs'Min
 
 -- Contract translation
 -----------------------
@@ -161,60 +162,62 @@ dual :: Variance -> Variance
 dual Plus = Minus
 dual Minus = Plus
 
-cTrans :: Variance -> H.Expression -> H.Contract -> Fresh [F.Formula]
-cTrans _ e H.Any = return [Top]
+cTrans :: Variance -> H.Expression -> H.Contract -> Fresh F.LabeledFormula
+cTrans v e c = F.LabeledFormula "hack2" <$> cTrans' v e c
 
-cTrans v e (H.Pred x p) =  do
+cTrans' _ e H.Any = return Top
+
+cTrans' v e (H.Pred x p) =  do
   let  p' = H.subst e x p
   eT <- eTrans e
   p'T <- eTrans p'
[…]

### 2011-11-09 contracts/ThmProver.hs
Recommit: Added support for labeled formulas.  Labels appear in generated theory.
Now the axiom names give some indication of where they came from.
Still have long ways to go in terms of useful names, but now the
machinery is there.  Joint implementation with Dimitrios: blame him
for any bugs :)

I removed the extra axioms for constructors, which i will add back
next.

@@ -12,7 +12,7 @@ data ThmProverOpt = T { path :: FilePath
                       }
 
 data Theory = Theory { 
-      showFormula :: F.Formula -> String
+      showFormula :: F.LabeledFormula -> String
     , header :: [H.DefGeneral] -> String
     , fileExtension :: String
     , footer :: String
@@ -29,12 +29,14 @@ fof = Theory {
       }
 smt2 :: Theory
 smt2 = Theory {
-           showFormula = F.toSMTLIB
+           showFormula = F.toSMTLIB . unlabel
          , header = F.showDefsSMTLIB
          , fileExtension = "smt2"
          , footer = "(check-sat)"
          }
-                    
+
+unlabel (LabeledFormula _ e) = e
+
 provers :: [ThmProver]
 provers = [ ("equinox", T "equinox"
                           []

### 2011-11-09 contracts/Translation.hs

@@ -84,7 +84,7 @@ dPtr f vs = if null vs then Top else F.And [eq1Min]--,eq2Min]
 -- | Same as dPtr, but for recursive occurances of f.
 dPtrRec f@(Var v) = dPtr $ Rec v
 
-dTrans :: H.Definition -> Fresh [F.Formula]
+dTrans :: H.Definition -> Fresh [F.LabeledFormula]
 dTrans (H.Let f vs e) = do
   eT <- eTrans e
   let vsN = map (Named . Var) vs
@@ -94,8 +94,8 @@ dTrans (H.Let f vs e) = do
       eq    = F.Forall vs $ full :=: eT
       eqMin = F.Forall vs $ F.Min full :=>: (full :=: eT)
 
-  return [eqMin
-         ,dPtr fV vs, dPtrRec fV vs]
+  return $ map (F.LabeledFormula f)
+             [eqMin ,dPtr fV vs, dPtrRec fV vs]
 
 
 -- Recall that the patterns 'pes' has the form [([Variable],Expression)].
@@ -151,7 +151,8 @@ dTrans (H.LetCase f vs e pes) = do
       eqs = (eq1++[eq2,eq3])
       eqs'    = map (F.Forall (vs++zs)) eqs
       eqs'Min = map (\eq -> F.Forall (vs++zs) $ F.Min full :=>: F.And [F.Min eT,eq]) eqs
-  return $ dPtr fV vs : dPtrRec fV vs : eqs'Min
+
+  return $ map (F.LabeledFormula f) $ dPtr fV vs : dPtrRec fV vs : eqs'Min
 
 -- Contract translation
 -----------------------
@@ -161,60 +162,62 @@ dual :: Variance -> Variance
 dual Plus = Minus
 dual Minus = Plus
 
-cTrans :: Variance -> H.Expression -> H.Contract -> Fresh [F.Formula]
-cTrans _ e H.Any = return [Top]
+cTrans :: Variance -> H.Expression -> H.Contract -> Fresh F.LabeledFormula
+cTrans v e c = F.LabeledFormula "hack2" <$> cTrans' v e c
 
-cTrans v e (H.Pred x p) =  do
+cTrans' _ e H.Any = return Top
+
+cTrans' v e (H.Pred x p) =  do
   let  p' = H.subst e x p
   eT <- eTrans e
   p'T <- eTrans p'
[…]

### 2011-11-09 paper.tex
Paper: clarified the incompleteness of the recursion translation.
The reduction translation is incomplete for proofs that require more than one
unfolding of the recursive definition.  I added an example of this, and
an example showing that the obvious "fix" is unsound.

@@
 Where $M - f $ means the content of the module $M$ without $f$'s
-definition and $f$'s contract. XXX, TODO Stress that it's not always enough
-and that we may have to unroll several times!\note[nc]{??? Why? Need an example here.}
-
+definition and $f$'s contract. However, this translation is not
+complete: some proofs need to unroll the definition more than
+once!\note[nc]{For a simple example, consider a contract that is
+  proved by a finite number of reductions, but not by induction.  E.g.
+}
+  \begin{verbatim}
+  isZero y = case y of Z -> True | S _ -> False
+  eqNat x y = case x of
+    Z -> isZero y
+    S x' -> eqNatAux x' y
+  eqNatAux x y = case y of
+    Z -> False
+    S y' -> eqNat x y'
+
+  f x = case x of Z -> True | S x' -> f x'
+  f ::: (CF&&{x: eqNat x (S Z)}) -> {y:y} 
+  \end{verbatim}
+\note[nc]{
+  Now, the correct way to prove $f$'s contract is two unfoldings and
+  reduction: from $eqNat ~x ~(S ~Z)$ we get $x = S ~Z$, and $f~(S~Z) =
+  f~Z = \true$.  However, since we translate $f$ to
+}
+  \begin{verbatim}
+  f x = case x of Z -> True | S x' -> f_rec x'
+  \end{verbatim}
+\note[nc]{
+  we get $f~(S~Z) = \rec{f}~Z$ and we're stuck.
+
+  That example is very contrived, in that the much strong contract $f
+  ::: \cfc \to \{y:y\}$ is provable by induction in our translation,
+  but the same thing can happen in more realistic examples.  For
+  example, if you try to prove $add ::: x:(\cfc \land \{x:gt ~x ~Zero\})
+  \to y:\cfc \to \{z:gt ~z ~y\}$, then you eventually get to $gt
+  ~(S~(\rec{add} ~Z ~y)) ~y$ and then $gt ~(\rec{add} ~Z ~y') ~y'$, for
+  $y = S~y'$. But now you're stuck, because there is no reduction rule
+  for $\rec{add}$ allowing you to conclude $\rec{add} ~Z ~y' = y'$.
+
+  OK, so we might hope to simply fully define $\rec{add}$, so that
+  computations with $\rec{add}$ can be reduced.  That approach is
+  unsound.  It's obvious, because as part of the translation of $f :::
+  c$ we get $\rec{f} ::: c$ as an axiom, and so if $f ::: c$ doesn't
+  actually hold, then $\rec{f} ::: c$ is contradictory.  For
+  concreteness, here's simple example. Consider $f$ as above, but
+  change the contract to $c := \cfc \to \{y: not ~y\}$.  Then $f$
+  doesn't satisfy $c$, and a context that assumes $\rec{f}$ satisfies
+  $c$ is contradictory when $\rec{f}$ is fully defined, by
+  e.g. reducing $\rec{f}~Z = \true$.
+ }
 \subsection{Module checking}

### 2011-11-09 contracts/FOL.hs
Revert "Added support for labeled formulas.  Labels appear in generated theory."
This reverts commit 4079c6481b3a45c9a417f443b4a49db136d8f965.

@@ -10,9 +10,6 @@ import Data.List (intercalate)
 
 type Term = Expression
 type Formula = MetaFormula Term
-type LabeledFormula = MetaLabeledFormula Formula
-data MetaLabeledFormula a = LabeledFormula Label a deriving (Show, Eq, Functor)
-type Label = String
 
 infix 7 :<=>:
 infix 7 :=>:
@@ -37,16 +34,10 @@ data MetaFormula a = Forall [Name] (MetaFormula a)
 -- min = Pred "min"
 
 -- forall a . x && y --> (forall a . x) && (forall a . y)
-
-splitOnAndLabeled :: LabeledFormula -> [LabeledFormula]
-splitOnAndLabeled (LabeledFormula lbl f)
-  = [LabeledFormula (lbl++show i) f' | (i,f') <- zip [1..] (splitOnAnd f)]
-
 splitOnAnd :: Formula -> [Formula]
 splitOnAnd (Forall xs (And fs)) = map (Forall xs) fs
 splitOnAnd (Forall xs f) = map (Forall xs) $ splitOnAnd f
 splitOnAnd (And fs) = concatMap splitOnAnd fs
-splitOnAnd Top = []
 splitOnAnd f = [f]
 
 trivializeMin :: Formula -> Formula
@@ -90,15 +81,11 @@ removeConstants (And fs) = if any (==Bottom) fs' then Bottom else f
           f = if null fs'' then Top else And fs''
 removeConstants f = f
 
+simplify = filter (/= Top) . splitOnAnd . removeConstants -- . trivializeMin
 
 
-
-simplify :: LabeledFormula -> [LabeledFormula]
-simplify = splitOnAndLabeled . fmap removeConstants -- . trivializeMin
-
-
-toTPTP :: LabeledFormula -> String
-toTPTP (LabeledFormula l f) = header ++ "\n" ++ go [] f ++ "\n" ++ footer
+toTPTP :: Formula -> String
+toTPTP f = header ++ "\n" ++ go [] f ++ "\n" ++ footer
   -- XXX, MAYBE TODO: add better header or comments. Right now the first "axiom"
   -- below is the name of the axiom.  The TPTP format also allows optional 4th and
   -- 5th fields for comments.  Would be nice to see each formula labeled with its
@@ -109,7 +96,7 @@ toTPTP (LabeledFormula l f) = header ++ "\n" ++ go [] f ++ "\n" ++ footer
   -- light on documentation, but the source might be worth a look.  It
[…]

### 2011-11-09 contracts/ThmProver.hs

@@ -12,7 +12,7 @@ data ThmProverOpt = T { path :: FilePath
                       }
 
 data Theory = Theory { 
-      showFormula :: F.LabeledFormula -> String
+      showFormula :: F.Formula -> String
     , header :: [H.DefGeneral] -> String
     , fileExtension :: String
     , footer :: String
@@ -29,14 +29,12 @@ fof = Theory {
       }
 smt2 :: Theory
 smt2 = Theory {
-           showFormula = F.toSMTLIB . unlabel
+           showFormula = F.toSMTLIB
          , header = F.showDefsSMTLIB
          , fileExtension = "smt2"
          , footer = "(check-sat)"
          }
-
-unlabel (LabeledFormula _ e) = e
-
+                    
 provers :: [ThmProver]
 provers = [ ("equinox", T "equinox"
                           []

### 2011-11-09 contracts/Translation.hs

@@ -84,7 +84,7 @@ dPtr f vs = if null vs then Top else F.And [eq1Min]--,eq2Min]
 -- | Same as dPtr, but for recursive occurances of f.
 dPtrRec f@(Var v) = dPtr $ Rec v
 
-dTrans :: H.Definition -> Fresh [F.LabeledFormula]
+dTrans :: H.Definition -> Fresh [F.Formula]
 dTrans (H.Let f vs e) = do
   eT <- eTrans e
   let vsN = map (Named . Var) vs
@@ -94,8 +94,8 @@ dTrans (H.Let f vs e) = do
       eq    = F.Forall vs $ full :=: eT
       eqMin = F.Forall vs $ F.Min full :=>: (full :=: eT)
 
-  return $ map (F.LabeledFormula f)
-             [eqMin ,dPtr fV vs, dPtrRec fV vs]
+  return [eqMin
+         ,dPtr fV vs, dPtrRec fV vs]
 
 
 -- Recall that the patterns 'pes' has the form [([Variable],Expression)].
@@ -151,8 +151,7 @@ dTrans (H.LetCase f vs e pes) = do
       eqs = (eq1++[eq2,eq3])
       eqs'    = map (F.Forall (vs++zs)) eqs
       eqs'Min = map (\eq -> F.Forall (vs++zs) $ F.Min full :=>: F.And [F.Min eT,eq]) eqs
-
-  return $ map (F.LabeledFormula f) $ dPtr fV vs : dPtrRec fV vs : eqs'Min
+  return $ dPtr fV vs : dPtrRec fV vs : eqs'Min
 
 -- Contract translation
 -----------------------
@@ -162,62 +161,60 @@ dual :: Variance -> Variance
 dual Plus = Minus
 dual Minus = Plus
 
-cTrans :: Variance -> H.Expression -> H.Contract -> Fresh F.LabeledFormula
-cTrans v e c = F.LabeledFormula "hack2" <$> cTrans' v e c
+cTrans :: Variance -> H.Expression -> H.Contract -> Fresh [F.Formula]
+cTrans _ e H.Any = return [Top]
 
-cTrans' _ e H.Any = return Top
-
-cTrans' v e (H.Pred x p) =  do
+cTrans v e (H.Pred x p) =  do
   let  p' = H.subst e x p
   eT <- eTrans e
   p'T <- eTrans p'
[…]

### 2011-11-09 contracts/Translation.hs
Added (commented out) new projection axiom.
The axiom slows down equinox, and i'm not sure we gained
anything from it.

@@ -228,10 +228,21 @@ phi_project (H.Data _ dns) = map f dns where
   f (c,a,_) =
     let xs = makeVars a "X"
         xsN = map (Named . Var) xs
-        full k = F.FullApp (Proj k c) [F.FullApp (Con c) xsN]
+        fullProj i e = F.FullApp (Proj i c) e
+        fullProjK i = fullProj i [F.FullApp (Con c) xsN]
     -- XXX: the paper has 'min(c(xs))', but that can't be right ?
-    in F.Forall xs $ F.And [F.Min (full k) :=>: (full k :=: x)
-                           | (x,k) <- zip xsN [1..a]]
+        projectCorrect = F.Forall xs $ F.And [F.Min (fullProjK i) :=>: (fullProjK i :=: x)
+                                             | (x,i) <- zip xsN [1..a]]
+        -- XXX,DESIGN CHOICE: no mins. This axiom slowed down Equinox,
+        -- do we really want it?
+        --
+        -- forall y. (forall xs. K xs /= y) -> K_i y = UNR
+        projectWrong i = F.Forall [y] $ (F.Forall xs $ yN :/=: F.FullApp (Con c) xsN)
+                         :=>: (fullProj i [yN] :=: unr)
+        y = "ZDEF"
+        yN = Named . Var $ y
+        projectWrongs = F.And [projectWrong i | i <- [1..a]]
+    in projectCorrect --F.And [projectCorrect, projectWrongs]
 
 -- Axiom: Term constructors have disjoint ranges (Phi_2 in paper).
 phi_disjoint (H.Data _ dns) = map f $ zip dns (tail dns) where

### 2011-11-28 contracts/Translation.hs
10 character fix for a 5 hour bug :P
I placed the min to restrictively on phi_project, resulting in the inability
to prove that constructors are invertible, unless you're already interested
in their projection.  Figured this out because I tried to do a proof on
paper that we expected the theorem prover to do, and then got stuck
when i needed to prove 'Zero = X_' from 'Succ Zero = Succ X_'
and 'min(Succ Zero)'.

Ouch.

@@ -261,11 +261,17 @@ phi_project (H.Data _ dns) = map f dns where
   f (c,a,_) =
     let xs = makeVars a "X"
         xsN = map (Named . Var) xs
+        fullC = F.FullApp (Con c) xsN
         fullProj i e = F.FullApp (Proj i c) e
-        fullProjK i = fullProj i [F.FullApp (Con c) xsN]
+        fullProjK i = fullProj i [fullC]
     -- XXX: the paper has 'min(c(xs))', but that can't be right ?
-        projectCorrect = F.Forall xs $ F.And [F.Min (fullProjK i) :=>: (fullProjK i :=: x)
-                                             | (x,i) <- zip xsN [1..a]]
+
+    -- UPDATE: actually, it is :P interest in the constructor
+    -- application lets you invert!
+        projectCorrect = F.Forall xs $
+          F.And [ F.Or[F.Min (fullProjK i),F.Min(fullC)] 
+                  :=>: (fullProjK i :=: x)
+                | (x,i) <- zip xsN [1..a]]
         -- XXX,DESIGN CHOICE: no mins. This axiom slowed down Equinox,
         -- do we really want it?
         --

### 2011-11-30 contracts/FOL.hs
Use 'exists xs. phi' instead of '~ forall xs. ~ phi'.

@@ -30,6 +30,7 @@ trivializeMin :: Formula -> Formula
 trivializeMin = t where
   t f = case f of
     Forall xs f -> Forall xs (t f)
+    Exists xs f -> Exists xs (t f)
     f1 :=>: f2  -> t f1 :=>: t f2
     f1 :<=>: f2 -> t f1 :<=>: t f2
     Not f       -> Not $ t f
@@ -39,8 +40,12 @@ trivializeMin = t where
     f           -> f
 
 removeConstants :: Formula -> Formula
+-- Assuming a non-empty domain here.
 removeConstants (Forall [] f) = removeConstants f
+removeConstants (Exists [] f) = removeConstants f
 removeConstants (Forall xs f) = if f' == Top then Top else Forall xs f'
+  where f' = removeConstants f
+removeConstants (Exists xs f) = if f' == Top then Top else Exists xs f'
   where f' = removeConstants f
 removeConstants (f1 :=>: f2) = case (removeConstants f1, removeConstants f2) of
                                  (Top, f)    -> f
@@ -101,12 +106,15 @@ toTPTP (LabeledFormula l f) = fof
         -- n-ary function application
         fun qs f args = f <> nary goTerm qs comma args
 
+        quantifier qs q xs f = parens $ hang (text q <+> goQList xs <+> text ":")
+                                             2 (go (xs++qs) f)
+
         prepunctuate sep [d] = [d]
         prepunctuate sep (d:ds) = d : map (sep <+>) ds
         -- 'go qs f' converts formula 'f' to TPTP syntax, assuming
         -- 'qs' are the names of the quantified variables in 'f'.
-        go qs (Forall xs f) = parens $ (text "!" <+> goQList xs <+> text ":")
-                              $$ nest 2 (go (xs++qs) f)
+        go qs (Forall xs f) = quantifier qs "!" xs f
+        go qs (Exists xs f) = quantifier qs "?" xs f
 
         go qs (f1 :=>: f2) = naryF qs "=>" [f1,f2]
         go qs (f1 :<=>: f2) = naryF qs "<=>" [f1,f2]

### 2011-11-30 contracts/FOLTypes.hs

@@ -16,6 +16,7 @@ type Label = String
 infix 7 :<=>:
 infix 7 :=>:
 data MetaFormula a = Forall [Name] (MetaFormula a)
+                   | Exists [Name] (MetaFormula a)
                    | (MetaFormula a) :=>: (MetaFormula a)
                    | (MetaFormula a) :<=>: (MetaFormula a)
                    | Not (MetaFormula a)
@@ -34,4 +35,3 @@ data MetaFormula a = Forall [Name] (MetaFormula a)
                    deriving (Show,Eq,Functor)
 -- cf = Pred "cf"
 -- min = Pred "min"
-exists xs phi = Not $ Forall xs $ Not phi

### 2011-11-30 contracts/Translation.hs

@@ -179,7 +179,7 @@ dTrans (H.Let f vs ce) = do
           -- substitute the fresh pattern variables before recursing.
           ceT <- go (vs'++fvs) $ H.substsCE (zip vs'N vs) ce
           return (F.Forall vs' $        eT :/=: fullC
-                 ,F.exists vs' $ F.And [eT :=: fullC, ceT])
+                 ,F.Exists vs' $ F.And [eT :=: fullC, ceT])
     (nonConCases,conCases) <- unzip <$> mapM conCase pces
     let conCaseIneqs = [ F.Not eq | F.And (eq:_) <- conCases ]
         badCase = F.And [eT :=: bad, full :=: bad]

### 2011-11-30 contracts/FOL.hs
Added a simplifier that flattens nested n-ary ops of the same type.
E.g. 'And[f1,And[f2,f3],f4]' ==> 'And[f1,f2,f3,f4]' for 'f1,f2,f3,f4' atomic.

This would probably be less painful if I knew generic programming: I wrote
the same function twice, specialized to 'And' and 'Or', and both include
a "deep apply" style recursion where the "boring" cases simply recurse
into the subformulas.

@@ -39,6 +39,35 @@ trivializeMin = t where
     Min t       -> Top
     f           -> f
 
+-- | Replace e.g. 'And(fs1++[And fs]++fs2)' with 'And(fs1++fs++fs2)'.
+flattenNAryOps :: Formula -> Formula
+flattenNAryOps = goAnd . goOr where
+  me = flattenNAryOps
+  -- XXX: can constructor-generic programming be used here? And what
+  -- about a generic map over recursive subcomponents to effect the
+  -- "boring" cases?
+  goAnd (And fs) = And $ concat $ map (go . me) fs where
+    go (And fs) = fs
+    go f        = [f]
+  goAnd (Or fs) = Or $ map me fs
+  goAnd (Not f) = Not $ me f
+  goAnd (Forall xs f) = Forall xs $ me f
+  goAnd (Exists xs f) = Exists xs $ me f
+  goAnd (f1 :=>: f2)  = me f1 :=>: me f2
+  goAnd (f1 :<=>: f2) = me f1 :<=>: me f2
+  goAnd f             = f
+
+  goOr  (Or  fs) = Or  $ concat $ map (go . me) fs where
+    go (Or  fs) = fs
+    go f        = [f]
+  goOr  (And fs) = And $ map me fs
+  goOr  (Not f)  = Not $ me f
+  goOr  (Forall xs f) = Forall xs $ me f
+  goOr  (Exists xs f) = Exists xs $ me f
+  goOr  (f1 :=>: f2)  = me f1 :=>: me f2
+  goOr  (f1 :<=>: f2) = me f1 :<=>: me f2
+  goOr  f             = f
+
 removeConstants :: Formula -> Formula
 -- Assuming a non-empty domain here.
 removeConstants (Forall [] f) = removeConstants f
@@ -74,7 +103,10 @@ removeConstants f = f
 
 
 simplify :: Conf -> LabeledFormula -> [LabeledFormula]
-simplify cfg = splitOnAndLabeled . fmap removeConstants . maybeTrivializeMin
+simplify cfg = splitOnAndLabeled
+             . fmap flattenNAryOps
+             . fmap removeConstants
+             . maybeTrivializeMin
  where
   maybeTrivializeMin = if no_min cfg then fmap trivializeMin else id
 

### 2011-11-30 contracts/Options.hs
Renamed 'Check.hs' to 'Main.hs' and renamed executable to 'hcc'.
"hcc" is short for "Haskell Contract Checker".

@@ -52,7 +52,7 @@ getOpts = cmdArgs $ Conf
 ++". The prover configurations are in ThmProver.hs.")
 
   , idirs = ["."] -- Always check current dir first.
-    &= explicit -- Don't guess names for this option
+    &= explicit -- Don't guess switch names for this option
     &= name "idir" &= name "i"
     &= typDir
     &= help "Add DIR to the paths searched for imports. This option may be specified multiple times. Import dirs are searched in the order specified, with an implicit \".\" (current dir) first."
@@ -71,8 +71,8 @@ getOpts = cmdArgs $ Conf
     &= typFile
   } &= 
   verbosity &=
-  program "Check" &=
-  helpArg [explicit, name "h", name "help"] &= -- Avoid stupid '-?' default.
+  program "hcc" &=
+  helpArg [explicit, name "h", name "help"] &= -- Avoid stupid '-?' default switch.
   summary "Haskell Contract Checker, V 1/0" &=
   details
   -- XXX, MAYBE TODO: support module names on command line, in addition to paths.

### 2011-11-30 paper/paper.tex
Reorganized source code and udpated READMEs.
Now have src/, paper/, egs/, and Folkung/ (equinox) at top level.

Improved .gitignore too.

--- 2011-11-09 paper.tex
@@
       \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
-$\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
+$\min{a}$ or $\min{sel_{j,K_i}(a)}$?
+UPDATE: either!}}\\
       \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
@@
   \end{center}
-  \caption{$\ktrans{}$ -- Data type translation}
+  \caption{$\ktrans{}$ -- Data type translation\note[nc]{Need
+      $\min{a}$ to be sufficient for unfolding in
+      $\phi_\text{projection}$ to get invertability of constructors!
+      On the other hand, in a translation of case expressions based on
+      projectors, we probably need $\min{sel_{j,K_i}(a)}$ to be
+      sufficient for unfolding.  So, we make both sufficient in the
+      implementation: $\min{a} \lor \min{sel_{j,K_i}(a)}$.}}
   \label{ktrans}

### 2011-12-02 src/FOL.hs
Added '--unrolls N' to add N additional def unrollings for checked funs.
Improved labels on generated formulas along the way.

Now, I just need to implement an example :)

@@ -19,8 +19,9 @@ unlabel (LabeledFormula _ e) = e
 -- forall a . x && y --> (forall a . x) && (forall a . y)
 splitOnAndLabeled :: LabeledFormula -> [LabeledFormula]
 splitOnAndLabeled (LabeledFormula lbl f)
-  = [LabeledFormula (lbl++show i) f' | (i,f') <- zip [1..] (splitOnAnd f)]
-
+  = [LabeledFormula (lbl++i) f' | (i,f') <- zip labelExtensions (splitOnAnd f)]
+ where
+  labelExtensions = "":map (("__splitOnAnd_"++) . show) [1..]
 splitOnAnd :: Formula -> [Formula]
 splitOnAnd (Forall xs (And fs)) = map (Forall xs) fs
 splitOnAnd (Forall xs f) = map (Forall xs) $ splitOnAnd f
@@ -84,6 +85,12 @@ simplify cfg = splitOnAndLabeled
  where
   maybeTrivializeMin = if no_min cfg then trivializeMin else id
 
+named2TPTP :: Named -> String
+named2TPTP (Var v) = v
+named2TPTP (Con v) = "c_"++v -- "'" ++ v ++ "'"
+named2TPTP (Rec v) = v++"__r"
+named2TPTP (Proj i v) = "'"++v++"__p"++show i++"'"
+named2TPTP (Unroll i v) = v++"__u"++show i
 
 toTPTP :: LabeledFormula -> Doc
 toTPTP (LabeledFormula l f) = fof
@@ -144,10 +151,8 @@ toTPTP (LabeledFormula l f) = fof
         goTerm qs (FullApp f []) = goNamed qs f
         goTerm qs (FullApp f as) = fun qs (goFull f) as
 
-        goNamed qs (Var v) = goVar qs v
-        goNamed qs (Con v) = text $ "c_" ++ v -- "'" ++ v ++ "'"
-        goNamed qs (Rec v) = text $ v ++ "__R"
-        goNamed qs (Proj i v) = text $ "'"++v++"__"++show i++"'"
+        goNamed qs (Var v) = text $ named2TPTP $ Var $ goVar qs v
+        goNamed _  vN      = text $ named2TPTP vN
 
         -- Uppercase a list of quantified variables.
         goQList xs = brackets $ hsep $ punctuate comma $ map (text . uppercase) xs
@@ -156,7 +161,7 @@ toTPTP (LabeledFormula l f) = fof
         -- so no 'qs' here.
[…]

### 2011-12-02 src/Haskell.hs

@@ -14,6 +14,7 @@ getName (Var v) = v
 getName (Con v) = v
 getName (Rec v) = v
 getName (Proj _ v) = v
+getName (Unroll _ v) = v
 
 tls2Name :: TopLevelStatement -> Name
 tls2Name (Def (Let f _ _))         = f
@@ -29,20 +30,16 @@ arities ds = concatMap go ds
         go (DataType (Data _ vacs)) = [(v,a) | (v,a,_) <- vacs]
         go _ = []
 
+-- | "Appify" all 'Expression's in a term.
 appify :: [Arity] -> GenericT
 appify = gfmap . appifyExpr
-
--- XXX, TODO: rename
--- | Return the arity of a name
-lookupT :: Named -> [Arity] -> Maybe Int
-lookupT v as = (lookup . getName) v as where
 
 -- takes a program and a list of arities for each definition
 -- returns the same program but using full application wherever possible
 appifyExpr :: [Arity] -> Expression -> Expression
 appifyExpr a e = go e []
   -- 'a' is the arities, 'args' is the arguments to the enclosing applications.
-  where go (n@(Named v) :@: e) args = case lookupT v a of
+  where go (n@(Named v) :@: e) args = case lookup (getName v) a of
           Just k -> if length args' == k
                     then FullApp v args'
                     else apps (n : e' : args)

### 2011-12-02 src/Main.hs

@@ -114,7 +114,7 @@ checkFile cfg f = do
   -- change how most formulas are generated, since most formulas are
   -- generated using full application where possible. If appification
   -- were disabled, the formulas relating full and partial application
-  -- could be elided (e.g. 'dPtr').
+  -- could be elided (e.g. 'ptrAxiom').
   let prog' = appify (arities prog) prog
       sccs = orderedChecks prog'
       -- If the set of functions to check is restricted, then check
@@ -152,7 +152,7 @@ check cfg f prog (checks,deps) | all null contracts = return True
           $ assert (deps' == nub deps')
           $ assert (null (checks' `intersect` deps'))
           $ showFormula thy $ simplify cfg
-                              =<< trans checks' deps'
+                              =<< trans (unrolls cfg) checks' deps'
       tmpFile = takeFileName f ++
                 "." ++ intercalate "-" fs ++
                 "." ++ fileExtension thy

### 2011-12-02 src/Options.hs

@@ -18,6 +18,7 @@ data Conf = Conf { keep_tmps    :: Bool
                  , type_check   :: Bool
 
                  , no_min       :: Bool
+                 , unrolls      :: Int
                  , file         :: FilePath
                  } deriving (Show, Data, Typeable)
 
@@ -64,6 +65,9 @@ getOpts = cmdArgs $ Conf
 
   , no_min = def
     &= help "Don't use the 'min' predicate in the translation. This should degrade performance, but makes the resulting theory file much easier to read, and can be used to debug changes to 'min' placement, e.g. to check if they are too restrictive."
+
+  , unrolls = def
+    &= help "Specify the number of additional unrollings to perform when translating function definitions."
 
   , file = def
     &= argPos 0

### 2011-12-02 src/Translation.hs

@@ -56,7 +56,9 @@ eTrans = return
 -- forall [x1, ..., xn]. f(x1, ..., xn) = f@x1@...@xn
 --
 -- NB: appification breaks this, so don't 'appify' it!
-dPtr f vs = if null vs then Top else F.And [eq1Min]--,eq2Min]
+ptrAxiom vs f = F.LabeledFormula 
+  ("ptrAxiom__"++F.named2TPTP f)
+  (if null vs then Top else F.And [eq1Min]) --,eq2Min]
   -- 'null' check above to avoid pointless 'f = f'.
   where vsN = map (Named . Var) vs
         fN = Named f
@@ -89,9 +91,6 @@ dPtr f vs = if null vs then Top else F.And [eq1Min]--,eq2Min]
         fullCFMin = F.Forall vs $ allCF :=>: (F.Min full :=>: F.CF full)
         eq2Min = F.Min fN :=>: (fullCFMin :<=>: F.CF fN) -- XXX, ???: is this min right?
 
--- | Same as dPtr, but for recursive occurances of f.
-dPtrRec f@(Var v) = dPtr $ Rec v
-
 -- | Shorthand for a common idiom.
 nv = Named . Var
 -- | Return a fresh variable name, based on 'v', which does not occur
@@ -112,16 +111,25 @@ ceTrans (H.Case e pces) = H.Case <$> eTrans e <*> mapM pceTrans pces where
   pceTrans (p,ce) = do ceT <- ceTrans ce
                        return (p, ceT)
 
--- | Translate a definition.
-dTrans :: H.Definition -> Fresh [F.LabeledFormula]
-dTrans (H.Let f vs ce) = do
+
+
+-- | Translate a definition 'Let _ vs ce', using the given 'Named'
+-- 'fV' to name it.  The actual 'Let' is not passed, just the
+-- components.
+--
+-- You supply 'fV' because in the case of multiple unrollings we need
+-- to give the function a different name.
+--
+-- For a non-unrolled 'dTrans' you pass 'Var f' for 'fV', assuming
+-- the 'Definition' is 'Let f vs ce'.
+dTrans :: H.Named -> [Name] -> H.Case -> Fresh [F.LabeledFormula]
+dTrans fV vs ce = do
   rhs <- go vs ce
   let eq = F.Forall vs $ F.Min full :=>: rhs
-  return $ map (F.LabeledFormula f)
-               [eq, dPtr fV vs, dPtrRec fV vs]
[…]

### 2011-12-02 src/Types/Haskell.hs

@@ -23,6 +23,8 @@ data Named
    -- to enforce this.
    | Rec Name -- ^ Recursive version of a function
    | Proj Int Name -- ^ Projector for a term constructor.
+   | Unroll Int Name -- ^ An unrolling of a function.  Used for
+                     -- multiple unrolling support.
    -- There is no 'Full' because full application is
    -- determined by context.
    deriving (Eq,Ord,Show,Data,Typeable)

### 2011-12-02 paper/paper.tex
Paper: Updated to reflect the multiple unrollings implementation.

@@
 
-$$\trans{M - f}, \dtrans{f=e[f/\rec{f}]}, \ctrans{\rec{f} \in c} \vdash \ctrans{f \in c}$$
-
+$$\trans{M - f}, \dtrans{f=e[f/\rec{f}]}, \ctrans{\rec{f} \in c[\rec{f}/f]}^- \vdash \ctrans{f \in c}^+$$
+\note[nc]{The LHS used to have $\ctrans{\rec{f} \in c}^-$, 
+not $\ctrans{\rec{f} \in c[\rec{f}/f]}^-$, 
+but the code has always implemented the latter. Which is preferable???}
 Where $M - f $ means the content of the module $M$ without $f$'s
@@
  }
+
+ So, we provide an option to control the number, $N$, of unrollings.
+ First, note that recursive anonymous functions, and hence unrolling,
+ only apply to checked functions in the SCC; the dependencies of the
+ SCC being checked are not anonymized.  If $fs$ are the functions
+ being checked, then for each $f xs = e \in fs$ we generate
+ \[
+   \bigwedge_{i=0,N+1} \dtrans{f_i ~xs = e[g_{i+1}/g]_{g \in fs}},
+ \]
+ where $g_0 = g$, $g_{N+1} = \rec{g}$, and $\{g_i\}_{i = 1 \text{ to } N }$ 
+are new symbols, for each $g \in fs$.  Note that all functions
+in $fs$ receive the same amount, $N$ levels, of unrolling.  We could
+provide a per function unrolling value, but that would complicate
+things, and it's not clear there would be any benefit.  The only
+concern with the uniform unrolling is the complexity of the generated
+theory.
+
 \subsection{Module checking}

### 2011-12-06 paper/paper.tex
Removed extraneous 'min's from premise in phi_project per discussion with Koen.

@@
     \[  \begin{array}{rcl}
-      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{???
-$\min{a}$ or $\min{sel_{j,K_i}(a)}$?
-UPDATE: either!}}\\
+      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{ $\min{a}$, not $\min{sel_{j,K_i}(a)}$.}\\
       \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
@@
       $\phi_\text{projection}$ to get invertability of constructors!
-      On the other hand, in a translation of case expressions based on
-      projectors, we probably need $\min{sel_{j,K_i}(a)}$ to be
-      sufficient for unfolding.  So, we make both sufficient in the
-      implementation: $\min{a} \lor \min{sel_{j,K_i}(a)}$.}}
+      I thought we needed $\min{sel_{j,K_i}(a)}$ to be sufficient when
+      we use a projection based translation of case expressions, but
+      as Koen pointed out this is not true.
+      }}
   \label{ktrans}

### 2011-12-06 src/Translation.hs

@@ -279,9 +279,6 @@ tTrans d@(H.Data nm _) = do
   where ptr (H.Data _ cas) = [ptrAxiom (makeVars a "X") (Con c) | (c,a,_) <- cas]
 
 -- | Axiom: Term constructors are invertable (Phi_1 in paper).
---
--- XXX, DESIGN CHOICE: this axiom isn't used, although it could be
--- used to eliminate some quantified variables in other axioms.
 phi_project (H.Data t dns) = map f dns where
   f (c,a,_) =
     let xs = makeVars a "X"
@@ -294,7 +291,7 @@ phi_project (H.Data t dns) = map f dns where
     -- UPDATE: actually, it is :P interest in the constructor
     -- application lets you invert!
         projectCorrect = F.Forall xs $
-          F.And [ F.Or[F.Min (fullProjK i),F.Min(fullC)] 
+          F.And [ F.Min(fullC) -- F.Or[F.Min (fullProjK i),F.Min(fullC)] 
                   :=>: (fullProjK i :=: x)
                 | (x,i) <- zip xsN [1..a]]
         -- XXX,DESIGN CHOICE: no mins. This axiom slowed down Equinox,

### 2011-12-06 src/Translation.hs
Contract translation now collects arrows and uses a single universal quantifier.
E.g. for

  [| e in x1:c1 -> ... -> xn:cn -> c |]

we used to get

  forall x1. min (e x1) -> [| x1 in c1 |] -> ...
             -> forall xn. min(e x1 ... xn) -> [| xn in cn |]
                           -> [| e x1 ... xn in c |]

but now we get

  forall x1 ... xn. min (e x1 x2 ... xn)
                    -> [| x1 in c1 |] /\ ... /\ [| xn in cn |]
                    -> [| e x1 ... xn in c |]

This is important for Skolemizing goal formulas, because we need to turn

  ~ forall x1 ... xn. phi

into

  ~ phi [ s1

@@ -236,7 +236,30 @@ cTrans' v e (H.Pred x p) =  do
     Plus  -> return $ F.And [F.Min(eT),            F.Min(p'T)] :=>: plain
     Minus -> return $        F.Min(eT)  :=>: F.And [F.Min(p'T),      plain]
 
-cTrans' v e (H.Arr mx c1 c2) = do
+cTrans' v e c@(H.Arr mx c1 c2) = do
+  eT <- eTrans e
+  (xs,cs,c') <- collectArrows c
+  let xsN = map nv xs
+  let app = H.apps (eT : xsN)
+  csT <- sequence [ cTrans' (dual v) xN c | xN <- xsN | c <- cs ]
+  c'T <- cTrans' v app c'
+  return $
+    -- XXX, DESIGN CHOICE: the single 'min' constraint here is only
+    -- adequate because we have a 'min(f x) -> min(f)' axiom (???).
+    -- Moreover, we may get this min anyway in c'T.
+    F.Forall xs $ F.Min app :=>: (F.And csT :=>: c'T)
+ where
+  -- Collect all the variables and contracts for a sequence of
+  -- arrows. Returns '(variables, variable contracts, conclusion contract)'.
+  collectArrows (H.Arr mx c1 c2) = do
+    x <- maybe fresh return mx
+    (xs, cs, c') <- collectArrows c2
+    return $ (x:xs, c1:cs, c')
+  collectArrows c' = return ([],[],c')
+{-
+  -- Simpler cTrans', as in paper, where the arrow sequence is not
+  -- collected.
+
   -- Parser inserts 'Nothing' for unnamed arrow arguments.
   x <- maybe fresh return mx
   let xN = Named $ Var x
@@ -246,6 +269,7 @@ cTrans' v e (H.Arr mx c1 c2) = do
   -- XXX, DESIGN CHOICE: need for 'min' constraint here depends on
   -- whether we have a 'min(f x) -> min(f)' axiom.
   return $ F.Forall [x] $ F.Min app :=>: (f1 :=>: f2)
+-}
 
 cTrans' v e (H.And c1 c2) = do
   f1 <- cTrans' v e c1

### 2011-12-06 src/FOL.hs
Translation now skolemizes outermost existentials in goal.
And they get printed with the leading a_, so Equinox should treat
them specially.

I have not looked at the models yet.

@@ -87,6 +87,7 @@ simplify cfg = splitOnAndLabeled
 
 named2TPTP :: Named -> String
 named2TPTP (Var v) = v
+named2TPTP (Skolem v) = "a_"++v
 named2TPTP (Con v) = "k_"++v -- "'" ++ v ++ "'"
 named2TPTP (Rec v) = v++"_r"
 named2TPTP (Proj i v) = "p" ++ show i ++ "_" ++ v

### 2011-12-06 src/Haskell.hs

@@ -6,6 +6,7 @@ import Data.Maybe (isJust)
 
 import Types.Haskell
 import Types.Translation (Arity)
+import qualified Types.FOL as F
 import Generics (gfmap,GenericT)
 
 -- | Projector for Named
@@ -61,10 +62,11 @@ appifyExpr a e = go e []
 
 -- | Perform many substitutions, rightmost first.
 substs :: [(Expression, Name)] -> Expression -> Expression
-substs subs e    = foldr (uncurry subst) e subs
-substsC subs c   = foldr (uncurry substC) c subs
+-- XXX: this is screaming "type class"
+substs   subs e  = foldr (uncurry subst)   e  subs
+substsC  subs c  = foldr (uncurry substC)  c  subs
 substsCE subs ce = foldr (uncurry substCE) ce subs
-
+substsF  subs f  = foldr (uncurry substF)  f  subs
 -- | 'subst e1 y e2' = e2[e1/y]
 --
 -- NB: only Var, and not Con or Rec, can be substituted for.  The idea
@@ -103,7 +105,23 @@ substCE e y (Case e' pces) = Case (subst e y e') (map substP pces) where
                            then pce
                            else ((c,vs), substCE e y ce)
 
-
+-- | Substitution for formulas.
+substF :: Expression -> Name -> F.Formula -> F.Formula
+substF e x f = go f where
+  goT = subst e x
+  go f@(F.Forall xs f') = if x `elem` xs then f else F.Forall xs $ go f'
+  go f@(F.Exists xs f') = if x `elem` xs then f else F.Exists xs $ go f'
+  go (f1 F.:=>: f2) = go f1 F.:=>: go f2
+  go (f1 F.:<=>: f2) = go f1 F.:<=>: go f2
+  go (F.Not f) = F.Not $ go f
+  go (F.Or fs) = F.Or $ map go fs
+  go (F.And fs) = F.And $ map go fs
+  go F.Top = F.Top
+  go F.Bottom = F.Bottom
+  go (t1 F.:=: t2) = goT t1 F.:=: goT t2
+  go (t1 F.:/=: t2) = goT t1 F.:/=: goT t2
+  go (F.CF t) = F.CF $ goT t
+  go (F.Min t) = F.Min $ goT t
 
 -- | Generate a Haskell function 'f' from a contract 'c', s.t. if 'f'
 -- type checks then 'c' would too.

### 2011-12-06 src/Translation.hs

@@ -12,7 +12,7 @@ import Types.Translation
 import Types.ThmProver (Conf(..))
 
 import Control.Monad.State
-import Data.List (partition, intercalate)
+import Data.List (partition, intercalate, foldl')
 import Data.Char (toUpper)
 import Control.Applicative
 import Control.Exception (assert)
@@ -489,10 +489,21 @@ trans cfg' checks deps = evalState result startState
     -- because we do a refutation proof.
     goalFormula <- do
       goals <- gets getGoals
+      -- XXX, HERE: only skolemize when the engine is Equinox.  Also,
+      -- generated lemmas when the engine is Coq.
       let labels = map F.getLabel goals
           formulas = map F.getFormula goals
           label = intercalate "_" labels
-          formula = F.Not $ F.And formulas
+          sv = H.Named . H.Skolem
+          -- Freshen the foralls because we will replace them with
+          -- with skolem variables. NB: these foralls are really
+          -- exists', because the goal is negated.
+          formulas' = fst $ foldl' freshen ([],[]) formulas where
+            (phis,fvs) `freshen` (F.Forall xs phi) = (phi':phis, fvs'++fvs) where
+              fvs' = makeFVs fvs xs
+              phi' = H.substsF (zip (map sv fvs') xs) phi
+            (phis,fvs) `freshen` f = (f:phis,fvs)
+          formula = F.Not $ F.And formulas'
       return $ F.LabeledFormula label formula
 
     return . concat $ 

### 2011-12-06 src/Types/Haskell.hs

@@ -18,6 +18,7 @@ type Name = String
 -- application, is the special case.
 data Named
    = Var Name -- ^ Regular variable, including functions.
+   | Skolem Name -- ^ Skolemized variable (we don't have Skolem functions).
    | Con Name -- ^ Constructor
    -- The rest are only relevant to FOL? Could use GADT tricks
    -- to enforce this.

### 2011-12-16 src/Translation.hs
Renamed 'phi_project' to 'phi_injective' and 'phi_total' to 'phi_lazy'.
Injectivity is the key here, and I inly understood this
recently.  I'm going to emphasize it in the paper.

'lazy' is better than 'total', because the point is that constructors
don't scrutinize their arguments, and is the term already used in the
paper.

@@ -317,15 +317,15 @@ tTrans d@(H.Data nm cas) = do
   phi_cfd <- phi_cf d
   ptr_d <- sequence [ ptrAxiom (makeVars a "X") (Con c)
                     | (c,a,_) <- cas ]
-  let phis = concat [ phi_project d
+  let phis = concat [ phi_injective d
                     , phi_disjoint d
-                    , phi_total d]
+                    , phi_lazy d]
              ++ phi_cfd
              ++ ptr_d
   return phis
 
 -- | Axiom: Term constructors are invertable (Phi_1 in paper).
-phi_project (H.Data t dns) = map f dns where
+phi_injective (H.Data t dns) = map f dns where
   f (c,a,_) =
     let xs = makeVars a "X"
         xsN = map (Named . Var) xs
@@ -349,7 +349,7 @@ phi_project (H.Data t dns) = map f dns where
         y = "ZDEF"
         yN = Named . Var $ y
         projectWrongs = F.And [projectWrong i | i <- [1..a]]
-    in F.LabeledFormula (F.axiom $ "phi_project__"++t++"__"++c) 
+    in F.LabeledFormula (F.axiom $ "phi_injective__"++t++"__"++c) 
          projectCorrect --F.And [projectCorrect, projectWrongs]
 
 -- Axiom: Term constructors have disjoint ranges (Phi_2 in paper).
@@ -390,12 +390,12 @@ phi_cf (H.Data t dns) = mapM f dns where
                        :=>: (F.And [F.CF x | x <- xsN]))
     return $ F.LabeledFormula (F.axiom $ "phi_cf__"++t++"__"++c) $ F.And [phi,cfc]
 -- Axiom: Term constructors are total/lazy (Phi_4 in paper).
-phi_total (H.Data t dns) = map f dns where
+phi_lazy (H.Data t dns) = map f dns where
   f (c,a,_) =
     let xs = makeVars a "X"
         xsN = map (Named . Var) xs
         full = F.FullApp (Con c) xsN
-    in F.LabeledFormula (F.axiom $ "phi_total__"++t++"__"++c) $
+    in F.LabeledFormula (F.axiom $ "phi_lazy__"++t++"__"++c) $
          F.Forall xs $ F.Min full :=>:
            F.And [full :/=: unr
                  ,full :/=: bad]

### 2011-12-16 paper/paper.tex
Changed projection syntax and name of datatype translation.

@@
 \newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
-\newcommand{\ktrans}[1]{\mathcal{K} \llbracket #1 \rrbracket}
+\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
 \newcommand{\ctrans}[1]{\mathcal{C} \llbracket #1 \rrbracket}
@@
 \renewcommand{\min}[1]{\mbox{min}(#1)}
+\newcommand{\proj}[2]{\pi_{#1}^{#2}}% the #1^th projection of constructor #2
 
@@
   doesn't seem to play any role in formal part of this paper (besides
-  being specified in axioms $\phi_\text{projection}$). In the current
+  being specified in axioms $\phi_\text{injective}$). In the current
   implementation it's used to avoid introducing existentially
@@
 \begin{equation*}
-\forall xs,y,ys.~sel_{1,Cons}~Cons~y~ys = xs \implies xs = y 
+\forall xs,y,ys.~\proj{1}{Cons}~Cons~y~ys = xs \implies xs = y 
 \end{equation*}
 \begin{equation*}
-\forall xs,y,ys.~sel_{2,Cons}~Cons~y~ys = xs \implies xs = ys 
+\forall xs,y,ys.~\proj{2}{Cons}~Cons~y~ys = xs \implies xs = ys 
 \end{equation*}\note[nc]{???
-Why not the simpler \[\forall y,ys.~sel_{2,Cons}~Cons~y~ys = ys\]?
+Why not the simpler \[\forall y,ys.~\proj{2}{Cons}~Cons~y~ys = ys\]?
 Same question in Figure \ref{dtrans}.
@@
   \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
-  \land xs \neq Cons~(sel_{1,Cons}~xs)~(sel_{2,Cons}~xs) \to  \notag \\
+  \land xs \neq Cons~(\proj{1}{Cons}~xs)~(\proj{2}{Cons}~xs) \to  \notag \\
   notnull(xs) = \unr 
@@
 \[
-\forall xs.~ xs=Cons~(sel_{Cons,1}~xs)~(sel_{Cons,2}~xs) \implies head~xs = sel_{Cons,1}~xs
+\forall xs.~ xs=Cons~(\proj{Cons}{1}~xs)~(\proj{Cons}{2}~xs) \implies head~xs = \proj{Cons}{1}~xs
 \]?  Because it's ugly?  Would Equinox care?}
@@
  \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
- \land xs \neq Cons~(sel_{1,Cons}~xs)~(sel_{2,Cons}~xs) \to  \notag \\
+ \land xs \neq Cons~(\proj{1}{Cons}~xs)~(\proj{2}{Cons}~xs) \to  \notag \\
  head~xs = \unr
@@
 A (possibly open) expression $e$ is \emph{syntactically safe} iff $\bad \not
-\in_s e$. Similarly a context $\mathcal{C}$ is \emph{syntactically safe} iff $\bad
-\not \in_s \mathcal{C}$.
+\in_s e$. Similarly a context $\mathcal{C}[\cdot]$ is \emph{syntactically safe} iff $\bad
+\not \in_s \mathcal{C}[\cdot]$.
 \end{definition}
@@
 The notation $\bad \notin_s e$ means that $\bad$ does not appear
-anywhere in $e$\note[nc]{,including the definitions of the function symbols used in $e$}, similarly for $\bad \not \in_s \mathcal{C}$. For
+anywhere in $e$ or $e$'s (transitive) dependencies.  For
 example, $Just ~3$ is syntactically safe whereas $Just ~\bad$ is not.
@@
 An expression $e$ is said to be \emph{crash-free} iff 
-for all $\mathcal{C}$, $\bad \not \in_s \mathcal{C}$ and $\emptyset \vdash
-\mathcal{C} \llbracket e \rrbracket :: ()$ implies $\mathcal{C} \llbracket e \rrbracket \not \to^* \bad$
+for all $\mathcal{C}[\cdot]$, $\bad \not \in_s \mathcal{C}[\cdot]$ and $\emptyset \vdash
+\mathcal{C}[e] :: ()$ implies $\mathcal{C}[e] \not \to^* \bad$
 \end{definition}
-The notation $\emptyset \vdash \mathcal{C} \llbracket e \rrbracket :: ()$ means that
-$\mathcal{C} \llbracket e \rrbracket$ is closed and well-typed.  Note
+The notation $\emptyset \vdash \mathcal{C}[e] :: ()$ means that
+$\mathcal{C}[e]$ is closed and well-typed.  Note
 that there are crash-free expressions that are not syntactically safe,
@@
     \dtrans{def} &\to& \phi~(\mbox{formula}) \\
-    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
+    \ttrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
     \ctrans{f \in c} &\to& \phi~(\mbox{formula}) \\
@@
   \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
-  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ktrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
+  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
   \dtrans{f \in c} &=& \ctrans{f \in c}\\
@@
 
-\subsection{$\ktrans{}$ -- Datatypes}
+\subsection{$\ttrans{}$ -- Datatypes}
 We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
-$$\ktrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+$$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 
 \begin{itemize}
-\item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $sel_{k,K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
-  on its $k$-th component, $k = 1\text{ to }a_i$.
+\item $(\phi_\text{injective})$ For each $K_i$ of arity $a_i$ we introduce selectors
+  $\proj{k}{K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
+  on its $k$-th component, for $k = 1\text{ to }a_i$.
 \item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
@@
 \begin{figure*}
-  $$ \ruleform{\ktrans{data~def} = \Phi} $$
-$$\ktrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+  $$ \ruleform{\ttrans{data~def} = \Phi} $$
+$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
  \hspace{5.6cm}where
@@
     \[  \begin{array}{rcl}
-      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{ $\min{a}$, not $\min{sel_{j,K_i}(a)}$.}\\
+      \phi_\text{injective} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(a) \text{\note[nc]{ $\min{a}$, not $\min{\proj{j}{K_i}(a)}$.}}\\
       \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
@@
   \end{center}
-  \caption{$\ktrans{}$ -- Data type translation\note[nc]{Need
+  \caption{$\ttrans{}$ -- Data type translation\note[nc]{Need
       $\min{a}$ to be sufficient for unfolding in
-      $\phi_\text{projection}$ to get invertability of constructors!
-      I thought we needed $\min{sel_{j,K_i}(a)}$ to be sufficient when
+      $\phi_\text{injective}$ to get invertability of constructors!
+      I thought we needed $\min{\proj{j}{K_i}(a)}$ to be sufficient when
       we use a projection based translation of case expressions, but
@@
 positions in an implication: $- \to +$.  The goal of the $\min$'s is to limit the
-applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ktrans{}$), 
+applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
 in order to prune 

### 2011-12-19 paper/paper.tex
Paper: cleaned up the description of H' and the data-type translation.

@@
 
-The syntax of $\mathcal{H}'$ is defined in figure~\ref{hprime-stx}. A module
-is a list of toplevel definitions, claims that functions statisfy
-contracts and data definitions.
+The syntax of $\hprime$ is defined in figure~\ref{hprime-stx}. A
+module is a list of toplevel definitions, claims that functions
+statisfy contracts and data definitions.  The $\hprime$ language we
+present is minimalistic, but the implementation supports a little
+more.
 
 \begin{itemize}
-\item There's no $\lambda$-abstraction, because we can always lift
-  them to a toplevel declaration.
-\item We do not allow nested case expressions, because once again, we
-  can always lift them to the toplevel.
-\item Until section~\ref{ho} we will only consider full application of
-  functions ($\full{f}(x,y)$),\note[nc]{
-
-Not true: the definition of satisfaction for arrow contracts
+\item $\hprime$ doesn't include $\lambda$-abstractions, only top-level
+  function definitions, because nested $\lambda$s can be lifted to the
+  top-level.
+\item $\hprime$ doesn't include nested case expressions, because
+  nested case expressions can be lifted to the top-level.  The
+  implementation actually supports nested case expressions.
+
+\note[nc]{
+Not true that we only consider full application of functions: the definition of satisfaction for arrow contracts
 $x:c_1 \to c_2$ already includes partial applications.  We could add a
@@
 \begin{eqnarray*}
-&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
+&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c}^v \\
 &&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
-&&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
-              \lor \ctrans{\full{f}(x_1, \ldots, x_n) \in c}
+&&\quad   \to \left( \bigwedge_{i=1}^n \ctrans{x_i \in c_i}^{\bar{v}} \right)
+              \to \ctrans{\full{f}(x_1, \ldots, x_n) \in c}^v
 \end{eqnarray*}
@@
   \[  \begin{array}{rcl}
-    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t) \mid \min{t} \\
+    \phi &:=& \forall x.\phi \mid \exists x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t) \mid \min{t} \\
     && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
@@
 \begin{itemize}
-\item $(\phi_\text{injective})$ For each $K_i$ of arity $a_i$ we introduce selectors
-  $\proj{k}{K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
-  on its $k$-th component, for $k = 1\text{ to }a_i$.
+\item $(\phi_\text{injective})$ Constructors are injective.  That is, for
+  each term constructor $K$,
+  if $K e_1 \cdots e_n = K e'_1 \cdots e'_n$, then $e_i = e'_i$ for
+  all $i$.  We axiomitize this via projections $\proj{i}{K}$: for each $i$,
+  $\proj{i}{K}(K e_1 \cdots e_n) = e_i$.
 \item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
@@
   \item $\gets$: if we pack crash-free values in a data constructor,
-    the resulting value is crash-free. \note[nc]{We use the contract
-    translation for arrows, and this translation is principled.}
+    the resulting value is crash-free.  For $K$ has arity $n$
+    we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.
+
   \item $\to$: a value $t$ of type $T$\note[nc]{
@@
     } is crash-free implies
-    that every value packed in it is crash-free. Recall that one can
-    define projection on any argument of a value of type $t$. So if
-    the $k$-th argument of $t$ is not crash-free, then the $k$-th
-    projection is a crash-free context that throws an expression that
-    is not crash-free. \note[nc]{XXX: This translation is ad-hoc.}
+    that every value packed in it is crash-free, since we can use
+    a case expression to extract any packed value.
+    \note[nc]{XXX: This translation is ad-hoc.}
 
@@
   if the first argument of the pair is crash-free. The second argument
-  being crash-free or not doesn't matter.
+  is irrelevant.
 

### 2011-12-19 src/Translation.hs
cTrans: no min's for arrows, and stronger predicate in 'Minus' case.

@@ -251,10 +251,14 @@ cTrans' v e (H.Pred x p) =  do
   eT <- eTrans e
   p'T <- eTrans p'
   -- XXX, DESIGN CHOICE: could also do 'F.Or[p'T :=: unr, p'T :=: true]'
-  let plain = F.And [F.Or [(eT :=: unr), F.And [p'T :/=: bad, p'T :/=: false]]]
+  --
+  -- Or, use 'in {unr,true}' in for assumptions and 'not-in
+  -- {bad,false}' for goals.
+  let plainM = F.Or [(eT :=: unr), F.Or  [p'T :=:  unr, p'T :=:  true ]]
+  let plainP = F.Or [(eT :=: unr), F.And [p'T :/=: bad, p'T :/=: false]]
   case v of
-    Plus  -> return $ F.And [F.Min(eT),            F.Min(p'T)] :=>: plain
-    Minus -> return $        F.Min(eT)  :=>: F.And [F.Min(p'T),      plain]
+    Plus  -> return $ F.And [F.Min(eT),             F.Min(p'T)] :=>: plainP
+    Minus -> return $        F.Min(eT)  :=>: F.And [F.Min(p'T),      plainM]
 
 cTrans' v e c@(H.Arr mx c1 c2) = do
   eT <- eTrans e
@@ -267,7 +271,11 @@ cTrans' v e c@(H.Arr mx c1 c2) = do
     -- XXX, DESIGN CHOICE: the single 'min' constraint here is only
     -- adequate because we have a 'min(f x) -> min(f)' axiom (???).
     -- Moreover, we may get this min anyway in c'T.
-    F.Forall xs $ F.Min app :=>: (F.And csT :=>: c'T)
+    --
+    -- UPDATE: getting rid of the 'min(f x)'
+    --
+    -- F.Forall xs $ F.Min app :=>: (F.And csT :=>: c'T)
+    F.Forall xs $ F.And csT :=>: c'T
  where
   -- Collect all the variables and contracts for a sequence of
   -- arrows. Returns '(variables, variable contracts, conclusion contract)'.
@@ -307,6 +315,7 @@ cTrans' v e (H.CF) = do
   -- the 'Minus' case.
   return $ case v of
     Plus  -> F.Min eT :=>: F.CF eT
+    --Minus -> F.Min eT :=>: F.CF eT
     Minus -> F.CF eT
 
 -- Data decl translation

### 2011-12-20 paper/paper.tex
Paper: removed most mention of old translation.
The fancy figures still have the old trans.  The plain text figures
use the new translation.

@@
 
-The syntax of $\hprime$ is defined in figure~\ref{hprime-stx}. A
+The syntax of $\hprime$ is defined in figure~\ref{fig:hprime-syntax}. A
 module is a list of toplevel definitions, claims that functions
@@
   \begin{array}{rclr}
-    mod &:=& def_1,\dots,def_n &\\
+    \hprime module &:=& def_1,\dots,def_n &\\
     def &\in& \mbox{Definition} & \\
@@
 \caption{Syntax of the language $\hprime$}
-\label{hprime-stx}
+\label{fig:hprime-syntax}
 \end{figure}
@@
 Contract syntax is described in figure \ref{cont-stx}. The predicates
-we use in our contracts can be any boolean $\hprime$ expression. We
+$p$ in contracts range over boolean $\hprime$ expressions. We
 only consider pairs of contract for simplicity, although there is no
@@
 anywhere in $e$ or $e$'s (transitive) dependencies.  For
-example, $Just ~3$ is syntactically safe whereas $Just ~\bad$ is not.
+example, $Just ~3$ is syntactically safe whereas $Just ~jb$ is 
+not, for $jb$ defined in a separate declaration as $jb = Just~\bad$.
 
@@
 forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
-\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} The expression language in our logic is \hprime.
+\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} 
+The expression language in our logic is the expression language of 
+\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
+corresponds to a binary function symbol in FOL, which we denote by juxtaposition.
 
@@
 \note[nc]{%
-Figure~\ref{fig:newCTrans} contains the new min contract translation, and
-Figure~\ref{fig:naiveCTrans} contains the naive contract translation
+Figure~\ref{fig:newCTrans} contains the $\min$-translation of contracts, and
+Figure~\ref{fig:naiveCTrans} contains the naive contract translation,
 with no $\min$'s.  The min translation reduces to the naive translation when
@@
 plus comes from the negative and positive
-positions in an implication: $- \to +$.  The goal of the $\min$'s is to limit the
+positions in an implication: $- \to +$.  The goal of the $\min$'s twofold: 
+(1) to allow finite counter models when the theory 
+does not prove the contracts; (2) to limit the
 applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
-in order to prune 
-the search space.  So, $\min$'s
+in order to prune the proof search space.  The $\min$'s
 appear as obligations (the premises in implications) 
@@
 that we're interested in a particular application.  
-Once that obligation is satisfied, interest ($\min{e}$) in the
+Once that $\min$ obligation is satisfied, interest ($\min{e}$) in the
 scrutinee of the case match may be concluded, because the function
@@
 Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
-Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
+Figure~\ref{fig:newCTrans}.  When proving that $e \in \{x|p\}$, we get to
 assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
@@
 \begin{verbatim}
-[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
-[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
-[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}
-
+-- Need 'min(e)' and 'min(p[e/x])' here to unfold defs.
+[| e ::: {x|p} |]+ = min(e) /\ min(p[e/x]) ->
+                     (e /= unr -> p[e/x] not-in {false, bad})
+-- Need 'min(e)' here to guard 'min(p[e/x])'.
+-- Need 'min(p[e/x])' here to unfold defs.
+[| e ::: {x|p} |]- = min(e) -> min(p[e/x]) /\
+                     (e /= unr -> p[e/x]     in {true, unr})
+
+ -- Need 'min(e)' here to unfold defs.
 [[e in cf]]+ := min(e) ->  cf(e)
-[[e in cf]]- := [min(e) ->] cf(e)
-
--- 'min's optional if a
--- 'forall f,x. min(f x) -> min(f)' axiom
--- is added to prelude.
-[[e in x:c1 -> c2]]+
-  := forall x. [min(e x) ->] [[x in c1]]- -> [[e x in c2]]+   
-[[e in x:c1 -> c2]]-
-  := forall x. [min(e x) ->] [[x in c1]]+ -> [[e x in c2]]-
+[[e in cf]]- := cf(e)
+
+ -- No auxiliary mins.
+[[e in x:c1 -> c2]]v
+  := forall x. [[x in c1]](dual v) -> [[e x in c2]]v
 \end{verbatim}
@@
 
-We give in figure~\ref{ctrans} the translation of contract satisfaction.
-\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
-c$. We have to do that because one is not the negation of the other,
-even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{
-
-The contract translation, $\ctrans{}$, introduces a new predicate,
-$\min{}$, and contract non-satisfaction, $f \notin c$.  The formula
-$\min{e}$, short for ``(a)m in(terested) in $e$'' (?), is used to
-guide the proof search, by restricting proofs to follow the evaluation
-order when instantiating hypotheses (Is this meaningful? Is it even
-correct?).  To understand the translation intuitively, simply ignore
-all $\min{}$'s (I.e., assume $\forall x. \min{x}$).  The
-non-satisfaction form, $f \notin c$, is introduced to make the
-$\min{}$'s appear positively in translations in negative/contravariant
-positions; Ignoring $\min{}$'s, we have $\ctrans{f \notin c} \equiv
-\lnot \ctrans{f \in c}.$ (Well, after I change the definitions of
-$\ctrans{}$ for $e \in \{x | p\}$ and $e \notin \{x|p\}$ \ldots)
-
-Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
-$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
-  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
-\to \cfc$.  If we define 
-\protect \[ 
-\ctrans{g \in x:c1 \to c2} :=
-\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
-\]
-then we get stuck, because we must prove $\min{Z}$ to satisfy
-$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
-definition).  (UPDATE: Actually, this is probably just more evidence
-that $\ctrans{e \in \cfc}$ is defined incorrectly: we were already
-worried about that $\min{}$, because it stops us from proving
-$\cf{Z}$, even when no arrows are involved).  On the other hand, if we
-define the arrow case as in Figure~\ref{ctrans}, namely
-\protect \[
-\ctrans{g \in x:c1 \to c2} 
-:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
-\]
-then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
-$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$. UPDATE: when you
-notice that that implies $\min{Z} \to \cf{Z}$, (and the reverse follows from EM), the next 
-paragraph might become a little dubious.  I still think the second form 
-is more intuitive.).
-
-The irrelevance of $\lnot \min{Z}$ above leads to the question, can we
-do away with the $\min{}$'s altogether in such positions?  It turns
-out the answer is ``No'', when we consider proving $g ~(g ~Z) \in
-\cfc$, but I have some ideas for an alternative translation where the
-answer is ``Yes''.  But first $g ~(g ~Z)$: to use $g$'s contract
-assumption we need to prove $\lnot \min{g ~Z} \lor \cf{g ~Z}$, and to
-prove $\cf{g ~Z}$ we in turn need to use $g$'s contract again.  To use
-$g$'s contract again we need $\min{g ~Z}$, and so $\min{}$ is relevant
-again.  The solution is to invoke excluded middle on $\min{g ~Z}$:
-either $\min{g ~Z}$, and we're done, or $\lnot \min{g ~Z}$, in which
-case we satisfy $g$'s preconditions \emph{without} showing $g$'s
-argument to be crash free.  This seems a little surprising to me,
-making a translation with no $\min{}$ here even more appealing.
-
-Here's a start at such a translation.  I
-think it's more natural to think about variance, so write
-$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
-a system that works for these $\cfc$ examples is
-\protect \begin{eqnarray*}
-\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
-\ctrans{x \in \cfc}^- &:=& \cf{x}\\
-\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
-  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
-\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
-  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
-\end{eqnarray*}
 Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
@@
 So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
-This time the proof proceeds more naturally (no EM).  
 From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
-$\min{g ~Z} \to \cf{g ~Z}$ (UPDATE: this is actually the same as 
-before, up to using EM to rewrite the implication).
+$\min{g ~Z} \to \cf{g ~Z}$
 But that follows by instantiating $g$'s contract again, and 
@@
 
-Notice the assymetry between the positive and negative arrow
-translations: the positive case has no $\min{f ~x}$ assumption.  I
-left this $\min{}$ out because it's already in the 
-conclusion of the $\cf$ example, but I certainly haven't
-worked enough examples to justify it.  Moreover, this fragment
-(no $e \in \{x|p\}$ yet) already fails to have a standard
+Going back to the current translation, we consider some of its
+metatheoretic properties.
+\protect \begin{enumerate}
+\item A standard
   property of bidirectional systems: that you can check/verify what you assume/use.
-In other words, we can't prove that, forall $e$ and $c$,
+  In other words, forall $e$ and $c$,
   \protect \[
@@
   \]
-The trouble seems to come down to needing to prove $\min{e ~x}$ implies 
-$\min{e}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
-assumption to the positive arrow translation to recover the
- ``negative implies positive'' property. (UPDATE: the symmetric $\min{}$
-may also be necessary when reasoning about functions that return functions: if
-$f ~x = g$ then $f ~x ~y = g ~y$, but we'd need a $\min{f ~x}$ to unfold $f$'s def).
-
-Going back to the current translation, we consider some of its
-metatheoretic properties.
-\protect \begin{enumerate}
-\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
-  translation, in the sense that $\ctrans{e \in c}$ should imply
-  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
-  predicate case so that the implication holds (e.g. by more accurately
-  modeling the operation semantics of contracts), and add a symmetric 
-  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
-  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
-  not imply $\ctrans{x \in \cfc}$.
 \item The ``fancy'' $\min{}$-ified translation should 
   degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
-  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
+  This is trivially true if we define the ``naive'' translatino as the ``min'' translation
   plus that axiom.  In symbols, we have
@@
   evaulation strategy, and it's far from obvious that all proofs can proceed this way.
-
 \end{enumerate}
 
-Concerns about the current translation
-\protect \begin{enumerate}
-\item The changes needed to make the first property above hold (a $\min{}$ in the negative
-  arrow case and symmetric predicate contract cases).
-  I had trouble proving
-  \protect \[
-  (\lambda x. \text{case $x$ of $\true \to \true$; $\false \to \bad$})
-  \in \{x|x==\true\} \to \cfc
-  \]
-  , after adding $p = \unr$ to $\ctrans{x \in \{x|p\}}$, but I need to think about it more.
-
-\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
-  it's impossble to prove $\ctrans{\true \in \cfc}$, because you can't prove
-  $\min{\true}$ a priori.  On the other hand, we don't need any extra
-  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
-  very simple examples (checked on the simple
-  example that $id \in \cfc \to \cfc$). (UPDATE: removing the $\min{x} \land$ in $cfc$, 
-  or changing it to $\min{x} \to$, is probably the solution).
-
-\item A related question, is what do we feed to Equinox?  Equinox is a 
-  refutation prover, so to prove $\Gamma \vdash \phi$ we ask if 
-  $\Gamma,\lnot \phi \vdash \bot$.  So, for us, when $\phi$ is $\ctrans{e \in c}$, do
-  give Equinox $\Gamma,\lnot \ctrans{e \in c}$ or $\Gamma, \ctrans{e \notin c}$.  I think
-  we want the former, since ignoring Equinox our goal is $\Gamma \vdash \ctrans{e \in c}$.
-\end{enumerate}
-}
-
-\note[nc]{Here's a more fleshed out alternate translation, and a
-  comparison with the current translation.}
-\small
-\begin{verbatim}
-new translation
-===============
-
--- you can assume both min()s when proving e in {x|p} 
-[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
--- you need to prove min(e) but then get min(p[e/x])
-[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
--- could also write this as 'e /= unr -> p[e/x] in {true,unr}'
-[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}
-
-[[e in cf]]+ := min(e) -> cf(e)
--- i think we can add a 'min(e) ->' here without losing anything.
-[[e in cf]]- :=           cf(e)
-
--- should add a 'min(e x) ->' here as discussed above.
-[[e in x:c1 -> c2]]+
-  := forall x.             [[x in c1]]- -> [[e x in c2]]+   
-[[e in x:c1 -> c2]]-
-  := forall x. min(e x) -> [[x in c1]]+ -> [[e x in c2]]-
-
-old translation (written in implication style)
-==============================================
-
-[[e in {x|p}]]       
-  := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
-not[[e notin {x|p}]] 
-  := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)
-
-[[e in cf]]       := min(e) /\ cf(e)
-not[[e notin cf]] := min(e) -> cf(e)
-
-[[e in x:c1 -> c2]] 
-  := forall x. min(e x) -> not[[x notin c1]] -> [[e x in c2]]
-not[[e notin x:c1 -> c2]]
-  := forall x.             [[x in c1]] -> not[[e x notin c2]]
-
-updated cf and {x|p} rules for old translation
-==============================================
-
-[[e in {x|p}]]       
-  := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
-not[[e notin {x|p}]] 
-  := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})
-
--- what we actually discussed was 
--- adding a top level min(e) assumption, but
--- removing the min(e) goal has a similar effect.
-[[e in cf]]       :=           cf(e)
-not[[e notin cf]] := min(e) -> cf(e)
-\end{verbatim}
-\note[nc]{And what do we see?
-If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
-then the ``new'' translation looks almost the same,
-except it has more $\min{}$'s in the predicate rules.  So, have we learned
-anything? Probably:
-\protect \begin{enumerate}
-  \item We're more confident about our suggested changes to the old translation,
-        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
-  \item The new translation is more intuitive / we might understand how Koen came up
-        with the old translation.
-  \item We can answer the question about what we give as input to Equinox: for the
-        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
-        That is consistent with giving Equinox $\in$ axioms.  
-        What to give as the goal is maybe
-        less obvious, since Equinox is a refutation prover, but consider: ignoring the
-        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
+  the theory $T$ should consist of $\in^-$ axioms.
+  We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
   But
   that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.
-        And going back
-        to the old translation, the analogy is 
-        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
-\end{enumerate}
-
-The predicate case for the new translation came from an example:
-}
-\begin{verbatim}
-try:
-
-[[e in {x|p} ]]+ = min(e) -> min(p[e/x]) -> [[e in {x|p}]]-
-[[e in {x|p} ]]- = e = unr \/ p[e/x] in {true, unr}
-
-example:
-
-e := (\x.case x of True -> True) -- so e false = bad
-c := {x|x==True} -> CF
-[[e in c]]+ = forall x. min(e x) 
-              -> [[x in {x|x==True}]]- -> [[e x in CF]]+
-            = forall x. min(e x) 
-              -> (x = unr \/ p in {true,unr})
-              -> (min(e x) -> CF(e x))
-
-min(e x) -> min(x) /\ (
-         \/ (x = bad /\ (e x) = bad) 
-         \/ (x = True /\ (e x) = true) 
-         \/ (x = False /\ (e x) = bad)
- -- better to add the redundant x /= bad, true, false here?
-         \/ (e x) = unr
-         )
-
-if x = unr then (e x) = unr and so CF(unr) and done
-
-if x /= unr /\ p = true then we're stuck, since we can't 
-unroll the definition of ==true in p
-
-so, change the defs:
-
-[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
-[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
-[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}
-
-starting again, we have min(x) from D[[]] so we get
-
-min(p) /\ (x = unr \/ p in {true,unr})
-
-now
-
- - x = unr gives (e x) = unr as before
-
- - for x/=unr and p in {true,unr} we unroll def (==true) 
-   using min(p), for p = ((==True) x):
-
-   min((==True) x) -> min(x) /\ ( -- e' is (==True) here
-         \/ (x = bad /\ (e' x) = bad) 
-         \/ (x = True /\ (e' x) = true) 
-         \/ (x = False /\ (e' x) = bad)
-         \/ (e' x) = unr
-         )
-
-for p = unr, we get x notin {bad,true,false}, 
-and hence (e x) = unr, and we're done.
-
-for p = true, we get x = true, 
-and hence (e x) = true, and we're done.
-\end{verbatim}
-\note[nc]{That example also has an alternate version of the
- definition translation for functions defined by case
-matching.  It's a little more concise than the old translation.
-}
+
 \begin{verbatim}

### 2011-12-20 paper/paper.tex
Paper: cleaned up / removed many things, started removing the 'a' let-binding.

@@
 
+  The following argument has major problems:\begin{enumerate}
+  \item it's not clear what ``terminating terms'' means, since our
+    language is lazy.
+  \item consistency of $\trans{}^+$ (checking) does not imply
+    consistency of the $\trans{}^-$ (assuming), e.g. because the function
+    definitions change (we only use anonymous recursive functions in
+    $\trans{}^+$).  Do we need to argue that our induction is sound? We
+    can't simply define assumed ($\trans{}^-$) functionss as $\unr$,
+    since e.g. $\unr \neq S ~x$ for all $x$.
+  \item the presentation of $\cfc$ is wrong: asking if something is
+    equal to $\bad$ only works in a strict language.
+  \end{enumerate}
+
   The carrier consists of all closed (i.e. variable free / grounded)
@@
 
-\note[nc]{
-Not true that we only consider full application of functions: the definition of satisfaction for arrow contracts
+  \note[nc]{ UPDATE: actually, this is true now, but we still want to
+    talk about collapsing arrows, because the implementation does
+    that?
+
+Not true that we only consider full application of functions: 
+the definition of satisfaction for arrow contracts
 $x:c_1 \to c_2$ already includes partial applications.  We could add a
@@
 
+\subsection{BAD and UNR}
+There are two types of problematic expressions: those that cannot
+happen during a run of the program and those that can.  We call
+infinite loops, and expressions of the first type, such as type
+errors, \emph{unreachable} and equate them to the special value $\unr$
+in our FOL theory.\footnote{
+%
+In particular, ill-typed expressions are $\unr$, not $\bad$
+%
+} We call expressions of the second type \emph{bad} and equate them to
+the special value $\bad$ in our FOL theory.
+
+Although we only consider well-typed programs, 
+our FOL theory is not typed,\footnote{
+%
+  We could encode types with predicates.  Will Sonnex (Zeno) did this 
+  when encoding examples in ACL2s, in order to improve performance.  E.g.
+  \[
+  \forall x,y. isNat(x) \land isNat(y) \to isNat(add(x,y)) \land \text{ standard stuff}.
+  \]
+  Things might become more complicated for higher order functions, e.g.
+  \begin{align*}
+  \forall f,xs,a,b.& isType(a) \land isType(b) \\
+                   &\land hasType(f,arrowType(a,b)) \\
+                   &\land hasType(xs,listType(a))\\
+                   &\to hasType(map(f,xs),listType(b)) \land \text{ standard stuff}.
+  \end{align*}
+%
+}
+and so quantifiers in FOL range over \emph{all}\footnote{
+%
+  Well, technically, in \hprime we have both $f$ and $\full{f}$ for
+  each declared function $f$, and the quantifier only includes $f$.
+%
+} expressions in \hprime.
+This is OK, because the axioms are designed so that scrutinizing an
+ill-typed constructor-head results in $\unr$.  On the other hand,
+some ill-typed expressions are not $\unr$, e.g. we can prove
+that $tail~(Cons~Nil~Z) = Z$.\footnote{
+%
+  Basically, if we discover that an expression is ill-typed, then we return $\unr$, but
+  we do this lazily.
+%
+}
+
 \subsection{Contracts}
 Contract syntax is described in figure \ref{cont-stx}. The predicates
-$p$ in contracts range over boolean $\hprime$ expressions. We
+$p$ in predicate contracts range over boolean $\hprime$ expressions. We
 only consider pairs of contract for simplicity, although there is no
@@
 
-\subsection{BAD and UNR}
-Consider the following piece of code:
-\begin{verbatim}
-a = 0 + True
-
-b ::: CF
-b = undefined
-
-c = error "foo"
-\end{verbatim}
-
-\begin{itemize}
-\item $a$ is ill-typed
-\item $b$'s implementation is not correct wrt its contract
-\item $c$ goes through the whole toolchain (compiler, typechecker, contractchecker)
-\end{itemize}
-
-One thing to notice is that $a$ and $b$ are things that ``should not
-happen'' but are caught statically whereas $c$ should not happen but
-can only be dealt with dynamically.\note[nc]{
-I find this example confusing: \texttt{error} calls should be just as
-easy to find as \texttt{undefined} calls statically.}
-
-We can now define two types of problematic expressions: those that
-cannot happen during a run of the program and those that can.
-Expressions of the first type are called unreachable (and equated to
-the special value $\unr$ in our first-order theory)\note[nc]{
-%
-In particular, ill-typed expressions are $\unr$, not $\bad$
-%
-}, whilst
-expressions of the latter type are called bad (and equated to the
-special value $\bad$).
-
-We said earlier that we only considered syntactically correct and
-well-typed programs as input. That implies that the ``$a$'' case
-cannot happen. But given that our first-order logic is not typed, the
-theorem prover may decide to instanciate a variable with an ill-typed
-value! \change[nc]{In order to prevent this, we will need to encode some basic
-type-checking mecanism directly in our first-order theory.}{This is OK, because the axioms are designed so that scrutinizing an ill-typed constructor-head results in $\unr$.
-On the other hand, some ill-typed expressions are not $\unr$, e.g.
-we can prove that $tail~(Cons~Nil~Z) = Z$.}
-
 \subsection{First-order logic with equality}
-We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e.
-forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
-\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} 
+We use first-order logic with equality, defined in figure \ref{fol-stx}.\footnote{
+%
+I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e.
+forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.
+%
+}
 The expression language in our logic is the expression language of 
@@
 \subsection{$\dtrans{}$ -- Definitions}
-We give in figure~\ref{dtrans} the two translations of function
-definitions.
-
-figure~\ref{dtrans} gives the translation of function not defined by
-pattern matching, which is really easy: we just have to state the
-equality between le left-hand side and the right-hand side.
-
-Translating definitions that use pattern-matching is more challenging
-and is described in figure~\ref{dtrans}.  The first line says that
-when applied to an argument that matches a pattern of the case
-expression, we should equate the function call to the corresponding
-expression. The second line states that if the pattern-matching failed
-or if we pattern-matched on $\bad$ then the result should be
-$\unr$.\note[nc]{This text does not agree with the figure.  In
-  particular, the axiom says nothing about the result of scrutinizing
-  $\bad$.  I think the correct behavior is that scrutinizing $\bad$
-  results in $\bad$, \emph{not} $\unr$.}
-
+We give in figure~\ref{dtrans} the translation of function
+definitions. The first line says that when applied to an argument that
+matches a pattern of the case expression, we should equate the
+function call to the corresponding expression. The second line states
+that if the pattern-matching failed or if we pattern-matched on $\bad$
+then the result should be $\unr$.\note[nc]{This text does not agree
+  with the figure.  In particular, the axiom says nothing about the
+  result of scrutinizing $\bad$.  I think the correct behavior is that
+  scrutinizing $\bad$ results in $\bad$, \emph{not} $\unr$.}
+
+\newcommand{\fullfxs}{\full{f}(\vec{x})}
 \begin{figure*}
@@
   \dtrans{f \in c} &=& \ctrans{f \in c}\\
-  \dtrans{f~\vec{x} = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
-  \full{f}(\vec{x})) \to a = e \\
+  \dtrans{f~\vec{x} = e} &=& \forall \vec{x}.\min{\fullfxs} 
+                                              \to \fullfxs = e \\
    \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
-     \forall \vec{x},a.~(\mbox{min}(a) \land a = \full{f}(\vec{x})) &\to&
-     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
-       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
-     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& a = e_1 \\
+     \forall \vec{x}. \min{\fullfxs} &\to&
+     \multicolumn{3}{c}{  \min{e} \land (e = \texttt{BAD} \lor
+       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor \fullfxs = \texttt{UNR})} \\
+     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& \fullfxs = e_1 \\
      &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& a = e_n \\
+     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& \fullfxs = e_n \\
    \end{array} 
@@
   \item $\gets$: if we pack crash-free values in a data constructor,
-    the resulting value is crash-free.  For $K$ has arity $n$
-    we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.
-
-  \item $\to$: a value $t$ of type $T$\note[nc]{
-    ``type'' is used loosely here: it means the constructor head is of type $T$.
+    the resulting value is crash-free.  If $K$ has arity $n$
+    then we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.
+
+  \item $\to$: a value $t$ of type $T$\footnote{
+    ``type'' is used loosely here: it means the constructor 
+      head is from the type $T$.
     } is crash-free implies
@@
   Note that this is not true for functions: a function is not required
-  to use all of its arguments. \texttt{fst} is crash-free if and only
+  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
   if the first argument of the pair is crash-free. The second argument
@@
 
-\item $(\phi_\text{lazy})$ No full application of a constructor $K_i$ is unreachable or crashing.
-\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
-  $\full{K}_i(\vec{x}) \neq \bad$. It is already implied by the fact that
-  $\left(\forall j. \cf{x_j}\right) \to \cf{\full{K}_i(\vec{x})}$.
+\item $(\phi_\text{lazy})$ No full application of a constructor
+  $K_i$ is unreachable or crashing.
 \end{itemize}

### 2011-12-21 paper/paper.tex
Paper: improved description of phi_injective.

@@
 \subsection{$\ttrans{}$ -- Datatypes}
-We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
+We break down the translation for datatypes in four parts, described in Figure~\ref{ktrans}
 $$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
@@
 \begin{itemize}
-\item $(\phi_\text{injective})$ Constructors are injective.  That is, for
-  each term constructor $K$,
-  if $K e_1 \cdots e_n = K e'_1 \cdots e'_n$, then $e_i = e'_i$ for
-  all $i$.  We axiomitize this via projections $\proj{i}{K}$: for each $i$,
-  $\proj{i}{K}(K e_1 \cdots e_n) = e_i$.
-\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
+\item $(\phi_\text{injective})$ Constructors are injective.  That is,
+  for each term constructor $K$, if $K e_1 \cdots e_n = K e'_1 \cdots
+  e'_n$, then $e_i = e'_i$ for all $i$.  We axiomitize this via
+  projections $\proj{i}{K}$: for each $i$, $\proj{i}{K}(K e_1 \cdots
+  e_n) = e_i$.  NB: unlike for user defined functions, the $\min{}$
+  constraint on unfolding projections is on the \emph{argument}, and
+  not the \emph{function application} itself.  I.e., instead of
+  $\min{\proj{i}{K}(K e_1 \cdots e_n)}$ we have $\min{K e_1 \cdots
+    e_n}$.  We guard with the latter, not the former, because the
+  primary purpose of projectors is to axiomitize the injectivity of
+  constructors.  E.g., given $S~x_1 = S~x_2$, and $\min{S~x_1}$, we
+  want to conclude that $x_1 = x_2$.  We won't, in general, have
+  $\min{\proj{1}{S}(S~x_1)}$ here, so such a min-constraint would
+  block us.
+
+  We also use projectors in some versions of the case-expression
+  translation, to Skolemize, but this is just a happy coincidence.  In
+  particular, user programs will never contain explicit projections.
+  If a user wants to project they need to write a projection function
+  using pattern matching.
+\item $(\phi_\text{disjoint})$ Distinct term constructors have
+  disjoint ranges.  Note that we don't axiomitize disjointness of term
+  constructors from distinct \emph{data types}.
 \item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:

### 2011-12-21 paper/paper.tex
Paper: updated the ttrans figure.

@@
 
-  We also use projectors in some versions of the case-expression
+  We also use projections in some versions of the case-expression
   translation, to Skolemize, but this is just a happy coincidence.  In
-  particular, user programs will never contain explicit projections.
-  If a user wants to project they need to write a projection function
-  using pattern matching.
+  particular, user programs will never contain projections.  If a user
+  wants to project then they need to write there own projection
+  function using pattern matching.
 \item $(\phi_\text{disjoint})$ Distinct term constructors have
@@
 
-  \item $\to$: a value $t$ of type $T$\footnote{
+  \item $\to$: if a value $t$ of type $T$\footnote{
     ``type'' is used loosely here: it means the constructor 
       head is from the type $T$.
-    } is crash-free implies
-    that every value packed in it is crash-free, since we can use
+    } is crash-free then
+    every value packed in it is crash-free, since we can use
     a case expression to extract any packed value.
@@
 
-  Note that this is not true for functions: a function is not required
+  Note that this is not true for arbitrary functions: a function is not required
   to use all of its arguments. For example, \texttt{fst} is crash-free if and only
@@
   \begin{center}
-    \[  \begin{array}{rcl}
-      \phi_\text{injective} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(a) \text{\note[nc]{ $\min{a}$, not $\min{\proj{j}{K_i}(a)}$.}}\\
-      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
-      \phi_\text{cf} &=& 
-        \bigwedge_{1 \leq i \leq n}~\ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
-          \land \forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) 
-                  \to \cf{a} \to (\bigwedge_{1 \leq j \leq a_i} \cf{x_j}) \\
-      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
-    \end{array} \]
+    \newcommand{\fullkxs}[1]{\full{K}_#1(\vec{x})}
+    \begin{align*}
+      \phi_\text{injective} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
+      \phi_\text{disjoint} &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
+      \phi_\text{cf} &= 
+        \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
+          \land \forall \vec{x}.~\min{\fullkxs i}
+                  \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
+      \phi_\text{lazy} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
+    \end{align*}
   \end{center}
-  \caption{$\ttrans{}$ -- Data type translation\note[nc]{Need
-      $\min{a}$ to be sufficient for unfolding in
-      $\phi_\text{injective}$ to get invertability of constructors!
-      I thought we needed $\min{\proj{j}{K_i}(a)}$ to be sufficient when
-      we use a projection based translation of case expressions, but
-      as Koen pointed out this is not true.
-      }}
+  \caption{$\ttrans{}$ -- Data type translation}
   \label{ktrans}

### 2011-12-21 src/Translation.hs
Added 'min's in negative 'CF' translation.
This gives a significant performance increase for the
'yes/add-symmetric.hs' example: 48s to 30s!

@@ -312,11 +312,13 @@ cTrans' v e (H.Or c1 c2) = do
 cTrans' v e (H.CF) = do
   eT <- eTrans e
   -- XXX, DESIGN CHOICE: the 'min' constraint should be optional in
-  -- the 'Minus' case.
+  -- the 'Minus' case.  However, the 'min' version is significantly
+  -- faster for the 'yes/add-symmetric.hs' example: full check goes
+  -- from 48s without 'min' to 30s with 'min'.
   return $ case v of
     Plus  -> F.Min eT :=>: F.CF eT
-    --Minus -> F.Min eT :=>: F.CF eT
-    Minus -> F.CF eT
+    Minus -> F.Min eT :=>: F.CF eT
+    --Minus -> F.CF eT
 
 -- Data decl translation
 ------------------------
@@ -394,9 +396,26 @@ phi_cf (H.Data t dns) = mapM f dns where
        -- translation, by treating the constructor as a function
        -- (which, of course, it is).
     cfc <- F.appifyF =<< cTrans' Minus cN contract
+    -- The 'min' here is essential: w/o it the 'lem_eqNat_trans'
+    -- is not provable (after 4 minutes), and Paradox does not
+    -- find counter models to 'lem_add_symmetric', with
+    -- 'lem_eqNat_trans ::: CF^3 -> CF' removed.  I waited up
+    -- to size 9 counter model attempts.
+    --
+    -- This is interesting: I've been arguing that we don't need
+    -- to guard the conclusion of 'CF', since 'CF' doesn't let us
+    -- unfold defs.  But here, in 'phi', we only conclude 'CF',
+    -- and yet it causes problems when unguarded.
     let phi = F.Forall xs $ F.Min full :=>:
                        (F.CF full
                        :=>: (F.And [F.CF x | x <- xsN]))
+    -- This 'phi'' is more conservative than 'phi', but it also
+    -- performs worse on the 'yes/add-symmetric.hs', by two to four
+    -- seconds. I guess it's possible it could perform *better* on
+    -- other tests???
+    let phi' = F.Forall xs $ F.Min full :=>:
+                       (F.CF full
+                       :=>: (F.And [F.Min x :=>: F.CF x | x <- xsN]))
     return $ F.LabeledFormula (F.axiom $ "phi_cf__"++t++"__"++c) $ F.And [phi,cfc]
 -- Axiom: Term constructors are total/lazy (Phi_4 in paper).
 phi_lazy (H.Data t dns) = map f dns where

### 2011-12-21 paper/paper.tex
Paper: updating phi_cf section per experiments.

@@
   \item $\gets$: if we pack crash-free values in a data constructor,
-    the resulting value is crash-free.  If $K$ has arity $n$
-    then we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.
-
-  \item $\to$: if a value $t$ of type $T$\footnote{
-    ``type'' is used loosely here: it means the constructor 
-      head is from the type $T$.
-    } is crash-free then
-    every value packed in it is crash-free, since we can use
-    a case expression to extract any packed value.
-    \note[nc]{XXX: This translation is ad-hoc.}
-
-  Note that this is not true for arbitrary functions: a function is not required
+    the resulting value is crash-free.  If $K$ has arity $n$ then we
+    assert $\ctrans{K \in \cfc^n \to \cfc}^-$.  I.e., we simply treat
+    K as a strict \cfc function.
+
+  \item $\to$: if a value $t$ of type $T$\footnote{ ``type'' is used
+      loosely here: it means the constructor head is from the type
+      $T$.  } is crash-free then every value packed in it is
+    crash-free, since we can use a case expression to extract any
+    packed value.  More precisely, we assert
+    \[
+    \forall \vec{x}. \min{K~\vec{x}} \land \cf{K~\vec{x}}
+    \to \bigwedge_{1 \le i \le n} \cf{x_i}.
+    \]
+    Unlike the $\gets$-direction of the translation, where we use
+    $\ctrans{}$ (principled), this direction of the translation is
+    ad-hoc.  Interestingly, dropping the $\min$ here causes
+    problems. The \texttt{lem\_eqNat\_trans} in
+    \texttt{yes/add-symmetric.hs} becomes unprovable (or at least,
+    really slow), and Paradox is no longer able to find small
+    counter-models to \texttt{lem\_add\_symmetric} when
+    \texttt{lem\_eqNat\_trans}'s $\cfc^3 \to \cfc$ contract is removed.
+
+  Note that $\phi_\text{cf}$ is not true for arbitrary functions: a function is not required
   to use all of its arguments. For example, \texttt{fst} is crash-free if and only
@@
   \begin{center}
-    \newcommand{\fullkxs}[1]{\full{K}_#1(\vec{x})}
+    \newcommand{\fullkxs}[1]{\full{K}_#1~\vec{x}}
     \begin{align*}

### 2011-12-21 paper/paper.tex
Paper: updated '[| e ::: CF |]v' per experiments.

@@
 
- -- Need 'min(e)' here to unfold defs.
-[[e in cf]]+ := min(e) ->  cf(e)
-[[e in cf]]- := cf(e)
+-- Need 'min(e)' here in the (+)-case to unfold defs.
+-- The 'min(e)' in the (-)-case doesn't seem strictly
+-- necessary, but it gives a big performance improvement for
+-- 'yes/add-symmetric.hs', from 48s to 30s for a full check.
+[[e in cf]]v := min(e) ->  cf(e)
 

### 2011-12-22 paper/paper.tex
Paper: high level description of 'min' and many smaller edits.
Still need to motivate the particular axioms, and explain the proof/use
distinction in contract translation.

@@
 
+\newcommand{\phiLazy}{\phi_\text{lazy}}
+\newcommand{\phiCF}{\phi_\text{CF}}
+\newcommand{\phiInjective}{\phi_\text{injective}}
+\newcommand{\phiDisjoint}{\phi_\text{disjoint}}
+
 \newcommand{\designChoice}{(DESIGN CHOICE/OPTIONAL)}
@@
 Nil \neq \unr 
-\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phi_\text{lazy}$) say that constructors are ``lazy'' or ``total''.}
+\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phiLazy$) say that constructors are ``lazy'' or ``total''.}
 
@@
   doesn't seem to play any role in formal part of this paper (besides
-  being specified in axioms $\phi_\text{injective}$). In the current
+  being specified in axioms $\phiInjective$). In the current
   implementation it's used to avoid introducing existentially
@@
 Why not the simpler \[\forall y,ys.~\proj{2}{Cons}~Cons~y~ys = ys\]?
-Same question in Figure \ref{dtrans}.
+Same question in Figure \ref{fig:dtrans}.
 (Equinox does have a built in notion of equality, e.g.
@@
 The expression language in our logic is the expression language of 
-\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
+\hprime (Figure~\ref{fig:hprime-syntax}).\footnote{
+  %
+  Note that this does \emph{not} include \texttt{case} expressions.
+  %
+}  Note that curried application in \hprime
 corresponds to a binary function symbol in FOL, which we denote by juxtaposition.
@@
 \section{Translations}
-For an overview of the different translations we define, see figure~\ref{trans}
+For an overview of the different translations we define, see
+Figure~\ref{fig:all-trans}.
 
@@
   \caption{Translations}
-  \label{trans}
+  \label{fig:all-trans}
 \end{figure}
+
+\subsection{The $\min$ predicate}
+
+We introduce a predicate $\min$, short for ``am interested'', and use
+it both to guide proof search and guarantee (?) the existence of
+finite models.  Dimitrios says that such use of a predicate, at least
+the guiding of proof search part, is related to the concept of
+``triggers''.  We are particularly interested in finite
+\emph{counter}-models, which exist when the contracts aren't provable
+(i.e. when our refutation theory is counter satisfiable).
+
+The $\min$ predicate guides proof search by restricting the use of
+axioms. Axioms are guarded by $\min$ preconditions, and in some cases
+introduce new $\min$s.  The rough idea is that $\min$ corresponds to
+evaluation: you have $\min(e)$ when you force the value of $e$ in our
+lazy language.  For example, the axioms for function definitions in
+Figure~\ref{fig:dtrans} require interest $\min{f x}$ to unfold the
+definition of $f$, but then introduce $\min{e}$ if $f$ is defined by
+\texttt{case}-matching $e$, because \texttt{case}-matching forces the
+scrutinee.
+
+The $\min$ predicate guarantees finite models by preventing us from
+proving all models are infinite (recall the completeness theorem for
+FOL).  Without $\min$, the axiomitization of any infinite datatype $T$
+would allow us to prove $T$, and hence the domain of our models, was
+infinite.  Namely, we axiomitize the injectivity and disjointness of
+$T$'s term constructors, and without $\min$ these axioms prove that
+inductive term constructors are injective, but not surjective, and
+hence the domain is infinite.  However, we guard these axioms, namely
+$\phiInjective$ and $\phiDisjoint$ in Figure~\ref{fig:dtrans}, with
+$\min$ predicates, to restrict their use to interesting constructor
+applications.
 
@@
 \subsection{$\dtrans{}$ -- Definitions}
-We give in figure~\ref{dtrans} the translation of function
+We give in figure~\ref{fig:dtrans} the translation of function
 definitions. The first line says that when applied to an argument that
@@
                                               \to \fullfxs = e \\
-   \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
-     \forall \vec{x}. \min{\fullfxs} &\to&
-     \multicolumn{3}{c}{  \min{e} \land (e = \texttt{BAD} \lor
-       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor \fullfxs = \texttt{UNR})} \\
-     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& \fullfxs = e_1 \\
-     &\land& & \cdots & \\
-     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& \fullfxs = e_n \\
+   \dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}
+    &=&
+    \begin{array}{ll}
+      \forall \vec{x}. \min{\fullfxs} 
+      &\to \min{e} \\
+      &\land \left(
+        \bigwedge_i \left(
+                      \forall \vec{x_i}.~e = \full{K}_i(\vec{x_i}) \to \fullfxs = e_i
+                    \right)
+             \right) \\
+      &\land \left( e = \bad \to \fullfxs = \bad \right) \\
+      &\land \left( e = \bad \lor 
+                    \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i}) \right) 
+                    \lor \fullfxs = \unr
+             \right) \\
     \end{array} 
@@
 \]
-\caption{$\dtrans{}$ -- Defintions translation}
- \label{dtrans}
- \note[nc]{??? Why do we introduce the extra quantified variable $a$,
-   representing the result (``answer''?), instead of just using the result directly
-   (which is logically equivalent)? Same question for the next
-   Figure.  Does this make a difference to Equinox?}
+\caption{$\dtrans{}$ -- Definition translation}
+ \label{fig:dtrans}
 \end{figure*}
@@
 We break down the translation for datatypes in four parts, described in Figure~\ref{ktrans}
-$$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+$$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phiInjective \land \phiDisjoint \land \phiCF \land \phiLazy$$
 
 \begin{itemize}
-\item $(\phi_\text{injective})$ Constructors are injective.  That is,
+\item $(\phiInjective)$ Constructors are injective.  That is,
   for each term constructor $K$, if $K e_1 \cdots e_n = K e'_1 \cdots
@@
   function using pattern matching.
-\item $(\phi_\text{disjoint})$ Distinct term constructors have
+\item $(\phiDisjoint)$ Distinct term constructors have
   disjoint ranges.  Note that we don't axiomitize disjointness of term
   constructors from distinct \emph{data types}.
-\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
+\item $(\phiCF)$ Then, we have to give crash-freeness conditions for each $K_i$:
   Notice that we have a equivalence.
@@
 
-  Note that $\phi_\text{cf}$ is not true for arbitrary functions: a function is not required
+  Note that $\phiCF$ is not true for arbitrary functions: a function is not required
   to use all of its arguments. For example, \texttt{fst} is crash-free if and only
@@
 
-\item $(\phi_\text{lazy})$ No full application of a constructor
+\item $(\phiLazy)$ No full application of a constructor
   $K_i$ is unreachable or crashing.
@@
   $$ \ruleform{\ttrans{data~def} = \Phi} $$
-$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
+$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phiInjective \land \phiDisjoint \land \phiCF \land \phiLazy$$
  \hspace{5.6cm}where
@@
     \begin{align*}
-      \phi_\text{injective} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
-      \phi_\text{disjoint} &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
-      \phi_\text{cf} &= 
+      \phiInjective &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
+      \phiDisjoint &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
+      \phiCF &= 
         \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
@@
                   \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
-      \phi_\text{lazy} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
+      \phiLazy &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
     \end{align*}
@@
 As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
-in Figure~\ref{dtrans}.  Definition translations are axioms, 
+in Figure~\ref{fig:dtrans}.  Definition translations are axioms, 
 so the the initial $\min{\full{f}(\vec{x})}$
@@
 \begin{verbatim}
-[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
-[[e in cf]]         := cf(e)
-[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
+[|e in {x|p}|]      := e /= unr -> p[e/x] in {true, unr}
+[|e in cf|]         := cf(e)
+[|e in x:c1 -> c2|] := forall x. [|x in c1|] -> [|e x in c2|]
 \end{verbatim}
@@
 -- Need 'min(e)' and 'min(p[e/x])' here to unfold defs.
-[| e ::: {x|p} |]+ = min(e) /\ min(p[e/x]) ->
+[| e in {x|p} |]+ = min(e) /\ min(p[e/x]) ->
                       (e /= unr -> p[e/x] not-in {false, bad})
@@
 -- Need 'min(p[e/x])' here to unfold defs.
-[| e ::: {x|p} |]- = min(e) -> min(p[e/x]) /\
+[| e in {x|p} |]- = min(e) -> min(p[e/x]) /\
                       (e /= unr -> p[e/x]     in {true, unr})
@@
 -- 'yes/add-symmetric.hs', from 48s to 30s for a full check.
-[[e in cf]]v := min(e) ->  cf(e)
+[|e in cf|]v := min(e) ->  cf(e)
 
  -- No auxiliary mins.
-[[e in x:c1 -> c2]]v
-  := forall x. [[x in c1]](dual v) -> [[e x in c2]]v
+[|e in x:c1 -> c2|]v
+  := forall x. [|x in c1|](dual v) -> [|e x in c2|]v
 \end{verbatim}
@@
 
-Going back to the current translation, we consider some of its
-metatheoretic properties.
-\protect \begin{enumerate}
+Metatheoretic properties of the translation
+\begin{enumerate}
 \item A standard
@@
   assumptions on $\Gamma$ and $e \in c$,
-  \protect \[
+  \[
     \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
@@
 
-  \item \add[nc]{state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
-This is already present in the implementation, and assumed in the
-consistency proof.}
-
-  \item give formulae for the boolean datatype, equivalent to
-translating
-\begin{verbatim}
-data Bool = True 0 | False 0
-\end{verbatim}
+  \item state that $\unr$ is crash-free: $\cf{\unr}$.
 
@@
 \begin{figure}\label{fig:prelude}
+  The implementation enables
   \begin{itemize}
@@
     \item $\cf{\unr}$
-    \item $\dtrans{data~Bool = \true~0 | \false~0}$
     \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$
+  \end{itemize}
+  and disables
+  \begin{itemize}
     \item \designChoice\\ $\forall x. \min{\bad ~x} \to \bad ~x = \bad$ 
@@
     \item \designChoice\\ $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$
-    \item 
   \end{itemize}

### 2011-12-22 paper/paper.tex
Paper: 'min' and lemmas.
A little more on 'min', and a description of lemmas. The
lemma description is short, and Koen has a much more complete
description in the google doc.

@@
 We introduce a predicate $\min$, short for ``am interested'', and use
-it both to guide proof search and guarantee (?) the existence of
+it both to guide proof search and guarantee (?)\footnote{
+  % 
+  Do we really guarantee finite models?  I think Koen said ``yes'',
+  but I don't see how to prove this.  Namely, I don't see how to prove
+  we never have $S : M \to M$ for $M \subseteq \{ x \mid \min{x} \}$ and
+  $M \neq \emptyset$.
+
+  I think I wrote down an example where we have $\min{Z}$,
+  $\min{S~Z}$, $\min{S~(S~Z)}$, etc, but it seems I'd need an
+  induction axiom to prove that $S$ really took $M$ to $M$, for $M$
+  the naturals.
+  % 
+} the existence of
 finite models.  Dimitrios says that such use of a predicate, at least
@@
 
-The $\min$ predicate guarantees finite models by preventing us from
+The $\min$ predicate guarantees (?) finite models by preventing us from
 proving all models are infinite (recall the completeness theorem for
@@
 applications.
+
+The $\min$ predicate complicates the translation of contracts,
+requiring us to define separate contract translations for goals and
+axioms.  The only actual difference is for predicate-contract, but we
+have considered many minor variations.
 
@@
 
-\item $(\phiLazy)$ No full application of a constructor
-  $K_i$ is unreachable or crashing.
+\item $(\phiLazy)$ No full application of a constructor $K_i$ is
+  unreachable or crashing.  In other words, term constructors don't
+  scrutinize their arguments / are lazy.
 \end{itemize}
@@
 \subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
-\note[nc]{%
 Figure~\ref{fig:newCTrans} contains the $\min$-translation of contracts, and
@@
 plus comes from the negative and positive
-positions in an implication: $- \to +$.  The goal of the $\min$'s twofold: 
+positions in an implication
+\[
+- \to +.
+\]
+The goal of the $\min$'s twofold: 
 (1) to allow finite counter models when the theory 
@@
 we guard it with $\min{e}$.
-}
 
@@
 it's natural to think of contract assumptions in the theory as
-negative, and contract goals as positive.  An example is helpful:
-\protect \begin{eqnarray*}
-\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
-&&\text{whereas}\\
-\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
-\end{eqnarray*}
-Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
+negative, and contract goals as positive. So, the theory
+consists of $\ctrans{}^-$ axioms, so call it $\Gamma^{\in^-}$.
+We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
+But that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$,
+and that is the refutation we feed to the theorem prover.
+
+An example: 
+given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
 want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
@@
 From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
-$\min{g ~Z} \to \cf{g ~Z}$
+$\min{g ~Z} \to \cf{g ~Z}$.
 But that follows by instantiating $g$'s contract again, and 
-then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.
+then using $\phiCF$ for $Z$, which is $\ctrans{Z \in \cfc}^-$, 
+i.e. $\min{Z} \to \cf{Z}$, exactly what we need.
 
@@
 
-  the theory $T$ should consist of $\in^-$ axioms.
-  We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
-  But
-  that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.
-
-\begin{verbatim}
-In any case, another question: 
-does equinox distinguish between D[[]] like
-
-  (x = Nil /\ f x = eNil)
-  \/
-  (exists y,ys. x = Cons(y,ys) /\ f x = eCons)
-
-and D[[]] like
-
-  (x = Nil /\ f x = eNil)
-  \/
-  (x = Cons(cons_1 x,cons_2 x)
-   /\ f x = eCons[cons_1 x/y, cons_2 x/ys])
-
-the latter is nice because we don't need 
-to invent names (consider
-
-  f x = case x of
-          K x -> x
-
-we can't write
-
-  (exists x. x = K x /\ f x = x)
-
-since the two x's equated are not the same!). 
-the former is nice because it's more concise?
-\end{verbatim}
-
 \begin{figure*}
@@
 
+\section{Implementation}
+
+\subsection{Case translations}
+XXX: describe the various case translations here.
+
+\begin{verbatim}
+In any case, another question: 
+does equinox distinguish between D[[]] like
+
+  (x = Nil /\ f x = eNil)
+  \/
+  (exists y,ys. x = Cons(y,ys) /\ f x = eCons)
+
+and D[[]] like
+
+  (x = Nil /\ f x = eNil)
+  \/
+  (x = Cons(cons_1 x,cons_2 x)
+   /\ f x = eCons[cons_1 x/y, cons_2 x/ys])
+
+the latter is nice because we don't need 
+to invent names (consider
+
+  f x = case x of
+          K x -> x
+
+we can't write
+
+  (exists x. x = K x /\ f x = x)
+
+since the two x's equated are not the same!). 
+the former is nice because it's more concise?
+\end{verbatim}
+
+\section{Lemmas}
+
+It's possible to encode, prove, and apply lemmas in our system,
+without adding any extensions.  The application of lemmas can change
+the termination behavior of programs (although so far I think we've
+only used lemmas in other lemmas), and it's up to the user to look out
+for this.
+
+Before presenting the solution we have to understand the problem: the
+FOL prover requires finite axiomitizations, and hence there is no
+general induction support (FOL does not allow quantification over
+formulas, and so induction can only be axiomitized with a schema).
+Each inductive proof of a universal property $\forall x. \phi(x)$
+requires its own induction axiom, and the system does not attempt to
+discover these (a separate problem).  When checking the contract of a
+recursive function, we axiomitize induction for that function,
+following the structure of its recursion.  But a contract may depend
+on a lemma, itself requiring an inductive proof, and so we need a
+facility for introduces auxiliary inductions.  Of course, we also need
+to be able to state the lemmas themselves, because the system does not
+discover them.
+
+Our solution is to define ``dummy'' recursive functions, whose
+recursion structure follows the induction needed to prove the lemma.
+We don't actually care what these functions compute, so we define a
+unit type, $QED$, and return that in the base cases.  E.g., if we want
+to prove that $Z$ is a right identity for addition, then we can define
+a dummy function \texttt{lem\_add\_Z\_r}
+\begin{verbatim}
+lem_add_Z_r x = case x of
+  Zero -> QED
+  Succ x_ -> lem_add_Z_r x_
+
+\end{verbatim}
+and give it a contract stating the property in question
+\begin{verbatim}
+lem_add_Z_r ::: x:CF -> CF&&{qed: x `eqNat` (x `add` Zero)}
+\end{verbatim}
+
+To use the lemma we, need to have $\min$ on it's instantiation.  To
+get this we introduce another dummy function, $using$,
+s.t. $e `using` l$ scrutinizes $l$ and returns $e$.  I.e.
+\begin{verbatim}
+e `using` lem = case lem of
+  QED -> e
+\end{verbatim}
+The \texttt{case} match in $using$ introduces $\min$ for the lemma,
+allowing it's contracts to be unfolded.
+
+XXX: Koen has a long and detailed writeup of lemmas in the his google
+doc.
+
 \section{Correctness of the translation}
@@
 $$ \cf{f} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{\full{f}(x_1,\dots,x_n)}$$
-$$\forall g,x.~\cf{f} \land \cf{x} \to \cf{g~x}$$
-\caption{Encoding of higher-orderness}
+$$\forall g,x.~\cf{g} \land \cf{x} \to \cf{g~x}$$
+\caption{Encoding of higher-orderness.
+%
+XXX: need 'min's.
+%
+XXX: i don't think the second and third formulas are implemented.}
 \label{ho-fig}
